### 2021-05-25 paper/msc.tex
Copy over the partII paper from https://github.com/bueler/mg-glaciers.

\documentclass[letterpaper,final,12pt,reqno]{amsart}

\usepackage[total={6.3in,9.2in},top=1.1in,left=1.1in]{geometry}

\usepackage{times,bm,bbm,empheq,fancyvrb,graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{longtable}
\usepackage{booktabs}

\usepackage[within=section]{newfloat}

\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing}

\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt}

% hyperref should be the last package we load
\usepackage[pdftex,
colorlinks=true,
plainpages=false, % only if colorlinks=true
linkcolor=blue,   % ...
citecolor=Red,    % ...
urlcolor=black    % ...
]{hyperref}

\renewcommand{\baselinestretch}{1.05}

\allowdisplaybreaks[1]  % allow display breaks in align environments, if they avoid major underfulls

\newtheoremstyle{claim}% name
  {5pt}% space above
  {5pt}% space below
  {\itshape}% body font
  {}% indent amount
  {\itshape}% theorem head font
  {.}% punctuation after theorem head
  {.5em}% space after theorem head
  {\thmname{#1}\thmnumber{ #2}\thmnote{ (#3)}}% theorem head spec
\theoremstyle{claim}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\newcommand{\eps}{\epsilon}
\newcommand{\RR}{\mathbb{R}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\hbn}{\hat{\mathbf{n}}}

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

\newcommand{\bxi}{\bm{\xi}}

\newcommand{\bzero}{\bm{0}}

\newcommand{\rhoi}{\rho_{\text{i}}}

\newcommand{\ip}[2]{\left<#1,#2\right>}

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

% numbering
\setcounter{tocdepth}{3}
\makeatletter
\def\l@subsection{\@tocline{2}{0pt}{4pc}{5pc}{}}
\makeatother

\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}
\numberwithin{theorem}{section}

\DeclareFloatingEnvironment[name=Pseudocode]{pcode}

\begin{document}
\title[Multilevel computation of glacier geometry from Stokes dynamics]{Multilevel computation of glacier geometry \\ from Stokes dynamics}

\author{Ed Bueler}

\author{Lawrence Mitchell}

\begin{abstract} FIXME MCD for steady and evolving geometry with Glen-Stokes dynamics
\end{abstract}

\maketitle

%\tableofcontents

\thispagestyle{empty}
%\bigskip

\section{Introduction} \label{sec:intro}

FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020}

\section{The steady ice geometry problem} \label{sec:stokesgeometry}

The standard model for determining the geometry of glaciers is based in part upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.  This Glen-Stokes ice-dynamics model can only determine the free-surface glacier geometry when it is combined with the surface kinematical equation (SKE) and a form of climate input, namely the rate of accumulation of snow or ablation, i.e.~melting and runoff.  In this section we state the strong form of this ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).  The emphasis here is on the complementarity-problem nature of the SIGP, often not explicit in the glaciers literature (e.g.~as observed by \cite{SchoofHewitt2013}).  We then define its weak form, noting along the way several unknown aspects of the theory.  Section \ref{sec:evolution} gives the straightfoward extension of the steady case to a time-discretized evolving-geometry model, the implicit ice geometry problem (IIGP).

The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ (Figure \ref{fig:stokesdomain}).  We denote the map-plane variables as $x$ when $d=1$ and $x,y$ when $d=2$.  On $\Omega$ we assume that a climatic mass balance (CMB) function $a(x,y)$ is defined at every point, whether or not ice is present at that location.  In ice-free areas this function could be described as the ``potential'' CMB, namely the annual balance of snow accumulation and potential melt if ice were present \cite{Cogleyetal2011}, e.g.~as computed by energy balance in a climate or weather model.  Here we assume $a$ has units of ice thickness per time, equivalently ice volume per area (in $\Omega$) per time, compatible with our assumption of constant ice density below.  Also we assume there is a bed elevation function $b(x,y)$ defined everywhere on $\Omega$, which determines the topography upon which the glacier sits.  The functions $a$ and $b$ are the data (inputs) of the SIGP.

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.75\textwidth]{genfigs/stokesdomain.pdf}
\end{center}
\caption{In the steady ice geometry problem (SIGP) the CMB $a$ (arrows; downward $=$ accumulation) and bed elevation $b$ (solid) are given on a fixed map-plane region $\Omega \subset \RR^d$.  The solution includes a surface elevation $s$ (dashed) on $\Omega$; note $s=b$ where ice-free.  The ice velocity $\bu$ and pressure $p$ are defined in the icy domain: $\Lambda_s = \{(x,y,z)\,:\,b(x,y) < z < s(x,y)\} \subset \RR^{d+1}$.}
\label{fig:stokesdomain}
\end{figure}

The Glen-Stokes dynamical model, the so-called stress balance, for ice only applies in the icy domain in $\RR^{d+1}$.  We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption this this icy domain has a well-defined upper surface elevation, a function $s(x,y)$.  That is, we assume there are \emph{no overhangs}, and that the under-side of the ice is in contact with the bed $b$.  We then define $s$ everywhere in $\Omega$ by extending with $s=b$ where ice is absent, thus $s\ge b$ applies on $\Omega$.  Note that $s$ is part of the model solution; it is not given data.

However, as a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.  There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that $s$ is well-defined.  Furthermore the same concern applies to each time step of an evolving model; the model does not stop overhangs from appearing at the margin.  However, because overhangs are small features in large glaciers and ice sheets, essentially all modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.  An exception is \cite{PralongFunk2005}, who furthermore suggest a potentially well-posed continuum model, a nontrivial extension of the one here, in which serac and ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion.  This extended model also explains one of our numerical boundary variants, a short, fixed-height cliff (Section \ref{sec:mcdstokes}), but overhangs are not allowed in any of our numerical constructions.

Based on the assumption of a well-defined upper surface, we define the (solution) extent of the ice as the open set
\begin{equation}
\Lambda_s = \{(x,y,z)\,|\,(x,y) \in \Omega \,\text{ and }\, b(x,y) < z < s(x,y)\}  \subset \RR^{d+1}, \label{eq:lambdas}
\end{equation}
where $z$ is vertically-upward.  (If $d=1$ then coordinates are ``$(x,z)$'' on $\Lambda_s$.)  The map-plane region $\Omega$ need not be connected or simply-connected, but even if so the solution set $\Lambda_s$ need not be.  However, $\Lambda_s$ has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).

For a given geometry $\Lambda_s$, we model the ice as a very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.  Allowing any Glen exponent $n\ge 1$, the equations inside $\Lambda_s$ are
\begin{align}
- \nabla \cdot \tau + \nabla p &= \rhoi \bg &&\text{\emph{stress balance}} \label{eq:forcebalance} \\
\nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
\tau &= B_n |D\bu|^{(1/n) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
\end{align}
The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.  Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} is an aspect of our mass conservation model.  As seen below, mass conservation on the free surface of $\Lambda_s$ requires an additional equation \eqref{eq:ske}.

Regarding tensors and their notation, recall that the (Cauchy) stress tensor $\sigma$ decomposes into the deviatoric part $\tau$ minus the pressure, i.e.~$\sigma = \tau - p\,I$, so equation \eqref{eq:forcebalance} simply says $-\Div \sigma = \rhoi \bg$.  The strain rate tensor $D\bu$ is the symmetric part of $\grad \bu$, $D\bu = \frac{1}{2} \left(\grad\bu + \grad\bu^\top\right)$, and the tensor norm used in \eqref{eq:viscflowlaw} satisfies $|D\bu|^2 = \frac{1}{2} (D\bu)_{ij} (D\bu)_{ij}$.  Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} implies that $\tau$ is also symmetric with trace zero, thus that $p=-(d+1)^{-1} \trace \sigma$.

In computations we will use constants $n=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.  The ice hardness $B_n$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996}.

For the $n=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $n>1$ imply an effective viscosity function of $|D\bu|$.  This would thus be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity
\begin{equation}
\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/n)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used here.  Also we eliminate $\tau$ and rewrite \eqref{eq:forcebalance} as $- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p = \rhoi \mathbf{g}$, with a vector $p$-Laplacian as the leading-order term.

Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding ice.  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.  On the upper free surface we set a condition of zero applied stress,
\begin{equation}
\left(2 \nu_\eps D\bu - pI\right) \bn_s = \bzero  \qquad \qquad \text{\emph{top} } \overline{\partial} \Lambda_s \label{eq:topbc}
\end{equation}
where $\bn_s$ is any normal to $\overline{\partial} \Lambda_s$.  (Providing a nonzero atmospheric pressure at the surface is straightforward, but it has no important consequence.)  On the base we impose no slip:
\begin{equation}
\bu = \bzero  \qquad\qquad \text{\emph{base} } \underline{\partial} \Lambda_s. \label{eq:basebc}
\end{equation}

Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, and the surface gradient $\grad s$ may become singular there, but the top and bottom surfaces are assumed to meet at the margin (Figure \ref{fig:stokesdomain}).  An alternative geometry with a small cliff is also considered (Section \ref{sec:fe}).  Real glacier margins may indeed occur as fracture-generated cliffs \cite{PralongFunk2005}, but fracture processes are not modeled here.

The well-posedness of the above dynamics model, for fixed domain $\Lambda_s$, is proven by \cite{JouvetRappaz2011}, and the solution is a unique pair $(\bu,p)$ defined on $\Lambda_s$.  (The function spaces are identified below when stating the weak formulation.)  However, the simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.  Indeed, the above equations make no reference to the climate input function $a$, so we need another ``equation'', which we will see is actually an inequality.

\newcommand{\bus}{\bu|_s}

Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we also extend the surface normal $\bn_s$, thus we set
\begin{equation}
\bn_s = \left<-s_x,-s_y,1\right> \label{eq:surfacenormal}
\end{equation}
so that the (un-normalized) surface normal is specified almost everywhere on $\Omega$.  (By contrast, note that the dynamical boundary condition \eqref{eq:topbc} applies only on top of the ice.)  Furthermore we extend the surface value of the velocity to all of $\Omega$:
\begin{equation}
\bus(x,y) = \begin{cases} \bu(x,y,s(x,y)), & s(x,y) > b(x,y), \\
                     \bzero, & \text{elsewhere}. \end{cases} \label{eq:surfacevelocity}
\end{equation}
(The movement of the solid earth would determine $\bu|_b \ne 0$, thus $\bu|_s$ could be nonzero in ice-free areas, and a more-complete model could include this small effect.)

The steady-state surface kinematical equation (SKE) \cite[see equation (5.21)]{GreveBlatter2009} is the needed additional aspect of mass conservation:
\begin{equation}
\bus \cdot \bn_s + a = 0 \qquad \text{\emph{on the ice}}. \label{eq:ske}
\end{equation}
Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time.  (Some references use a distinct quantity, namely the thickness added perpendicularly to the ice surface \cite{GreveBlatter2009}, which is significantly different if the gradient $\grad s$ is large.)

While the involved quantities are defined everywhere, kinematical balance \eqref{eq:ske} applies only \emph{on} the ice.  However $\bus \cdot \bn_s + a \le 0$ everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.  In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.  Recall that an NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements $x\ge 0$, $F(x)\ge 0$, and $x F(x)=0$, where $F:V\to V$ \cite{FacchineiPang2003}.

We may now state the strong form of the SIGP as follows:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn_s &= \bzero && \text{on $\overline{\partial} \Lambda_s$} \notag \\
\bu &= \bzero && \text{on $\underline{\partial} \Lambda_s$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but it is coupled to the dynamical boundary value problem formed by the last four statements.

The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$.  However, as the domain on which $\bu,p$ are defined is only known via the solution elevation $s$, \eqref{eq:strongform} is at best an incomplete description.  The weak form in the next section, using a $s\mapsto \bu|_s$ solution operator, will partially address this concern.  Whether in strong or weak form, this inequality-constrained system of PDEs has a largely-unknown theory regarding well-posedness and solution regularity.  However, certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, and uniqueness in the flat bed case.

In Sections \ref{sec:fe}--\ref{sec:mcdstokes} we will construct and demonstrate a robust multilevel scheme for the numerical solution of the SIGP weak form (Section \ref{sec:weakido}).  As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.  By contrast, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.

\section{Weak form and the ice dynamics operator} \label{sec:weakido}

Recall that the weak form of a PDE states that a certain functional, constructed by multiplying by a test function and integrating by parts, is zero.  Both the solution and the test functions come from identified function spaces, namely Sobolev spaces \cite{Evans2010}.  Here we denote by $W^{\text{k},\text{r}}$ the Sobolev space of functions on which have $k$ derivatives which are $r$th-power integrable.

The weak form of the fixed-geometry Glen-Stokes model for ice, which extends the better-known linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.  Suppose that $\Lambda \subset \RR^{d+1}$ is fixed domain on which we prescribe a Dirchlet condition $\bu=0$ on a portion of the boundary, and where the remainder of the boundary (Neumann boundary) is stress-free.  Let $\pp=(1/n)+1$ as before, and let $\qq=(1-\pp^{-1})^{-1}=n+1$ be the conjugate exponent, so that $\pp^{-1}+\qq^{-1}=1$. (Note $\pp=4/3$ and $\qq = 4$ if $n=3$.)  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions which are zero along the Dirichlet boundary.  Let
\begin{equation}
\mathcal{M}_{\text{dyn}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs, the dynamical variables.  The Glen-Stokes weak form is the statement that $(\bu,p) \in \mathcal{M}_{\text{dyn}}$ should satisfy
\begin{equation}
F_{\text{dyn}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_{\text{dyn}}$; as before the effective viscosity $\nu_\eps$ is defined by \eqref{eq:regeffvisc}.  Jouvet and Rappaz \cite{JouvetRappaz2011} have proven that this formulation is well-posed if $\Lambda$ is bounded, the Neumann boundary is $C^1$, and the Dirichlet boundary is nonempty.  In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a coercive functional over the divergence-free subspace, and also that there is a unique pressure $p\in L^\qq(\Lambda)$.

Going beyond predetermined ice geometry, we are interested in finding a triple $s,\bu,p$ which solves the weak form of \eqref{eq:strongform}, where $s$ is the surface elevation.  Suppose $\Omega \subset \RR^d$ is the map-plane domain, and suppose the bed elevation $b$ is in $C^1(\overline{\Omega}) \subset W^{1,\qq}(\Omega)$.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the set of admissible surface elevations of our isolated glacier or ice sheet.  This set is closed and convex.

As noted earlier it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}.  In fact we suppose when needed that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has piecewise-$C^1$ boundary.} \label{eq:quixotic}
\end{equation}
This quixotic assumption connects the known theory of shallow ice sheet well-posedness \cite{JouvetBueler2012}, which motivates definition \eqref{eq:Kconstraintset}, to the widely-used numerical construction of evolving ice sheet geometries from Glen-Stokes dynamics \cite{Jouvetetal2008,Lengetal2012}.  Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory presented in the current paper, which we cannot offer at this time, and subject to the no-overhangs hypothesis which might follow from the inclusion of non-fluid fracture processes (previous section).

If $\partial \Lambda$ is piecewise $C^1$ then the Sobolev space $W^{1,\pp}(\Lambda)$ supports a linear trace operator $T:W^{1,\pp}(\Lambda) \to L^\pp(\partial \Lambda)$ which defines $T(f) = f|_{\partial \Lambda}$ \cite[Section 5.5]{Evans2010}.

With the trace in mind we define the ice dynamics operator (IDO), a fundamental concept for our work (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, extended by zero to all of $\Omega$ (as in \eqref{eq:surfacevelocity}):
\begin{equation}
\Phi(s) = \begin{cases} \bu\big|_{\overline{\partial} \Lambda_s} \cdot \bn_s & s(x,y) > b(x,y), \\
                        \bzero & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Note that the output $\Phi(s)$ may jump at the glacier margin; we only expect $\Phi(s) \in L^\pp(\Omega)$.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$; the thin line is $s+\Phi(s)$.  Both input and output are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may re-write the strong form \eqref{eq:strongform} as an NCP but with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
- \Phi(s) - a &\ge 0 \notag \\
(s - b) (- \Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$, the data $a,b$ are clear, and the solution is the ice surface elevation $z=s(x,y)$.  The velocity and pressure $\bu,p$ are hidden variables within the evaluation of $\Phi(s)$.

It remains to write NCP \eqref{eq:idostrongform} in weak form.  For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we define a functional which is nonlinear in $s$:
\begin{equation}
F(s)[r] = - \int_\Omega (\Phi(s) + a)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity to compute the IDO \eqref{eq:ido}, and then integrating the SKE quantity $\Phi(s) + a$ against the test function $r$.  Implicit in formula \eqref{eq:sigpfunctional} is the assumption that $\Phi(s) + a$ is sufficiently regular to be in the dual space of $W^{1,\qq}(\Omega)$, but this allows significant singularities.

The SIGP weak form is a variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge 0 \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.  Here the computation of $F$ in \eqref{eq:sigpfunctional} is non-local and expensive; to evaluate $F(s)[r]$ one must solve the Glen-Stokes problem on the given $z=s(x,y)$ geometry.

To the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise as a constrained minimization.  The general-bed, steady SIA model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.  Despite this situation, to understand \eqref{eq:sigpweakform} one may analogize to minimization over the closed and convex set $\mathcal{K}$, regarding $F(s)$ as a gradient-like vector.  Definition \eqref{eq:sigpfunctional} can be interpreted as a dual pairing of $F(s)$ and $r$, thus roughly an inner product, in which case the VI says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.  In geometrical analogy, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.

Thus the SIGP couples the weak forms \eqref{eq:glenstokesweak} and \eqref{eq:sigpweakform}.  It therefore has three fundamental nonlinearities, namely the Glen power-law rheology, the inequality constraint, and the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.  Observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}, and also that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

In order to numerically solve \eqref{eq:sigpweakform} we will use the (Gateaux) derivative of the functional $F$ in \eqref{eq:sigpfunctional}, which requires differentiating the IDO $\Phi$ in \eqref{eq:ido} as well.  The efficient computation of these derivatives will also involve an important approximation.

To compute the derivative of $F$ from \eqref{eq:surfacenormal} and \eqref{eq:sigpfunctional}, suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  (In other words, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  Then
\begin{equation}
F'(s,t)[r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s,t)\, r \,dx dy \label{eq:sigpfunctionalderiv}
\end{equation}
where a product-rule computation gives
\begin{align}
\Phi'(s,t) &= \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv} \\
    &= - \left(\lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} - \bus}{\eps}\right) \cdot \bn_s + \Phi(s) \cdot \left<t_x,t_y,0\right>. \notag
\end{align}
In this computation note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.  Observe that $F'$ is straightforward to compute from $\Phi'$.  We will assume that $\Phi'(s,t)$ is in the dual space of $W^{1,\qq}(\Omega)$, thus that \eqref{eq:sigpfunctionalderiv} is finite, but again this allows substantial irregularity.

The new functionals $F'(s,t)$ and $\Phi'(s,t)$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$.  This requires $t\ge 0$ only on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{D}_+ = \{t \in W^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
Then $\Phi'$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.

In the next two sections we construct an iterative, multilevel finite element solver for the SIGP.  However, a significant issue arises from formula \eqref{eq:idoderiv}, namely that to compute the difference $\bu|_{s+\eps t} - \bus$, two Glen-Stokes solutions are seemingly required, namely over the ice domains $\Lambda_{s+\eps t}$ and $\Lambda_s$.  The next section includes an approximation scheme which avoids such large numbers of Glen-Stokes solves.

\section{Finite element discretization} \label{sec:fe}

Suppose $\mathcal{T}$ is a triangulation of $\Omega$, which we assume is polygonal.  Based on its low expected regularity at the ice margin, assume that surface elevations $s\in \mathcal{K}$ are represented using the piecewise-linear ($P_1$) finite element (FE) space \cite{Elmanetal2014}
\begin{equation}
\mathcal{V}^h = \{s:\Omega \to \RR \,:\, s|_T \text{ is linear for each } T \in \mathcal{T}, \text{ and $s$ is continuous}\}. \label{eq:feP1}
\end{equation}
(Note \cite{JouvetBueler2012} make the same choice for the SIA problem.)  Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $P_1$ interpolant of $b$, and let
\begin{equation}
\mathcal{K}^h = \{r \in \mathcal{V}^h \,:\, r \ge b^h \text{ and } r|_{\partial\Omega} = b^h|_{\partial\Omega}\}, \qquad \mathcal{D}_+^h = \{t \in \mathcal{V}^h \,:\, t \ge 0\}. \label{eq:feKD}
\end{equation}

With these FE choices, the discrete SIGP is to compute $s^h \in \mathcal{K}^h$ such that
\begin{equation}
F^h(s^h)[r - s^h] \ge 0 \quad \text{for all } r \in \mathcal{K}^h  \label{eq:fesigpweakform}
\end{equation}
where $F^h(s)[r] = - \int_\Omega (\Phi^h(s) + a)\, r \,dx dy$, essentially as before, but with $\Phi^h$ an FE approximation of the IDO $\Phi$.  (We will compute $\Phi^h$ using a extruded-mesh approximation of the Glen-Stokes problem \eqref{eq:glenstokesweak}, about which we will say more below.)  We suppose that the discrete SIGP is well-posed, subject to the caveats in the last section regarding overhangs at the margin.

Let $p$ denote an arbitrary node of the mesh $\mathcal{T}$, $(x_p,y_p)$ the coordinates of that node, and $\psi_p \in \mathcal{V}^h$ the corresponding $P_1$ hat function such that $\psi_p(x_q,y_q)=\delta_{pq}$ \cite{Elmanetal2014}.  Note that, because $\psi_p \in \mathcal{D}_+^h$, if $r \in \mathcal{K}^h$ then $r + c \psi_p \in \mathcal{K}^h$ for all $c\ge b(x_p,y_p) - r(x_p,y_p)$.

FIXME use Firedrake \cite{Alnaesetal2014,Rathgeberetal2016} with extruded meshes \cite{Gibsonetal2019,McRaeetal2016} for computing $\Phi^h$

\section{Smoothers} \label{sec:smoothers}

As a first step toward solving \eqref{eq:fesigpweakform} we now propose (slow) solvers which will become smoothers in our multilevel method (next section)  The smoothers are the projected nonlinear Gauss-Seidel and projected nonlinear Jacobi (PNJ) methods \cite{KinderlehrerStampacchia1980}.  Given a current surface elevation $s\in \mathcal{K}^h$, these methods sweep through the nodes in a fixed order, solving one-dimensional VI problems, as follows, and updating $s$.  For node $p$ let $\omega_p^s = b(x_p,y_p) - s(x_p,y_p)$.  The one-dimensional VI at $p$ implied by \eqref{eq:fesigpweakform} is to find $c \ge \omega_p^s$ such that
\begin{equation}
F^h(s+c\psi_p)[(s+\gamma \psi_p) - (s+c \psi_p)] \ge 0 \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \omega_p^s$.  Defining $F_p^h(c) = F^h(s+c\psi_p)[\psi_p]$, \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
\begin{equation}
(\gamma - c) F_p^h(c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s$ is then updated $s \gets s + c \psi_p$ by the smoother.  (Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.)  Each pointwise update is done immediately by the (serial) PNGS smoother, while the (parallel) PNJ smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $p$ before the updates.

If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.  The simplest approximation scheme is thus to linearize $F_p^h(c)$, that is, smooth using one projected Newton step for the one-dimensional VI problem at each node.  To implement this strategy, note that \eqref{eq:sigpfunctionalderiv} and \eqref{eq:idoderiv} imply
\begin{equation}
F_p^h(c) = F^h(s+c\psi_p)[\psi_p] \approx F^h(s)[\psi_p] + c\, (F^h)'(s,\psi_p)[\psi_p] = \alpha c + \beta \label{eq:fepointwisefunctionalapprox}
\end{equation}
where
\begin{align}
\alpha &= \int_\Omega \left(- \left(\lim_{\eps\to 0^+} \frac{\bu^h|_{s+\eps \psi_p} - \bu^h|_s}{\eps}\right) \cdot \bn_s + \Phi^h(s) \cdot \left<(\psi_p)_x,(\psi_p)_y,0\right>\right)\, \psi_p \,dx dy \label{eq:fepointwisealpha} \\
\beta &= - \int_\Omega (\Phi^h(s) + a)\, \psi_p \,dx dy \label{eq:fepointwisebeta}
\end{align}
Each smoother application will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation.  In fact, supposing we have computed $\bu^h|_s$ and $\Phi^h(s)$ then several terms in these formulas are straightforward FE computations over the support of $\psi_p$.

However, if we wish to compute $\alpha$ in \eqref{eq:fepointwisealpha}, even approximately, then it remains to compute a difference quotient in \eqref{eq:fepointwisealpha}, which in turn requires evaluating a perturbed-geometry surface velocity $\bu^h|_{s+\eps \psi_p}$.  The expensive direct way of doing this is to solve a Glen-Stokes problem for the perturbed upper-surface geometry $z=s + \eps \psi_p$.

Thus we seek a less expensive way to estimate the perturbed velocity field for a small ``bump'' $\eps\psi_p$ added to the surface elevation $s$.  For small $\eps$ the nonlinearities of the Glen-Stokes problem should be modest, but furthermore the most important change to the velocity and pressure fields from this bump would be through the perturbed gravitational load of the ice in the bump.  Thus FIXME

FIXME Pseudocode \ref{pc:pnj} for PNJ with bump approximation

\begin{pcode}[ht]
\begin{pseudo*}
\pr{pnj}(j,g^j,y^j,N^j,\ell^j,\phi^j,\id{newtonits}=2,\id{omega}=0.67)\text{:} \\+
    \ct{check admissibility: $y^j \ge \phi^j$} \\
    for $k=1,\dots,\id{newtonits}$ \\+
        $\rho_p(c) := N^j(g^j+y^j+c\psi_p^j)[\psi_p^j] - \ell^j[\psi_p^j]$ \\
        $r_p, \delta_p = \rho_p(0), \rho_p'(0)$ \qquad\qquad\qquad\qquad \ct{compute and save for all $p$}\\
        for $p=1,\dots,m_j$ \\+
            $c = \pr{pointupdate}(r_p, \delta_p,y_p,\phi_p,\ell^j[\psi_p^j])$ \\
            $y_p \gets y_p + \id{omega}\,c$
\end{pseudo*}
\caption{Projected nonlinear Jacobi as a smoother.}
\label{pc:pnj}
\end{pcode}

\section{Multilevel constraint decomposition} \label{sec:mcdstokes}

FIXME CP residual norm

FIXME MCD theory

FIXME nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-vcycle}(J,w^J,N,\ell^J,\chi^J,\id{down}=1,\id{coarse}=1,\id{up}=0)\text{:} \\+
    $g^J = w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smoother}}^{\text{\id{down}}}(j,g^j,y^j,N^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{K}^j$} \\
      $F = N^j(g^j+y^j) - \ell^j$ \\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = N^{j-1}(g^{j-1}) - R F$ \\-
    $y^0 = 0$ \\
    $\text{\pr{smoother}}^{\text{\id{coarse}}}(0,g^0,y^0,N^0,\ell^0,\chi^0)$ \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smoother}}^{\text{\id{up}}}(j,g^j,z^j,N^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\-
    return $z^J$
\end{pseudo*}
\caption{Nonlinear MCD V-cycle.}
\label{pc:mcdn-vcycle}
\end{pcode}

FIXME we will iterate \pr{mcdn-vcycle} until the CP residual norm has been reduced by a tolerance factor, as in Pseudocode \ref{pc:mcdn-solver}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-solver}(J,w^J,N,a,b,\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
    $\ell^J = \ip{a}{\cdot}$ \\
    $F(z)[\cdot] := N^J(z)[\cdot] - \ell^J[\cdot]$ \\
    $r_0=\|\hat\bF(w^J)\|$ \\
    for $s=1,\dots,\id{cyclemax}$ \\+
        $\chi^J = b^J - w^J$ \\
        $w^J\gets w^J+\pr{mcdn-vcycle}(J,w^J,N,\ell^J,\chi^J)$ \\
        if $\|\hat\bF(w^J)\| \le \id{rtol} \, r_0$ \\+
            break \\--
\end{pseudo*}
\caption{The SIGP is solved by iterating \pr{mcdn-vcycle}.}
\label{pc:mcdn-solver}
\end{pcode}

FIXME PETSc \cite{Balayetal2020,Bueler2021} used to solve the dynamics problem and thereby evaluate the residual $F$

FIXME in this paper the Glen-Stokes problem is solved by Newton linearization and (parallel) direct solution of the Newton step equations; however, geometric multigrid is available already for Glen-Stokes dynamics \cite{IsaacStadlerGhattas2015}; see also \cite{BrownSmithAhmadia2013} and \cite{Tuminaroetal2016} which solve a first-order approximation of the Glen-Stokes ice sheet problem by geometric and algebraic multigrid, respectively

FIXME convergence results; scaling results

\section{Multilevel methods for evolving geometry} \label{sec:evolution}

FIXME time-dependent runs

\section*{Acknowledgments}  Thanks to David Maxwell for suggestions on the formulation and well-posedness of the model.

\small

\bigskip
\bibliography{msc}
\bibliographystyle{siam}

\appendix

\section{Glossary of acronyms} \label{app:glossary}

\renewcommand{\arraystretch}{1.1}
\begin{longtable}{l|ll}
\toprule
\textbf{Acronym} {\Large$\strut$} & \textbf{Definition} & \textbf{Reference} \\ \hline
CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
FE & finite element & Section \ref{sec:fe} \\
IDO & ice dynamics operator & Section \ref{sec:weakido}, eqn.~\eqref{eq:ido} \\
IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
PNJ & projected, nonlinear Jacobi & Section \ref{sec:smoothers} \\
SIA & shallow ice approximation & Section \ref{sec:intro} \\
SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, eqn.~\eqref{eq:strongform} \\
SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, eqn.~\eqref{eq:ske} \\
VI & variational inequality & Section \ref{sec:weakido} \\
WU & work units & Section \ref{sec:mcdstokes} \\ % final \\ required
\bottomrule
\caption{Glossary of acronyms used in this paper.}
\label{tab:acronyms}
\end{longtable}

\end{document}

### 2021-05-25 paper/genfigs/stokesdomain.py

@@ -1,0 +1,80 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib.pyplot as plt
+from writeout import writeout
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["mathtext.fontset"] = "cm"
+
+fsize=18.0
+smallfsize=14.0
+extrasmallfsize=12.0
+bigfsize=20.0
+
+def genbasicfig(xshift=0.0):
+    x = np.linspace(0.0,10.0,1001)
+    # bed elevation
+    b = 0.07*(x-3.0)**2 + 0.2*np.sin(2.0*x) - 0.1
+    plt.plot(x + xshift, b, 'k', lw=2.5)
+    # current thickness for Omega^{n-1}
+    h0 = 3.0
+    L = 3.0
+    firstshape = h0*(-0.2 + np.sqrt(np.maximum(0.0,1.0 - (x-5)**2/L**2)))
+    thk = np.maximum(0.0, firstshape)
+    # surface
+    s = b + thk
+    offset = 0.1
+    plt.plot(x + xshift, s + offset, 'k--', lw=3.0)
+    return x + xshift, s, b
+
+def drawclimate(x,s):
+    plt.text(x[10], s[10]+2.7, r'$a(x,y)$', fontsize=bigfsize, color='k')
+    for j in range(10):
+        xarr = x[50+100*j]
+        if j>0:
+            magarr = 0.6*np.sin(np.pi/2 + 0.6*xarr)
+        else:
+            magarr = 0.05
+        plt.arrow(xarr, s.max()+0.3, 0.0, magarr,
+                  lw=1.5, head_width=0.1, color='k')
+
+# domain notation figure
+plt.figure(figsize=(10,5.5))
+x, s, b = genbasicfig()
+plt.text(x[600] - 1.0, b[600] + 0.4 * s[600], r'$\Lambda_s$',
+         fontsize=bigfsize, color='k')
+drawclimate(x,s)
+# mark top surface
+plt.text(x[250], s[250] + 1.5, r'$s(x,y)$', fontsize=bigfsize, color='k')
+# mark bottom surface
+plt.text(x[650], b[650] - 0.5, r'$b(x,y)$', fontsize=bigfsize, color='k')
+# show \Omega
+yR = min(b) - 0.5
[…]

### 2021-05-25 paper/genfigs/writeout.py

@@ -1,0 +1,11 @@
+# module provides writeout()
+
+import matplotlib.pyplot as plt
+
+_SHOW = False
+def writeout(outname):
+    if _SHOW:
+        plt.show()
+    else:
+        print('writing file ' + outname)
+        plt.savefig(outname, bbox_inches='tight')

### 2021-05-25 py/basesmoother.py
Copy over the partII Python codes from https://github.com/bueler/mg-glaciers.

@@ -1,0 +1,80 @@
+'''Module for abstract base class SmootherObstacleProblem.'''
+
+__all__ = ['SmootherObstacleProblem']
+
+from abc import ABC, abstractmethod
+import numpy as np
+
+class SmootherObstacleProblem(ABC):
+    '''Abstact base class for a smoother on an obstacle problem.  Works on
+    any mesh of class MeshLevel1D.'''
+
+    def __init__(self, args, admissibleeps=1.0e-10):
+        self.args = args
+        self.admissibleeps = admissibleeps
+        self.inadmissible = 0  # count of repaired admissibility violations
+        self.name = 'base'
+
+    def _checkrepairadmissible(self, mesh, w, phi):
+        '''Check and repair feasibility.'''
+        for p in range(1, mesh.m+1):
+            if w[p] < phi[p] - self.admissibleeps:
+                if self.args.printwarnings:
+                    print('WARNING: repairing inadmissible w[%d]=%e < phi[%d]=%e on level %d (m=%d)' \
+                          % (p, w[p], p, phi[p], mesh.j, mesh.m))
+                w[p] = phi[p]
+                self.inadmissible += 1
+
+    def _sweepindices(self, mesh, forward=True):
+        '''Generate indices for sweep.'''
+        if forward:
+            ind = range(1, mesh.m+1)    # 1,...,m
+        else:
+            ind = range(mesh.m, 0, -1)  # m,...,1
+        return ind
+
+    def shownonzeros(self, z):
+        '''Print a string indicating locations where array z is zero.'''
+        Jstr = ''
+        for k in range(len(z)):
+            Jstr += '_' if z[k] == 0.0 else '*'
+        print('  %d nonzeros: ' % sum(z > 0.0) + Jstr)
+
+    def inactiveresidualnorm(self, mesh, w, r, phi, ireps=50.0):
+        '''Compute the norm of the residual values at nodes where the constraint
+        is NOT active.  Where the constraint is active the residual F(w) in the
+        complementarity problem is allowed to have any positive value; only the
+        residual at inactive nodes is relevant to convergence.'''
+        F = r.copy()
+        F[w < phi + ireps] = np.minimum(F[w < phi + ireps], 0.0)
+        return mesh.l2norm(F)
+
+    def smoother(self, iters, mesh, w, ell, phi):
+        '''Apply iters sweeps of obstacle-problem smoother on mesh to modify w in place.  Alternate directions (unless overridden).'''
+        forward = True
[…]

### 2021-05-25 py/experiments/poissonL.py

@@ -1,0 +1,53 @@
+#!/usr/bin/env python3
+
+# FIXME this code relates to a suggestion by Colin Cotter, not followed up,
+# about subclassing DirichletBC to allow interior imposition of trivialized
+# equations, like Dirichlet boundary conditions area handled
+
+from firedrake import *
+
+class InteriorBC(DirichletBC):
+    #@cached_property
+    def nodes(self):
+        dm = self.function_space().mesh().topology_dm
+        section = self.function_space().dm.getDefaultSection()
+        nodes = []
+        for sd in as_tuple(self.sub_domain):
+            nfaces = dm.getStratumSize(FACE_SETS_LABEL, sd)
+            faces = dm.getStratumIS(FACE_SETS_LABEL, sd)
+            if nfaces == 0:
+                continue
+            for face in faces.indices:
+                if dm.getLabelValue("interior_facets", face) < 0:
+                    continue
+                closure, _ = dm.getTransitiveClosure(face)
+                for p in closure:
+                    dof = section.getDof(p)
+                    offset = section.getOffset(p)
+                    nodes.extend((offset + d) for d in range(dof))
+        return np.unique(np.asarray(nodes, dtype=IntType))
+
+mesh = UnitSquareMesh(8, 8)
+
+W = FunctionSpace(mesh, 'Lagrange', degree=1)
+f_rhs = Constant(1.0)
+u = Function(W)
+v = TestFunction(W)
+F = (dot(grad(u), grad(v)) - f_rhs * v) * dx
+
+g_bdry = Constant(0.0)
+bdry_ids = (1, 2, 3, 4)   # all four sides of boundary
+bc = DirichletBC(W, g_bdry, bdry_ids)
+
+# substituting this throws error; it does not have same calling signature as DirichletBC()?
+# bc = InteriorBC(W, g_bdry, bdry_ids)
+
+solve(F == 0, u, bcs = [bc], options_prefix = 's',
+      solver_parameters = {'snes_type': 'ksponly',
+                           'ksp_type': 'preonly',
+                           'pc_type': 'cholesky'})
+
+solnnorm = sqrt(assemble(dot(u,u) * dx))
+PETSc.Sys.Print('solution norm |u|_2 = %.3e' % solnnorm)
+u.rename('u')
+File('soln.pvd').write(u)

### 2021-05-25 py/experiments/test_extruded_top.py

@@ -1,0 +1,54 @@
+#!/usr/bin/env python3
+
+# sent as MFE to Firedrake slack on 9 May
+# Lawrence responded it is unsupported ... but then on slack 13 May he patched
+#   firedrake/src/firedrake/cython/extrusion_numbering.pyx
+# and now it is supported
+
+from firedrake import *
+
+# supported extruded mesh with 1-element columns
+#         6 --11
+#         |    |
+#         5 --10
+#         |    |
+#         4 -- 9
+#         |    |
+#    1 -- 3 -- 8 --13
+#    |    |    |    |
+#    0 -- 2 -- 7 --12
+print('1-element columns')
+basemesh = UnitIntervalMesh(3)
+layermap = [[0,1], [0,4], [0, 1]]
+mesh = ExtrudedMesh(basemesh, layers=layermap, layer_height=0.25)
+P1 = FunctionSpace(mesh, 'Lagrange', 1)
+bottombc = DirichletBC(P1, 1.0, 'bottom')
+print(bottombc.nodes) # [0 2 7 12]
+topbc = DirichletBC(P1, 1.0, 'top')
+print(topbc.nodes) # [1 3 4 5 6 8 9 10 11 13]
+
+# NOW supported extruded mesh with zero-element columns
+#         5 --10
+#         |    |
+#         4 -- 9
+#         |    |
+#         3 -- 8
+#         |    |
+#         2 -- 7
+#         |    |
+#    0 -- 1 -- 6 -- 11   (note 1,6 are both top and bottom)
+print('0-element columns')
+basemesh = UnitIntervalMesh(3)
+layermap = [[0,0], [0,4], [0, 0]]
+mesh = ExtrudedMesh(basemesh, layers=layermap, layer_height=0.25)
+P1 = FunctionSpace(mesh, 'Lagrange', 1)
+bottombc = DirichletBC(P1, 1.0, 'bottom')
+print(bottombc.nodes) # [0 1 6 11]
+topbc = DirichletBC(P1, 1.0, 'top')
+# following is now fixed:
+#topbc.nodes  # FAILS: "AssertionError: Not expecting negative number of layers"
+print(topbc.nodes) # [1 2 3 4 5 6 7 8 9 10]
+
+# use Paraview point inspector for another confirmation
+#f = Function(P1)
+#File('foo.pvd').write(f)

### 2021-05-25 py/experiments/top_dofs_1d.py

@@ -1,0 +1,70 @@
+#!/usr/bin/env python3
+
+'''For an extruded mesh over a 1D interval base mesh, with some empty columns,
+loop over the base mesh cells to get the top cells and their top nodes.
+This code results from major slack help from Lawrence Mitchell.'''
+
+from firedrake import *
+
+# define base mesh and get its coordinates
+bm = UnitIntervalMesh(6)
+W = FunctionSpace(bm, 'Lagrange', 1)  # P1
+bm_cell_node_map = W.cell_node_map().values
+x = SpatialCoordinate(bm)  # does not unpack, thus x[0] below
+bm_xf = Function(W).interpolate(x[0])
+
+# picture of extruded mesh, with its node indexing:
+#              10--14
+#               |   |
+#       3---6   9--13
+#       |   |   |   |
+#       2---5---8--12  16--18
+#       |   |   |   |   |   |
+#   0---1---4---7--11--15--17   <-- base mesh of 6 intervals
+#     0   1   2   3   4   5     <-- numbering of base mesh cells
+mesh = ExtrudedMesh(bm, layers=[[0,0], [0,2], [0,1], [0,3], [0,0], [0,1]],
+                    layer_height=0.3)
+
+# NOTE mesh.cell_set.layers_array is a singleton  [[0,n],]  if
+# "mesh = ExtrudedMesh(bm, layers=n)", so generally check mesh.variable_layers
+assert mesh.variable_layers
+
+# use mesh coordinate functions so I can check the above picture
+x, z = SpatialCoordinate(mesh)
+V = FunctionSpace(mesh, 'Lagrange', 1)  # Q1
+xf = Function(V).interpolate(x)
+zf = Function(V).interpolate(z)
+
+# map from *lowest* extruded-mesh cells into extruded-mesh node indices
+# WARNING: here cell_node_map[j] is invalid nonsense for j=0,4 empty columns!
+cell_node_map = V.cell_node_map().values
+#print(cell_node_map)
+node_offset = V.cell_node_map().offset  # node offset as you go up a column
+
+# to get nodes on top cells, get the cell-wise indexing scheme
+section, iset, facets = V.cell_boundary_masks
+# facets ordered with sides first, then bottom, then top
+off = section.getOffset(facets[-1])    # -2 for "bottom"
+dof = section.getDof(facets[-1])
+indices = iset[off:off+dof]
+
[…]

### 2021-05-25 py/smoother.py

@@ -1,0 +1,287 @@
+'''Module for SmootherStokes class derived from SmootherObstacleProblem.'''
+
+import numpy as np
+import firedrake as fd
+from basesmoother import SmootherObstacleProblem
+
+secpera = 31556926.0        # seconds per year
+
+def extend(mesh, f):
+    '''On an extruded mesh extend a function f(x,z), already defined on the
+    base mesh, to the mesh using the 'R' constant-in-the-vertical space.'''
+    Q1R = fd.FunctionSpace(mesh, 'P', 1, vfamily='R', vdegree=0)
+    fextend = fd.Function(Q1R)
+    fextend.dat.data[:] = f.dat.data_ro[:]
+    return fextend
+
+def D(w):
+    return 0.5 * (fd.grad(w) + fd.grad(w).T)
+
+class SmootherStokes(SmootherObstacleProblem):
+    '''To evaluate the residual this Jacobi smoother solves the Stokes problem
+    for the given geometry by creating an extruded mesh.'''
+
+    def __init__(self, args, admissibleeps=1.0e-10):
+        super().__init__(args, admissibleeps=admissibleeps)
+        # smoother name
+        self.name = 'SmootherStokes'
+        # physical parameters
+        self.g = 9.81                    # m s-2
+        self.rhoi = 910.0                # kg m-3
+        self.nglen = 3.0
+        self.A3 = 1.0e-16 / secpera      # Pa-3 s-1;  EISMINT I ice softness
+        self.B3 = self.A3**(-1.0/3.0)    # Pa s(1/3);  ice hardness
+        # used in Stokes solver
+        self.Dtyp = 1.0 / secpera        # s-1
+        self.sc = 1.0e-7                 # velocity scale for symmetric scaling
+        # parameters for initial condition, a Bueler profile; see van der Veen
+        #   (2013) section 5.3
+        self.buelerL = 10.0e3       # half-width of sheet
+        self.buelerH0 = 1000.0      # center thickness
+        self.Gamma = 2.0 * self.A3 * (self.rhoi * self.g)**self.nglen
+        self.Gamma /= (self.nglen + 2.0)
+        # we store the basemesh info and the bed elevation
+        self.basemesh = None
+        self.mx = None
+        self.b = None
+        self.saveflag = False
+        self.savename = None
+
+    def savestatenextresidual(self, name):
+        '''On next call to residual(), save the state.'''
+        self.saveflag = True
+        self.savename = name
[…]

### 2021-05-25 py/steady.py

@@ -1,0 +1,162 @@
+#!/usr/bin/env python3
+'''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
+
+# TODO:
+#   1. implement Jacobi smoother by SLOW method, getting derivative of SKE
+#      w.r.t. a surface displacement by finite difference
+#   2. copy partI/mcdn.py and build it out
+
+import sys
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+from firedrake import *
+
+from meshlevel import MeshLevel1D
+from smoother import SmootherStokes
+#from mcdn import mcdnsolver
+
+parser = argparse.ArgumentParser(description='''
+Solve the 1D obstacle problem for the steady geometry of a glacier using
+Stokes dynamics:  For given climate a(x) and bed elevation b(x),
+find s(x) in the closed, convex subset
+    K = {r | r >= b}
+so that the variational inequality (VI) holds,
+    F(s)[r-s] >= 0   for all r in K.
+Note s solves the surface kinematical equation, as an interior PDE, in the
+inactive set {x | s(x) > b(x)}.
+
+Solution is by the nonlinear (FAS) extension of the multilevel constraint
+decomposition (MCD) method of Tai (2003).  See the simpler cases, including
+the SIA, in part I (Bueler, 2022).
+
+Initial implementation generates Bueler profile geometry as initial state
+and then tries to converge from there.
+
+References:
+  * Bueler, E. (2022). Geometric multigrid for glacier modeling I:
+    New concepts and algorithms.  In preparation.
+  * Tai, X.-C. (2003). Rate of convergence for some constraint
+    decomposition methods for nonlinear variational inequalities.
+    Numer. Math. 93 (4), 755--786.
+''',
+    formatter_class=argparse.RawTextHelpFormatter,
+    allow_abbrev=False,  # bug in python 3.8 causes this to be ignored
+    add_help=False)
+adda = parser.add_argument
+adda('-alpha', type=float, metavar='X', default=1000.0,  # FIXME sensitive
+    help='Richardson iteration parameter (default=1000.0)')
+adda('-coarse', type=int, default=1, metavar='N',
+     help='smoother sweeps on coarsest grid (default=1)')
+adda('-cyclemax', type=int, default=100, metavar='N',
[…]

### 2021-05-25 py/testit.sh

@@ -1,0 +1,43 @@
+#!/bin/bash
+
+# A script to run regression tests.  Use "make test" in the subdirectories,
+# which runs this script as follows:
+#    ./testit.sh PROGRAM OPTS PROCESSES TESTNUM DESCRIPTION
+
+rm -f maketmp tmp difftmp
+
+make $1 > maketmp 2>&1;
+
+grep warning maketmp
+
+CURRDIR=${PWD##*/}
+
+if [ $3 -eq 1 ]; then
+    CMD="./$1 $2"
+else
+    CMD="mpiexec -n $3 ./$1 $2"
+fi
+
+if [[ ! -f output/$1.test$4 ]]; then
+    echo "FAIL: Test #$4 of $CURRDIR/$1  ($5)"
+    echo "       command = '$CMD'"
+    echo "       OUTPUT MISSING"
+
+else
+
+    $CMD > tmp
+
+    diff output/$1.test$4 tmp > difftmp
+
+    if [[ -s difftmp ]] ; then
+       echo "FAIL: Test #$4 of $CURRDIR/$1  ($5)"
+       echo "       command = '$CMD'"
+       echo "       diffs follow:"
+       cat difftmp
+    else
+       echo "PASS: Test #$4 of $1  ($5)"
+       rm -f maketmp tmp difftmp
+    fi
+
+fi
+

### 2021-05-26 paper/msg.tex
Correct a naming typo.
Edits regarding the weak form and the FE approximation.

--- 2021-05-25 paper/msc.tex
@@
 
-Recall that the weak form of a PDE states that a certain functional, constructed by multiplying by a test function and integrating by parts, is zero.
-Both the solution and the test functions come from identified function spaces, namely Sobolev spaces \cite{Evans2010}.
-Here we denote by $W^{\text{k},\text{r}}$ the Sobolev space of functions on which have $k$ derivatives which are $r$th-power integrable.
+The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.
+Then we will address the weak form of the SIGP.
 
-The weak form of the fixed-geometry Glen-Stokes model for ice, which extends the better-known linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.
-Suppose that $\Lambda \subset \RR^{d+1}$ is fixed domain on which we prescribe a Dirchlet condition $\bu=0$ on a portion of the boundary, and where the remainder of the boundary (Neumann boundary) is stress-free.
-Let $\pp=(1/n)+1$ as before, and let $\qq=(1-\pp^{-1})^{-1}=n+1$ be the conjugate exponent, so that $\pp^{-1}+\qq^{-1}=1$.
-(Note $\pp=4/3$ and $\qq = 4$ if $n=3$.)  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions which are zero along the Dirichlet boundary.
-Let
+Denote by $W^{\text{k},\text{r}}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.
+Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).
+Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and let $\qq=(1-\pp^{-1})^{-1}=n+1$ be the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.
+Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and
 \begin{equation}
-\mathcal{M}_{\text{dyn}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
+\mathcal{M}_{\text{GS}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
 \end{equation}
-be the (mixed) space of admissible velocity and pressure pairs, the dynamical variables.  The Glen-Stokes weak form is the statement that $(\bu,p) \in \mathcal{M}_{\text{dyn}}$ should satisfy
+be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes weak form is the statement that $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy
 \begin{equation}
-F_{\text{dyn}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
+F_{\text{GS}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
 \end{equation}
-for all $(\bv,q) \in \mathcal{M}_{\text{dyn}}$; as before the effective viscosity $\nu_\eps$ is defined by \eqref{eq:regeffvisc}.
-Jouvet and Rappaz \cite{JouvetRappaz2011} have proven that this formulation is well-posed if $\Lambda$ is bounded, the Neumann boundary is $C^1$, and the Dirichlet boundary is nonempty.
-In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a coercive functional over the divergence-free subspace, and also that there is a unique pressure $p\in L^\qq(\Lambda)$.
+for all $(\bv,q) \in \mathcal{M}_{\text{GS}}$.
 
-Going beyond predetermined ice geometry, we are interested in finding a triple $s,\bu,p$ which solves the weak form of \eqref{eq:strongform}, where $s$ is the surface elevation.
-Suppose $\Omega \subset \RR^d$ is the map-plane domain, and suppose the bed elevation $b$ is in $C^1(\overline{\Omega}) \subset W^{1,\qq}(\Omega)$.
+Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary is $C^1$.
+In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p\in L^\qq(\Lambda)$.
+
+However, we seek to go beyond this dynamical problem on predetermined ice geometry.
+We are interested in finding a triple $s,\bu,p$ which solves \eqref{eq:strongform}, or rather its weak form which we now propose.
+Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.
+Because of the $C^1$ boundary, the Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines $f|_{\partial \Omega}$ for $f \in W^{1,\qq}(\Omega)$ \cite[Section 5.5]{Evans2010}.
 Let
@@
 \end{equation}
-be the set of admissible surface elevations of our isolated glacier or ice sheet.  This set is closed and convex.
+be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.
+However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
+\begin{equation}
+s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ free surface.} \label{eq:quixotic}
+\end{equation}
+Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.
+However, the existing theory of shallow ice sheet well-posedness \cite{JouvetBueler2012} includes a version of definition \eqref{eq:Kconstraintset},\footnote{In \cite{JouvetBueler2012} the $2\qq/(\qq-1)=8/3$ power of the ice thickness is in $W^{1,\qq}(\Omega)$.} and in that shallow theory $s$ solves an elliptic equation and is thus smooth, at least interior to the free boundary.
 
-As noted earlier it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}.  In fact we suppose when needed that
+We may now define the ice dynamics operator (IDO), a map using the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).
+For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
 \begin{equation}
-s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has piecewise-$C^1$ boundary.} \label{eq:quixotic}
+\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
+                        \bzero, & \text{otherwise}, \end{cases} \label{eq:ido}
 \end{equation}
-This quixotic assumption connects the known theory of shallow ice sheet well-posedness \cite{JouvetBueler2012}, which motivates definition \eqref{eq:Kconstraintset}, to the widely-used numerical construction of evolving ice sheet geometries from Glen-Stokes dynamics \cite{Jouvetetal2008,Lengetal2012}.
-Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory presented in the current paper, which we cannot offer at this time, and subject to the no-overhangs hypothesis which might follow from the inclusion of non-fluid fracture processes (previous section).
-
-If $\partial \Lambda$ is piecewise $C^1$ then the Sobolev space $W^{1,\pp}(\Lambda)$ supports a linear trace operator $T:W^{1,\pp}(\Lambda) \to L^\pp(\partial \Lambda)$ which defines $T(f) = f|_{\partial \Lambda}$ \cite[Section 5.5]{Evans2010}.
-
-With the trace in mind we define the ice dynamics operator (IDO), a fundamental concept for our work (Figure \ref{fig:idoaction}).
-For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, extended by zero to all of $\Omega$ (as in \eqref{eq:surfacevelocity}):
-\begin{equation}
-\Phi(s) = \begin{cases} \bu\big|_{\overline{\partial} \Lambda_s} \cdot \bn_s & s(x,y) > b(x,y), \\
-                        \bzero & \text{otherwise}. \end{cases} \label{eq:ido}
-\end{equation}
-Note that the output $\Phi(s)$ may jump at the glacier margin; we only expect $\Phi(s) \in L^\pp(\Omega)$.
+where $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
+Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, thus permitting substantial irregularities.
 
@@
 
-Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may re-write the strong form \eqref{eq:strongform} as an NCP but with a hidden dynamical problem:
+Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may immediately re-write the \eqref{eq:strongform} in a simpler strong form, namely an NCP with a hidden dynamical problem:
 \begin{align}
@@
 \end{align}
-Here the statements hold on all of $\Omega$, the data $a,b$ are clear, and the solution is the ice surface elevation $z=s(x,y)$.
-The velocity and pressure $\bu,p$ are hidden variables within the evaluation of $\Phi(s)$.
+Here the statements hold on all of $\Omega$ and the data $a,b$ are clear.  The solution is the ice surface $z=s(x,y)$, but $\bu,p$ are hidden variables within the evaluation of $\Phi(s)$.
 
-It remains to write NCP \eqref{eq:idostrongform} in weak form.  For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we define a functional which is nonlinear in $s$:
+For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a functional which is nonlinear in $s$:
 \begin{equation}
@@
 \end{equation}
-One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity to compute the IDO \eqref{eq:ido}, and then integrating the SKE quantity $\Phi(s) + a$ against the test function $r$.
-Implicit in formula \eqref{eq:sigpfunctional} is the assumption that $\Phi(s) + a$ is sufficiently regular to be in the dual space of $W^{1,\qq}(\Omega)$, but this allows significant singularities.
+One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the SKE quantity $\Phi(s) + a$ against the test function $r$.
+Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s) + a$ has minimal regularity; it need only be in the dual space of $W^{1,\qq}(\Omega)$.
 
-The SIGP weak form is a variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
+Finally, the SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
 \begin{equation}
@@
 This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.
-Here the computation of $F$ in \eqref{eq:sigpfunctional} is non-local and expensive; to evaluate $F(s)[r]$ one must solve the Glen-Stokes problem on the given $z=s(x,y)$ geometry.
+Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.
 
-To the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise as a constrained minimization.
-The general-bed, steady SIA model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
-Despite this situation, to understand \eqref{eq:sigpweakform} one may analogize to minimization over the closed and convex set $\mathcal{K}$, regarding $F(s)$ as a gradient-like vector.
-Definition \eqref{eq:sigpfunctional} can be interpreted as a dual pairing of $F(s)$ and $r$, thus roughly an inner product, in which case the VI says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.
-In geometrical analogy, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
+Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.
+VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.
+Specifically, in geometrical analogy, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
 
-Thus the SIGP couples the weak forms \eqref{eq:glenstokesweak} and \eqref{eq:sigpweakform}.
-It therefore has three fundamental nonlinearities, namely the Glen power-law rheology, the inequality constraint, and the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
-Observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}, and also that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.
+It is well-known that some VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
+The general-bed, steady shallow (SIA) model is not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
 
-In order to numerically solve \eqref{eq:sigpweakform} we will use the (Gateaux) derivative of the functional $F$ in \eqref{eq:sigpfunctional}, which requires differentiating the IDO $\Phi$ in \eqref{eq:ido} as well.
-The efficient computation of these derivatives will also involve an important approximation.
+The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through the definition of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
+\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
+\begin{enumerate}
+\item the Glen power-law rheology,
+\item the inequality constraint, and
+\item the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
+\end{enumerate}
+Regarding (\emph{ii}), observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}.
+Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.
 
-To compute the derivative of $F$ from \eqref{eq:surfacenormal} and \eqref{eq:sigpfunctional}, suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
-(In other words, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  Then
+To numerically solve \eqref{eq:sigpweakform} we will need the (Gateaux) derivative of $F$, which requires differentiating the $\Phi$ as well.
+Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
+(Both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional},
 \begin{equation}
-F'(s,t)[r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s,t)\, r \,dx dy \label{eq:sigpfunctionalderiv}
+F'(s)[t,r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s)[t]\, r \,dx dy \label{eq:sigpfunctionalderiv}
 \end{equation}
-where a product-rule computation gives
+where a product-rule computation on \eqref{eq:ido} gives
 \begin{align}
-\Phi'(s,t) &= \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv} \\
+\Phi'(s)[t] &= \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv} \\
     &= - \left(\lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} - \bus}{\eps}\right) \cdot \bn_s + \Phi(s) \cdot \left<t_x,t_y,0\right>. \notag
 \end{align}
-In this computation note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.
-Observe that $F'$ is straightforward to compute from $\Phi'$.
-We will assume that $\Phi'(s,t)$ is in the dual space of $W^{1,\qq}(\Omega)$, thus that \eqref{eq:sigpfunctionalderiv} is finite, but again this allows substantial irregularity.
 
-The new functionals $F'(s,t)$ and $\Phi'(s,t)$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$.
-This requires $t\ge 0$ only on the (active) set where $s=b$.
+Observe that $F'$ is straightforward to compute from $\Phi'$, and we will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.
+In \eqref{eq:idoderiv} note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.
+
+The new functionals $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, and this clearly requires $t\ge 0$ on the (active) set where $s=b$.
 However, to simplify our considerations we define an $s$-independent set
 \begin{equation}
-\mathcal{D}_+ = \{t \in W^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
+\mathcal{D}_+ = \{t \in W^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}, \label{eq:infdefectset}
 \end{equation}
-Then $\Phi'$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
+and $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$.
+Thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
 
-In the next two sections we construct an iterative, multilevel finite element solver for the SIGP.
-However, a significant issue arises from formula \eqref{eq:idoderiv}, namely that to compute the difference $\bu|_{s+\eps t} - \bus$, two Glen-Stokes solutions are seemingly required, namely over the ice domains $\Lambda_{s+\eps t}$ and $\Lambda_s$.
-The next section includes an approximation scheme which avoids such large numbers of Glen-Stokes solves.
+In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.
+When we approximate the true derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because two Glen-Stokes solutions over the ice domains $\Lambda_{s+\eps t}$ and $\Lambda_s$ are seemingly required.
+While this would imply a very large numbers of expensive dynamical solves, in Section \ref{sec:smoothers} we propose a less-expensive approximation.
 
@@
 Suppose $\mathcal{T}$ is a triangulation of $\Omega$, which we assume is polygonal.
-Based on its low expected regularity at the ice margin, assume that surface elevations $s\in \mathcal{K}$ are represented using the piecewise-linear ($P_1$) finite element (FE) space \cite{Elmanetal2014}
-\begin{equation}
-\mathcal{V}^h = \{s:\Omega \to \RR \,:\, s|_T \text{ is linear for each } T \in \mathcal{T}, \text{ and $s$ is continuous}\}. \label{eq:feP1}
-\end{equation}
-(Note \cite{JouvetBueler2012} make the same choice for the SIA problem.)  Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $P_1$ interpolant of $b$, and let
+(If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on their low expected regularity at the ice margin, we will represent the surface elevations $s\in \mathcal{K}$ using the piecewise-linear ($P_1$) finite element (FE) space $\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega)$.
+Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.
+
+Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$, and define two closed and convex subsets of $\mathcal{V}^h$:
 \begin{equation}
@@
 \end{equation}
+Let $(x_j,y_j)$ denote the coordinates an arbitrary node of the mesh $\mathcal{T}$, and $\psi_j \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.
 
-With these FE choices, the discrete SIGP is to compute $s^h \in \mathcal{K}^h$ such that
+With these FE choices, the discrete SIGP computes the FE approximation $s^h \in \mathcal{K}^h$ satisfying
 \begin{equation}
-F^h(s^h)[r - s^h] \ge 0 \quad \text{for all } r \in \mathcal{K}^h  \label{eq:fesigpweakform}
+F^h(s^h)[r - s^h] \ge 0 \quad \text{for all } r \in \mathcal{K}^h , \label{eq:fesigpweakform}
 \end{equation}
-where $F^h(s)[r] = - \int_\Omega (\Phi^h(s) + a)\, r \,dx dy$, essentially as before, but with $\Phi^h$ an FE approximation of the IDO $\Phi$.
-(We will compute $\Phi^h$ using a extruded-mesh approximation of the Glen-Stokes problem \eqref{eq:glenstokesweak}, about which we will say more below.)  We suppose that the discrete SIGP is well-posed, subject to the caveats in the last section regarding overhangs at the margin.
+where $F^h(s^h)[r] = - \int_\Omega (\Phi^h(s^h) + a)\, r \,dx dy$; compare \eqref{eq:sigpfunctional}.
+Here $\Phi^h$ is an FE approximation of the IDO $\Phi$; see below.
+We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.
 
-Let $p$ denote an arbitrary node of the mesh $\mathcal{T}$, $(x_p,y_p)$ the coordinates of that node, and $\psi_p \in \mathcal{V}^h$ the corresponding $P_1$ hat function such that $\psi_p(x_q,y_q)=\delta_{pq}$ \cite{Elmanetal2014}.
-Note that, because $\psi_p \in \mathcal{D}_+^h$, if $r \in \mathcal{K}^h$ then $r + c \psi_p \in \mathcal{K}^h$ for all $c\ge b(x_p,y_p) - r(x_p,y_p)$.
+All FE computations in this paper use the Python library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.
+For $s^h \in \mathcal{K}^h$ we will compute $\Phi^h(s^h)$, namely the numerical Glen-Stokes \eqref{eq:glenstokesweak} surface velocity associated to the surface elevation $s^h$, using a extruded-mesh approximation of $\Lambda_s^h \subset \RR^{d+1}$, the icy domain as defined by \eqref{eq:lambdas}.
+Specifically, if $d=2$ and if $\mathcal{T}$ is a triangulation of $\Omega \subset \RR^d$, and given $s^h$ and $b^h$ from $\mathcal{V}^h$, we apply the Firedrake concept of an extruded mesh \cite{Gibsonetal2019,McRaeetal2016} to give a prism element decomposition of $\Lambda_s^h$.
+If $d=1$ then the extruded mesh has quadrilateral elements; a sketch is shown in Figure \ref{fig:extruded}.
 
-FIXME use Firedrake \cite{Alnaesetal2014,Rathgeberetal2016} with extruded meshes \cite{Gibsonetal2019,McRaeetal2016} for computing $\Phi^h$
+\begin{figure}[t]
+\begin{center}
+FIXME %\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
+\end{center}
+\caption{Given a triangulation ($d=2$) or interval decomposition ($d=1$ as shown) of $\Omega$, to compute $\Phi^h(s^h)$ for $s^h \in \mathcal{K}^h$ we use a Firedrake extruded mesh of prisms or quadrilateral, respectively, to discretize $\Lambda_s^h$.}
+\label{fig:extruded}
+\end{figure}
+
+FIXME computing $\Phi^h(s)$
 
@@
 
-As a first step toward solving \eqref{eq:fesigpweakform} we now propose (slow) solvers which will become smoothers in our multilevel method (next section)  The smoothers are the projected nonlinear Gauss-Seidel and projected nonlinear Jacobi (PNJ) methods \cite{KinderlehrerStampacchia1980}.
-Given a current surface elevation $s\in \mathcal{K}^h$, these methods sweep through the nodes in a fixed order, solving one-dimensional VI problems, as follows, and updating $s$.
-For node $p$ let $\omega_p^s = b(x_p,y_p) - s(x_p,y_p)$.
+As a first step toward solving \eqref{eq:fesigpweakform} we propose (slow) solvers which will be smoothers in our multilevel method (next section)  The smoothers are the projected nonlinear Gauss-Seidel and projected nonlinear Jacobi (PNJ) methods \cite{KinderlehrerStampacchia1980}.
+In fact we will describe both smoothers but only implement PNJ.
+
+Given a current surface elevation $s\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems, as follows, and updating $s$.
+Note that, because $\psi_p \in \mathcal{D}_+^h$, if $r \in \mathcal{K}^h$ then $r + c \psi_p \in \mathcal{K}^h$ for all $c\ge b(x_p,y_p) - r(x_p,y_p)$.
+Therefore let $\omega_p^s = b(x_p,y_p) - s(x_p,y_p)$.
 The one-dimensional VI at $p$ implied by \eqref{eq:fesigpweakform} is to find $c \ge \omega_p^s$ such that
@@
 \bigskip
-\bibliography{msc}
+\bibliography{msg}
 \bibliographystyle{siam}

### 2021-05-26 paper/msg.tex
This figure shows how I imagine the extruded mesh will be used.

@@
 For $s^h \in \mathcal{K}^h$ we will compute $\Phi^h(s^h)$, namely the numerical Glen-Stokes \eqref{eq:glenstokesweak} surface velocity associated to the surface elevation $s^h$, using a extruded-mesh approximation of $\Lambda_s^h \subset \RR^{d+1}$, the icy domain as defined by \eqref{eq:lambdas}.
-Specifically, if $d=2$ and if $\mathcal{T}$ is a triangulation of $\Omega \subset \RR^d$, and given $s^h$ and $b^h$ from $\mathcal{V}^h$, we apply the Firedrake concept of an extruded mesh \cite{Gibsonetal2019,McRaeetal2016} to give a prism element decomposition of $\Lambda_s^h$.
+Specifically, if $d=2$ and if $\mathcal{T}$ is a triangulation of $\Omega \subset \RR^d$, and given $s^h$ and $b^h$ from $\mathcal{V}^h$, we apply the Firedrake concept of an extruded mesh \cite{McRaeetal2016} to give a prism element decomposition of $\Lambda_s^h$.
 If $d=1$ then the extruded mesh has quadrilateral elements; a sketch is shown in Figure \ref{fig:extruded}.
@@
 \begin{center}
-FIXME %\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
+\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
 \end{center}
-\caption{Given a triangulation ($d=2$) or interval decomposition ($d=1$ as shown) of $\Omega$, to compute $\Phi^h(s^h)$ for $s^h \in \mathcal{K}^h$ we use a Firedrake extruded mesh of prisms or quadrilateral, respectively, to discretize $\Lambda_s^h$.}
+\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh.
+In the ``pinched'' extrusion (left), $s^h$ defines where the trace of $\bu^h$ is evaluated (strong).
+In the ``cliffs'' extrusion (right) all elements have a minimum thickness $H_{\text{min}} / m_z$, so the trace evaluation may be above the original $s^h$ (dotted).}
 \label{fig:extruded}

### 2021-05-26 paper/genfigs/extruded.py

@@ -1,0 +1,52 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib.pyplot as plt
+from writeout import writeout
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["mathtext.fontset"] = "cm"
+
+x = np.linspace(0.0,10.0,11)
+# following data was generated by stokesdomain.py with print(b[::100]) etc.
+b = np.array([0.5300000000000001, 0.36185948536513635, -0.18136049906158566, -0.1558830996397852, 0.16787164932467638, 0.07119577782212608, 0.4226854163999131, 1.218121471138974, 1.592419336666987, 2.269802550645665, 3.5125890501455257])
+s = np.array([0.5300000000000001, 0.36185948536513635, -0.18136049906158566, 1.4801848778600042, 2.3962987740708668, 2.4711957778221265, 2.6511125411461034, 2.5, 1.592419336666987, 2.269802550645665, 3.5125890501455257])
+
+def isgap(a, b):
+    return abs(a - b) > 1.0e-4
+
+def plotextruded(x, b, s, mz=5, Hmin=0.0):
+    '''Plot the extruded mesh.  mz is the number of element layers.  Hmin > 0
+    produces a cliff at the margin.  Make the top of the top element strong.'''
+    assert all(s >= b)
+    plt.plot(x, b, 'k-')
+    dz = np.maximum(s - b, Hmin) / mz # everywhere at least Hmin/mz
+    # plot verticals
+    for j in range(len(x)):
+        if isgap(b[j],s[j]) or (j > 0 and isgap(b[j-1],s[j-1])) \
+                            or (j < len(x) - 1 and isgap(b[j+1],s[j+1])):
+            snew = b[j] + mz * dz[j]
+            plt.plot([x[j], x[j]], [b[j], snew], 'k-')
+    # plot element tops
+    for j in range(len(x)-1):
+        if isgap(b[j],s[j]) or isgap(b[j+1],s[j+1]):
+            for k in range(mz):
+                zl = b[j] + (k+1) * dz[j]
+                zr = b[j+1] + (k+1) * dz[j+1]
+                if k + 1 == mz:  # show strong top
+                    plt.plot([x[j], x[j+1]], [zl, zr], 'k-', lw=3.0)
+                else:
+                    plt.plot([x[j], x[j+1]], [zl, zr], 'k-')
[…]

### 2021-05-26 paper/genfigs/stokesdomain.py

@@ -8,8 +8,6 @@ plt.rcParams["font.family"] = "serif"
 plt.rcParams["mathtext.fontset"] = "cm"
 
 fsize=18.0
-smallfsize=14.0
-extrasmallfsize=12.0
 bigfsize=20.0
 
 def genbasicfig(xshift=0.0):

### 2021-05-26 paper/msg.tex
Discuss the difference quotient for differentiating Phi.

@@
 
-\newcommand{\ip}[2]{\left<#1,#2\right>}
+\newcommand{\ip}[2]{\left(#1,#2\right)}
 
@@
 \begin{equation}
-F(s)[r] = - \int_\Omega (\Phi(s) + a)\, r \,dx dy. \label{eq:sigpfunctional}
+F(s)[r] = - \ip{\Phi(s)}{r} = - \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
 \end{equation}
-One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the SKE quantity $\Phi(s) + a$ against the test function $r$.
-Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s) + a$ has minimal regularity; it need only be in the dual space of $W^{1,\qq}(\Omega)$.
+One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.
+Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ has minimal regularity; it need only be in the dual space of $W^{1,\qq}(\Omega)$.
 
@@
 \begin{equation}
-F(s)[r - s] \ge 0 \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
+F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
 \end{equation}
@@
 \end{equation}
-where a product-rule computation on \eqref{eq:ido} gives
-\begin{align}
-\Phi'(s)[t] &= \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv} \\
-    &= - \left(\lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} - \bus}{\eps}\right) \cdot \bn_s + \Phi(s) \cdot \left<t_x,t_y,0\right>. \notag
-\end{align}
+where by definition
+\begin{equation}
+\Phi'(s)[t] = \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}. \label{eq:idoderiv}
+\end{equation}
+In \eqref{eq:idoderiv} note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.
+We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.
+Observe that $F'$ is straightforward to compute from $\Phi'$.
 
-Observe that $F'$ is straightforward to compute from $\Phi'$, and we will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.
-In \eqref{eq:idoderiv} note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.
-
-The new functionals $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, and this clearly requires $t\ge 0$ on the (active) set where $s=b$.
+Note $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, and this clearly requires $t\ge 0$ on the (active) set where $s=b$.
 However, to simplify our considerations we define an $s$-independent set
@@
 \end{equation}
-and $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$.
-Thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
+so $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
+
+We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, but only in the situation where $t \in \mathcal{D}_+$ has small support.
+(It will be a finite element basis function; see the next section.)  For this it is relevant to say more about the difference, as it depends on whether one is in the support of $t$ or not:
+\begin{align*}
+\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s &= \left(\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_{s+\eps t}\right) + \left(\bus \cdot \bn_{s+\eps t} - \bus \cdot \bn_s\right) \\
+    &= \begin{cases}
+           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
+           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b \text{ and } t > 0 \\
+           0, & s=b \text{ and } t = 0
+                 \end{cases}
+\end{align*}
+Observe that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$, i.e.~that the dynamical part of the SKE vanishes for small ice masses on bare ground, is a necessary condition for differentiability of $\Phi$.
+It is reasonable to suppose this is so when there is no sliding.
 

### 2021-05-26 paper/msg.tex
Edits on the FE section, and on the smoother section.

@@
 
-FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020}
+FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020};  The computations in this paper use the Python FE library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.
 
@@
 \Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
-                        \bzero, & \text{otherwise}, \end{cases} \label{eq:ido}
+                        0, & \text{otherwise}, \end{cases} \label{eq:ido}
 \end{equation}
@@
 
-Suppose $\mathcal{T}$ is a triangulation of $\Omega$, which we assume is polygonal.
-(If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on their low expected regularity at the ice margin, we will represent the surface elevations $s\in \mathcal{K}$ using the piecewise-linear ($P_1$) finite element (FE) space $\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega)$.
+Assume $\Omega \subset \RR^d$ is polygonal and suppose $\mathcal{T}$ is a triangulation of $\Omega$.
+(If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
+\begin{equation}
+\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
+\end{equation}
 Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.
 
-Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$, and define two closed and convex subsets of $\mathcal{V}^h$:
+Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$.  From $b^h$ we define the closed and convex admissible subset of $\mathcal{V}^h$:
 \begin{equation}
-\mathcal{K}^h = \{r \in \mathcal{V}^h \,:\, r \ge b^h \text{ and } r|_{\partial\Omega} = b^h|_{\partial\Omega}\}, \qquad \mathcal{D}_+^h = \{t \in \mathcal{V}^h \,:\, t \ge 0\}. \label{eq:feKD}
+\mathcal{K}^h = \{r^h \in \mathcal{V}^h \,:\, r^h \ge b^h \text{ and } r^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
 \end{equation}
-Let $(x_j,y_j)$ denote the coordinates an arbitrary node of the mesh $\mathcal{T}$, and $\psi_j \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.
+For $s^h\in \mathcal{K}^h$ we also define the numerical icy domain as in \eqref{eq:lambdas}:
+\begin{equation}
+\Lambda_{s^h} = \{(x,y,z)\,:\,b^h(x,y) < z < s^h(x,y)\} \subset \RR^{d+1}.  \label{eq:felambdas}
+\end{equation}
+Also, a triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in the $s^h$ geometry if $s^h=b^h$ at every vertex of $T$, and icy otherwise.
 
-With these FE choices, the discrete SIGP computes the FE approximation $s^h \in \mathcal{K}^h$ satisfying
-\begin{equation}
-F^h(s^h)[r - s^h] \ge 0 \quad \text{for all } r \in \mathcal{K}^h , \label{eq:fesigpweakform}
-\end{equation}
-where $F^h(s^h)[r] = - \int_\Omega (\Phi^h(s^h) + a)\, r \,dx dy$; compare \eqref{eq:sigpfunctional}.
-Here $\Phi^h$ is an FE approximation of the IDO $\Phi$; see below.
-We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.
-
-All FE computations in this paper use the Python library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.
-For $s^h \in \mathcal{K}^h$ we will compute $\Phi^h(s^h)$, namely the numerical Glen-Stokes \eqref{eq:glenstokesweak} surface velocity associated to the surface elevation $s^h$, using a extruded-mesh approximation of $\Lambda_s^h \subset \RR^{d+1}$, the icy domain as defined by \eqref{eq:lambdas}.
-Specifically, if $d=2$ and if $\mathcal{T}$ is a triangulation of $\Omega \subset \RR^d$, and given $s^h$ and $b^h$ from $\mathcal{V}^h$, we apply the Firedrake concept of an extruded mesh \cite{McRaeetal2016} to give a prism element decomposition of $\Lambda_s^h$.
-If $d=1$ then the extruded mesh has quadrilateral elements; a sketch is shown in Figure \ref{fig:extruded}.
+Given $s^h$ and $b^h$ from $\mathcal{V}^h$, we construct a Firedrake extruded mesh \cite{McRaeetal2016} to decompose $\Lambda_{s^h}$.
+If $d=2$ then the extruded mesh consists of triangular prisms, while if $d=1$ it has quadrilateral elements.
+The extruded mesh construction starts from the triangulation $\mathcal{T}$ of $\Omega \subset \RR^d$, now called the ``base mesh''.
+As shown in Figure \ref{fig:extruded}, each icy $T$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.
+However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.
 
@@
 \end{center}
-\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh.
-In the ``pinched'' extrusion (left), $s^h$ defines where the trace of $\bu^h$ is evaluated (strong).
-In the ``cliffs'' extrusion (right) all elements have a minimum thickness $H_{\text{min}} / m_z$, so the trace evaluation may be above the original $s^h$ (dotted).}
+\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.
+In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold); see \eqref{eq:ido}.
+In the ``cliffs'' extrusion (right) all elements have a minimum thickness and the trace evaluation (bold) may occur above $s^h$ (dotted).}
 \label{fig:extruded}
@@
 
-FIXME computing $\Phi^h(s)$
+We will compare two extrusion modes.
+In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly, prism (quadilateral) elements at the ice margin degenerate; such elements have empty facets.
+In the alternate ``cliffs'' extrusion, each element over an icy $T$ is nondegenerate.
+Specifically, if $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in experiments, then each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$.
+In this extrusion, a stress-free condition is applied on the cliffs  in the Glen-Stokes problem \eqref{eq:glenstokesweak}; they are treated the same way as the top surface.
+
+Let $\Phi^h:\mathcal{K}^h \to (\mathcal{V}^h)'$ be the FE approximation of the IDO $\Phi$.
+By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.
+That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, but on $\overline{\partial} \Lambda_{s^h}$, and then extend by zero to all of $\Omega$.
+In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.
+
+We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}, and a Newton iteration, with direct solution of the step equations and back-tracking line search, to resolve the power-law nonlinearity.
+As stated in Section \ref{sec:intro}, note that our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme (next section) are needed to solve the SIGP \eqref{eq:sigpweakform}.
+Overall optimality of the solution algorithm would also need an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
+
+Finally, the discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
+\begin{equation}
+F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h , \label{eq:fesigpweakform}
+\end{equation}
+where $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.
+We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.
 
@@
 
-As a first step toward solving \eqref{eq:fesigpweakform} we propose (slow) solvers which will be smoothers in our multilevel method (next section)  The smoothers are the projected nonlinear Gauss-Seidel and projected nonlinear Jacobi (PNJ) methods \cite{KinderlehrerStampacchia1980}.
-In fact we will describe both smoothers but only implement PNJ.
+In this section we propose solvers for \eqref{eq:fesigpweakform}.
+While they are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
+We will describe both projected nonlinear Gauss-Seidel and Jacobi methods \cite{KinderlehrerStampacchia1980}, but only implement the latter.
 
-Given a current surface elevation $s\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems, as follows, and updating $s$.
-Note that, because $\psi_p \in \mathcal{D}_+^h$, if $r \in \mathcal{K}^h$ then $r + c \psi_p \in \mathcal{K}^h$ for all $c\ge b(x_p,y_p) - r(x_p,y_p)$.
-Therefore let $\omega_p^s = b(x_p,y_p) - s(x_p,y_p)$.
-The one-dimensional VI at $p$ implied by \eqref{eq:fesigpweakform} is to find $c \ge \omega_p^s$ such that
+To describe these methods suppose $(x_i,y_i)$ denotes an arbitrary vertex of $\mathcal{T}$, and $\psi_i \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.
+Also let
 \begin{equation}
-F^h(s+c\psi_p)[(s+\gamma \psi_p) - (s+c \psi_p)] \ge 0 \label{eq:fepointwiseviEARLY}
+\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}.  \label{eq:feDplus}
 \end{equation}
-for all $\gamma \ge \omega_p^s$.  Defining $F_p^h(c) = F^h(s+c\psi_p)[\psi_p]$, \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
+Note that, because $\psi_i \in \mathcal{D}_+^h$, if $r^h \in \mathcal{K}^h$ then $r + c \psi_i \in \mathcal{K}^h$ for all $c\ge \omega_i^s$ where $\omega_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+
+Given a current surface elevation $s^h\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
+The one-dimensional VI at node $i$, derived from \eqref{eq:fesigpweakform}, finds $c \ge \omega_i^{s^h}$ such that
 \begin{equation}
-(\gamma - c) F_p^h(c) \ge 0. \label{eq:fepointwisevi}
+F^h(s^h+c\psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
 \end{equation}
-Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s$ is then updated $s \gets s + c \psi_p$ by the smoother.
-(Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.)  Each pointwise update is done immediately by the (serial) PNGS smoother, while the (parallel) PNJ smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $p$ before the updates.
+for all $\gamma \ge \omega_i^{s^h}$.  Let
+\begin{equation}
+\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i}, \label{eq:ferhoi}
+\end{equation}
+a point-wise residual function.  Then \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
+\begin{equation}
+(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
+\end{equation}
+Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
+Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.
 
-If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.
+Each pointwise update is done immediately by the (serial) Gauss-Seidel smoother, while the (parallel) Jacobi smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $i$ before then doing all the updates.
+In linear problems it is known that Gauss-Seidel generally converges in about half as many iterations \cite{Greenbaum1997}, but here there is a more glaring difference between the two iterations.
+Gauss-Seidel would be an extraordinarily expensive approach because a new Glen-Stokes solution would be required at every node of $\mathcal{T}$.
+We will prefer the Jacobi strategy; see below.
+
+FIXME If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.
 The simplest approximation scheme is thus to linearize $F_p^h(c)$, that is, smooth using one projected Newton step for the one-dimensional VI problem at each node.

### 2021-05-27 paper/msg.tex
Minor: fiddle with glossary.

@@
 \renewcommand{\arraystretch}{1.1}
-\begin{longtable}{l|ll}
+\begin{longtable}{l|l|l}
 \toprule
@@
 FE & finite element & Section \ref{sec:fe} \\
-IDO & ice dynamics operator & Section \ref{sec:weakido}, eqn.~\eqref{eq:ido} \\
+IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\
 IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
-MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
+MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes}, Pseudocodes \ref{pc:mcdn-vcycle} and \ref{pc:mcdn-solver} \\
 NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
-PNJ & projected, nonlinear Jacobi & Section \ref{sec:smoothers} \\
+PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\
 SIA & shallow ice approximation & Section \ref{sec:intro} \\
-SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, eqn.~\eqref{eq:strongform} \\
-SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, eqn.~\eqref{eq:ske} \\
+SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, equation \eqref{eq:strongform} \\
+SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, equation \eqref{eq:ske} \\
 VI & variational inequality & Section \ref{sec:weakido} \\

### 2021-05-27 paper/msg.tex
Edits on smoother text.

@@
 We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, but only in the situation where $t \in \mathcal{D}_+$ has small support.
-(It will be a finite element basis function; see the next section.)  For this it is relevant to say more about the difference, as it depends on whether one is in the support of $t$ or not:
-\begin{align*}
-\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s &= \left(\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_{s+\eps t}\right) + \left(\bus \cdot \bn_{s+\eps t} - \bus \cdot \bn_s\right) \\
+(It will be a finite element basis function; see the next section.)  For this it is relevant to say more about the difference, as it depends on whether the point is in the support of $t$ or not:
+\begin{align}
+\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s &= \left(\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_{s+\eps t}\right) + \left(\bus \cdot \bn_{s+\eps t} - \bus \cdot \bn_s\right) \label{eq:differencecases} \\
     &= \begin{cases}
@@
            0, & s=b \text{ and } t = 0
-                 \end{cases}
-\end{align*}
-Observe that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$, i.e.~that the dynamical part of the SKE vanishes for small ice masses on bare ground, is a necessary condition for differentiability of $\Phi$.
-It is reasonable to suppose this is so when there is no sliding.
+                 \end{cases} \notag
+\end{align}
+Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$, i.e.~the dynamical part of the SKE vanishes for small ice masses on bare ground.
+It is reasonable to suppose this is so when such small ice masses do not slide.
 
@@
 In this section we propose solvers for \eqref{eq:fesigpweakform}.
-While they are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
-We will describe both projected nonlinear Gauss-Seidel and Jacobi methods \cite{KinderlehrerStampacchia1980}, but only implement the latter.
+We describe certain projected and nonlinear iterations, variants of the Gauss-Seidel and Jacobi methods, which are suitable for VIs \cite{KinderlehrerStampacchia1980}.
+While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
 
-To describe these methods suppose $(x_i,y_i)$ denotes an arbitrary vertex of $\mathcal{T}$, and $\psi_i \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.
-Also let
+Given a current surface elevation $s^h\in \mathcal{K}^h$, these smoothers sweep through the vertices ($P_1$ nodes) in $\mathcal{T}$, in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
+Denote an arbitrary vertex of $\mathcal{T}$ by $(x_i,y_i)$ and let $\psi_i \in \mathcal{V}^h$ be the corresponding hat function, with $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.
+Also let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}$.
+Note that, because $\psi_i \in \mathcal{D}_+^h$, $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \omega_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+
+The one-dimensional VI at node $i$ is derived from \eqref{eq:fesigpweakform}.  It finds $c \ge \omega_i^{s^h}$ such that
 \begin{equation}
-\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}.  \label{eq:feDplus}
+F^h(s_c)[r_\gamma - s_c] \ge \ip{a}{r_\gamma - s_c} \label{eq:fepointwiseviEARLY}
 \end{equation}
-Note that, because $\psi_i \in \mathcal{D}_+^h$, if $r^h \in \mathcal{K}^h$ then $r + c \psi_i \in \mathcal{K}^h$ for all $c\ge \omega_i^s$ where $\omega_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
-
-Given a current surface elevation $s^h\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
-The one-dimensional VI at node $i$, derived from \eqref{eq:fesigpweakform}, finds $c \ge \omega_i^{s^h}$ such that
+for all $\gamma \ge \omega_i^{s^h}$, with $r_\gamma = s^h+\gamma \psi_i$ and $s_c = s^h+c \psi_i$.  Note \eqref{eq:fepointwiseviEARLY} simplifies substantially.  Let
 \begin{equation}
-F^h(s^h+c\psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
+\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i} \label{eq:ferhoi}
 \end{equation}
-for all $\gamma \ge \omega_i^{s^h}$.  Let
-\begin{equation}
-\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i}, \label{eq:ferhoi}
-\end{equation}
-a point-wise residual function.  Then \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
+be the point-wise residual function.  Then \eqref{eq:fepointwiseviEARLY} is equivalent
 \begin{equation}
@@
 \end{equation}
-Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
-Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.
+Supposing \eqref{eq:fepointwisevi} is well-posed, after solving it for $c$ the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
+Observe that admissibility is preserved because the updated surface elevation is also in $\mathcal{K}^h$.
 
-Each pointwise update is done immediately by the (serial) Gauss-Seidel smoother, while the (parallel) Jacobi smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $i$ before then doing all the updates.
+If each pointwise update is done immediately then the smoother is of Gauss-Seidel (GS) type, sometimes called a serial or multiplicative iteration.
+By contrast, the (parallel, additive) Jacobi smoother computes solutions at all nodes $i$ before doing all the updates.
 In linear problems it is known that Gauss-Seidel generally converges in about half as many iterations \cite{Greenbaum1997}, but here there is a more glaring difference between the two iterations.
-Gauss-Seidel would be an extraordinarily expensive approach because a new Glen-Stokes solution would be required at every node of $\mathcal{T}$.
-We will prefer the Jacobi strategy; see below.
+The GS iteration would be extraordinarily expensive because a new Glen-Stokes solution would be required at every node of $\mathcal{T}$, so we will only implement the Jacobi version.
 
-FIXME If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.
-The simplest approximation scheme is thus to linearize $F_p^h(c)$, that is, smooth using one projected Newton step for the one-dimensional VI problem at each node.
-To implement this strategy, note that \eqref{eq:sigpfunctionalderiv} and \eqref{eq:idoderiv} imply
+If $\rho_i(s^h; c)$ were linear and increasing in $c$ then the solution to \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_i^{s^h}\right\}$ if $\rho_i(s^h; c) = \alpha c + \beta$ with $\alpha > 0$.
+In fact $\rho_i(s^h; c)$ is nonlinear, but the simplest approximation scheme is to linearize $\rho_i(s^h; c)$, that is, to do one projected Newton step for the one-dimensional VI problem \eqref{eq:fepointwisevi} at each node.
+
+To implement this strategy we approximate
 \begin{equation}
-F_p^h(c) = F^h(s+c\psi_p)[\psi_p] \approx F^h(s)[\psi_p] + c\, (F^h)'(s,\psi_p)[\psi_p] = \alpha c + \beta \label{eq:fepointwisefunctionalapprox}
+\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, (F^h)'(s^h)[\psi_i,\psi_i]. \label{eq:rhoapprox}
 \end{equation}
-where
+If we partition the support of $\psi_i$ into where there is ice and not,
+\begin{equation}
+\theta_i = \{\psi_i > 0\} \cap \{s^h > b^h\}, \qquad {\hat\theta}_i = \{\psi_i > 0\} \setminus \theta_i,  \label{eq:thetasupport}
+\end{equation}
+then by \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute with local integrals
+\begin{equation}
+\rho_i(s^h; 0) = F^h(s^h)[\psi_i] - \ip{a}{\psi_i} = - \int_{\theta_i} \bu|_{s^h} \cdot \bn_{s^h}\, \psi_i\,dx\,dy - \ip{a}{\psi_i}  \label{eq:rhozero}
+\end{equation}
+and
 \begin{align}
-\alpha &= \int_\Omega \left(- \left(\lim_{\eps\to 0^+} \frac{\bu^h|_{s+\eps \psi_p} - \bu^h|_s}{\eps}\right) \cdot \bn_s + \Phi^h(s) \cdot \left<(\psi_p)_x,(\psi_p)_y,0\right>\right)\, \psi_p \,dx dy \label{eq:fepointwisealpha} \\
-\beta &= - \int_\Omega (\Phi^h(s) + a)\, \psi_p \,dx dy \label{eq:fepointwisebeta}
+(F^h)'(s^h)[\psi_i,\psi_i] &= - \int_{\theta_i} \lim_{\eps\to 0^+} \frac{(\bu|_{s^h+\eps\psi_i} - \bu|_{s^h}) \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i\,dx\,dy \label{eq:rholocalderiv} \\
+  &\qquad - \int_{{\hat\theta}_i} \lim_{\eps\to 0^+} \frac{\bu|_{s^h+\eps\psi_i} \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i\,dx\,dy.  \notag
 \end{align}
-Each smoother application will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation.
-In fact, supposing we have computed $\bu^h|_s$ and $\Phi^h(s)$ then several terms in these formulas are straightforward FE computations over the support of $\psi_p$.
 
-However, if we wish to compute $\alpha$ in \eqref{eq:fepointwisealpha}, even approximately, then it remains to compute a difference quotient in \eqref{eq:fepointwisealpha}, which in turn requires evaluating a perturbed-geometry surface velocity $\bu^h|_{s+\eps \psi_p}$.
-The expensive direct way of doing this is to solve a Glen-Stokes problem for the perturbed upper-surface geometry $z=s + \eps \psi_p$.
+If $\alpha=(F^h)'(s^h)[\psi_i,\psi_i]$ is positive then the solution of \eqref{eq:fepointwisevi} is
+\begin{equation}
+  c = \max\left\{-\frac{\rho_i(s^h; 0)}{\alpha}, \omega_i^{s^h}\right\}, \label{eq:fepointwisevisolution}
+\end{equation}
+which generalizes a well-known formula for the classical obstacle problem \cite[formula (4.4), for example]{GraeserKornhuber2009}.
+However, if $\alpha\le 0$ then the model has degenerated in the sense that the linearized operator is not acting elliptically with respect to that perturbation.
+In that case we make no change to $s^h$; we set $c=0$.
+While one can prove in the shallow, flat-bed model that the analogous $\alpha$ alpha quantity is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, which follows since the equation is elliptic, the full Glen-Stokes model allows no such guarantee to our knowledge.
 
-Thus we seek a less expensive way to estimate the perturbed velocity field for a small ``bump'' $\eps\psi_p$ added to the surface elevation $s$.
-For small $\eps$ the nonlinearities of the Glen-Stokes problem should be modest, but furthermore the most important change to the velocity and pressure fields from this bump would be through the perturbed gravitational load of the ice in the bump.
-Thus FIXME
+We will approximate the limits in \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} by using the obvious difference quotients, i.e.~by removing ``$\lim_{\eps\to 0^+}$'' and choosing $\eps$ to be small, and we set $\eps$ to one meter in computations.
+However, we need to make further approximations to the computation of the divisor $\alpha$, i.e.~in \eqref{eq:rholocalderiv}, if we want a practical solver.
+
+Note that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
+Thus $\bu|_{s^h}$ and $\bu|_{s^h+\eps\psi_i}$ are the results of separate Glen-Stokes solves, over $\Lambda_{s^h}$ and $\Lambda_{s^h+\eps\psi_i}$, respectively.
+Clearly, each smoother application, a sweep over all nodes in $\mathcal{T}$, will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak}, namely solving with the geometry determined by the current surface elevation $s^h$.
+However, formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated are prohibitively expensive; a Stokes solve is required  for each mesh node.
+
+Thus we seek a less expensive way to estimate the surface velocity for a perturbed surface elevation with a small ``bump'' $\eps\psi_i$.
+The nonlinearities in this Glen-Stokes problem should be unimportant, and it would seem that the most important change to the velocity and pressure fields from a small bump perturbation is through the perturbed gravitational load applied to the linearized problem around the state $s^h$.
+(Versus the additional effect of a perturbed normal direction on the surface, for example.)
+
+In fact, let us fix a current surface elevation $s^h$ and assume that problem \eqref{eq:glenstokesweak} for $\Lambda_{s^h}$ yields solution $\bu^h,p^h$ at the convergence of its (Newton) iteration.
+Noting that we are using a stable mixed space for this Glen-Stokes problem (Section \ref{sec:fe}), the final step in the iteration can be regarded as providing the solution of a linear system
+\begin{equation}
+    \begin{bmatrix} A & B^\top \\
+                    B & 0      \end{bmatrix}
+    \begin{bmatrix} \bu^h \\ p^h \end{bmatrix}
+    = \begin{bmatrix} \rhoi \bg \\ 0 \end{bmatrix}.  \label{eq:system}
+\end{equation}
+where we note that the matrix depends on $s^h$.
+(Even the size of $K$ depends on $s^h$ because it is used to determine which $T\in\mathcal{T}$ get extruded icy columns.)  Also denote the perturbed problem, for surface $s^h + \eps \psi_i$, using an $\eps$ subscript.
+
+We assert that the perturbed-geometry Glen-Stokes solution is approximated by re-using the unperturbed matrix but adding the mass of the bump $\eps \psi_i$ to the right-hand side:
+\begin{equation}
+\begin{bmatrix} A_\eps & B_\eps^\top \\
+                B_\eps & 0      \end{bmatrix}
+\begin{bmatrix} \bu_\eps^h \\ p_\eps^h \end{bmatrix}
+    = \begin{bmatrix} \rhoi \bg \\ 0 \end{bmatrix}
+\qquad \approx \qquad
+\begin{bmatrix} A & B^\top \\
+                B & 0      \end{bmatrix}
+\begin{bmatrix} \bu_\eps^h \\ p_\eps^h \end{bmatrix}
+    = \begin{bmatrix}  \rhoi (1 + \eps \tau_i)\bg  \\ 0 \end{bmatrix}.  \label{eq:systemmassanalogy}
+\end{equation}
+Here $\tau_i$ is a positive $P_1$ function on $\Lambda_{s^h}$, that is, on the extruded $d+1$-dimensional mesh, with support only at the top-surface node with base mesh index $i$, and with magnitude such that if $\eps=1$ meter then
+\begin{equation}
+  \int_{\Lambda_{s^h}} \tau_i\,dx\,dy\,dz = \int_\Omega \psi_i\,dx\,dy . \label{eq:massanalogy}
+\end{equation}
+In words, we approximate the Glen-Stokes solution $\bu_\eps^h,p_\eps^h$ for the perturbed geometry by re-using the same (i.e.~$s^h$) geometry but solving the final linear system using a right-hand-side perturbation of the equivalent mass to the geometry perturbation.
+This is shown in Figure \ref{fig:massanalogy}.
+
+\begin{figure}[t]
+\begin{center}
+FIXME %\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
+\end{center}
+\caption{A sketch of the mass-perturbation approximation in equations \eqref{eq:systemmassanalogy} and \eqref{eq:massanalogy}.}
+\label{fig:massanalogy}
+\end{figure}
 

### 2021-05-28 py/smoother.py
Maintenance and clean up.

@@ -73,6 +73,9 @@ class SmootherStokes(SmootherObstacleProblem):
 
     def savestate(self, mesh, u, p, kres):
         ''' Save state and diagnostics into .pvd file.'''
+        assert self.saveflag == True
+        assert self.savename is not None
+        assert len(self.savename) > 0
         nu, tau = self.stresses(mesh, u)
         u *= secpera
         u.rename('velocity (m a-1)')
@@ -126,10 +129,12 @@ class SmootherStokes(SmootherObstacleProblem):
 
     def residual(self, mesh1d, s, ella):
         '''Compute the residual functional, namely the surface kinematical
-        residual for the entire domain, for a given iterate s.  Note mesh1D is
+        residual for the entire domain, for a given iterate s.  In symbols
+        matching the paper, returns  r = F(s)[.] - ella(.)  where . ranges over
+        hat functions on mesh1d.  Note mesh1D is
         a MeshLevel1D instance and ella(.) = <a(x),.> is a source term in V^j'.
-        This residual evaluation requires setting up an (x,z) Firedrake mesh,
-        starting from a (stored) base mesh and then using extrusion.  The icy
+        This residual evaluation requires setting up an (x,z) Firedrake mesh
+        by extrusion of a (stored) base mesh.  The icy
         columns get their height from s, with minimum height args.Hmin.  By
         default the extruded mesh has empty (0-element) columns if ice-free
         according to s.  If args.padding==True then the whole extruded mesh has
@@ -150,7 +155,8 @@ class SmootherStokes(SmootherObstacleProblem):
         # extrude the mesh, to temporary total height 1.0
         mz = self.args.mz
         if self.args.padding:
-            assert self.args.Hmin > 0.0, 'padding requires positive thickness'
+            assert self.args.Hmin > 0.0, \
+                'padding requires minimum positive thickness'
             mesh = fd.ExtrudedMesh(self.basemesh, mz, layer_height=1.0 / mz)
             if firstcall:
                 print('            extruded mesh: padded, %d x %d elements' \
@@ -166,7 +172,7 @@ class SmootherStokes(SmootherObstacleProblem):
                                    layer_height=1.0 / mz)
             if firstcall:
                 icycount = sum(icyelement)
-                print('            extruded mesh: %d x %d icy and %d ice-free elements' \
[…]

### 2021-05-28 py/steady.py

@@ -27,15 +27,14 @@ Note s solves the surface kinematical equation, as an interior PDE, in the
 inactive set {x | s(x) > b(x)}.
 
 Solution is by the nonlinear (FAS) extension of the multilevel constraint
-decomposition (MCD) method of Tai (2003).  See the simpler cases, including
-the SIA, in part I (Bueler, 2022).
+decomposition (MCD) method of Tai (2003), or by sweeps of the smoother.
 
 Initial implementation generates Bueler profile geometry as initial state
 and then tries to converge from there.
 
 References:
-  * Bueler, E. (2022). Geometric multigrid for glacier modeling I:
-    New concepts and algorithms.  In preparation.
+  * Bueler, E. and Mitchell, L. (2022). Multilevel computation of glacier
+    geometry from Stokes dynamics.  In preparation.
   * Tai, X.-C. (2003). Rate of convergence for some constraint
     decomposition methods for nonlinear variational inequalities.
     Numer. Math. 93 (4), 755--786.
@@ -44,46 +43,58 @@ References:
     allow_abbrev=False,  # bug in python 3.8 causes this to be ignored
     add_help=False)
 adda = parser.add_argument
-adda('-alpha', type=float, metavar='X', default=1000.0,  # FIXME sensitive
-    help='Richardson iteration parameter (default=1000.0)')
 adda('-coarse', type=int, default=1, metavar='N',
-     help='smoother sweeps on coarsest grid (default=1)')
+     help='smoother sweeps on coarsest grid (default=%(default)s)')
 adda('-cyclemax', type=int, default=100, metavar='N',
-     help='maximum number of (multilevel) cycles (default=100)')
+     help='maximum number of (multilevel) cycles (default=%(default)s)')
 adda('-domainlength', type=float, default=30.0e3, metavar='L',
-     help='solve on [0,L] (default L=30 km)')
+     help='solve on [0,L] (default=%(default)s m)')
 adda('-down', type=int, default=0, metavar='N',
-     help='smoother sweeps before coarse-mesh correction (default=0)')
+     help='smoother sweeps before coarse-mesh correction (default=%(default)s)')
 adda('-eps', type=float, metavar='X', default=1.0e-2,  # FIXME sensitive
-    help='regularization used in viscosity (default=10^{-2})')
+    help='regularization used in viscosity (default=%(default)s)')
 adda('-Hmin', type=float, metavar='X', default=0.0,
[…]

### 2021-05-28 py/steady.py
Cannot check for unknown options this way ... need s_... solver control to pass through.

@@ -86,12 +86,6 @@ args, unknown = parser.parse_known_args()
 if args.steadyhelp:
     parser.print_help()
     sys.exit(0)
-# allow -help for PETSc help, but otherwise fail on unknown options
-if unknown:
-    if len(unknown) > 1 or unknown[0] != '-help':
-        print('ERROR: unknown option')
-        parser.print_help()
-        sys.exit(0)
 
 if args.padding:
     assert args.Hmin > 0.0, 'padding requires minimum positive thickness'

### 2021-05-28 py/smoother.py
Implement the slow Jacobi smoother.

@@ -18,8 +18,9 @@ def D(w):
     return 0.5 * (fd.grad(w) + fd.grad(w).T)
 
 class SmootherStokes(SmootherObstacleProblem):
-    '''To evaluate the residual this Jacobi smoother solves the Stokes problem
-    for the given geometry by creating an extruded mesh.'''
+    '''Smoother for solving the steady-geometry Stokes problem.  Generates an
+    extruded mesh for each residual evaluation.  Implements projected
+    nonlinear versions of the Richardson and Jacobi smoothers.'''
 
     def __init__(self, args, admissibleeps=1.0e-10):
         super().__init__(args, admissibleeps=admissibleeps)
@@ -237,23 +238,53 @@ class SmootherStokes(SmootherObstacleProblem):
         if currentr is None:
             currentr = self.residual(mesh1d, s, ella)
         if self.args.smoother == 'richardson':
-            self.richardsonsweep(s, currentr, phi)
+            self.richardsonsweep(s, phi, currentr)
         elif self.args.smoother == 'jacobislow':
-            raise NotImplementedError('-smoother jacobislow not implemented')
+            self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
         mesh1d.WU += 1
         return self.residual(mesh1d, s, ella)
 
-    def richardsonsweep(self, s, r, phi):
+    def richardsonsweep(self, s, phi, r):
         '''Do in-place projected nonlinear Richardson smoothing on s(x):
-            s <- max(s - omega * r, phi)'''
+            s <- max(s - omega * r, phi)
+        User must adjust omega to reasonable level.  (Do this with SIA-type
+        stability criterion argument.)'''
         np.maximum(s - self.args.omega * r, phi, s)
 
-    def jacobislowsweep(self, s, r, phi):
-        '''Do in-place projected nonlinear Jacobi smoothing on s(x):
-            s_i <- max(s_i - omega * r_i / d_i, phi_i)
+    def jacobislowsweep(self, mesh1d, s, ella, phi, r,
+                        eps=1.0, info=False, dump=False):
+        '''Do in-place projected nonlinear Jacobi smoothing on s(x)
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
-        by VERY SLOW finite differencing of residual calculations.'''
-        raise NotImplementedError('-smoother jacobislow not implemented')
+        by VERY SLOW finite differencing of expensive residual calculations.
+        If d_i > 0 then
[…]

### 2021-05-28 py/steady.py
Regression test for jacobislow.

@@ -2,9 +2,14 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   1. implement Jacobi smoother by SLOW method, getting derivative of SKE
-#      w.r.t. a surface displacement by finite difference
+#   1. implement gsslow
 #   2. copy partI/mcdn.py and build it out
+
+# RUNS:
+#   1. actually converges
+#      ./steady.py -sweepsonly -J JJ -smoother jacobislow -Hmin HH
+#      with JJ=1,2,3,4 and HH=0,10,100;  HH=0 seems as good as any
+#      also trying on JJ=5,6 and HH=0
 
 import sys
 import argparse
@@ -67,7 +72,7 @@ adda('-o', metavar='FILEROOT', type=str, default='',
      help='save .pvd and final image in .png to FILEROOT.*')
 adda('-oimage', metavar='FILE', type=str, default='',
      help='save final image, e.g. .pdf or .png')
-adda('-omega', type=float, metavar='X', default=1000.0,  # FIXME sensitive
+adda('-omega', type=float, metavar='X', default=1.0,  # FIXME sensitive
     help='scale by this factor in smoother iteration (default=%(default)s)')
 adda('-padding', action='store_true', default=False,
      help='put Hmin thickness of ice in ice-free locations')

### 2021-05-28 paper/msg.tex
Minor reword.

@@
 In that case we make no change to $s^h$; we set $c=0$.
-While one can prove in the shallow, flat-bed model that the analogous $\alpha$ alpha quantity is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, which follows since the equation is elliptic, the full Glen-Stokes model allows no such guarantee to our knowledge.
+Because the interior PDE in the shallow, flat-bed model is elliptic, one can prove that the analogous $\alpha$ is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, but the full Glen-Stokes model here allows no such guarantee to our knowledge.
 

### 2021-05-28 py/smoother.py
Implement slow version of GS.  Reduces iterations over jacobislow, but a bit slower in fact.

@@ -241,6 +241,8 @@ class SmootherStokes(SmootherObstacleProblem):
             self.richardsonsweep(s, phi, currentr)
         elif self.args.smoother == 'jacobislow':
             self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
+        elif self.args.smoother == 'gsslow':
+            self.gsslowsweep(mesh1d, s, ella, phi, currentr)
         mesh1d.WU += 1
         return self.residual(mesh1d, s, ella)
 
@@ -252,7 +254,7 @@ class SmootherStokes(SmootherObstacleProblem):
         np.maximum(s - self.args.omega * r, phi, s)
 
     def jacobislowsweep(self, mesh1d, s, ella, phi, r,
-                        eps=1.0, info=False, dump=False):
+                        eps=1.0, dump=False):
         '''Do in-place projected nonlinear Jacobi smoothing on s(x)
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
         by VERY SLOW finite differencing of expensive residual calculations.
@@ -264,8 +266,6 @@ class SmootherStokes(SmootherObstacleProblem):
         snew = s.copy()
         negd = []
         for j in range(1, len(s)-1): # loop over interior points
-            if info:
-                print('    perturb at j=%d ... ' % j, end='')
             sperturb = s.copy()
             sperturb[j] += eps
             if dump:
@@ -274,17 +274,39 @@ class SmootherStokes(SmootherObstacleProblem):
             d = (rperturb[j] - r[j]) / eps
             if d > 0.0:
                 snew[j] = max(s[j] - self.args.omega * r[j] / d, phi[j])
-                if info:
-                    print('d[%d] = %.4e > 0.0' % (j,d))
             else:
                 snew[j] = phi[j]
                 negd.append(j)
-                if info:
-                    print('d[%d] = %.4e <= 0.0 ... setting s[j] = phi[j]' \
-                          % (j,d))
         print('    jacobislow negd = ', end='')
         print(negd)
         s[:] = snew[:] # in-place copy
+
+    def gsslowsweep(self, mesh1d, s, ella, phi, r,
+                    eps=1.0, dump=False):
+        '''Do in-place projected nonlinear Gauss-Seidel smoothing on s(x)
+        where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
+        by VERY SLOW finite differencing of expensive residual calculations.
+        If d_i > 0 then
[…]

### 2021-05-28 py/steady.py

@@ -78,7 +78,7 @@ adda('-padding', action='store_true', default=False,
      help='put Hmin thickness of ice in ice-free locations')
 adda('-printwarnings', action='store_true', default=False,
      help='print pointwise feasibility warnings')
-adda('-smoother', choices=['richardson', 'jacobislow'],
+adda('-smoother', choices=['richardson', 'jacobislow', 'gsslow'],
      metavar='X', default='richardson',
      help='smoother (default=%(default)s)')
 adda('-steadyhelp', action='store_true', default=False,

### 2021-05-28 py/steady.py
Regression test for GS.

@@ -2,14 +2,14 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   1. implement gsslow
-#   2. copy partI/mcdn.py and build it out
+#   1. copy partI/mcdn.py and build it out
 
 # RUNS:
 #   1. actually converges
 #      ./steady.py -sweepsonly -J JJ -smoother jacobislow -Hmin HH
 #      with JJ=1,2,3,4 and HH=0,10,100;  HH=0 seems as good as any
-#      also trying on JJ=5,6 and HH=0
+#      also JJ=5
+#      also works with -smoother gsslow
 
 import sys
 import argparse

### 2021-05-29 paper/msg.tex
Edits on smoother text.  Added PNGS_SLOW Pseudocode.

@@
 \caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$; the thin line is $s+\Phi(s)$.
-Both input and output are defined on all of $\Omega$.}
+Both the input and output of $\Phi$ are defined on all of $\Omega$.}
 \label{fig:idoaction}
@@
 
-In this section we propose solvers for \eqref{eq:fesigpweakform}.
-We describe certain projected and nonlinear iterations, variants of the Gauss-Seidel and Jacobi methods, which are suitable for VIs \cite{KinderlehrerStampacchia1980}.
+In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations, variants of the Gauss-Seidel and Jacobi methods, which are suitable for VIs \cite{KinderlehrerStampacchia1980}.
 While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
+However, because of the non-locality of the SIGP functional $F^h$, we must make careful approximations in order to retain adequate per-iteration performance.
 
-Given a current surface elevation $s^h\in \mathcal{K}^h$, these smoothers sweep through the vertices ($P_1$ nodes) in $\mathcal{T}$, in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
+Given a current surface elevation $s^h\in \mathcal{K}^h$, our smoothers will sweep through the $m$ vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
 Denote an arbitrary vertex of $\mathcal{T}$ by $(x_i,y_i)$ and let $\psi_i \in \mathcal{V}^h$ be the corresponding hat function, with $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.
 Also let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}$.
-Note that, because $\psi_i \in \mathcal{D}_+^h$, $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \omega_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+Note that $\psi_i \in \mathcal{D}_+^h$ and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
 
-The one-dimensional VI at node $i$ is derived from \eqref{eq:fesigpweakform}.  It finds $c \ge \omega_i^{s^h}$ such that
+The one-dimensional VI at node $i$ is derived from \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for scalars $\gamma,c$.
+Specifically, we seek $c \ge \beta_i^{s^h}$ such that
 \begin{equation}
-F^h(s_c)[r_\gamma - s_c] \ge \ip{a}{r_\gamma - s_c} \label{eq:fepointwiseviEARLY}
+F^h(s_c)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
 \end{equation}
-for all $\gamma \ge \omega_i^{s^h}$, with $r_\gamma = s^h+\gamma \psi_i$ and $s_c = s^h+c \psi_i$.  Note \eqref{eq:fepointwiseviEARLY} simplifies substantially.  Let
+for all $\gamma \ge \beta_i^{s^h}$.  Let
 \begin{equation}
@@
 \end{equation}
-be the point-wise residual function.  Then \eqref{eq:fepointwiseviEARLY} is equivalent
+be the point-wise residual function and note \eqref{eq:fepointwiseviEARLY} simplifies into
 \begin{equation}
@@
 \end{equation}
-Supposing \eqref{eq:fepointwisevi} is well-posed, after solving it for $c$ the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
+After solving \eqref{eq:fepointwisevi} for $c$ the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
 Observe that admissibility is preserved because the updated surface elevation is also in $\mathcal{K}^h$.
 
-If each pointwise update is done immediately then the smoother is of Gauss-Seidel (GS) type, sometimes called a serial or multiplicative iteration.
-By contrast, the (parallel, additive) Jacobi smoother computes solutions at all nodes $i$ before doing all the updates.
-In linear problems it is known that Gauss-Seidel generally converges in about half as many iterations \cite{Greenbaum1997}, but here there is a more glaring difference between the two iterations.
-The GS iteration would be extraordinarily expensive because a new Glen-Stokes solution would be required at every node of $\mathcal{T}$, so we will only implement the Jacobi version.
+If $\rho_i(s^h; c)$ were linear and increasing in $c$ then the solution to \eqref{eq:fepointwisevi} would be given by a simple and well-known formula \cite[formula (4.4), for example]{GraeserKornhuber2009}, namely $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ where $\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$.
+In fact $\rho_i(s^h; c)$ is nonlinear, but we adopt a simple approximation scheme for a smoother.
+That is, we linearize
+\begin{equation}
+\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0) \label{eq:rhoapprox}
+\end{equation}
+and do one projected Newton step for problem \eqref{eq:fepointwisevi} at each node where $\rho_i'(s^h; 0) > 0$.
+Because the interior PDE in the shallow, flat-bed model is elliptic, one can prove that the Jacobian diagonal entry in that case is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, but the full Glen-Stokes model here allows no such guarantee to our knowledge.
+As the simplest-possible rule, if $\rho_i'(s^h; 0) \le 0$, so we know that the model has degenerated at the point in the sense that the linearized operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.
+% FIXME THIS MUST BE TOO AGGRESSIVE?
 
-If $\rho_i(s^h; c)$ were linear and increasing in $c$ then the solution to \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_i^{s^h}\right\}$ if $\rho_i(s^h; c) = \alpha c + \beta$ with $\alpha > 0$.
-In fact $\rho_i(s^h; c)$ is nonlinear, but the simplest approximation scheme is to linearize $\rho_i(s^h; c)$, that is, to do one projected Newton step for the one-dimensional VI problem \eqref{eq:fepointwisevi} at each node.
+Note that the derivative, which is also a Jacobian diagonal entry, can be approximated by a finite difference,
+\begin{equation}
+\rho_i'(s^h; 0) = (F^h)'(s^h)[\psi_i,\psi_i] \approx \frac{\rho_i(s^h; \eps) - \rho_i(s^h; 0)}{\eps}.  \label{eq:rhofd}
+\end{equation}
+Use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.
 
-To implement this strategy we approximate
+If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative.
+By contrast, a (parallel, additive) Jacobi smoother would compute and store solutions $c$ at each node $i$ before doing any updates.
+In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely the expense of yet more residual evaluations.
+Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}, in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.
+
+\begin{pcode}[ht]
+\begin{pseudo*}
+\pr{pngs\_slow}(s^h,b^h,a,\id{eps}=1.0\,m,\id{omega}=1)\text{:} \\+
+    \ct{check admissibility: $s^h \ge b^h$} \\
+    function $\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i}$ \\
+    for $i = 0,\dots,m-1$ \\+
+        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{FD for Jacobian diagonal entry} \\
+        $\beta_i = b^h(x_i,y_i) - s^h(x_i,y_i)$ \qquad\qquad \ct{pointwise defect obstacle} \\
+        if $J_{ii} > 0$ \\+
+            $(s^h)_i \gets \max\{(s^h)_i - \id{omega}\,\rho_i(s^h; 0) / \alpha_i, \beta_i\}$ \\-
+        else \\+
+            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
+\end{pseudo*}
+\caption{Projected nonlinear GS iteration, an expensive in-place solver (smoother) using a finite-difference derivative.
+Note that $2m$ solutions of Glen-Stokes problem \eqref{eq:glenstokesweak} are required per application.}
+\label{pc:pngsslow}
+\end{pcode}
+
+To build a better smoother we return to the computation of the Jacobian diagonal entry $\rho_i'(s^h; 0)$.  Suppose we partition the support of $\psi_i$ into where there is ice and not,
 \begin{equation}
-\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, (F^h)'(s^h)[\psi_i,\psi_i]. \label{eq:rhoapprox}
+\theta_i = \{\psi_i > 0\} \cap \{s^h > b^h\}, \qquad {\hat\theta}_i = \{\psi_i > 0\} \setminus \theta_i.  \label{eq:thetasupport}
 \end{equation}
-If we partition the support of $\psi_i$ into where there is ice and not,
+Suppose, for the next two formulas, that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
+Then by \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute with local integrals
 \begin{equation}
-\theta_i = \{\psi_i > 0\} \cap \{s^h > b^h\}, \qquad {\hat\theta}_i = \{\psi_i > 0\} \setminus \theta_i,  \label{eq:thetasupport}
-\end{equation}
-then by \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute with local integrals
-\begin{equation}
-\rho_i(s^h; 0) = F^h(s^h)[\psi_i] - \ip{a}{\psi_i} = - \int_{\theta_i} \bu|_{s^h} \cdot \bn_{s^h}\, \psi_i\,dx\,dy - \ip{a}{\psi_i}  \label{eq:rhozero}
+\rho_i(s^h; 0) = F^h(s^h)[\psi_i] - \ip{a}{\psi_i} = - \int_{\theta_i} (\bu|_{s^h} \cdot \bn_{s^h}- a)\, \psi_i  \label{eq:rhozero}
 \end{equation}
@@
 \begin{align}
-(F^h)'(s^h)[\psi_i,\psi_i] &= - \int_{\theta_i} \lim_{\eps\to 0^+} \frac{(\bu|_{s^h+\eps\psi_i} - \bu|_{s^h}) \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i\,dx\,dy \label{eq:rholocalderiv} \\
-  &\qquad - \int_{{\hat\theta}_i} \lim_{\eps\to 0^+} \frac{\bu|_{s^h+\eps\psi_i} \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i\,dx\,dy.  \notag
+\rho_i'(s^h; 0) &= (F^h)'(s^h)[\psi_i,\psi_i]  \label{eq:rholocalderiv} \\
+  &= - \int_{\theta_i} \lim_{\eps\to 0^+} \frac{(\bu|_{s^h+\eps\psi_i} - \bu|_{s^h}) \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i - \int_{{\hat\theta}_i} \lim_{\eps\to 0^+} \frac{\bu|_{s^h+\eps\psi_i} \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i.
+\notag
 \end{align}
 
-If $\alpha=(F^h)'(s^h)[\psi_i,\psi_i]$ is positive then the solution of \eqref{eq:fepointwisevi} is
-\begin{equation}
-  c = \max\left\{-\frac{\rho_i(s^h; 0)}{\alpha}, \omega_i^{s^h}\right\}, \label{eq:fepointwisevisolution}
-\end{equation}
-which generalizes a well-known formula for the classical obstacle problem \cite[formula (4.4), for example]{GraeserKornhuber2009}.
-However, if $\alpha\le 0$ then the model has degenerated in the sense that the linearized operator is not acting elliptically with respect to that perturbation.
-In that case we make no change to $s^h$; we set $c=0$.
-Because the interior PDE in the shallow, flat-bed model is elliptic, one can prove that the analogous $\alpha$ is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, but the full Glen-Stokes model here allows no such guarantee to our knowledge.
+Clearly, any smoother application, which is a sweep over all nodes in $\mathcal{T}$, will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak}, namely solving with the geometry determined by the current surface elevation $s^h$.
+However, formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated are prohibitively expensive because a Stokes solve is required  for each mesh node.
+Thus we seek a less expensive way to estimate the surface velocity for a perturbed surface elevation with a small ``bump'' $\eps\psi_i$.
+The change to the nonlinearities in this Glen-Stokes problem should be small, and it would seem that the most important change to the velocity and pressure fields from a small bump perturbation is through the perturbed gravitational load applied to the linearized problem around the state $s^h$.
+That is, we suppose other effect such as a perturbed normal direction on the surface are significantly smaller.
 
-We will approximate the limits in \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} by using the obvious difference quotients, i.e.~by removing ``$\lim_{\eps\to 0^+}$'' and choosing $\eps$ to be small, and we set $\eps$ to one meter in computations.
-However, we need to make further approximations to the computation of the divisor $\alpha$, i.e.~in \eqref{eq:rholocalderiv}, if we want a practical solver.
-
-Note that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
-Thus $\bu|_{s^h}$ and $\bu|_{s^h+\eps\psi_i}$ are the results of separate Glen-Stokes solves, over $\Lambda_{s^h}$ and $\Lambda_{s^h+\eps\psi_i}$, respectively.
-Clearly, each smoother application, a sweep over all nodes in $\mathcal{T}$, will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak}, namely solving with the geometry determined by the current surface elevation $s^h$.
-However, formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated are prohibitively expensive; a Stokes solve is required  for each mesh node.
-
-Thus we seek a less expensive way to estimate the surface velocity for a perturbed surface elevation with a small ``bump'' $\eps\psi_i$.
-The nonlinearities in this Glen-Stokes problem should be unimportant, and it would seem that the most important change to the velocity and pressure fields from a small bump perturbation is through the perturbed gravitational load applied to the linearized problem around the state $s^h$.
-(Versus the additional effect of a perturbed normal direction on the surface, for example.)
-
-In fact, let us fix a current surface elevation $s^h$ and assume that problem \eqref{eq:glenstokesweak} for $\Lambda_{s^h}$ yields solution $\bu^h,p^h$ at the convergence of its (Newton) iteration.
+FIXME Let us fix a current surface elevation $s^h$ and assume that problem \eqref{eq:glenstokesweak} for $\Lambda_{s^h}$ yields solution $\bu^h,p^h$ at the convergence of its (Newton) iteration.
 Noting that we are using a stable mixed space for this Glen-Stokes problem (Section \ref{sec:fe}), the final step in the iteration can be regarded as providing the solution of a linear system
@@
 FE & finite element & Section \ref{sec:fe} \\
+GS & Gauss-Seidel & Section \ref{sec:smoothers} \\
 IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\

### 2021-05-30 py/basesmoother.py
1mm is much better than 50 m as a CP residual norm criterion

@@ -40,7 +40,7 @@ class SmootherObstacleProblem(ABC):
             Jstr += '_' if z[k] == 0.0 else '*'
         print('  %d nonzeros: ' % sum(z > 0.0) + Jstr)
 
-    def inactiveresidualnorm(self, mesh, w, r, phi, ireps=50.0):
+    def inactiveresidualnorm(self, mesh, w, r, phi, ireps=0.001):
         '''Compute the norm of the residual values at nodes where the constraint
         is NOT active.  Where the constraint is active the residual F(w) in the
         complementarity problem is allowed to have any positive value; only the

### 2021-05-30 py/steady.py
Update tests and notes thereon.

@@ -5,10 +5,10 @@
 #   1. copy partI/mcdn.py and build it out
 
 # RUNS:
-#   1. actually converges
+#   actually converges
 #      ./steady.py -sweepsonly -J JJ -smoother jacobislow -Hmin HH
-#      with JJ=1,2,3,4 and HH=0,10,100;  HH=0 seems as good as any
-#      also JJ=5
+#      with JJ=1,2,3,4,5 and HH=0,10,100;  HH=0 seems as good as any
+#      also sort of works with JJ=6 when HH=100 and -mz 8
 #      also works with -smoother gsslow
 
 import sys

### 2021-05-30 paper/msg.tex
More stuff on smoothers, and start to describe MCDN.

@@
 \newcommand{\rr}{{\text{r}}}
+
+% norm |||x|||
+\newcommand{\vertiii}[1]{{\left\vert\kern-0.25ex\left\vert\kern-0.25ex\left\vert #1 \right\vert\kern-0.25ex\right\vert\kern-0.25ex\right\vert}}
 
@@
 \end{align}
-Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$, i.e.~the dynamical part of the SKE vanishes for small ice masses on bare ground.
-It is reasonable to suppose this is so when such small ice masses do not slide.
+Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.
+In physical terms this means that the dynamical part of the SKE vanishes for small ice masses on bare ground, and it is reasonable to suppose this is so if the ice cannot slide.
 
@@
 Note that $\psi_i \in \mathcal{D}_+^h$ and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+One can call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009}.
 
@@
 \begin{pseudo*}
-\pr{pngs\_slow}(s^h,b^h,a,\id{eps}=1.0\,m,\id{omega}=1)\text{:} \\+
+\pr{pngs\_slow}(s^h,b^h,\id{eps}=1.0,\id{omega}=1.0)\text{:} \\+
     \ct{check admissibility: $s^h \ge b^h$} \\
-    function $\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i}$ \\
     for $i = 0,\dots,m-1$ \\+
         $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{FD for Jacobian diagonal entry} \\
-        $\beta_i = b^h(x_i,y_i) - s^h(x_i,y_i)$ \qquad\qquad \ct{pointwise defect obstacle} \\
-        if $J_{ii} > 0$ \\+
-            $(s^h)_i \gets \max\{(s^h)_i - \id{omega}\,\rho_i(s^h; 0) / \alpha_i, \beta_i\}$ \\-
+        if $\alpha_i > 0$ \\+
+            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
+            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
         else \\+
@@
 \end{pseudo*}
-\caption{Projected nonlinear GS iteration, an expensive in-place solver (smoother) using a finite-difference derivative.
-Note that $2m$ solutions of Glen-Stokes problem \eqref{eq:glenstokesweak} are required per application.}
+\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.
+Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of \pr{pngs\_slow}.}
 \label{pc:pngsslow}
@@
 \begin{pseudo*}
-\pr{pnj}(j,g^j,y^j,N^j,\ell^j,\phi^j,\id{newtonits}=2,\id{omega}=0.67)\text{:} \\+
-    \ct{check admissibility: $y^j \ge \phi^j$} \\
-    for $k=1,\dots,\id{newtonits}$ \\+
-        $\rho_p(c) := N^j(g^j+y^j+c\psi_p^j)[\psi_p^j] - \ell^j[\psi_p^j]$ \\
-        $r_p, \delta_p = \rho_p(0), \rho_p'(0)$ \qquad\qquad\qquad\qquad \ct{compute and save for all $p$}\\
-        for $p=1,\dots,m_j$ \\+
-            $c = \pr{pointupdate}(r_p, \delta_p,y_p,\phi_p,\ell^j[\psi_p^j])$ \\
-            $y_p \gets y_p + \id{omega}\,c$
+\pr{pnj}(s^h,b^h,\id{omega}=1.0)\text{:} \\+
+    \ct{check admissibility: $s^h \ge b^h$} \\
+    evaluate $\{\rho_i(s^h; 0)\}_{i=0}^{m-1}$  \qquad\qquad \ct{and save the final Newton step linear system} \\
+    for $i = 0,\dots,m-1$ \\+
+        $\alpha_i = (A^{-1})_{ii}$  \qquad\qquad \ct{FIXME: give correct Jacobian diagonal entry} \\
+        if $\alpha_i > 0$ \\+
+            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\--
+    for $i = 0,\dots,m-1$ \\+
+        if $\alpha_i > 0$ \\+
+            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
+        else \\+
+            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
 \end{pseudo*}
-\caption{Projected nonlinear Jacobi as a smoother.}
+\caption{Projected nonlinear Jacobi smoother using a bump approximation for the Jacobian diagonal entry $\alpha_i$.
+Problem \eqref{eq:glenstokesweak} is solved only once per application of \pr{pnj}, but additional linear algebra is needed to compute $\alpha_i$.}
 \label{pc:pnj}
@@
 
+We finish this section with some observations which are supported by the results reported in Section \ref{sec:results}.
+Constructing an efficient smoother is the most difficult part of building an effective multilevel scheme for numerically solving the SIGP.
+In the time-stepping approach to steady state, used by almost all existing models, the difficulty is reflected in the nontrivial determination of a valid conditionally-stable time step criterion for the evolving geometry model.
+Here the concern is associated to the ``ellipticity'' of the coupled equations for the SIGP, reflected in whether the diagonal Jacobian entry $\alpha_i$, used in the above pointwise smoothers, is indeed positive.
+
+In fact, suppose $s^h$ gives the $P_1$ geometry of the ice which solves the discrete SIGP.
+For each icy node $i$ in $\Omega$, i.e.~such that $(s^h)_i>(b^h)_i$, one can see that $\alpha_i>0$ if and only if the addition of a thin (e.g.~one meter) layer of ice to surface of the ice, covering the vicinity of node $i$ (e.g.~the support of $\psi_i$) will cause a dynamical response in which the surface goes down.
+One can show this is so for the SIA theory \cite{JouvetBueler2012}, but, even for nonsliding ice, we know of no proof that the Glen-Stokes SIGP always gives $\alpha_i>0$ at icy locations.
+It follows that pointwise smoothers like \pr{pngs\_slow} and \pr{pnj} are inherently fragile across a full range of glacier geometries.
+
 \section{Multilevel constraint decomposition} \label{sec:mcdstokes}
 
-FIXME CP residual norm
+In this section we propose a multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles of a new method descended from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).
+Our nonlinear variant of MCD, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
+
+We need notation for multiple mesh levels.
+Suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, the coarse level.
+Suppose $\{\mathcal{T}^j\}_{j=0}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
+(If $d=1$ this becomes simple halving of each interval.)  We seek a solution of \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.
+Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and $m_j$ be the number of nodes in, and dimension of, $\mathcal{V}^j$.
+
+Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.
+On this fine level the admissible surface elevations form a closed and convex set $\mathcal{L}^J = \{r^J \ge b^J, r^J|_{\partial\Omega} = b^J|_{\partial\Omega}\} \subset \mathcal{V}^J$.
+However, instead of defining such admissible sets on each level, presumably via coarse interpolants of $b^J$, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition starting from the defect obstacle for the current fine-level iterate $s^J$.
+That is, suppose FIXME
+
+Before describing the MCDN approach we should first be clear about when an iteration has converged.
+For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
+\begin{equation}
+\vertiii{F^j}_{(s^j,\varphi^j)} = \left(\sum_{s_i > \varphi_i} |F[\psi_i]|^2 + \sum_{s_i = \varphi_i} |\min\{F[\psi_i],0\}|^2\right)^{1/2} \label{eq:cpnorm}
+\end{equation}
+be the ``CP residual norm'' associated to the finite-dimensional complementarity problem (CP) $s^j \ge \varphi^j$, $F(s^j) \ge 0$, and $(s^j-\varphi^j) F(s^j) = 0$.
+Note $s^j \in \mathcal{V}^j$ solves this CP if and only if $s^j$ is admissible and $\vertiii{F(s^j)}_{(s^j,\varphi^j)}=0$.
+As shown in Pseudocode \ref{pc:mcdn-solver} we will iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.
+
+\begin{pcode}[ht]
+\begin{pseudo*}
+\pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
+    function $\rho(s)[r] = F^J(s)[r] - \ip{a}{r}$ \qquad\qquad \ct{residual of \eqref{eq:fesigpweakform}} \\
+    $\rho_0=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
+    for $k=1,\dots,\id{cyclemax}$ \\+
+        $\chi^J = b^J - s^J$ \qquad\qquad \ct{fine-level defect obstacle} \\
+        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,\chi^J)$ \\
+        $\rho_k=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
+        if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ \\+
+            break \\--
+\end{pseudo*}
+\caption{The SIGP is solved by iterating V-cycles (Pseudocode \ref{pc:mcdn-vcycle}) until the CP residual norm \eqref{eq:cpnorm} is small.}
+\label{pc:mcdn-solver}
+\end{pcode}
 
@@
 \end{pseudo*}
-\caption{Nonlinear MCD V-cycle.}
+\caption{FIXME Nonlinear MCD V-cycle.}
 \label{pc:mcdn-vcycle}
@@
 
-FIXME we will iterate \pr{mcdn-vcycle} until the CP residual norm has been reduced by a tolerance factor, as in Pseudocode \ref{pc:mcdn-solver}
+\section{Results for steady geometry} \label{sec:results}
 
-\begin{pcode}[ht]
-\begin{pseudo*}
-\pr{mcdn-solver}(J,w^J,N,a,b,\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
-    $\ell^J = \ip{a}{\cdot}$ \\
-    $F(z)[\cdot] := N^J(z)[\cdot] - \ell^J[\cdot]$ \\
-    $r_0=\|\hat\bF(w^J)\|$ \\
-    for $s=1,\dots,\id{cyclemax}$ \\+
-        $\chi^J = b^J - w^J$ \\
-        $w^J\gets w^J+\pr{mcdn-vcycle}(J,w^J,N,\ell^J,\chi^J)$ \\
-        if $\|\hat\bF(w^J)\| \le \id{rtol} \, r_0$ \\+
-            break \\--
-\end{pseudo*}
-\caption{The SIGP is solved by iterating \pr{mcdn-vcycle}.}
-\label{pc:mcdn-solver}
-\end{pcode}
-
-FIXME PETSc \cite{Balayetal2020,Bueler2021} used to solve the dynamics problem and thereby evaluate the residual $F$
-
-FIXME in this paper the Glen-Stokes problem is solved by Newton linearization and (parallel) direct solution of the Newton step equations; however, geometric multigrid is available already for Glen-Stokes dynamics \cite{IsaacStadlerGhattas2015}; see also \cite{BrownSmithAhmadia2013} and \cite{Tuminaroetal2016} which solve a first-order approximation of the Glen-Stokes ice sheet problem by geometric and algebraic multigrid, respectively
+FIXME in this paper the Glen-Stokes problem is solved by Newton linearization and (parallel) direct solution of the Newton step equations; Firedrake and PETSc \cite{Balayetal2020,Bueler2021} used to solve this dynamics problem and thereby evaluate the residual $F$; geometric multigrid schemes exist for the  Glen-Stokes problem \cite{IsaacStadlerGhattas2015}; see also \cite{BrownSmithAhmadia2013} and \cite{Tuminaroetal2016} which solve a first-order approximation of the Glen-Stokes ice sheet problem by geometric and algebraic multigrid, respectively
 
@@
 CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
+FAS & full approximation scheme & Section \ref{sec:mcdstokes} \\
 FE & finite element & Section \ref{sec:fe} \\
@@
 IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
-MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes}, Pseudocodes \ref{pc:mcdn-vcycle} and \ref{pc:mcdn-solver} \\
+MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
+MCDN & multilevel constraint decomposition (nonlinear) & Section \ref{sec:mcdstokes}, Pseudocode \ref{pc:mcdn-vcycle} \\
 NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\

### 2021-05-30 paper/msg.tex
Add a basic MCD figure.
Edits section 2.

@@
 
+\newcommand{\bus}{\bu|_s}
+
 % norm |||x|||
@@
 The standard model for determining the geometry of glaciers is based in part upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.
-This Glen-Stokes ice-dynamics model can only determine the free-surface glacier geometry when it is combined with the surface kinematical equation (SKE) and a form of climate input, namely the rate of accumulation of snow or ablation, i.e.~melting and runoff.
-In this section we state the strong form of this ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).
-The emphasis here is on the complementarity-problem nature of the SIGP, often not explicit in the glaciers literature (e.g.~as observed by \cite{SchoofHewitt2013}).
+This Glen-Stokes ice-dynamics model can only determine the free-surface glacier geometry when it is combined with the surface kinematical equation (SKE) and a climate input, namely the balance of the rates of accumulation of snow and of ablation, i.e.~melting and runoff.
+In this section we state the strong form of the resulting ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).
+The emphasis here is on the complementarity-problem nature of the SIGP, which is often not explicit in the glaciers literature (e.g.~as observed by \cite{SchoofHewitt2013}).
 We then define its weak form, noting along the way several unknown aspects of the theory.
@@
 
-The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ (Figure \ref{fig:stokesdomain}).
-We denote the map-plane variables as $x$ when $d=1$ and $x,y$ when $d=2$.
-On $\Omega$ we assume that a climatic mass balance (CMB) function $a(x,y)$ is defined at every point, whether or not ice is present at that location.
-In ice-free areas this function could be described as the ``potential'' CMB, namely the annual balance of snow accumulation and potential melt if ice were present \cite{Cogleyetal2011}, e.g.~as computed by energy balance in a climate or weather model.
-Here we assume $a$ has units of ice thickness per time, equivalently ice volume per area (in $\Omega$) per time, compatible with our assumption of constant ice density below.
-Also we assume there is a bed elevation function $b(x,y)$ defined everywhere on $\Omega$, which determines the topography upon which the glacier sits.
+The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ for $d=1,2$ (Figure \ref{fig:stokesdomain}).
+The coordinates on $\Omega$ are denoted as $x$ when $d=1$ and $x,y$ when $d=2$.
+On $\Omega$ we assume that a climatic mass balance (CMB) \cite{Cogleyetal2011} function $a(x,y)$ is defined at every point, whether or not ice is present at that location.
+In ice-free areas this function is the ``potential'' CMB, namely the annual balance of snow accumulation and melt if ice were present, e.g.~as computed by energy balance in a climate or weather model.
+Here we assume $a$ has units of ice thickness per time, compatible with our assumption of constant ice density (below).
+Next we assume there is a bed elevation function $b(x,y)$ defined everywhere on $\Omega$, which determines the topography upon which the glacier sits.
 The functions $a$ and $b$ are the data (inputs) of the SIGP.
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{genfigs/stokesdomain.pdf}
+\includegraphics[width=0.7\textwidth]{genfigs/stokesdomain.pdf}
 \end{center}
-\caption{In the steady ice geometry problem (SIGP) the CMB $a$ (arrows; downward $=$ accumulation) and bed elevation $b$ (solid) are given on a fixed map-plane region $\Omega \subset \RR^d$.
-The solution includes a surface elevation $s$ (dashed) on $\Omega$; note $s=b$ where ice-free.
-The ice velocity $\bu$ and pressure $p$ are defined in the icy domain: $\Lambda_s = \{(x,y,z)\,:\,b(x,y) < z < s(x,y)\} \subset \RR^{d+1}$.}
+\caption{In the SIGP the CMB $a$ (arrows; down for accumulation) and bed elevation $b$ (solid) are given data on a map-plane region $\Omega \subset \RR^d$.
+The solution includes a surface elevation $s$ (dashed) on $\Omega$, such that $s\ge b$, plus the ice velocity $\bu$ and pressure $p$ on the icy domain $\Lambda_s = \{b < z < s\} \subset \RR^{d+1}$.}
 \label{fig:stokesdomain}
@@
 
-The Glen-Stokes dynamical model, the so-called stress balance, for ice only applies in the icy domain in $\RR^{d+1}$.
+The Glen-Stokes dynamical model, the so-called stress balance, only applies in the icy domain in $\RR^{d+1}$.
 We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption this this icy domain has a well-defined upper surface elevation, a function $s(x,y)$.
 That is, we assume there are \emph{no overhangs}, and that the under-side of the ice is in contact with the bed $b$.
-We then define $s$ everywhere in $\Omega$ by extending with $s=b$ where ice is absent, thus $s\ge b$ applies on $\Omega$.
+We then define $s$ everywhere in $\Omega$ by extending with $s=b$ where ice is absent, thus $s\ge b$ applies on all of $\Omega$.
 Note that $s$ is part of the model solution; it is not given data.
@@
 There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that $s$ is well-defined.
-Furthermore the same concern applies to each time step of an evolving model; the model does not stop overhangs from appearing at the margin.
+Furthermore the same concern applies to each time step of an evolving model; actual glaciers can develop overhangs at the margin.
 However, because overhangs are small features in large glaciers and ice sheets, essentially all modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.
-An exception is \cite{PralongFunk2005}, who furthermore suggest a potentially well-posed continuum model, a nontrivial extension of the one here, in which serac and ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion.
-This extended model also explains one of our numerical boundary variants, a short, fixed-height cliff (Section \ref{sec:mcdstokes}), but overhangs are not allowed in any of our numerical constructions.
+An exception is \cite{PralongFunk2005}, who suggest a potentially well-posed continuum model, a nontrivial extension of the one here, in which serac and ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion.
+Though marginal cliffs may exist in our model (Section \ref{sec:fe}), overhangs are never allowed.
 
-Based on the assumption of a well-defined upper surface, we define the (solution) extent of the ice as the open set
+Based on a well-defined upper surface, the icy domain is the open set
 \begin{equation}
@@
 where $z$ is vertically-upward.
-(If $d=1$ then coordinates are ``$(x,z)$'' on $\Lambda_s$.)  The map-plane region $\Omega$ need not be connected or simply-connected, but even if so the solution set $\Lambda_s$ need not be.
-However, $\Lambda_s$ has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).
+(Coordinates are denote $(x,z)$ on $\Lambda_s$ when $d=1$.)  The domain $\Lambda_s$ has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).
+Again note $\Lambda_s$ is part of the solution.
 
-For a given geometry $\Lambda_s$, we model the ice as a very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.
+We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.
 Allowing any Glen exponent $n\ge 1$, the equations inside $\Lambda_s$ are
@@
 The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.
-Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} is an aspect of our mass conservation model.
-As seen below, mass conservation on the free surface of $\Lambda_s$ requires an additional equation \eqref{eq:ske}.
+Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation.
 
@@
 The strain rate tensor $D\bu$ is the symmetric part of $\grad \bu$, $D\bu = \frac{1}{2} \left(\grad\bu + \grad\bu^\top\right)$, and the tensor norm used in \eqref{eq:viscflowlaw} satisfies $|D\bu|^2 = \frac{1}{2} (D\bu)_{ij} (D\bu)_{ij}$.
-Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} implies that $\tau$ is also symmetric with trace zero, thus that $p=-(d+1)^{-1} \trace \sigma$.
+Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} then implies that $\tau$ is also symmetric with trace zero, thus that $p=-(d+1)^{-1} \trace \sigma$.
 
 In computations we will use constants $n=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.
-The ice hardness $B_n$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996}.
+The ice hardness $B_n$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.
 
 For the $n=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $n>1$ imply an effective viscosity function of $|D\bu|$.
-This would thus be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity
+This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
 \begin{equation}
@@
 \end{equation}
-where $\pp=(1/n)+1$, with $\pp=4/3$ in computations.
-The constant $D_0$ defines a strain-rate scale for glacier flow; $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used here.
-Also we eliminate $\tau$ and rewrite \eqref{eq:forcebalance} as $- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p = \rhoi \mathbf{g}$, with a vector $p$-Laplacian as the leading-order term.
+where $\pp=(1/n)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.
 
-Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding ice.
-In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.
-On the upper free surface we set a condition of zero applied stress,
+Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding glaciers.
+(Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.
+On the base we impose no slip:
 \begin{equation}
-\left(2 \nu_\eps D\bu - pI\right) \bn_s = \bzero  \qquad \qquad \text{\emph{top} } \overline{\partial} \Lambda_s \label{eq:topbc}
+\bu = \bzero  \qquad\qquad \text{\emph{base} } \Gamma_0. \label{eq:basebc}
 \end{equation}
-where $\bn_s$ is any normal to $\overline{\partial} \Lambda_s$.
-(Providing a nonzero atmospheric pressure at the surface is straightforward, but it has no important consequence.)  On the base we impose no slip:
+On the other sub-aerial surfaces we set a condition of zero applied stress,
 \begin{equation}
-\bu = \bzero  \qquad\qquad \text{\emph{base} } \underline{\partial} \Lambda_s. \label{eq:basebc}
+\left(2 \nu_\eps D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
 \end{equation}
+where $\bn$ is any normal to $\overline{\partial} \Lambda$.
+(A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, but that $\grad s$ may become singular there.
 
-Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, and the surface gradient $\grad s$ may become singular there, but the top and bottom surfaces are assumed to meet at the margin (Figure \ref{fig:stokesdomain}).
-An alternative geometry with a small cliff is also considered (Section \ref{sec:fe}).
-Real glacier margins may indeed occur as fracture-generated cliffs \cite{PralongFunk2005}, but fracture processes are not modeled here.
-
-The well-posedness of the above dynamics model, for fixed domain $\Lambda_s$, is proven by \cite{JouvetRappaz2011}, and the solution is a unique pair $(\bu,p)$ defined on $\Lambda_s$.
-(The function spaces are identified below when stating the weak formulation.)  However, the simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.
-Indeed, the above equations make no reference to the climate input function $a$, so we need another ``equation'', which we will see is actually an inequality.
-
-\newcommand{\bus}{\bu|_s}
-
-Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we also extend the surface normal $\bn_s$, thus we set
+The simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.
+However, the above equations make no reference to the climate input function $a$, so we need another ``equation'', one which is actually an inequality.
+Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we set
 \begin{equation}
@@
 \end{equation}
-so that the (un-normalized) surface normal is specified almost everywhere on $\Omega$.
-(By contrast, note that the dynamical boundary condition \eqref{eq:topbc} applies only on top of the ice.)  Furthermore we extend the surface value of the velocity to all of $\Omega$:
+as an (un-normalized) top-surface normal almost everywhere on $\Omega$.  Furthermore we extend the surface value of the velocity to all of $\Omega$:
 \begin{equation}
@@
 \end{equation}
-(The movement of the solid earth would determine $\bu|_b \ne 0$, thus $\bu|_s$ could be nonzero in ice-free areas, and a more-complete model could include this small effect.)
+A discontinuity of this function is allowed at the ice margin.
 
-The steady-state surface kinematical equation (SKE) \cite[see equation (5.21)]{GreveBlatter2009} is the needed additional aspect of mass conservation:
+The steady-state surface kinematical equation (SKE) \cite[equation (5.21)]{GreveBlatter2009} is the needed additional aspect of mass conservation:
 \begin{equation}
@@
 \end{equation}
-Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time.
-(Some references use a distinct quantity, namely the thickness added perpendicularly to the ice surface \cite{GreveBlatter2009}, which is significantly different if the gradient $\grad s$ is large.)
+Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time, while some references use a distinct quantity, namely the thickness added perpendicularly to the ice surface, significantly different if $\grad s$ is large.
 
-While the involved quantities are defined everywhere, kinematical balance \eqref{eq:ske} applies only \emph{on} the ice.
-However $\bus \cdot \bn_s + a \le 0$ everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.
+Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.
 In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.
-Recall that an NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements $x\ge 0$, $F(x)\ge 0$, and $x F(x)=0$, where $F:V\to V$ \cite{FacchineiPang2003}.
+Recall that an NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
+\begin{equation}
+x\ge 0, \quad F(x)\ge 0, \quad x F(x)=0 \label{eq:ncp}
+\end{equation}
+for all $x\in V$, where $F:V\to V$ \cite{FacchineiPang2003}.
 
@@
 
-Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary is $C^1$.
+Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary is $C^1$.
 In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p\in L^\qq(\Lambda)$.
@@
 
+FIXME Figure \ref{fig:decompclassical} shows such a decomposition
+
+% figure generated in mg-glaciers/py by using decomposition_plain() in visualize.py and then:
+% $ ./obstacle.py -J 5 -jcoarse 1 -irtol 1.0e-5 -monitor -diagnostics -random -randommodes 15 -o classical.pdf
+% to get decomp_classical.pdf
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{fixfigs/decompclassical.pdf}
+\end{center}
+\caption{An example $d=1$ and $J=4$ MCD decomposition of the defect obstacle $\chi^J$ for an iterate $s^J$.}
+\label{fig:decompclassical}
+\end{figure}
+
 Before describing the MCDN approach we should first be clear about when an iteration has converged.

### 2021-05-30 paper/genfigs/stokesdomain.py

@@ -8,7 +8,7 @@ plt.rcParams["font.family"] = "serif"
 plt.rcParams["mathtext.fontset"] = "cm"
 
 fsize=18.0
-bigfsize=20.0
+bigfsize=24.0
 
 def genbasicfig(xshift=0.0):
     x = np.linspace(0.0,10.0,1001)
@@ -27,7 +27,7 @@ def genbasicfig(xshift=0.0):
     return x + xshift, s, b
 
 def drawclimate(x,s):
-    plt.text(x[10], s[10]+2.7, r'$a(x,y)$', fontsize=bigfsize, color='k')
+    plt.text(x[10], s[10]+2.7, r'$a$', fontsize=bigfsize, color='k')
     for j in range(10):
         xarr = x[50+100*j]
         if j>0:
@@ -44,9 +44,9 @@ plt.text(x[600] - 1.0, b[600] + 0.4 * s[600], r'$\Lambda_s$',
          fontsize=bigfsize, color='k')
 drawclimate(x,s)
 # mark top surface
-plt.text(x[250], s[250] + 1.5, r'$s(x,y)$', fontsize=bigfsize, color='k')
+plt.text(x[250], s[250] + 0.7, r'$s$', fontsize=bigfsize, color='k')
 # mark bottom surface
-plt.text(x[650], b[650] - 0.5, r'$b(x,y)$', fontsize=bigfsize, color='k')
+plt.text(x[650], b[650] - 0.5, r'$b$', fontsize=bigfsize, color='k')
 # show \Omega
 yR = min(b) - 0.5
 plt.plot([min(x),max(x)],[yR,yR],color='k',lw=1.0)

### 2021-05-30 paper/msg.tex
Edits section 3.

@@
 In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.
-Recall that an NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
+An NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
 \begin{equation}
-x\ge 0, \quad F(x)\ge 0, \quad x F(x)=0 \label{eq:ncp}
+x\ge 0, \quad f(x)\ge 0, \quad x f(x)=0 \label{eq:ncp}
 \end{equation}
-for all $x\in V$, where $F:V\to V$ \cite{FacchineiPang2003}.
+for all $x\in V$, where $f:V\to V$ \cite{FacchineiPang2003}.
 
-We may now state the strong form of the SIGP as follows:
+We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
 \begin{align}
@@
 \nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
-\left(2 \nu_\eps D\bu - pI\right) \bn_s &= \bzero && \text{on $\overline{\partial} \Lambda_s$} \notag \\
-\bu &= \bzero && \text{on $\underline{\partial} \Lambda_s$} \notag
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag \\
+\bu &= \bzero && \text{on $\Gamma_0$} \notag
 \end{align}
-(Definition \eqref{eq:regeffvisc} is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but it is coupled to the dynamical boundary value problem formed by the last four statements.
+(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.
 
-The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$.
-However, as the domain on which $\bu,p$ are defined is only known via the solution elevation $s$, \eqref{eq:strongform} is at best an incomplete description.
-The weak form in the next section, using a $s\mapsto \bu|_s$ solution operator, will partially address this concern.
-Whether in strong or weak form, this inequality-constrained system of PDEs has a largely-unknown theory regarding well-posedness and solution regularity.
-However, certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, and uniqueness in the flat bed case.
+The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$ on $\Omega,\Lambda_s,\Lambda_s$, respectively.
+However, as the domain on which $\bu,p$ are defined is only known via the solution, \eqref{eq:strongform} is at best an incomplete description.
+The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will partially address this concern.
 
-In Sections \ref{sec:fe}--\ref{sec:mcdstokes} we will construct and demonstrate a robust multilevel scheme for the numerical solution of the SIGP weak form (Section \ref{sec:weakido}).
+However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unknown theory regarding well-posedness and solution regularity.
+Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.
+
+After stating the SIGP weak form next in Section \ref{sec:weakido}, in Sections \ref{sec:fe}--\ref{sec:results} we will construct and demonstrate a multilevel scheme for the problem.
 As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.
@@
 
-\section{Weak form and the ice dynamics operator} \label{sec:weakido}
+\section{The weak form using an ice dynamics operator} \label{sec:weakido}
 
-The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.
-Then we will address the weak form of the SIGP.
+The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here, after which we address the weak form of the SIGP.
 
-Denote by $W^{\text{k},\text{r}}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.
+Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.
 Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).
-Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and let $\qq=(1-\pp^{-1})^{-1}=n+1$ be the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.
-Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and
+Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=n+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.
+Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
 \begin{equation}
@@
 \end{equation}
-be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes weak form is the statement that $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy
+be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy the weak form
 \begin{equation}
@@
 
-Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary is $C^1$.
-In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p\in L^\qq(\Lambda)$.
+Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.
+In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p$.
 
-However, we seek to go beyond this dynamical problem on predetermined ice geometry.
-We are interested in finding a triple $s,\bu,p$ which solves \eqref{eq:strongform}, or rather its weak form which we now propose.
+However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.
 Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.
-Because of the $C^1$ boundary, the Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines $f|_{\partial \Omega}$ for $f \in W^{1,\qq}(\Omega)$ \cite[Section 5.5]{Evans2010}.
+The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.
 Let
@@
 be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.
+From now on we also assume that $n>1$.
+Because $\qq = n+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.
+
 However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
 \begin{equation}
-s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ free surface.} \label{eq:quixotic}
+s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
 \end{equation}
 Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.
-However, the existing theory of shallow ice sheet well-posedness \cite{JouvetBueler2012} includes a version of definition \eqref{eq:Kconstraintset},\footnote{In \cite{JouvetBueler2012} the $2\qq/(\qq-1)=8/3$ power of the ice thickness is in $W^{1,\qq}(\Omega)$.} and in that shallow theory $s$ solves an elliptic equation and is thus smooth, at least interior to the free boundary.
+However, the existing theory of well-posedness for the SIA model includes a version of definition \eqref{eq:Kconstraintset} \cite{JouvetBueler2012}, wherein a power of the ice thickness, $u=H^{2q/(q-1)}$, is in $\mathcal{K} = \{u \ge 0\} \subset W^{1,\qq}(\Omega)$, and furthermore in the SIA the surface $s$ solves an elliptic equation and is thus smooth on the set $\{s<b\} \subset \Omega$.
 
-We may now define the ice dynamics operator (IDO), a map using the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).
+We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.
 For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
@@
 \Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
-                        0, & \text{otherwise}, \end{cases} \label{eq:ido}
+                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
 \end{equation}
-where $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
-Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, thus permitting substantial irregularities.
+Here $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
+Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.
 
@@
 \end{center}
-\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$; the thin line is $s+\Phi(s)$.
+\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.
 Both the input and output of $\Phi$ are defined on all of $\Omega$.}
@@
 \end{equation}
-One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.
-Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ has minimal regularity; it need only be in the dual space of $W^{1,\qq}(\Omega)$.
+One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.
+Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ is in the dual space of $W^{1,\qq}(\Omega)$, so it possesses minimal regularity.
 
@@
 Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.
+In any case one can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
+In particular the strong SKE \eqref{eq:ske} holds on $s>b$.
 
 Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.
-VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.
-Specifically, in geometrical analogy, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
+Speaking geometrically, VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.
+Specifically, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
 
-It is well-known that some VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
-The general-bed, steady shallow (SIA) model is not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
+It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
+The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
 
-The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through the definition of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
+The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
 \renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
@@
 
-To numerically solve \eqref{eq:sigpweakform} we will need the (Gateaux) derivative of $F$, which requires differentiating the $\Phi$ as well.
-Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
-(Both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional},
+We will need the (Gateaux) derivative of $F$, and this requires differentiating $\Phi$ as well.
+Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
+(In the following, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional}, if $r\in W_{1,\qq}(\Omega)$ then
 \begin{equation}
@@
 \end{equation}
-In \eqref{eq:idoderiv} note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.
+In \eqref{eq:idoderiv} note that all quantities are extended to the whole of $\Omega$.
+
 We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.
-Observe that $F'$ is straightforward to compute from $\Phi'$.
-
-Note $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, and this clearly requires $t\ge 0$ on the (active) set where $s=b$.
+Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.
+In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.
 However, to simplify our considerations we define an $s$-independent set
 \begin{equation}
-\mathcal{D}_+ = \{t \in W^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}, \label{eq:infdefectset}
+\mathcal{D}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
 \end{equation}
-so $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
+It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
+We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{D}_+$ has small support.
 
-We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, but only in the situation where $t \in \mathcal{D}_+$ has small support.
-(It will be a finite element basis function; see the next section.)  For this it is relevant to say more about the difference, as it depends on whether the point is in the support of $t$ or not:
-\begin{align}
-\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s &= \left(\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_{s+\eps t}\right) + \left(\bus \cdot \bn_{s+\eps t} - \bus \cdot \bn_s\right) \label{eq:differencecases} \\
-    &= \begin{cases}
+In fact the meaning of the finite difference in \eqref{eq:idoderiv} depends on the support of $t$:
+\begin{equation}
+\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
            (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
-           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b \text{ and } t > 0 \\
-           0, & s=b \text{ and } t = 0
-                 \end{cases} \notag
-\end{align}
+           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
+           0, & s=b, t = 0.
+                 \end{cases} \label{eq:differencecases}
+\end{equation}
 Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.
-In physical terms this means that the dynamical part of the SKE vanishes for small ice masses on bare ground, and it is reasonable to suppose this is so if the ice cannot slide.
+In physical terms this means that the dynamical part of the SKE \eqref{eq:ske} vanishes for small ice masses on bare ground, which is reasonable to suppose if the ice cannot slide.
 
 In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.
-When we approximate the true derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because two Glen-Stokes solutions over the ice domains $\Lambda_{s+\eps t}$ and $\Lambda_s$ are seemingly required.
-While this would imply a very large numbers of expensive dynamical solves, in Section \ref{sec:smoothers} we propose a less-expensive approximation.
+When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because an extra solution of Glen-Stokes problem \eqref{eq:glenstokesweak} is required over the icy domain $\Lambda_{s+\eps t}$, but in Section \ref{sec:smoothers} we propose a less-expensive approximation.
 

### 2021-05-31 paper/msg.tex
Add notes to self.

@@
 
+FIXME TALK ABOUT IDO.  1 FORM IN SIA CASE. 2 LINEARIZATION DENSE.  3 SPECTRUM OF IDO DETERMINES STABILITY OF EXPLICIT TIME-STEPPING. 4 DIAGONAL OF IDO UNKNOWN BUT POSITIVE?
+
 Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may immediately re-write the \eqref{eq:strongform} in a simpler strong form, namely an NCP with a hidden dynamical problem:
@@
 
-Assume $\Omega \subset \RR^d$ is polygonal and suppose $\mathcal{T}$ is a triangulation of $\Omega$.
+Assume $\Omega \subset \RR^d$ is polygonal and that $\mathcal{T}$ is a triangulation of $\Omega$.
 (If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
@@
 
+FIXME SMOOTHER POSSIBILITIES.  1 COLORING (PSEUDO-COLORING WITH LATERAL SEPARATION OF A FEW ICE THICKNESSES).  2 PATCH-TYPE WHICH SOLVE LOW-DIMENSIONAL VI.
+
 \section{Multilevel constraint decomposition} \label{sec:mcdstokes}

### 2021-05-31 paper/msg.tex
Edits sections 4 and 5.

@@
 \begin{equation}
-\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
+\Phi(s) = \begin{cases} \bu|_s \cdot \bn_s, & s > b, \\
                         0, & \text{otherwise}. \end{cases} \label{eq:ido}
 \end{equation}
-Here $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
+Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
 Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.
@@
 Assume $\Omega \subset \RR^d$ is polygonal and that $\mathcal{T}$ is a triangulation of $\Omega$.
-(If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
+(If $d=1$ then $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
 \begin{equation}
@@
 
-Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$.  From $b^h$ we define the closed and convex admissible subset of $\mathcal{V}^h$:
+Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$, and define the following closed and convex admissible subset of $\mathcal{V}^h$:
 \begin{equation}
@@
 \end{equation}
-For $s^h\in \mathcal{K}^h$ we also define the numerical icy domain as in \eqref{eq:lambdas}:
-\begin{equation}
-\Lambda_{s^h} = \{(x,y,z)\,:\,b^h(x,y) < z < s^h(x,y)\} \subset \RR^{d+1}.  \label{eq:felambdas}
-\end{equation}
-Also, a triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in the $s^h$ geometry if $s^h=b^h$ at every vertex of $T$, and icy otherwise.
+Each $s^h\in \mathcal{K}^h$ defines a numerical icy domain $\Lambda_{s^h} \subset \RR^{d+1}$ as in \eqref{eq:lambdas}.
+A triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in $\Lambda_{s^h}$ if $s^h=b^h$ at every vertex of $T$, and icy otherwise.
 
-Given $s^h$ and $b^h$ from $\mathcal{V}^h$, we construct a Firedrake extruded mesh \cite{McRaeetal2016} to decompose $\Lambda_{s^h}$.
-If $d=2$ then the extruded mesh consists of triangular prisms, while if $d=1$ it has quadrilateral elements.
-The extruded mesh construction starts from the triangulation $\mathcal{T}$ of $\Omega \subset \RR^d$, now called the ``base mesh''.
-As shown in Figure \ref{fig:extruded}, each icy $T$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.
+Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$).
+The triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.
+As shown in Figure \ref{fig:extruded}, each icy $T \in \mathcal{T}$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.
 However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.
+An extruded mesh is regenerated each time $F^h(s^h)$ is evaluated, but the base mesh $\mathcal{T}$ is reused.
 
@@
 \end{center}
-\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.
-In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold); see \eqref{eq:ido}.
-In the ``cliffs'' extrusion (right) all elements have a minimum thickness and the trace evaluation (bold) may occur above $s^h$ (dotted).}
+\caption{We compute the IDO $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.
+In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold).
+In the ``cliffs'' extrusion (right) all elements have a minimum thickness so the trace evaluation may occur above $s^h$ (dotted).}
 \label{fig:extruded}
@@
 We will compare two extrusion modes.
-In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly, prism (quadilateral) elements at the ice margin degenerate; such elements have empty facets.
-In the alternate ``cliffs'' extrusion, each element over an icy $T$ is nondegenerate.
-Specifically, if $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in experiments, then each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$.
-In this extrusion, a stress-free condition is applied on the cliffs  in the Glen-Stokes problem \eqref{eq:glenstokesweak}; they are treated the same way as the top surface.
+In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly.
+It follows that the elements at the ice margin are degenerate; they have positive volume but zero height at some nodes or even empty (vertical) facets.
+In the alternate ``cliffs'' extrusion, each extruded element is nondegenerate.
+Specifically, each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$ where $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in computations.
+The surface $s^h$ remains unchanged (and continuous), so these cliffs are only relevent to problem \eqref{eq:glenstokesweak}.
 
@@
 By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.
-That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, but on $\overline{\partial} \Lambda_{s^h}$, and then extend by zero to all of $\Omega$.
+That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, and then extend by zero to all of $\Omega$.
 In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.
 
-We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}, and a Newton iteration, with direct solution of the step equations and back-tracking line search, to resolve the power-law nonlinearity.
-As stated in Section \ref{sec:intro}, note that our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme (next section) are needed to solve the SIGP \eqref{eq:sigpweakform}.
-Overall optimality of the solution algorithm would also need an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
+We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}.
+The power-law nonlinearity is resolved by using a Newton iteration, with direct solution of the step equations and back-tracking line search.
+As stated in Section \ref{sec:intro}, our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme are needed to solve the SIGP \eqref{eq:sigpweakform}.
+Overall optimality of the solution algorithm also requires an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
 
-Finally, the discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
+Finally, let $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.  The discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
 \begin{equation}
-F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h , \label{eq:fesigpweakform}
+F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h. \label{eq:fesigpweakform}
 \end{equation}
-where $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.
-We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.
+The only role of the extruded mesh in solving \eqref{eq:fesigpweakform} is in the evaluation of the IDO $\Phi^h(s^h)$.
+We believe that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin, and we have no proof.
 
@@
 
-In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations, variants of the Gauss-Seidel and Jacobi methods, which are suitable for VIs \cite{KinderlehrerStampacchia1980}.
+In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations which are variants of the Gauss-Seidel and Jacobi methods, but made suitable for VIs \cite{KinderlehrerStampacchia1980}.
+Given a current surface elevation $s^h\in \mathcal{K}^h$, these iterations sweep through the $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems.
 While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
-However, because of the non-locality of the SIGP functional $F^h$, we must make careful approximations in order to retain adequate per-iteration performance.
+Because of the non-locality and density of the IDO $\Phi^h$ (Section \ref{sec:weakido}), we must make careful approximations in order to retain adequate per-iteration performance.
 
-Given a current surface elevation $s^h\in \mathcal{K}^h$, our smoothers will sweep through the $m$ vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems and then updating $s^h$.
-Denote an arbitrary vertex of $\mathcal{T}$ by $(x_i,y_i)$ and let $\psi_i \in \mathcal{V}^h$ be the corresponding hat function, with $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.
-Also let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}$.
-Note that $\psi_i \in \mathcal{D}_+^h$ and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
-One can call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009}.
+Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.
+Let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.
+Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.
+Note that $\psi_i \in \mathcal{D}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.
 
-The one-dimensional VI at node $i$ is derived from \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for scalars $\gamma,c$.
+The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for some scalars $\gamma,c$.
 Specifically, we seek $c \ge \beta_i^{s^h}$ such that
 \begin{equation}
-F^h(s_c)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
+F^h(s^h+c \psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
 \end{equation}
@@
 \end{equation}
-be the point-wise residual function and note \eqref{eq:fepointwiseviEARLY} simplifies into
+be the point-wise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
 \begin{equation}
@@
 \end{equation}
-After solving \eqref{eq:fepointwisevi} for $c$ the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.
-Observe that admissibility is preserved because the updated surface elevation is also in $\mathcal{K}^h$.
+The smoother will approximately solve \eqref{eq:fepointwisevi} for $c$ the smoother, and then update $s^h \gets s^h + c \psi_i$.
+Observe that admissibility is preserved because $s^h+c \psi_i \in \mathcal{K}^h$ when $c \ge \beta_i^{s^h}$.
 
-If $\rho_i(s^h; c)$ were linear and increasing in $c$ then the solution to \eqref{eq:fepointwisevi} would be given by a simple and well-known formula \cite[formula (4.4), for example]{GraeserKornhuber2009}, namely $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ where $\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$.
-In fact $\rho_i(s^h; c)$ is nonlinear, but we adopt a simple approximation scheme for a smoother.
-That is, we linearize
+If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.
+In fact $\rho_i(s^h; c)$ is nonlinear, and we will linearize
 \begin{equation}
-\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0) \label{eq:rhoapprox}
+\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0), \label{eq:rhoapprox}
 \end{equation}
-and do one projected Newton step for problem \eqref{eq:fepointwisevi} at each node where $\rho_i'(s^h; 0) > 0$.
-Because the interior PDE in the shallow, flat-bed model is elliptic, one can prove that the Jacobian diagonal entry in that case is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, but the full Glen-Stokes model here allows no such guarantee to our knowledge.
-As the simplest-possible rule, if $\rho_i'(s^h; 0) \le 0$, so we know that the model has degenerated at the point in the sense that the linearized operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.
-% FIXME THIS MUST BE TOO AGGRESSIVE?
+and at each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.
+If $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.
+% FIXME TOO AGGRESSIVE?
 
-Note that the derivative, which is also a Jacobian diagonal entry, can be approximated by a finite difference,
+Because the interior PDE in the shallow, flat-bed model is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}.
+However, for the full Glen-Stokes model there is no such guarantee to our knowledge.
+
+Note that the derivative $\rho_i'(s^h; 0)$, which is also a diagonal entry of the Jacobian of $F^h$, can be approximated by a finite difference,
 \begin{equation}
@@
 \end{equation}
-Use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.
+Literal use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.
 
-If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative.
-By contrast, a (parallel, additive) Jacobi smoother would compute and store solutions $c$ at each node $i$ before doing any updates.
-In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely the expense of yet more residual evaluations.
-Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}, in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.
+If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node $i$ before doing any updates.
+In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely that GS imposes the expense of yet more residual evaluations.
+Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.
+It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.
+Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.
 
@@
     for $i = 0,\dots,m-1$ \\+
-        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{FD for Jacobian diagonal entry} \\
+        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{Jacobian diagonal entry} \\
         if $\alpha_i > 0$ \\+
@@
 \end{pseudo*}
-\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.
-Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of \pr{pngs\_slow}.}
+\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.}
 \label{pc:pngsslow}
@@
 \end{equation}
-Suppose, for the next two formulas, that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
-Then by \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute with local integrals
+Recall that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
+By \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute
 \begin{equation}
@@
 
-Clearly, any smoother application, which is a sweep over all nodes in $\mathcal{T}$, will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak}, namely solving with the geometry determined by the current surface elevation $s^h$.
-However, formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated are prohibitively expensive because a Stokes solve is required  for each mesh node.
-Thus we seek a less expensive way to estimate the surface velocity for a perturbed surface elevation with a small ``bump'' $\eps\psi_i$.
-The change to the nonlinearities in this Glen-Stokes problem should be small, and it would seem that the most important change to the velocity and pressure fields from a small bump perturbation is through the perturbed gravitational load applied to the linearized problem around the state $s^h$.
-That is, we suppose other effect such as a perturbed normal direction on the surface are significantly smaller.
+Clearly, any smoother application, which is a sweep over all nodes in $\mathcal{T}$, will require at least one Glen-Stokes velocity solution, namely from solving \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation $s^h$.
+Formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated require a Stokes solution $\bu|_{s^h+\eps\psi_i}$ for each mesh node.
+
+However, we may estimate the surface velocity for a such perturbed surface elevation, namely with a small ``bump'' $\eps\psi_i$, by replacing the geometrical bump with a local perturbation of the body force.
+That is, we may perturb the gravitational load relative to the solution which gave $s^h$, but without changing the geometry.
+The change to the nonlinearities in this Glen-Stokes problem should be small, and the most important change to the velocity and pressure fields from a small bump perturbation is through its weight.
+We suppose that other effects, such as a perturbed normal direction on the surface, are significantly smaller.
 
@@
 NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
+PNGS & projected, nonlinear Gauss-Seidel (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pngsslow} \\
 PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\

### 2021-06-01 paper/msg.tex
A bit more MCD theory.

@@
 \begin{equation}
-\mathcal{D}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
+\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
 \end{equation}
-It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.
-We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{D}_+$ has small support.
+It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
+We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{W}_+$ has small support.
 
@@
 Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.
-Let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.
+Let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.
 Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.
-Note that $\psi_i \in \mathcal{D}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+Note that $\psi_i \in \mathcal{W}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
 We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.
@@
 
-In this section we propose a multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles of a new method descended from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).
-Our nonlinear variant of MCD, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
+In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.
+The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).
+Our nonlinear variant, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
 
-We need notation for multiple mesh levels.
-Suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, the coarse level.
-Suppose $\{\mathcal{T}^j\}_{j=0}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
-(If $d=1$ this becomes simple halving of each interval.)  We seek a solution of \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.
-Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and $m_j$ be the number of nodes in, and dimension of, $\mathcal{V}^j$.
+Regarding notation for multiple mesh levels, suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, and call it the coarse level.
+For $J\ge 0$, suppose $\{\mathcal{T}^j\}_{j=1}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
+(We assume simple halving of each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and let $\mathcal{V}_0^j \subset \mathcal{V}^j$ be the $P_1$ functions which are zero on $\partial\Omega$.
+If $m_j$ is the number of interior nodes in $\mathcal{T}$ then $\dim(\mathcal{V}_0^j)=m_j$.
 
+We seek an admissible solution to \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.
 Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.
-On this fine level the admissible surface elevations form a closed and convex set $\mathcal{L}^J = \{r^J \ge b^J, r^J|_{\partial\Omega} = b^J|_{\partial\Omega}\} \subset \mathcal{V}^J$.
-However, instead of defining such admissible sets on each level, presumably via coarse interpolants of $b^J$, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition starting from the defect obstacle for the current fine-level iterate $s^J$.
-That is, suppose FIXME
+On this fine level the admissible surface elevations form a closed and convex set
+\begin{equation}
+\mathcal{K}^J = \{r^J \in \mathcal{V}^J \,:\, r^J \ge b^J, r^J|_{\partial\Omega} = b^J|_{\partial\Omega}\}.  \label{eq:singleadmissible}
+\end{equation}
+However, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition of problem \eqref{eq:fesigpweakform} based on a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.
+That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
+\begin{equation}
+\chi^J = b^J - s^J \in \mathcal{V}_0^J \label{eq:finedefectobstacle}
+\end{equation}
+be the associated defect obstacle, a nonpositive function on $\Omega$.
+Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.
 
-FIXME Figure \ref{fig:decompclassical} shows such a decomposition
+Also following \cite{GraeserKornhuber2009}, we define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$.
+Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and denote the level-$j$ nodes as $(x_i^j,y_i^j)$.
+If $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ for coefficients $z^j[i]\in \RR$ then
+\begin{equation}
+\mR z^j = \sum_{\ell=0}^{m_{j-1}-1} \max\left\{z^j[i] \,:\,\psi_\ell^{j-1}(x_i^j,y_i^j) \right\}\,\psi_\ell^{j-1}.  \label{eq:monotonerestriction}
+\end{equation}
+In words, the coefficients for $\mR z^j \in \mathcal{V}_0^j$ are computed by maximizing the finer-level nodal values over the (open) supports of each each coarser-level hat function.
+
+Now we can recursively define a defect obstacle on each level:
+\begin{equation}
+\chi^{j-1} = \mR \chi^j  \label{eq:recursivedefectobstacle}
+\end{equation}
+for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences of the defect obstacles,
+\begin{equation}
+\phi^j = \chi^j - \chi^{j-1}.  \label{eq:downobstacles}
+\end{equation}
+In terms of these functions, for $j=0,\dots,J$ we define down-constraint sets
+\begin{equation}
+\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}.  \label{eq:downconstraintset}
+\end{equation}
+and up-constraint sets
+\begin{equation}
+\mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}.  \label{eq:upconstraintset}
+\end{equation}
+This terminology of down- and up-constraint sets is explained by noting that the smoother will solve one-dimensional VIs on each level, as described in the previous section, where the admissible sets are different as one goes down and then up in a V-cycle.
 
@@
 \end{center}
-\caption{An example $d=1$ and $J=4$ MCD decomposition of the defect obstacle $\chi^J$ for an iterate $s^J$.}
+\caption{An example ($d=1$, $J=4$) of an MCD decomposition of the defect obstacle $\chi^J$.
+Defect obstacles $\chi^j$ are generated using the monotone restriction operator $\mR$ as in \eqref{eq:recursivedefectobstacle}.}
 \label{fig:decompclassical}
@@
 
-Before describing the MCDN approach we should first be clear about when an iteration has converged.
+FIXME MCD theory
+
+FIXME DEBUG THIS PSEUDOCODE nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}
+
+\begin{pcode}[ht]
+\begin{pseudo*}
+\pr{mcdn-vcycle}(J,s^J,b^J,\ell^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
+    $\chi^J = b^J - s^J$ \\
+    $g^J = s^J$ \\
+    for $j=J$ downto $j=1$ \\+
+      $\chi^{j-1} = \mR \chi^j$ \\
+      $\phi^j = \chi^j - \chi^{j-1}$ \\
+      $y^j = 0$ \\
+      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,g^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\
+      $f = F^j(g^j+y^j) - \ell^j$ \\
+      $g^{j-1} = \iR(g^j + y^j)$ \\
+      $\ell^{j-1} = F^{j-1}(g^{j-1}) - R f$ \\-
+    $y^0 = 0$ \\
+    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,g^0,\ell^0,\chi^0)$ \qquad \ct{smoothing in $\mathcal{U}^0$} \\
+    $z^0 = y^0$ \\
+    for $j=1$ to $j=J$ \\+
+      $z^j = P z^{j-1} + y^{j}$ \\
+      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,g^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{U}^j$} \\-
+    return $z^J$
+\end{pseudo*}
+\caption{FIXME Nonlinear MCD V-cycle.}
+\label{pc:mcdn-vcycle}
+\end{pcode}
+
+Finally, we need to be clear about when an iteration has converged.
 For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
@@
 \label{pc:mcdn-solver}
-\end{pcode}
-
-FIXME MCD theory
-
-FIXME nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}
-
-\begin{pcode}[ht]
-\begin{pseudo*}
-\pr{mcdn-vcycle}(J,w^J,N,\ell^J,\chi^J,\id{down}=1,\id{coarse}=1,\id{up}=0)\text{:} \\+
-    $g^J = w^J$ \\
-    for $j=J$ downto $j=1$ \\+
-      $\chi^{j-1} = \mR \chi^j$ \\
-      $\phi^j = \chi^j - P\chi^{j-1}$ \\
-      $y^j = 0$ \\
-      $\text{\pr{smoother}}^{\text{\id{down}}}(j,g^j,y^j,N^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{K}^j$} \\
-      $F = N^j(g^j+y^j) - \ell^j$ \\
-      $g^{j-1} = \iR(g^j + y^j)$ \\
-      $\ell^{j-1} = N^{j-1}(g^{j-1}) - R F$ \\-
-    $y^0 = 0$ \\
-    $\text{\pr{smoother}}^{\text{\id{coarse}}}(0,g^0,y^0,N^0,\ell^0,\chi^0)$ \\
-    $z^0 = y^0$ \\
-    for $j=1$ to $j=J$ \\+
-      $z^j = P z^{j-1} + y^{j}$ \\
-      $\text{\pr{smoother}}^{\text{\id{up}}}(j,g^j,z^j,N^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\-
-    return $z^J$
-\end{pseudo*}
-\caption{FIXME Nonlinear MCD V-cycle.}
-\label{pc:mcdn-vcycle}
 \end{pcode}
@@
 \begin{longtable}{l|l|l}
+\caption{Glossary of acronyms used in this paper.}
+\label{tab:acronyms} \\ % \\ REQUIRED HERE
 \toprule
@@
 \bottomrule
-\caption{Glossary of acronyms used in this paper.}
-\label{tab:acronyms}
 \end{longtable}

### 2021-06-01 talk/talk.tex
Skeleton of talk, but the title slide looks good.

\documentclass{beamer}

\usetheme{focus} % see https://github.com/elauksap/focus-beamertheme
% Add option [numbering=none] to disable the footer progress bar
% Add option [numbering=fullbar] to show the footer progress bar as always full with a slide count

\usepackage{booktabs} % Required for better table rules

\title{Evolution of Ice Sheet Geometry using Stokes Dynamics}

%\subtitle{Subtitle}

\author{Ed Bueler}

\titlegraphic{\includegraphics[width=0.55\textwidth]{figs/extruded.png} \\ \includegraphics[width=0.18\textwidth]{figs/uafbw.png}}

\institute{University of Alaska Fairbanks}

\date{\phantom{foo} \bigskip \bigskip \\ SIAM GS21 \\ with assistance from Lawrence Mitchell}

\begin{document}

\begin{frame}
	\maketitle
\end{frame}

\begin{frame}
	\tableofcontents
\end{frame}

\section{Section 1}

\begin{frame}{Typesetting and Math}
	The packages \texttt{inputenc} and \texttt{FiraSans}\footnote{\url{https://fonts.google.com/specimen/Fira+Sans}}\textsuperscript{,}\footnote{\url{http://mozilla.github.io/Fira/}} are used to properly set the main fonts.
	\vfill
	This theme provides styling commands to typeset \emph{emphasized}, \alert{alerted}, \textbf{bold}, \textcolor{example}{example text}, \dots
	\vfill
	\texttt{FiraSans} also provides support for mathematical symbols \cite{Bueler2021conservation}:
	\begin{equation*}
		e^{i\pi} + 1 = 0.
	\end{equation*}
\end{frame}

\section{Section 2}

\begin{frame}{Blocks}
	These blocks are part of 1 slide, to be displayed consecutively.
	\begin{block}{Block}
		Text.
	\end{block}
	\pause % Automatically creates a new "page" split between the above and above + below
	\begin{alertblock}{Alert block}
		Alert \alert{text}.
	\end{alertblock}
	\pause % Automatically creates a new "page" split between the above and above + below
	\begin{exampleblock}{Example block}
		Example \textcolor{example}{text}.
	\end{exampleblock}
\end{frame}

\begin{frame}{Columns}
	\begin{columns}
		\column{0.5\textwidth}
			This text appears in the left column and wraps neatly with a margin between columns.

		\column{0.5\textwidth}
			\includegraphics[width=\linewidth]{figs/extruded.png}
	\end{columns}
\end{frame}

\appendix

\begin{frame}{References}
	%\nocite{*} % Display all references regardless of if they were cited
	\bibliography{../paper/msg.bib}
	\bibliographystyle{plain}
\end{frame}

\end{document}

### 2021-06-01 talk/talk.tex
Start on slides.

@@
 \usepackage{booktabs} % Required for better table rules
+\usepackage{bm,times}
+
+\newcommand{\eps}{\epsilon}
+\newcommand{\RR}{\mathbb{R}}
+
+\newcommand{\grad}{\nabla}
+\newcommand{\Div}{\nabla\cdot}
+\newcommand{\trace}{\operatorname{tr}}
+
+\newcommand{\hbn}{\hat{\mathbf{n}}}
+
+\newcommand{\bb}{\mathbf{b}}
+\newcommand{\be}{\mathbf{e}}
+\newcommand{\bbf}{\mathbf{f}}
+\newcommand{\bg}{\mathbf{g}}
+\newcommand{\bn}{\mathbf{n}}
+\newcommand{\br}{\mathbf{r}}
+\newcommand{\bu}{\mathbf{u}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bw}{\mathbf{w}}
+\newcommand{\bx}{\mathbf{x}}
+
+\newcommand{\bF}{\mathbf{F}}
+\newcommand{\bV}{\mathbf{V}}
+\newcommand{\bX}{\mathbf{X}}
+
+\newcommand{\bxi}{\bm{\xi}}
+
+\newcommand{\bzero}{\bm{0}}
+
+\newcommand{\rhoi}{\rho_{\text{i}}}
+
+\newcommand{\ip}[2]{\left(#1,#2\right)}
+
+\newcommand{\mR}{R^{\bm{\oplus}}}
+\newcommand{\iR}{R^{\bullet}}
+
+\newcommand{\pp}{{\text{p}}}
+\newcommand{\qq}{{\text{q}}}
+\newcommand{\rr}{{\text{r}}}
+
+\newcommand{\bus}{\bu|_s}
 
@@
 
-\begin{frame}
-	\tableofcontents
+\begin{frame}{the steady ice geometry problem (SIGP)}
+
+\vspace{-2mm}
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figs/stokesdomain.png}
+\end{center}
+
+\vspace{-1mm}
+\begin{itemize}
+\item a non-shallow glacier interacting with the climate
+    \begin{itemize}
+    \item the simplest such problem (nonsliding, isothermal)
+    \item re~\emph{steady}: solve it and you can handle implicit steps too
+    \end{itemize}
+\item data given on fixed $\Omega \subset \RR^2$:
+    \begin{itemize}
+    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{steady climate}
+    \item bed elevation $b(x,y)$
+    \end{itemize}
+\item \emph{determine surface elevation $s(x,y)$ on $\Omega$ and domain $\Lambda_s \subset \RR^3$}
+\end{itemize}
 \end{frame}
 
-\section{Section 1}
+\begin{frame}{SIGP strong form}
 
-\begin{frame}{Typesetting and Math}
-	The packages \texttt{inputenc} and \texttt{FiraSans}\footnote{\url{https://fonts.google.com/specimen/Fira+Sans}}\textsuperscript{,}\footnote{\url{http://mozilla.github.io/Fira/}} are used to properly set the main fonts.
-	\vfill
-	This theme provides styling commands to typeset \emph{emphasized}, \alert{alerted}, \textbf{bold}, \textcolor{example}{example text}, \dots
-	\vfill
-	\texttt{FiraSans} also provides support for mathematical symbols \cite{Bueler2021conservation}:
-	\begin{equation*}
-		e^{i\pi} + 1 = 0.
-	\end{equation*}
+\begin{itemize}
+\item \alert{a nonlinear complementarity problem (NCP) coupled to the Glen-Stokes model for ice velocity $\bu$:}
+
+\vspace{-2mm}
+\begin{align*}
+s - b &\ge 0 && \text{on $\Omega$} \\
+- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
+(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} \\
+- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+\nabla \cdot \bu &= 0 && \text{''} \\
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \\
+\bu &= \bzero && \text{on $\Gamma_0$ (ice base)}
+\end{align*}
+
+    \begin{itemize}
+    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1=4/3$:
+      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
+    \item surface kinematical equation holds \emph{on the ice}: \quad $\bu|_s \cdot \bn_s + a=0$
+    \end{itemize}
+\end{itemize}
 \end{frame}
 
-\section{Section 2}
+\begin{frame}{what do existing ice sheet models do?}
 
-\begin{frame}{Blocks}
-	These blocks are part of 1 slide, to be displayed consecutively.
-	\begin{block}{Block}
-		Text.
-	\end{block}
-	\pause % Automatically creates a new "page" split between the above and above + below
-	\begin{alertblock}{Alert block}
-		Alert \alert{text}.
-	\end{alertblock}
-	\pause % Automatically creates a new "page" split between the above and above + below
-	\begin{exampleblock}{Example block}
-		Example \textcolor{example}{text}.
-	\end{exampleblock}
+FIXME screenshot from Seguinot
+
+\begin{itemize}
+\item almost no one is addressing the SIGP directly
+    \begin{itemize}
+    \item except \cite{WirbelJarosch2020}, but nonscalably and with ``fake ice'' extension
+    \item note scalable solution of fixed-geometry Glen-Stokes by Isaac and others \cite{IsaacStadlerGhattas2015}
+    \end{itemize}
+\item \emph{warning}: this free boundary problem is a barrier to exact numerical mass conservation, regardless of dynamical and/or numerical choices \cite{Bueler2021conservation}
+\item what are people doing?  \emph{explicit time-stepping with $s \ge b$ enforced by truncation}
+\item \alert{show movie clip}
+\end{itemize}
 \end{frame}
 
-\begin{frame}{Columns}
-	\begin{columns}
-		\column{0.5\textwidth}
-			This text appears in the left column and wraps neatly with a margin between columns.
+\begin{frame}{new method}
 
-		\column{0.5\textwidth}
-			\includegraphics[width=\linewidth]{figs/extruded.png}
-	\end{columns}
+\begin{itemize}
+\item FIXME \cite{BuelerMitchell2022}
+\end{itemize}
 \end{frame}
+
+%\begin{block}{Block}
+%\begin{alertblock}{Alert block}
+%	\pause % Automatically creates a new "page" split between the above and above + below
+%	\begin{exampleblock}{Example block}
+
+%\begin{frame}{Columns}
+%	\begin{columns}
+%		\column{0.5\textwidth}
+%			This text
+%		\column{0.5\textwidth}
+%			\includegraphics[width=\linewidth]{figs/extruded.png}
+%	\end{columns}
+%\end{frame}
 
@@
 \begin{frame}{References}
-	%\nocite{*} % Display all references regardless of if they were cited
 	\bibliography{../paper/msg.bib}

### 2021-06-01 paper/genfigs/stokesdomain.py

@@ -50,7 +50,7 @@ plt.text(x[650], b[650] - 0.5, r'$b$', fontsize=bigfsize, color='k')
 # show \Omega
 yR = min(b) - 0.5
 plt.plot([min(x),max(x)],[yR,yR],color='k',lw=1.0)
-plt.text(4.0,yR-0.4,r'$\Omega$',fontsize=fsize)
+plt.text(x[875],yR+0.2,r'$\Omega$',fontsize=fsize)
 plt.axis([0.0,10.0,yR-0.8,4.5])
 plt.axis('off')
 writeout('stokesdomain.pdf')

### 2021-06-01 talk/talk.tex
A bit more on slides.

@@
 
-\title{Evolution of Ice Sheet Geometry using Stokes Dynamics}
+\title{Evolution of \\ ice sheet geometry \\ using Stokes dynamics}
 
@@
 
-\titlegraphic{\includegraphics[width=0.55\textwidth]{figs/extruded.png} \\ \includegraphics[width=0.18\textwidth]{figs/uafbw.png}}
+\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/extruded.png} \\ \vspace{-3mm} \includegraphics[width=0.15\textwidth]{figs/uafbw.png}}
 
-\institute{University of Alaska Fairbanks}
+%\institute{University of Alaska Fairbanks}
 
-\date{\phantom{foo} \bigskip \bigskip \\ SIAM GS21 \\ with assistance from Lawrence Mitchell}
+\date{\phantom{foo} \bigskip \bigskip \bigskip \\ SIAM GS21 \\ with assistance from Lawrence Mitchell}
 
@@
     \begin{itemize}
-    \item the simplest such problem (nonsliding, isothermal)
-    \item re~\emph{steady}: solve it and you can handle implicit steps too
+    \item only the simplest such problem (nonsliding, isothermal)
+    \item re~\emph{steady}: solve this and you can handle implicit steps too
     \end{itemize}
@@
     \end{itemize}
-\item \emph{determine surface elevation $s(x,y)$ on $\Omega$ and domain $\Lambda_s \subset \RR^3$}
+\item determine: surface elevation $s(x,y)$ on $\Omega$ and domain $\Lambda_s \subset \RR^3$
 \end{itemize}
@@
 
-\begin{frame}{what do existing ice sheet models do?}
-
-FIXME screenshot from Seguinot
+\begin{frame}{existing ice sheet models}
 
 \begin{itemize}
-\item almost no one is addressing the SIGP directly
+\item almost no one is solving the SIGP directly, or implicit time-stepping
     \begin{itemize}
-    \item except \cite{WirbelJarosch2020}, but nonscalably and with ``fake ice'' extension
-    \item note scalable solution of fixed-geometry Glen-Stokes by Isaac and others \cite{IsaacStadlerGhattas2015}
+    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} \dots nonscalable, ``fake ice''
+    \item optimal fixed-geometry Stokes solver (Isaac and others \cite{IsaacStadlerGhattas2015})
     \end{itemize}
-\item \emph{warning}: this free boundary problem is a barrier to exact numerical mass conservation, regardless of dynamical and/or numerical choices \cite{Bueler2021conservation}
-\item what are people doing?  \emph{explicit time-stepping with $s \ge b$ enforced by truncation}
-\item \alert{show movie clip}
+%\item \emph{warning}: barrier to exact numerical mass conservation \cite{Bueler2021conservation}
+\item what are people doing?
+    \begin{itemize}
+    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation} (below)
     \end{itemize}
+\end{itemize}
+
+\medskip
+\begin{columns}
+    \column{0.8\textwidth}
+        \hfill \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
+    \column{0.2\textwidth}
+        \scriptsize \emph{show clip from Julien Seguinot using PISM}
+\end{columns}
 \end{frame}

### 2021-06-02 talk/talk.tex
Progress slides.

@@
 \usepackage{bm,times}
+
+\usepackage{tikz}
+\usetikzlibrary{decorations.pathreplacing,calc}
+
+\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
 
@@
 \begin{itemize}
-\item \alert{a nonlinear complementarity problem (NCP) coupled to the Glen-Stokes model for ice velocity $\bu$:}
+\item \alert{a nonlinear complementarity problem (NCP) coupled to the Glen-Stokes model for $\bu,p$:}
+
+\vspace{-5mm}
+
+\begin{align*}
+s - b &\ge 0 && \text{on $\Omega$} & \tikzmark{first} \\
+- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
+(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} & \tikzmark{second} \\
+- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+\nabla \cdot \bu &= 0 && \text{''} \\
+\bu &= \bzero && \text{on $\Gamma_0$ \quad (ice base)} \\
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
+\end{align*}
+
+\begin{tikzpicture}[overlay, remember picture]
+\draw[decoration={brace,amplitude=0.4em},decorate,ultra thick] (first.north east) node[right=8mm,below=3mm] {NCP} -- (second.east);
+\end{tikzpicture}
 
 \vspace{-2mm}
-\begin{align*}
-s - b &\ge 0 && \text{on $\Omega$} \\
-- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
-(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} \\
-- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
-\nabla \cdot \bu &= 0 && \text{''} \\
-\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \\
-\bu &= \bzero && \text{on $\Gamma_0$ (ice base)}
-\end{align*}
 
     \begin{itemize}
+    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \quad (normal to ice surface; upward)
+    \item surface kinematical equation holds \emph{on the ice}: \quad $\bu|_s \cdot \bn_s + a=0$
     \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1=4/3$:
       $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
-    \item surface kinematical equation holds \emph{on the ice}: \quad $\bu|_s \cdot \bn_s + a=0$
     \end{itemize}
@@
 \begin{itemize}
-\item almost no one is solving the SIGP directly, or implicit time-stepping
+\item almost no one is solving the SIGP directly, or doing implicit time-stepping
     \begin{itemize}
-    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} \dots nonscalable, ``fake ice''
-    \item optimal fixed-geometry Stokes solver (Isaac and others \cite{IsaacStadlerGhattas2015})
+    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} \dots nonscalable, w.~``fake ice''
     \end{itemize}
-%\item \emph{warning}: barrier to exact numerical mass conservation \cite{Bueler2021conservation}
 \item what are people doing?
@@
     \column{0.2\textwidth}
-        \scriptsize \emph{show clip from Julien Seguinot using PISM}
+        \scriptsize \emph{show clip of PISM run by Julien Seguinot}
 \end{columns}
+\end{frame}
+
+\begin{frame}{goals}
+
+\begin{itemize}
+\item FIXME
+\end{itemize}
+\end{frame}
+
+\begin{frame}{ice dynamics operator}
+
+\begin{itemize}
+\item FIXME
+\item optimal fixed-geometry Stokes solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
+\end{itemize}
 \end{frame}
@@
 \end{itemize}
+\end{frame}
+
+\begin{frame}{extra}
+ NCP is a barrier to exact numerical mass conservation \cite{Bueler2021conservation}
 \end{frame}
@@
 \begin{frame}{References}
-	\bibliography{../paper/msg.bib}
-	\bibliographystyle{plain}
+
+%\bibliography{../paper/msg.bib}
+%\bibliographystyle{plain}
+
+\begin{thebibliography}{1}
+\bibitem{Bueler2021conservation}
+E.~Bueler. \textbf{Conservation laws for free-boundary fluid layers.} {\em SIAM J. Appl. Math.}, to appear.
+
+\bibitem{BuelerMitchell2022}
+E.~Bueler and L.~Mitchell. \textbf{Multilevel computation of glacier geometry from Stokes dynamics.} in preparation
+
+\bibitem{IsaacStadlerGhattas2015}
+T.~Isaac, G.~Stadler, and O.~Ghattas.
+\textbf{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics.} {\em SIAM J.
+Sci.
+Comput.}, 37(6):B804--B833, 2015.
+
+\bibitem{WirbelJarosch2020}
+A.~Wirbel and A.~H.
+Jarosch.
+\textbf{Inequality-constrained free-surface evolution in a full {S}tokes ice flow model (\textit{evolve\_glacier v1.1}).} {\em Geoscientific Model Development}, 13(12):6425--6445, 2020.
+\end{thebibliography}
+
 \end{frame}

### 2021-06-02 talk/talk.tex
more slides

@@
 
-\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/extruded.png} \\ \vspace{-3mm} \includegraphics[width=0.15\textwidth]{figs/uafbw.png}}
+\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/titleextruded.png} \\ \vspace{-3mm} \includegraphics[width=0.15\textwidth]{figs/uafbw.png}}
 
@@
 
+\begin{frame}{surface kinematical equation}
+
+\begin{block}{steady surface kinematical equation (SKE)}
+$$\bu|_s \cdot \bn_s + a=0$$
+\end{block}
+
+\begin{itemize}
+\item $\bn_s = \left<-s_x,-s_y,1\right>$  \quad (normal to ice surface; upward)
+\item \emph{on} the ice surface, in steady-state there is balance (SKE) between motion and climatic mass balance
+\item \emph{off} the ice we know $a\le 0$ in steady state
+\item for all of this talk: $s=b$ where ice is not present
+\end{itemize}
+\end{frame}
+
 \begin{frame}{SIGP strong form}
@@
 \begin{itemize}
-\item \alert{a nonlinear complementarity problem (NCP) coupled to the Glen-Stokes model for $\bu,p$:}
+\item \alert{a nonlinear complementarity problem (NCP) for $s$ coupled to Stokes problem for $\bu,p$:}
 
@@
 \begin{align*}
-s - b &\ge 0 && \text{on $\Omega$} & \tikzmark{first} \\
+s - b &\ge 0 && \text{on $\Omega$} & \tikzmark{ncptop} \\
 - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
-(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} & \tikzmark{second} \\
-- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} & \tikzmark{ncpbot} \\
+- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} & \tikzmark{gstop} \\
 \nabla \cdot \bu &= 0 && \text{''} \\
 \bu &= \bzero && \text{on $\Gamma_0$ \quad (ice base)} \\
-\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} & \tikzmark{gsbot}
 \end{align*}
@@
 \begin{tikzpicture}[overlay, remember picture]
-\draw[decoration={brace,amplitude=0.4em},decorate,ultra thick] (first.north east) node[right=8mm,below=3mm] {NCP} -- (second.east);
+\draw[decoration={brace,amplitude=0.4em},decorate,ultra thick] ([xshift=-7mm]ncptop.north east) node[right=8mm,below=3mm] {NCP} -- ([xshift=-7mm]ncpbot.east);
 \end{tikzpicture}
 
-\vspace{-2mm}
+\begin{tikzpicture}[overlay, remember picture]
+\draw[decoration={brace,amplitude=0.4em},decorate,ultra thick] ([xshift=-7mm]gstop.north east) node[right=10mm,below=6mm] {Stokes} -- ([xshift=-7mm]gsbot.east);
+\end{tikzpicture}
+
+\vspace{-8mm}
 
     \begin{itemize}
-    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \quad (normal to ice surface; upward)
-    \item surface kinematical equation holds \emph{on the ice}: \quad $\bu|_s \cdot \bn_s + a=0$
-    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1=4/3$:
+    \item plus Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
       $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
     \end{itemize}
+\item SIGP is a free-boundary problem for $s$ defined on $\Omega$
 \end{itemize}
@@
     \end{itemize}
-\item what are people doing?
+\item what are people doing instead?
     \begin{itemize}
@@
 
-\begin{frame}{goals}
+\begin{frame}{coupling via operator}
+
+\begin{block}{ice dynamics operator}
+$\Phi : s \mapsto \bus \cdot \bn_s$
+
+\vspace{-3mm}
+
+\hfill \includegraphics[width=0.75\textwidth]{figs/idoaction.png}
+\end{block}
 
 \begin{itemize}
-\item FIXME
+\item to compute $\Phi(s)$, solve Stokes (weak form) problem
+    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0$$
+over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ and extract trace $\bus$
+
+\medskip
+    \begin{itemize}
+    \item optimal solver already exists for this problem (Isaac et al \cite{IsaacStadlerGhattas2015})
     \end{itemize}
+\item extend by zero so $\Phi(s)$ is defined on all of $\Omega$
+\end{itemize}
+
 \end{frame}
 
-\begin{frame}{ice dynamics operator}
+\begin{frame}{SIGP weak form is a variational inequality}
 
 \begin{itemize}
-\item FIXME
-\item optimal fixed-geometry Stokes solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
+\item using $\Phi$ gives clean NCP over $\Omega$:
+\begin{align*}
+s - b &\ge 0 \\
+- \Phi(s) - a &\ge 0 \\
+(s - b) (- \Phi(s) - a) &= 0
+\end{align*}
+
+\item well-known fact: \quad  NCP (strong form) $\leftrightarrow$ VI (weak form)
+\item find $s \in \mathcal{K} = \{r \ge b\}$ so that
+    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s} \hspace{10mm} (\ast)$$
+for all $r \in \mathcal{K}$, where
+    $$F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$$
+    \begin{itemize}
+    \item whether VI problem $(\ast)$ is well-posed is open, but the SIA version has an existence proof (Jouvet \& Bueler \cite{JouvetBueler2012})
+    \end{itemize}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{goal: robust and optimal SIGP solver}
+
+\begin{block}{goal}
+given triangulation $\mathcal{T}$ of $\Omega$, and $P_1$ FE space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, solve SIGP VI
+
+$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h}$$
+
+for admissible $s^h$ using solutions of Glen-Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s)=\bu|_s\cdot\bn_s$ and $F(s)[w] =- \int_\Omega \Phi(s)\, w $
+\end{block}
+
+\begin{itemize}
+\item seeking a solver which is:
+    \begin{itemize}
+    \item \alert{robust}
+    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
+    \end{itemize}
+\item optimal requires multigrid
 \end{itemize}
@@
 \end{itemize}
+
+\includegraphics[width=0.8\textwidth]{figs/extruded.png}
 \end{frame}
@@
 
+\bibitem{JouvetBueler2012}
+G.~Jouvet and E.~Bueler. \textbf{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation.} {\em SIAM J. Appl. Math.}, 72(4):1292--1314, 2012.
+
 \bibitem{WirbelJarosch2020}

### 2021-06-02 talk/talk.tex
A bit more.

@@
 \begin{itemize}
-\item a non-shallow glacier interacting with the climate
+\item a glacier which is in balance with its climate
     \begin{itemize}
-    \item only the simplest such problem (nonsliding, isothermal)
+    \item the simplest such problem (grounded, nonsliding, isothermal)
     \item re~\emph{steady}: solve this and you can handle implicit steps too
     \end{itemize}
-\item data given on fixed $\Omega \subset \RR^2$:
+\item steady climate data given on fixed $\Omega \subset \RR^2$:
     \begin{itemize}
-    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{steady climate}
+    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
     \item bed elevation $b(x,y)$
     \end{itemize}
-\item determine: surface elevation $s(x,y)$ on $\Omega$ and domain $\Lambda_s \subset \RR^3$
+\item determine: surface elevation $s(x,y)$ on $\Omega$, domain $\Lambda_s \subset \RR^3$
 \end{itemize}
@@
 \item \emph{off} the ice we know $a\le 0$ in steady state
-\item for all of this talk: $s=b$ where ice is not present
+\item for all of this talk:
+    \begin{itemize}
+    \item $s$ is assumed to be well-defined (no overhangs)
+    \item $s=b$ where ice is not present, so $s$ is defined on all of $\Omega$
+    \end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item \alert{a nonlinear complementarity problem (NCP) for $s$ coupled to Stokes problem for $\bu,p$:}
+\item the SIGP is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}
 
@@
 \begin{itemize}
-\item using $\Phi$ gives clean NCP over $\Omega$:
+\item using $\Phi$ gives a cleaner NCP over $\Omega$:
 \begin{align*}
@@
 \item well-known fact: \quad  NCP (strong form) $\leftrightarrow$ VI (weak form)
-\item find $s \in \mathcal{K} = \{r \ge b\}$ so that
-    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s} \hspace{10mm} (\ast)$$
-for all $r \in \mathcal{K}$, where
-    $$F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$$
+\item VI weak form: find $s \in \mathcal{K} = \{r \ge b\}$ so that
+    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
+for all $r \in \mathcal{K}$, where $F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$
+
+\medskip
     \begin{itemize}
-    \item whether VI problem $(\ast)$ is well-posed is open, but the SIA version has an existence proof (Jouvet \& Bueler \cite{JouvetBueler2012})
+    \item whether this VI is well-posed is open, but the SIA version has an existence proof (Jouvet \& Bueler \cite{JouvetBueler2012})
     \end{itemize}
@@
 
-\begin{block}{goal}
-given triangulation $\mathcal{T}$ of $\Omega$, and $P_1$ FE space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, solve SIGP VI
+\begin{block}{numerical goal}
+Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.  Solve VI
 
-$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h}$$
+$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
 
-for admissible $s^h$ using solutions of Glen-Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s)=\bu|_s\cdot\bn_s$ and $F(s)[w] =- \int_\Omega \Phi(s)\, w $
+to give admissible $s^h \in \mathcal{K}^h$.  Use the solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$ (and thus $F(s^h)$).
 \end{block}
@@
 \item optimal requires multigrid
+\item<2> \emph{for now these are aspirations only, but let's give it a shot}
 \end{itemize}

### 2021-06-03 talk/talk.tex
Fiddle with talk, and add some stuff.

@@
-\documentclass{beamer}
+\documentclass[usepdftitle=false]{beamer}
 
@@
 \usetikzlibrary{decorations.pathreplacing,calc}
+
+\usepackage{hyperref}
+\hypersetup{
+    pdftitle={Evolution of ice sheet geometry using Stokes dynamics}
+}
 
@@
     \end{itemize}
-\item steady climate data given on fixed $\Omega \subset \RR^2$:
+\item steady data given on fixed $\Omega \subset \RR^2$:
     \begin{itemize}
@@
 
-\begin{block}{steady surface kinematical equation (SKE)}
-$$\bu|_s \cdot \bn_s + a=0$$
-\end{block}
+\begin{itemize}
+\item glaciers flow, so there is an equation for the surface elevation
+\item the surface kinematical equation (SKE) holds \emph{on} the ice
+    $$\bu|_s \cdot \bn_s + a=0 \qquad \text{\emph{steady state}}$$
 
     \begin{itemize}
-\item $\bn_s = \left<-s_x,-s_y,1\right>$  \quad (normal to ice surface; upward)
-\item \emph{on} the ice surface, in steady-state there is balance (SKE) between motion and climatic mass balance
+    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
+    \end{itemize}
 \item \emph{off} the ice we know $a\le 0$ in steady state
+\end{itemize}
+
+\bigskip
+\begin{columns}
+    \column{0.6\textwidth}
+        \begin{itemize}
         \item for all of this talk:
@@
         \end{itemize}
+    \column{0.4\textwidth}
+        \includegraphics[width=1.05\textwidth]{figs/simpledomain.png}
+\end{columns}
 \end{frame}
@@
 \begin{itemize}
-\item the SIGP is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}
+\item the SIGP free-boundary problem determines $s$ on $\Omega$
+\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}
 
@@
       $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
+    \item \emph{count the 3 nonlinearities}
     \end{itemize}
-\item SIGP is a free-boundary problem for $s$ defined on $\Omega$
 \end{itemize}
@@
 \begin{itemize}
-\item almost no one is solving the SIGP directly, or doing implicit time-stepping
+\item almost no one is solving the SIGP directly
     \begin{itemize}
-    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} \dots nonscalable, w.~``fake ice''
+    \item or doing implicit time-stepping (\emph{basically the same as SIGP})
+    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} (\emph{semi-coupled, nonscalable, artificial ice layer})
     \end{itemize}
@@
     \begin{itemize}
-    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation} (below)
+    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
+    \item mostly with shallow approximations, e.g.~PISM run below
     \end{itemize}
@@
     \column{0.2\textwidth}
-        \scriptsize \emph{show clip of PISM run by Julien Seguinot}
+        \scriptsize \emph{show clip of Alpine run by Julien Seguinot}
 \end{columns}
@@
 
-\begin{block}{ice dynamics operator}
+\begin{block}{definition: ice dynamics operator}
 $\Phi : s \mapsto \bus \cdot \bn_s$
@@
 \begin{itemize}
-\item to compute $\Phi(s)$, solve Stokes (weak form) problem
+\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
     $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0$$
-over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ and extract trace $\bus$
 
-\medskip
     \begin{itemize}
-    \item optimal solver already exists for this problem (Isaac et al \cite{IsaacStadlerGhattas2015})
+    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz \cite{JouvetRappaz2011})
+    \item optimal solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
+    \item extract trace $\bus$, then $\Phi(s) = \bus \cdot \bn_s$, extend by zero to all of $\Omega$
     \end{itemize}
-\item extend by zero so $\Phi(s)$ is defined on all of $\Omega$
 \end{itemize}
@@
 
-\item well-known fact: \quad  NCP (strong form) $\leftrightarrow$ VI (weak form)
-\item VI weak form: find $s \in \mathcal{K} = \{r \ge b\}$ so that
+\item NCP (strong form) $\leftrightarrow$ VI (weak form)
+
+\begin{block}{SIGP weak form}
+find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
     $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
 for all $r \in \mathcal{K}$, where $F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$
+\end{block}
 
-\medskip
+\item well-posedness is open
     \begin{itemize}
-    \item whether this VI is well-posed is open, but the SIA version has an existence proof (Jouvet \& Bueler \cite{JouvetBueler2012})
+    \item existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
     \end{itemize}
+
 \end{itemize}
@@
 \begin{block}{numerical goal}
-Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.  Solve VI
+Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.
+Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$.
+Solve VI for $s^h \in \mathcal{K}^h$:
 
 $$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
-
-to give admissible $s^h \in \mathcal{K}^h$.  Use the solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$ (and thus $F(s^h)$).
 \end{block}
@@
     \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] optimal requires multigrid
         \end{itemize}
-\item optimal requires multigrid
-\item<2> \emph{for now these are aspirations only, but let's give it a shot}
+    \item<2> \emph{for now these are aspirations \dots}
+    \end{itemize}
 \end{itemize}
@@
 
-\begin{frame}{new method}
+\begin{frame}{what thing is $\Phi$?}
+
+\begin{block}{SIGP weak form}
+find $s \in \mathcal{K}$ so $F(s)[r - s] \ge \ip{a}{r-s}$ for all $r \in \mathcal{K}$
+\end{block}
+
+\begin{itemize}
+\item $F(s)[w] = - \int_\Omega \Phi(s)\, w$ is basically dual of $\Phi(s)=\bus\cdot\bn_s$
+\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$
+\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
+    \begin{itemize}
+    \item some kind of integral operator (\emph{convolve with Stokeslets})
+    \item $\Phi$ is concentrated near the diagonal, i.e.~short range
+    \end{itemize}
+\item yet it is a differential operator in the SIA case:
+\small
+    $$\Phi_{\text{SIA}}(s) = \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} + \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
+\normalsize
+\end{itemize}
+\end{frame}
+
+\begin{frame}{multilevel constraint decomposition}
 
@@
 \includegraphics[width=0.8\textwidth]{figs/extruded.png}
-\end{frame}
 
-\begin{frame}{extra}
- NCP is a barrier to exact numerical mass conservation \cite{Bueler2021conservation}
+\input{tikz/vcycle.tex}
 \end{frame}
@@
 \begin{frame}{References}
-
 %\bibliography{../paper/msg.bib}
@@
 \begin{thebibliography}{1}
-\bibitem{Bueler2021conservation}
-E.~Bueler. \textbf{Conservation laws for free-boundary fluid layers.} {\em SIAM J. Appl. Math.}, to appear.
-
 \bibitem{BuelerMitchell2022}
@@
 \bibitem{IsaacStadlerGhattas2015}
-T.~Isaac, G.~Stadler, and O.~Ghattas.
-\textbf{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics.} {\em SIAM J.
-Sci.
-Comput.}, 37(6):B804--B833, 2015.
+T.~Isaac, G.~Stadler, and O.~Ghattas.  \textbf{Solution of nonlinear Stokes equations \dots with application to ice sheet dynamics.} {\em SIAM J. Sci. Comput.}, 37(6):B804--B833, 2015.
 
@@
 
+\bibitem{JouvetRappaz2011}
+G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary Stokes problem arising in glaciology.} {\em Adv. Numer. Anal.}, 2011.
+
 \bibitem{WirbelJarosch2020}
-A.~Wirbel and A.~H.
-Jarosch.
-\textbf{Inequality-constrained free-surface evolution in a full {S}tokes ice flow model (\textit{evolve\_glacier v1.1}).} {\em Geoscientific Model Development}, 13(12):6425--6445, 2020.
+A.~Wirbel and A.~H. Jarosch. \textbf{Inequality-constrained free-surface evolution in a full {S}tokes ice flow model \dots} {\em Geosci. Model Dev.}, 13(12):6425--6445, 2020.
 \end{thebibliography}
-
 \end{frame}
 
+\begin{frame}{extra}
+
+NCP is a barrier to exact numerical mass conservation\footnote{E.~Bueler. \textbf{Conservation laws for free-boundary fluid layers.} \em{SIAM J.~Appl.~Math.}, to appear.}
+\end{frame}
 \end{document}

### 2021-06-03 py/smoother.py
Add -monitor and clean-up stdout.

@@ -147,7 +147,7 @@ class SmootherStokes(SmootherObstacleProblem):
         firstcall = (self.basemesh == None)
         if firstcall:
             self.mx = mesh1d.m + 1
-            print('residual(): base mesh of %d elements (intervals)' \
+            print('mesh: base of %d elements (intervals)' \
                   % self.mx)
             self.basemesh = fd.IntervalMesh(self.mx, length_or_left=0.0,
                                             right=mesh1d.xmax)
@@ -160,7 +160,7 @@ class SmootherStokes(SmootherObstacleProblem):
                 'padding requires minimum positive thickness'
             mesh = fd.ExtrudedMesh(self.basemesh, mz, layer_height=1.0 / mz)
             if firstcall:
-                print('            extruded mesh: padded, %d x %d elements' \
+                print('      extruded is padded, has %d x %d elements' \
                       % (self.mx, mz))
         else:
             layermap = np.zeros((self.mx, 2), dtype=int)  # [[0,0], ..., [0,0]]
@@ -173,7 +173,7 @@ class SmootherStokes(SmootherObstacleProblem):
                                    layer_height=1.0 / mz)
             if firstcall:
                 icycount = sum(icyelement)
-                print('            extruded mesh: %d x %d icy elements (%d ice-free base elements)' \
+                print('      extruded has %d x %d icy elements and %d ice-free base elements' \
                       % (icycount, mz, self.mx - icycount))
 
         # generate extruded mesh of height s(x)
@@ -238,12 +238,15 @@ class SmootherStokes(SmootherObstacleProblem):
         if currentr is None:
             currentr = self.residual(mesh1d, s, ella)
         if self.args.smoother == 'richardson':
-            self.richardsonsweep(s, phi, currentr)
+            negd = self.richardsonsweep(s, phi, currentr)
         elif self.args.smoother == 'jacobislow':
-            self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
+            negd = self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
         elif self.args.smoother == 'gsslow':
-            self.gsslowsweep(mesh1d, s, ella, phi, currentr)
+            negd = self.gsslowsweep(mesh1d, s, ella, phi, currentr)
         mesh1d.WU += 1
+        if self.args.monitor and len(negd) > 0:
+            print('      negative diagonal entries: ', end='')
+            print(negd)
[…]

### 2021-06-03 py/steady.py

@@ -66,6 +66,8 @@ adda('-jcoarse', type=int, default=0, metavar='J',
      help='coarse mesh is jth level (default jcoarse=0 gives 1 node)')
 adda('-J', type=int, default=3, metavar='J',
      help='fine mesh is Jth level (default J=%(default)s)')
+adda('-monitor', action='store_true', default=False,
+     help='show CP residual norm at each step')
 adda('-mz', type=int, default=4, metavar='MZ',
      help='number of (x,z) extruded mesh levels (default=%(default)s)')
 adda('-o', metavar='FILEROOT', type=str, default='',
@@ -145,16 +147,19 @@ if args.sweepsonly:
         obsprob.savestatenextresidual(args.o + '_0.pvd')
     r = obsprob.residual(mesh, s, ellf)
     normF0 = obsprob.inactiveresidualnorm(mesh, s, r, b)
-    print('0: %.4e' % normF0)
+    if args.monitor:
+        print('   0: %.4e' % normF0)
     for j in range(args.cyclemax):
         r = obsprob.smoothersweep(mesh, s, ellf, b, currentr=r)
         normF = obsprob.inactiveresidualnorm(mesh, s, r, b)
-        print('%d: %.4e' % (j+1, normF))
+        if args.monitor:
+            print('%4d: %.4e' % (j+1, normF))
         if normF < args.irtol * normF0:
-            print('smoother iteration CONVERGED by -irtol')
+            print('iteration CONVERGED at step %d by F<(%.2e)F0' \
+                  % (j+1, args.irtol))
             break
         elif normF > 100.0 * normF0:
-            print('smoother iteration DIVERGED detected')
+            print('iteration DIVERGED by F>100F0 at step %d' % (j+1))
             break
     if args.o:
         obsprob.savestatenextresidual(args.o + '_%d.pvd' % (j+1))

### 2021-06-03 py/smoother.py
Minor stdout fiddle.

@@ -96,7 +96,7 @@ class SmootherStokes(SmootherObstacleProblem):
         V = fd.VectorFunctionSpace(mesh, 'Lagrange', 2)
         W = fd.FunctionSpace(mesh, 'Lagrange', 1)
         if printsizes:
-            print('            sizes: n_u = %d, n_p = %d' % (V.dim(), W.dim()))
+            print('      dimensions n_u = %d, n_p = %d' % (V.dim(), W.dim()))
         Z = V * W
         up = fd.Function(Z)
         scu, p = fd.split(up)       # scaled velocity, unscaled pressure

### 2021-06-03 py/smoother.py
Factor methods out of residual evaluation.

@@ -128,38 +128,31 @@ class SmootherStokes(SmootherObstacleProblem):
         u *= sc
         return u, p
 
-    def residual(self, mesh1d, s, ella):
-        '''Compute the residual functional, namely the surface kinematical
-        residual for the entire domain, for a given iterate s.  In symbols
-        matching the paper, returns  r = F(s)[.] - ella(.)  where . ranges over
-        hat functions on mesh1d.  Note mesh1D is
-        a MeshLevel1D instance and ella(.) = <a(x),.> is a source term in V^j'.
-        This residual evaluation requires setting up an (x,z) Firedrake mesh
-        by extrusion of a (stored) base mesh.  The icy
-        columns get their height from s, with minimum height args.Hmin.  By
-        default the extruded mesh has empty (0-element) columns if ice-free
-        according to s.  If args.padding==True then the whole extruded mesh has
-        the same layer count.  If saveupname is a string then the Stokes
-        solution (u,p) is saved to that file.  The returned residual array is
-        defined on mesh1d and is in the dual space V^j'.'''
+    def createbasemesh(self, mesh1d):
+        '''Create a Firedrake interval base mesh matching mesh1d.
+        Also store the bed elevation.'''
+        self.mx = mesh1d.m + 1
+        self.basemesh = fd.IntervalMesh(self.mx, length_or_left=0.0,
+                                        right=mesh1d.xmax)
+        self.b = self.phi(mesh1d.xx())
 
-        # if needed, generate self.basemesh from mesh1d
-        firstcall = (self.basemesh == None)
-        if firstcall:
-            self.mx = mesh1d.m + 1
+    def extrudemesh(self, s, report=False):
+        '''Generate extruded mesh over self.basemesh, to height s.
+        The icy columns get their height from s, with minimum height
+        args.Hmin.  By default the extruded mesh has empty (0-element) columns
+        if ice-free according to s.  If args.padding==True then the whole
+        extruded mesh has the same layer count.'''
+        assert self.basemesh is not None
+        if report:
             print('mesh: base of %d elements (intervals)' \
                   % self.mx)
-            self.basemesh = fd.IntervalMesh(self.mx, length_or_left=0.0,
-                                            right=mesh1d.xmax)
-            self.b = self.phi(mesh1d.xx())
-
[…]

### 2021-06-03 py/smoother.py
Minor: Comments and naming.

@@ -129,19 +129,19 @@ class SmootherStokes(SmootherObstacleProblem):
         return u, p
 
     def createbasemesh(self, mesh1d):
-        '''Create a Firedrake interval base mesh matching mesh1d.
-        Also store the bed elevation.'''
+        '''Create a Firedrake interval base mesh matching mesh1d.  Also store
+        the bed elevation.'''
         self.mx = mesh1d.m + 1
         self.basemesh = fd.IntervalMesh(self.mx, length_or_left=0.0,
                                         right=mesh1d.xmax)
         self.b = self.phi(mesh1d.xx())
 
-    def extrudemesh(self, s, report=False):
-        '''Generate extruded mesh over self.basemesh, to height s.
-        The icy columns get their height from s, with minimum height
-        args.Hmin.  By default the extruded mesh has empty (0-element) columns
-        if ice-free according to s.  If args.padding==True then the whole
-        extruded mesh has the same layer count.'''
+    def extrudetogeometry(self, s, report=False):
+        '''Generate extruded mesh over self.basemesh, to height s.  The icy
+        columns get their height from s, with minimum height args.Hmin.  By
+        default the extruded mesh has empty (0-element) columns if ice-free
+        according to s.  If args.padding==True then the whole extruded mesh has
+        the same layer count.  Optional reporting of mesh stats.'''
         assert self.basemesh is not None
         if report:
             print('mesh: base of %d elements (intervals)' \
@@ -168,11 +168,11 @@ class SmootherStokes(SmootherObstacleProblem):
                 icycount = sum(icyelement)
                 print('      extruded has %d x %d icy elements and %d ice-free base elements' \
                       % (icycount, mz, self.mx - icycount))
-
-        # generate extruded mesh of height s(x)
+        # put s(x) into a Firedrake function on the base mesh
         P1base = fd.FunctionSpace(self.basemesh, 'Lagrange', 1)
         sbase = fd.Function(P1base)
         sbase.dat.data[:] = np.maximum(s, self.args.Hmin)
+        # change mesh height to s(x)
         x, z = fd.SpatialCoordinate(mesh)
         xxzz = fd.as_vector([x, extend(mesh, sbase) * z])
         coords = fd.Function(mesh.coordinates.function_space())
@@ -180,9 +180,10 @@ class SmootherStokes(SmootherObstacleProblem):
[…]

### 2021-06-03 py/smoother.py
Ability to save perturbations to u,p from +eps at list of 1D nodes.

@@ -244,6 +244,42 @@ class SmootherStokes(SmootherObstacleProblem):
         # include the climatic mass balance: - u|_s . n_s - a
         return mesh1d.ellf(r) - ella
 
+    def viewperturb(self, s, klist, eps=1.0):
+        '''For given s(x), compute velocity and pressure perturbations
+        from perturbation s[k] + eps, i.e. lifting surface by eps, at an
+        each interior node k in klist.  Saves to file self.savename, which
+        needs to be a .pvd file.'''
+        assert self.basemesh is not None
+        for k in klist:
+            assert 1 <= k <= len(s)-2  # only valid at interior nodes
+        assert self.savename is not None
+        assert len(self.savename) > 0
+        # solve the Glen-Stokes problem on the unperturbed extruded mesh
+        meshs = self.extrudetogeometry(s)
+        us, ps = self.solvestokes(meshs)
+        # solve on the PERTURBED extruded mesh
+        sP = s.copy()
+        for k in klist:
+            if s[k] > self.b[k] + 0.001:
+                sP[k] += eps
+            else:
+                print('WARNING: skipping bare-ground perturbation point k=%d' % k)
+        meshP = self.extrudetogeometry(sP)
+        uP, pP = self.solvestokes(meshP)
+        # compute difference as a function on the unperturbed mesh
+        V = fd.VectorFunctionSpace(meshs, 'Lagrange', 2)
+        W = fd.FunctionSpace(meshs, 'Lagrange', 1)
+        du = fd.Function(V)
+        du.dat.data[:] = uP.dat.data_ro[:] - us.dat.data_ro[:]
+        du *= secpera
+        du.rename('du (m a-1)')
+        dp = fd.Function(W)
+        dp.dat.data[:] = pP.dat.data_ro[:] - ps.dat.data_ro[:]
+        dp /= 1.0e5
+        dp.rename('dp (bar)')
+        print('saving perturbations du,dp to %s' % self.savename)
+        fd.File(self.savename).write(du,dp)
+
     def smoothersweep(self, mesh1d, s, ella, phi, currentr=None):
         '''Do in-place smoothing on s(x).  On input, set currentr to a vector
         to avoid re-computing the residual.  Computes and returns the residual

### 2021-06-03 py/steady.py

@@ -89,6 +89,8 @@ adda('-sweepsonly', action='store_true', default=False,
      help='do smoother sweeps as cycles, instead of multilevel')
 adda('-up', type=int, default=2, metavar='N',
      help='smoother sweeps after coarse-mesh correction (default=%(default)s)')
+adda('-viewperturb', type=int, default=None, nargs='+', metavar='N ...',
+     help='view u,p perturbations at these nodes to .pvd file; use with -o')
 args, unknown = parser.parse_known_args()
 if args.steadyhelp:
     parser.print_help()
@@ -96,6 +98,8 @@ if args.steadyhelp:
 
 if args.padding:
     assert args.Hmin > 0.0, 'padding requires minimum positive thickness'
+if args.viewperturb is not None:
+    assert len(args.o) > 0, 'use -view perturb with -o'
 
 # mesh hierarchy: a list of MeshLevel1D with indices [0,..,levels-1]
 assert args.J >= args.jcoarse >= 0
@@ -164,6 +168,9 @@ if args.sweepsonly:
     if args.o:
         obsprob.savestatenextresidual(args.o + '_%d.pvd' % (j+1))
     obsprob.residual(mesh, s, ellf)  # extra residual call
+    if args.viewperturb is not None:
+        obsprob.savename = args.o + '_perturb.pvd'
+        obsprob.viewperturb(s, args.viewperturb)
 else:
     raise NotImplementedError('MCDN not implemented')
 

### 2021-06-03 py/smoother.py
Also save perturbation of residual, which is what I really care about.

@@ -82,7 +82,7 @@ class SmootherStokes(SmootherObstacleProblem):
         u.rename('velocity (m a-1)')
         p /= 1.0e5
         p.rename('pressure (bar)')
-        kres.rename('kinetic residual (a=0)')
+        kres.rename('kinematic residual (a=0)')
         print('saving u,p,nu,tau,kres to %s' % self.savename)
         fd.File(self.savename).write(u,p,nu,tau,kres)
         self.saveflag = False
@@ -210,6 +210,17 @@ class SmootherStokes(SmootherObstacleProblem):
             f[bmcnm[cell,...]] = f_all[topind]
         return f
 
+    def kinematical(self, mesh, u):
+        ''' Evaluate kinematic part of residual from given velocity u, namely
+        as a field defined on the whole extruded mesh:
+            kres = u ds/dx - w.
+        Note n_s = <-s_x, 1> so this is <u,w> . n_s.'''
+        _, z = fd.SpatialCoordinate(mesh)
+        kres_ufl = u[0] * z.dx(0) - u[1]
+        Q1 = fd.FunctionSpace(mesh, 'Lagrange', 1)
+        kres = fd.Function(Q1).interpolate(kres_ufl)
+        return kres
+
     def residual(self, mesh1d, s, ella):
         '''Compute the residual functional, namely the surface kinematical
         residual for the entire domain, for a given iterate s.  In symbols
@@ -226,11 +237,8 @@ class SmootherStokes(SmootherObstacleProblem):
         # solve the Glen-Stokes problem on the extruded mesh
         mesh = self.extrudetogeometry(s, report=firstcall)
         u, p = self.solvestokes(mesh, printsizes=firstcall)
-        # evaluate kinematic part of residual  (note - u|_s . n_s = u ds/dx - w)
-        _, z = fd.SpatialCoordinate(mesh)
-        kres_ufl = u[0] * z.dx(0) - u[1]
-        Q1 = fd.FunctionSpace(mesh, 'Lagrange', 1)
-        kres = fd.Function(Q1).interpolate(kres_ufl)
+        # get kinematical part of residual
+        kres = self.kinematical(mesh, u)
         if self.saveflag:
             self.savestate(mesh, u, p, kres)
         # get kinematic residual r = - u|_s . n_s on z = s(x)
@@ -245,27 +253,30 @@ class SmootherStokes(SmootherObstacleProblem):
         return mesh1d.ellf(r) - ella
 
[…]

### 2021-06-04 py/steady.py
Update TODO list.

@@ -2,7 +2,10 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   1. copy partI/mcdn.py and build it out
+#   1. separate the initialization and source term (= problem) from the rest of the smoother
+#   2. widen the default sheet
+#   3. implement -smoother jacobicolor with default coloring mode being 3 ice thicknesses
+#   4. copy mg-glaciers/py/mcdn.py and build it out
 
 # RUNS:
 #   actually converges

### 2021-06-04 py/basesmoother.py
Factor the problem out of the smoother.

@@ -58,10 +58,6 @@ class SmootherObstacleProblem(ABC):
                 forward = not forward
 
     @abstractmethod
-    def initial(self, x):
-        '''Generate initial shape.'''
-
-    @abstractmethod
     def residual(self, mesh, w, ell):
         '''Compute the residual functional for given iterate w.  Note
         ell is a source term in V^j'.'''
@@ -70,11 +66,3 @@ class SmootherObstacleProblem(ABC):
     def smoothersweep(self, mesh, w, ell, phi, forward=True):
         '''Apply one sweep of obstacle-problem smoother on mesh to modify w in
         place.'''
-
-    @abstractmethod
-    def phi(self, x):
-        '''Evaluate obstacle phi at location(s) x.'''
-
-    @abstractmethod
-    def source(self, x):
-        '''Evaluate source function f at location(s) x.'''

### 2021-06-04 py/problem.py

@@ -1,0 +1,79 @@
+'''Dome ice problem.'''
+
+from abc import ABC
+import numpy as np
+
+# public physical parameters
+secpera = 31556926.0        # seconds per year
+g = 9.81                    # m s-2
+rhoi = 910.0                # kg m-3
+nglen = 3.0
+A3 = 1.0e-16 / secpera      # Pa-3 s-1;  EISMINT I ice softness
+B3 = A3**(-1.0/3.0)         # Pa s(1/3);  ice hardness
+
+class IceProblem(ABC):
+    '''Class for a particular ice problem.'''
+
+    def __init__(self, args):
+        self.args = args
+        # parameters for initial condition, a dome; see van der Veen
+        #   (2013) section 5.3
+        self.domeL = 10.0e3       # half-width of sheet
+        self.domeH0 = 1000.0      # center thickness
+        # derived constant
+        self.Gamma = 2.0 * A3 * (rhoi * g)**nglen
+        self.Gamma /= (nglen + 2.0)
+
+    def bed(self, x):
+        '''For now we have a flat bed.'''
+        return np.zeros(np.shape(x))
+
+    def source(self, x):
+        '''Continuous source term, the climatic mass balance (CMB).
+        See van der Veen (2013) equations (5.49) and (5.51).  Assumes x
+        is a numpy array.'''
+        n = nglen
+        invn = 1.0 / n
+        r1 = 2.0 * n + 2.0                   # e.g. 8
+        s1 = (1.0 - n) / n                   #     -2/3
+        C = self.domeH0**r1 * self.Gamma     # A_0 in van der Veen is Gamma here
+        C /= ( 2.0 * self.domeL * (1.0 - 1.0 / n) )**n
+        xc = self.args.domainlength / 2.0
+        X = (x - xc) / self.domeL          # rescaled coord
+        m = np.zeros(np.shape(x))
+        # usual formula for 0 < |X| < 1
+        zzz = (abs(X) > 0.0) * (abs(X) < 1.0)
+        if any(zzz):
+            Xin = abs(X[zzz])
+            Yin = 1.0 - Xin
+            m[zzz] = (C / self.domeL) \
+                     * ( Xin**invn + Yin**invn - 1.0 )**(n-1.0) \
+                     * ( Xin**s1 - Yin**s1 )
+        # fill singular origin with near value
+        if any(X == 0.0):
+            Xnear = 1.0e-8
+            Ynear = 1.0 - Xnear
+            m[X == 0.0] = (C / self.domeL) \
[…]

### 2021-06-04 py/smoother.py

@@ -3,8 +3,7 @@
 import numpy as np
 import firedrake as fd
 from basesmoother import SmootherObstacleProblem
-
-secpera = 31556926.0        # seconds per year
+from problem import secpera, g, rhoi, nglen, B3
 
 def extend(mesh, f):
     '''On an extruded mesh extend a function f(x,z), already defined on the
@@ -22,29 +21,17 @@ class SmootherStokes(SmootherObstacleProblem):
     extruded mesh for each residual evaluation.  Implements projected
     nonlinear versions of the Richardson and Jacobi smoothers.'''
 
-    def __init__(self, args, admissibleeps=1.0e-10):
+    def __init__(self, args, b, admissibleeps=1.0e-10):
         super().__init__(args, admissibleeps=admissibleeps)
+        self.b = b
         # smoother name
         self.name = 'SmootherStokes'
-        # physical parameters
-        self.g = 9.81                    # m s-2
-        self.rhoi = 910.0                # kg m-3
-        self.nglen = 3.0
-        self.A3 = 1.0e-16 / secpera      # Pa-3 s-1;  EISMINT I ice softness
-        self.B3 = self.A3**(-1.0/3.0)    # Pa s(1/3);  ice hardness
         # used in Stokes solver
         self.Dtyp = 1.0 / secpera        # s-1
         self.sc = 1.0e-7                 # velocity scale for symmetric scaling
-        # parameters for initial condition, a Bueler profile; see van der Veen
-        #   (2013) section 5.3
-        self.buelerL = 10.0e3       # half-width of sheet
-        self.buelerH0 = 1000.0      # center thickness
-        self.Gamma = 2.0 * self.A3 * (self.rhoi * self.g)**self.nglen
-        self.Gamma /= (self.nglen + 2.0)
         # we store the basemesh info and the bed elevation
         self.basemesh = None
         self.mx = None
-        self.b = None
         self.saveflag = False
         self.savename = None
 
@@ -62,9 +49,9 @@ class SmootherStokes(SmootherObstacleProblem):
         from the velocity solution.'''
         Q1 = fd.FunctionSpace(mesh,'Q',1)
         Du2 = self._regDu2(u)
-        r = 1.0 / self.nglen - 1.0
-        assert self.nglen == 3.0
[…]

### 2021-06-04 py/steady.py

@@ -2,7 +2,6 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   1. separate the initialization and source term (= problem) from the rest of the smoother
 #   2. widen the default sheet
 #   3. implement -smoother jacobicolor with default coloring mode being 3 ice thicknesses
 #   4. copy mg-glaciers/py/mcdn.py and build it out
@@ -21,6 +20,7 @@ import matplotlib.pyplot as plt
 from firedrake import *
 
 from meshlevel import MeshLevel1D
+from problem import IceProblem, secpera
 from smoother import SmootherStokes
 #from mcdn import mcdnsolver
 
@@ -111,16 +111,15 @@ hierarchy = [None] * (levels)             # list [None,...,None]
 for j in range(levels):
     hierarchy[j] = MeshLevel1D(j=j+args.jcoarse, xmax=args.domainlength)
 
-obsprob = SmootherStokes(args)
-
 # fine-level problem data
+problem = IceProblem(args)
 mesh = hierarchy[-1]
-phi = obsprob.phi(mesh.xx())
-ellf = mesh.ellf(obsprob.source(mesh.xx()))  # source functional ell[v] = <f,v>
-s = obsprob.initial(mesh.xx())
-#obsprob.shownonzeros(s)
+b = problem.bed(mesh.xx())
+ellf = mesh.ellf(problem.source(mesh.xx()))  # source functional ell[v] = <f,v>
+s = problem.initial(mesh.xx())
 
-b = mesh.zeros()
+# set-up smoother
+smooth = SmootherStokes(args, b)
 
 def output(filename, description):
     '''Either save result to an image file or use show().  Supply '' as filename
@@ -133,7 +132,6 @@ def output(filename, description):
 
 def final(mesh, s, cmb, filename=''):
     '''Generate graphic showing final iterate and CMB function.'''
-    secpera = 31556926.0
     mesh.checklen(s)
     xx = mesh.xx()
     xx /= 1000.0
@@ -151,14 +149,14 @@ def final(mesh, s, cmb, filename=''):
 
 if args.sweepsonly:
     if args.o:
-        obsprob.savestatenextresidual(args.o + '_0.pvd')
-    r = obsprob.residual(mesh, s, ellf)
-    normF0 = obsprob.inactiveresidualnorm(mesh, s, r, b)
+        smooth.savestatenextresidual(args.o + '_0.pvd')
[…]

### 2021-06-04 py/problem.py
Options -domeL,-domeH0 (and -domainlength) give control over problem.

@@ -16,10 +16,6 @@ class IceProblem(ABC):
 
     def __init__(self, args):
         self.args = args
-        # parameters for initial condition, a dome; see van der Veen
-        #   (2013) section 5.3
-        self.domeL = 10.0e3       # half-width of sheet
-        self.domeH0 = 1000.0      # center thickness
         # derived constant
         self.Gamma = 2.0 * A3 * (rhoi * g)**nglen
         self.Gamma /= (nglen + 2.0)
@@ -36,24 +32,24 @@ class IceProblem(ABC):
         invn = 1.0 / n
         r1 = 2.0 * n + 2.0                   # e.g. 8
         s1 = (1.0 - n) / n                   #     -2/3
-        C = self.domeH0**r1 * self.Gamma     # A_0 in van der Veen is Gamma here
-        C /= ( 2.0 * self.domeL * (1.0 - 1.0 / n) )**n
+        C = self.args.domeH0**r1 * self.Gamma # Gamma=A_0 in van der Veen
+        C /= ( 2.0 * self.args.domeL * (1.0 - 1.0 / n) )**n
         xc = self.args.domainlength / 2.0
-        X = (x - xc) / self.domeL          # rescaled coord
+        X = (x - xc) / self.args.domeL          # rescaled coord
         m = np.zeros(np.shape(x))
         # usual formula for 0 < |X| < 1
         zzz = (abs(X) > 0.0) * (abs(X) < 1.0)
         if any(zzz):
             Xin = abs(X[zzz])
             Yin = 1.0 - Xin
-            m[zzz] = (C / self.domeL) \
+            m[zzz] = (C / self.args.domeL) \
                      * ( Xin**invn + Yin**invn - 1.0 )**(n-1.0) \
                      * ( Xin**s1 - Yin**s1 )
         # fill singular origin with near value
         if any(X == 0.0):
             Xnear = 1.0e-8
             Ynear = 1.0 - Xnear
-            m[X == 0.0] = (C / self.domeL) \
+            m[X == 0.0] = (C / self.args.domeL) \
                           * ( Xnear**invn + Ynear**invn - 1.0 )**(n-1.0) \
                           * ( Xnear**s1 - Ynear**s1 )
         # extend by ablation
@@ -67,9 +63,9 @@ class IceProblem(ABC):
         n = nglen
         p1 = n / (2.0 * n + 2.0)                  # e.g. 3/8
         q1 = 1.0 + 1.0 / n                        #      4/3
[…]

### 2021-06-04 py/steady.py

@@ -2,7 +2,7 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   2. widen the default sheet
+#   2. widen the default sheet; consider -domainlength 60.0e3 -domeL 25.0e3
 #   3. implement -smoother jacobicolor with default coloring mode being 3 ice thicknesses
 #   4. copy mg-glaciers/py/mcdn.py and build it out
 
@@ -57,6 +57,10 @@ adda('-cyclemax', type=int, default=100, metavar='N',
      help='maximum number of (multilevel) cycles (default=%(default)s)')
 adda('-domainlength', type=float, default=30.0e3, metavar='L',
      help='solve on [0,L] (default=%(default)s m)')
+adda('-domeH0', type=float, default=1000.0, metavar='L',
+     help='center height of dome formula ice sheet (default=%(default)s m)')
+adda('-domeL', type=float, default=10.0e3, metavar='L',
+     help='half-width of dome formula ice sheet (default=%(default)s m)')
 adda('-down', type=int, default=0, metavar='N',
      help='smoother sweeps before coarse-mesh correction (default=%(default)s)')
 adda('-eps', type=float, metavar='X', default=1.0e-2,  # FIXME sensitive

### 2021-06-04 py/basesmoother.py
Replace unused _checkrepairadmissible() with _checkadmissible() which errors-out.

@@ -2,6 +2,7 @@
 
 __all__ = ['SmootherObstacleProblem']
 
+import sys
 from abc import ABC, abstractmethod
 import numpy as np
 
@@ -12,18 +13,15 @@ class SmootherObstacleProblem(ABC):
     def __init__(self, args, admissibleeps=1.0e-10):
         self.args = args
         self.admissibleeps = admissibleeps
-        self.inadmissible = 0  # count of repaired admissibility violations
         self.name = 'base'
 
-    def _checkrepairadmissible(self, mesh, w, phi):
-        '''Check and repair feasibility.'''
+    def _checkadmissible(self, mesh, w, phi):
+        '''Check admissibility and stop if not.'''
         for p in range(1, mesh.m+1):
             if w[p] < phi[p] - self.admissibleeps:
-                if self.args.printwarnings:
-                    print('WARNING: repairing inadmissible w[%d]=%e < phi[%d]=%e on level %d (m=%d)' \
-                          % (p, w[p], p, phi[p], mesh.j, mesh.m))
-                w[p] = phi[p]
-                self.inadmissible += 1
+                print('ERROR: inadmissible w[%d]=%e < phi[%d]=%e (m=%d)' \
+                      % (p, w[p], p, phi[p], mesh.m))
+                sys.exit(0)
 
     def _sweepindices(self, mesh, forward=True):
         '''Generate indices for sweep.'''

### 2021-06-04 py/steady.py
Better stdout.

@@ -152,6 +152,7 @@ def final(mesh, s, cmb, filename=''):
     output(filename, 'image of final iterate')
 
 if args.sweepsonly:
+    print('sweepsonly with smoother "%s" ...' % args.smoother)
     if args.o:
         smooth.savestatenextresidual(args.o + '_0.pvd')
     r = smooth.residual(mesh, s, ellf)
@@ -170,6 +171,8 @@ if args.sweepsonly:
         elif normF > 100.0 * normF0:
             print('iteration DIVERGED by F>100F0 at step %d' % (j+1))
             break
+        elif j + 1 == args.cyclemax:
+            print('iteration REACHED CYCLEMAX at step %d' % (j+1))
     if args.o:
         smooth.savestatenextresidual(args.o + '_%d.pvd' % (j+1))
     smooth.residual(mesh, s, ellf)  # extra residual call

### 2021-06-04 py/smoother.py
Implement smoother "jacobicolor" (by 3 ice thicknesses) which seems to work.

@@ -293,10 +293,12 @@ class SmootherStokes(SmootherObstacleProblem):
             currentr = self.residual(mesh1d, s, ella)
         if self.args.smoother == 'richardson':
             negd = self.richardsonsweep(s, phi, currentr)
-        elif self.args.smoother == 'jacobislow':
-            negd = self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
         elif self.args.smoother == 'gsslow':
             negd = self.gsslowsweep(mesh1d, s, ella, phi, currentr)
+        elif self.args.smoother == 'jacobislow':
+            negd = self.jacobislowsweep(mesh1d, s, ella, phi, currentr)
+        elif self.args.smoother == 'jacobicolor':
+            negd = self.jacobicolorsweep(mesh1d, s, ella, phi, currentr)
         mesh1d.WU += 1
         if self.args.monitor and len(negd) > 0:
             print('      negative diagonal entries: ', end='')
@@ -310,6 +312,32 @@ class SmootherStokes(SmootherObstacleProblem):
         stability criterion argument.)'''
         np.maximum(s - self.args.omega * r, phi, s)
         return []
+
+    def gsslowsweep(self, mesh1d, s, ella, phi, r,
+                    eps=1.0, dump=False):
+        '''Do in-place projected nonlinear Gauss-Seidel smoothing on s(x)
+        where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
+        by VERY SLOW finite differencing of expensive residual calculations.
+        If d_i > 0 then
+            s_i <- max(s_i - omega * r_i / d_i, phi_i)
+        but otherwise
+            s_i <- phi_i.'''
+        negd = []
+        for j in range(1, len(s)-1): # loop over interior points
+            sperturb = s.copy()
+            sperturb[j] += eps
+            if dump:
+                self.savestatenextresidual(self.args.o + '_gs_%d.pvd' % j)
+            rperturb = self.residual(mesh1d, sperturb, ella)
+            d = (rperturb[j] - r[j]) / eps
+            if d > 0.0:
+                s[j] = max(s[j] - self.args.omega * r[j] / d, phi[j])
+            else:
+                s[j] = phi[j]
+                negd.append(j)
+            # must recompute residual for s (nonlocal!)
+            r = self.residual(mesh1d, s, ella)
+        return negd
 
     def jacobislowsweep(self, mesh1d, s, ella, phi, r,
[…]

### 2021-06-04 py/steady.py

@@ -87,8 +87,8 @@ adda('-padding', action='store_true', default=False,
      help='put Hmin thickness of ice in ice-free locations')
 adda('-printwarnings', action='store_true', default=False,
      help='print pointwise feasibility warnings')
-adda('-smoother', choices=['richardson', 'jacobislow', 'gsslow'],
-     metavar='X', default='richardson',
+adda('-smoother', choices=['richardson', 'gsslow', 'jacobislow', 'jacobicolor'],
+     metavar='X', default='jacobicolor',
      help='smoother (default=%(default)s)')
 adda('-steadyhelp', action='store_true', default=False,
      help='print help for steady.py and end (vs -help for PETSc options)')

### 2021-06-04 py/steady.py
Update TODO list etc.

@@ -2,9 +2,11 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
-#   2. widen the default sheet; consider -domainlength 60.0e3 -domeL 25.0e3
-#   3. implement -smoother jacobicolor with default coloring mode being 3 ice thicknesses
-#   4. copy mg-glaciers/py/mcdn.py and build it out
+#   1. widen the default sheet?; consider -domainlength 60.0e3 -domeL 25.0e3
+#   2. replace meshlevel.py with use of Firedrake interval mesh (requires
+#      new implementation of monotone restriction); this would allow
+#      parallelization
+#   3. copy mg-glaciers/py/mcdn.py and build it out
 
 # RUNS:
 #   actually converges
@@ -12,6 +14,9 @@
 #      with JJ=1,2,3,4,5 and HH=0,10,100;  HH=0 seems as good as any
 #      also sort of works with JJ=6 when HH=100 and -mz 8
 #      also works with -smoother gsslow
+#   uses jacobicolor (faster than jacobislow) and works:
+#      for JJ in 3 4 5 6; do ./steady.py -sweepsonly -monitor -omega 0.7 -Hmin 20.0 -mz 8 -irtol 1.0e-2 -J $JJ; done
+#      but fails to converge for JJ=7
 
 import sys
 import argparse
@@ -37,7 +42,7 @@ inactive set {x | s(x) > b(x)}.
 Solution is by the nonlinear (FAS) extension of the multilevel constraint
 decomposition (MCD) method of Tai (2003), or by sweeps of the smoother.
 
-Initial implementation generates Bueler profile geometry as initial state
+Initial implementation generates a dome geometry on a flat bed as initial state
 and then tries to converge from there.
 
 References:

### 2021-06-04 py/steady.py
Factor out visualization, and cleanup.

@@ -21,13 +21,13 @@
 import sys
 import argparse
 import numpy as np
-import matplotlib.pyplot as plt
 from firedrake import *
 
 from meshlevel import MeshLevel1D
-from problem import IceProblem, secpera
+from problem import IceProblem
 from smoother import SmootherStokes
 #from mcdn import mcdnsolver
+from visualize import showiteratecmb
 
 parser = argparse.ArgumentParser(description='''
 Solve the 1D obstacle problem for the steady geometry of a glacier using
@@ -56,7 +56,7 @@ References:
     allow_abbrev=False,  # bug in python 3.8 causes this to be ignored
     add_help=False)
 adda = parser.add_argument
-adda('-coarse', type=int, default=1, metavar='N',
+adda('-coarsesweeps', type=int, default=1, metavar='N',
      help='smoother sweeps on coarsest grid (default=%(default)s)')
 adda('-cyclemax', type=int, default=100, metavar='N',
      help='maximum number of (multilevel) cycles (default=%(default)s)')
@@ -66,7 +66,7 @@ adda('-domeH0', type=float, default=1000.0, metavar='L',
      help='center height of dome formula ice sheet (default=%(default)s m)')
 adda('-domeL', type=float, default=10.0e3, metavar='L',
      help='half-width of dome formula ice sheet (default=%(default)s m)')
-adda('-down', type=int, default=0, metavar='N',
+adda('-downsweeps', type=int, default=0, metavar='N',
      help='smoother sweeps before coarse-mesh correction (default=%(default)s)')
 adda('-eps', type=float, metavar='X', default=1.0e-2,  # FIXME sensitive
     help='regularization used in viscosity (default=%(default)s)')
@@ -99,7 +99,7 @@ adda('-steadyhelp', action='store_true', default=False,
      help='print help for steady.py and end (vs -help for PETSc options)')
 adda('-sweepsonly', action='store_true', default=False,
      help='do smoother sweeps as cycles, instead of multilevel')
-adda('-up', type=int, default=2, metavar='N',
+adda('-upsweeps', type=int, default=2, metavar='N',
      help='smoother sweeps after coarse-mesh correction (default=%(default)s)')
 adda('-viewperturb', type=int, default=None, nargs='+', metavar='N ...',
[…]

### 2021-06-04 py/visualize.py

@@ -1,0 +1,32 @@
+'''Module for matplotlib visualization of glaciers.'''
+
+from problem import secpera
+import matplotlib.pyplot as plt
+
+def output(filename, description):
+    '''Either save result to an image file or use show().  Supply '' as filename
+    to use show().'''
+    if filename is None:
+        plt.show()
+    else:
+        print('saving %s to %s ...' % (description, filename))
+        plt.savefig(filename, bbox_inches='tight')
+
+def showiteratecmb(mesh, s, cmb, filename=''):
+    '''Generate graphic showing final iterate and CMB function.  mesh is of
+    type MeshLevel1D and s, cmb are vectors on the mesh.'''
+    mesh.checklen(s)
+    mesh.checklen(cmb)
+    xx = mesh.xx()
+    xx /= 1000.0
+    plt.figure(figsize=(15.0, 8.0))
+    plt.subplot(2,1,1)
+    plt.plot(xx, s, 'k', linewidth=4.0)
+    plt.xlabel('x (km)')
+    plt.ylabel('surface elevation (m)')
+    plt.subplot(2,1,2)
+    plt.plot(xx, cmb * secpera, 'r')
+    plt.grid()
+    plt.ylabel('CMB (m/a)')
+    plt.xlabel('x (km)')
+    output(filename, 'image of iterate and CMB')

### 2021-06-04 py/mcdn.py
Preliminary hook-up of MCDN.  Consistently use copy of b in MeshLevel1D.

@@ -1,0 +1,76 @@
+'''Module implementing a V-cycle of the nonlinear full approximation storage
+(FAS) extension of the multilevel constraint decomposition (MCD) method.'''
+
+__all__ = ['mcdnvcycle']
+
+import numpy as np
+
+def _levelreport(indent, j, m, sweeps):
+    indentprint(indent - j, 'level %d: %d sweeps over m=%d nodes' \
+                             % (j, sweeps, m))
+
+def _coarsereport(indent, m, sweeps):
+    indentprint(indent, 'coarsest: %d sweeps over m=%d nodes' \
+                         % (sweeps, m))
+
+def mcdnvcycle(args, smoother, hierarchy, s, ella, b, levels=None):
+    '''Apply one V-cycle of the MCDN method.  Input args is an options
+    dictionary with parameters.  Input smoother is of type
+    SmootherObstacleProblem.  Note hierarchy[j] is of type MeshLevel1D,
+    for j=0,...,args.J, but only args.jcoarse,...,args.J are actually used.
+    This method generates all the defect constraints hierarchy[j].chi for
+    j <= J, but hierarchy[j].b must
+    be defined for all mesh levels.  The input iterate w is in V^j and
+    linear functional ell is in V^J'.  The coarse solver is the same as
+    the smoother.'''
+
+    raise NotImplementedError('MCDN not implemented')
+    assert args.down >= 0 and args.up >= 0 and args.coarse >= 0
+
+    # set up on finest level
+    hierarchy[J].checklen(w)
+    hierarchy[J].checklen(ell)
+    hierarchy[J].ell = ell
+    hierarchy[J].g = w
+
+    # downward
+    for k in range(J, 0, -1):
+        # compute next defect constraint using monotone restriction
+        hierarchy[k-1].chi = hierarchy[k].mR(hierarchy[k].chi)
+        # define down-obstacle
+        phi = hierarchy[k].chi - hierarchy[k].cP(hierarchy[k-1].chi)
+        # smooth the correction y
+        if args.mgview:
+            _levelreport(levels-1, k, hierarchy[k].m, args.down)
+        hierarchy[k].y = hierarchy[k].zeros()
+        obsprob.smoother(args.down, hierarchy[k], hierarchy[k].y,
+                         hierarchy[k].ell, phi)
+        # update residual
+        wk = hierarchy[k].g + hierarchy[k].y
[…]

### 2021-06-04 py/meshlevel.py

@@ -29,6 +29,7 @@ class MeshLevel1D():
         else:
             self.mcoarser = None
         self.h = self.xmax / (self.m + 1)
+        self.b = None
         self.WU = 0
 
     def checklen(self, v, coarser=False):

### 2021-06-04 py/smoother.py

@@ -21,9 +21,8 @@ class SmootherStokes(SmootherObstacleProblem):
     extruded mesh for each residual evaluation.  Implements projected
     nonlinear versions of the Richardson and Jacobi smoothers.'''
 
-    def __init__(self, args, b, admissibleeps=1.0e-10):
+    def __init__(self, args, admissibleeps=1.0e-10):
         super().__init__(args, admissibleeps=admissibleeps)
-        self.b = b
         # smoother name
         self.name = 'SmootherStokes'
         # used in Stokes solver
@@ -119,11 +118,10 @@ class SmootherStokes(SmootherObstacleProblem):
         '''Create a Firedrake interval base mesh matching mesh1d.  Also store
         the bed elevation.'''
         self.mx = mesh1d.m + 1
-        mesh1d.checklen(self.b)
         self.basemesh = fd.IntervalMesh(self.mx, length_or_left=0.0,
                                         right=mesh1d.xmax)
 
-    def extrudetogeometry(self, s, report=False):
+    def extrudetogeometry(self, s, b, report=False):
         '''Generate extruded mesh over self.basemesh, to height s.  The icy
         columns get their height from s, with minimum height args.Hmin.  By
         default the extruded mesh has empty (0-element) columns if ice-free
@@ -144,7 +142,7 @@ class SmootherStokes(SmootherObstacleProblem):
                       % (self.mx, mz))
         else:
             layermap = np.zeros((self.mx, 2), dtype=int)  # [[0,0], ..., [0,0]]
-            thk = s - self.b
+            thk = s - b
             thkelement = ( (thk[:-1]) + (thk[1:]) ) / 2.0
             icyelement = (thkelement > self.args.Hmin + 1.0e-3)
             layermap[:,1] = mz * np.array(icyelement, dtype=int)
@@ -161,6 +159,7 @@ class SmootherStokes(SmootherObstacleProblem):
         sbase.dat.data[:] = np.maximum(s, self.args.Hmin)
         # change mesh height to s(x)
         x, z = fd.SpatialCoordinate(mesh)
+        # FIXME next line needs modification if b!=0
         xxzz = fd.as_vector([x, extend(mesh, sbase) * z])
         coords = fd.Function(mesh.coordinates.function_space())
         mesh.coordinates.assign(coords.interpolate(xxzz))
@@ -222,7 +221,7 @@ class SmootherStokes(SmootherObstacleProblem):
[…]

### 2021-06-04 py/steady.py

@@ -26,7 +26,7 @@ from firedrake import *
 from meshlevel import MeshLevel1D
 from problem import IceProblem
 from smoother import SmootherStokes
-#from mcdn import mcdnsolver
+from mcdn import mcdnvcycle
 from visualize import showiteratecmb
 
 parser = argparse.ArgumentParser(description='''
@@ -113,58 +113,67 @@ if args.padding:
 if args.viewperturb is not None:
     assert len(args.o) > 0, 'use -view perturb with -o'
 
+# problem and mesh hierarchy
+problem = IceProblem(args)
 if args.sweepsonly:
     # build fine-level mesh only
     finemesh = MeshLevel1D(j=args.J, xmax=args.domainlength)
+    finemesh.b = problem.bed(finemesh.xx())
 else:
-    # build mesh hierarchy: list of MeshLevel1D with indices [0,..,levels-1]
+    # build mesh hierarchy: list of MeshLevel1D with indices [0,..,J]
+    # (some coarse levels will be unused if jcoarse > 0)
     assert args.J >= args.jcoarse >= 0
-    levels = args.J - args.jcoarse + 1
-    hierarchy = [None] * (levels)             # list [None,...,None]
-    for j in range(levels):
-        hierarchy[j] = MeshLevel1D(j=j+args.jcoarse, xmax=args.domainlength)
-    finemesh = hierarchy[-1]
+    hierarchy = [None] * (args.J + 1)             # list [None,...,None]
+    for j in range(args.J + 1):
+        hierarchy[j] = MeshLevel1D(j=j, xmax=args.domainlength)
+        hierarchy[j].b = problem.bed(hierarchy[j].xx())
+    finemesh = hierarchy[args.J]
 
-# fine-level problem data
-problem = IceProblem(args)
-b = problem.bed(finemesh.xx())
+# fine-level data
 ella = finemesh.ellf(problem.source(finemesh.xx()))  # source ell[v] = <a,v>
 s = problem.initial(finemesh.xx())
 
 # set-up smoother
-smooth = SmootherStokes(args, b)
+smooth = SmootherStokes(args)
 
 # solve
 if args.sweepsonly:
     print('sweepsonly with smoother "%s" ...' % args.smoother)
-    if args.o:
-        smooth.savestatenextresidual(args.o + '_0.pvd')
-    r = smooth.residual(finemesh, s, ella)
-    normF0 = smooth.inactiveresidualnorm(finemesh, s, r, b)
-    if args.monitor:
[…]

### 2021-06-04 py/mcdn.py
Work toward MCDN (but not ready).

@@ -5,46 +5,56 @@ __all__ = ['mcdnvcycle']
 
 import numpy as np
 
+def _indentprint(n, s):
+    '''Print 2n spaces and then string s.'''
+    for _ in range(n):
+        print('  ', end='')
+    print(s)
+
 def _levelreport(indent, j, m, sweeps):
-    indentprint(indent - j, 'level %d: %d sweeps over m=%d nodes' \
-                             % (j, sweeps, m))
+    _indentprint(indent - j, 'level %d: %d sweeps over m=%d nodes' \
+                              % (j, sweeps, m))
 
 def _coarsereport(indent, m, sweeps):
-    indentprint(indent, 'coarsest: %d sweeps over m=%d nodes' \
-                         % (sweeps, m))
+    _indentprint(indent, 'coarsest: %d sweeps over m=%d nodes' \
+                          % (sweeps, m))
 
-def mcdnvcycle(args, smoother, hierarchy, s, ella, b, levels=None):
+def mcdnvcycle(args, smoother, hierarchy, s, ella):
     '''Apply one V-cycle of the MCDN method.  Input args is an options
     dictionary with parameters.  Input smoother is of type
     SmootherObstacleProblem.  Note hierarchy[j] is of type MeshLevel1D,
     for j=0,...,args.J, but only args.jcoarse,...,args.J are actually used.
     This method generates all the defect constraints hierarchy[j].chi for
-    j <= J, but hierarchy[j].b must
-    be defined for all mesh levels.  The input iterate w is in V^j and
-    linear functional ell is in V^J'.  The coarse solver is the same as
-    the smoother.'''
+    j <= J, but hierarchy[j].b must be already defined for all mesh levels.
+    The input iterate s is in V^j and linear functional ella is in V^J'.
+    The coarse solver is the same as the smoother.  Returns a correction
+    to s.'''
 
     raise NotImplementedError('MCDN not implemented')
-    assert args.down >= 0 and args.up >= 0 and args.coarse >= 0
+    assert args.downsweeps >= 0 and args.upsweeps >= 0 \
+           and args.coarsesweeps >= 0
 
     # set up on finest level
-    hierarchy[J].checklen(w)
-    hierarchy[J].checklen(ell)
-    hierarchy[J].ell = ell
-    hierarchy[J].g = w
+    hierarchy[args.J].checklen(s)
+    hierarchy[args.J].checklen(ella)
[…]

### 2021-06-04 py/steady.py

@@ -136,26 +136,25 @@ s = problem.initial(finemesh.xx())
 # set-up smoother
 smooth = SmootherStokes(args)
 
-# solve
-if args.sweepsonly:
-    print('sweepsonly with smoother "%s" ...' % args.smoother)
-else:
-    raise NotImplementedError('MCDN not implemented')
-
+# generate .pvd file with initial state (see first residual())
 if args.o:
     smooth.savestatenextresidual(args.o + '_0.pvd')
-r = smooth.residual(finemesh, s, ella)
+
+# solve
+if args.sweepsonly:
+    print('SWEEPSONLY with smoother "%s" ...' % args.smoother)
+else:
+    print('MCDN with smoother "%s" ...' % args.smoother)
+r = smooth.residual(finemesh, s, ella)  # generates file if -o above
 normF0 = smooth.inactiveresidualnorm(finemesh, s, r, finemesh.b)
 if args.monitor:
     print('   0: %.4e' % normF0)
-
 for j in range(args.cyclemax):
     if args.sweepsonly:
         r = smooth.smoothersweep(finemesh, s, ella, currentr=r)
     else:
-        s += mcdnvcycle(args, smooth, args.J, hierarchy,
-                        s, ella, levels=args.J+1)
-        r = smooth.residual(finemesh, s, ella)  # needed? return from MCDN?
+        s += mcdnvcycle(args, smooth, hierarchy, s, ella)
+        r = smooth.residual(finemesh, s, ella)  # needed? return it from MCDN?
     normF = smooth.inactiveresidualnorm(finemesh, s, r, finemesh.b)
     if args.monitor:
         print('%4d: %.4e' % (j+1, normF))
@@ -168,13 +167,16 @@ for j in range(args.cyclemax):
         break
     elif j + 1 == args.cyclemax:
         print('iteration REACHED CYCLEMAX at step %d' % (j+1))
+
+# output to .pvd
 if args.o:
     smooth.savestatenextresidual(args.o + '_%d.pvd' % (j+1))
-smooth.residual(finemesh, s, ella)  # extra residual call
-if args.viewperturb is not None:
-    smooth.savename = args.o + '_perturb.pvd'
-    smooth.viewperturb(s, args.viewperturb)
+    smooth.residual(finemesh, s, ella)  # extra residual call generates file
+    if args.viewperturb is not None:
+        smooth.savename = args.o + '_perturb.pvd'
+        smooth.viewperturb(s, args.viewperturb)
 
[…]

### 2021-06-05 talk/talk.tex
Change sign of Phi.

@@
 \begin{block}{definition: ice dynamics operator}
-$\Phi : s \mapsto \bus \cdot \bn_s$
+$\Phi : s \mapsto - \bus \cdot \bn_s$
 
@@
 \item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
-    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0$$
-
+    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
+extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
+\item regarding $F_{\Lambda_s}(\bu,p) = 0$:
     \begin{itemize}
@@
     \item optimal solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
-    \item extract trace $\bus$, then $\Phi(s) = \bus \cdot \bn_s$, extend by zero to all of $\Omega$
     \end{itemize}
@@
 s - b &\ge 0 \\
-- \Phi(s) - a &\ge 0 \\
-(s - b) (- \Phi(s) - a) &= 0
+\Phi(s) - a &\ge 0 \\
+(s - b) (\Phi(s) - a) &= 0
 \end{align*}
@@
     $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
-for all $r \in \mathcal{K}$, where $F(s)[w] = - \int_\Omega \Phi(s)\, w \,dx dy$
+for all $r \in \mathcal{K}$, where $F(s)[w] = \int_\Omega \Phi(s)\, w \,dx dy$
 \end{block}
 
-\item well-posedness is open
+\item well-posedness is completely open
     \begin{itemize}
-    \item existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
+    \item but existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
     \end{itemize}
@@
 Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.
-Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=\bu|_{s^h}\cdot\bn_{s^h}$.
+Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$.
 Solve VI for $s^h \in \mathcal{K}^h$:
@@
 \begin{itemize}
-\item $F(s)[w] = - \int_\Omega \Phi(s)\, w$ is basically dual of $\Phi(s)=\bus\cdot\bn_s$
+\item $F(s)[w] = \int_\Omega \Phi(s)\, w$ is dual of $\Phi(s)=-\bus\cdot\bn_s$
 \item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$
@@
     \begin{itemize}
-    \item some kind of integral operator (\emph{convolve with Stokeslets})
-    \item $\Phi$ is concentrated near the diagonal, i.e.~short range
+    \item some kind of integral operator (\emph{convolve with bdry Stokeslets})
+    \item $\Phi$ is concentrated near the diagonal, i.e.~short range?
     \end{itemize}
-\item yet it is a differential operator in the SIA case:
+\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
 \small
-    $$\Phi_{\text{SIA}}(s) = \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} + \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
+    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
 \normalsize

### 2021-06-05 paper/msg.tex
Edits on IDO material, including change of sign.

@@
 
+\newcommand{\nn}{{\text{n}}}
 \newcommand{\pp}{{\text{p}}}
@@
 We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.
-Allowing any Glen exponent $n\ge 1$, the equations inside $\Lambda_s$ are
+Allowing any Glen exponent $\nn\ge 1$, the equations inside $\Lambda_s$ are
 \begin{align}
@@
 \nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
-\tau &= B_n |D\bu|^{(1/n) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
+\tau &= B_\nn |D\bu|^{(1/\nn) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
 \end{align}
@@
 
-In computations we will use constants $n=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.
-The ice hardness $B_n$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.
+In computations we will use constants $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.
+The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.
 
-For the $n=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $n>1$ imply an effective viscosity function of $|D\bu|$.
+For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.
 This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
 \begin{equation}
-\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
+\nu_\eps = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
 \end{equation}
-where $\pp=(1/n)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.
+where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.
 
@@
 \nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
-\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag \\
-\bu &= \bzero && \text{on $\Gamma_0$} \notag
+\bu &= \bzero && \text{on $\Gamma_0$} \notag \\
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
 \end{align}
-(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.
+(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is included.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.
 
@@
 Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).
-Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=n+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.
+Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.
 Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
 \begin{equation}
-\mathcal{M}_{\text{GS}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
+\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
 \end{equation}
-be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy the weak form
+be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ should satisfy the weak form
 \begin{equation}
-F_{\text{GS}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
+F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
 \end{equation}
-for all $(\bv,q) \in \mathcal{M}_{\text{GS}}$.
+for all $(\bv,q) \in \mathcal{M}_\Lambda$.
 
@@
 be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.
-From now on we also assume that $n>1$.
-Because $\qq = n+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.
+From now on we also assume that $\nn>1$.
+Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.
+(Though it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, the constraint is surely correct.
+Existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)
 
-However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
+Recalling definition \eqref{eq:lambdas}, we further suppose that
 \begin{equation}
@@
 \end{equation}
-Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.
-However, the existing theory of well-posedness for the SIA model includes a version of definition \eqref{eq:Kconstraintset} \cite{JouvetBueler2012}, wherein a power of the ice thickness, $u=H^{2q/(q-1)}$, is in $\mathcal{K} = \{u \ge 0\} \subset W^{1,\qq}(\Omega)$, and furthermore in the SIA the surface $s$ solves an elliptic equation and is thus smooth on the set $\{s<b\} \subset \Omega$.
+Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.
+In the SIA model the surface $s$ solves an elliptic equation, so it is smooth on $\{s>b\} \subset \Omega$ if $b$ is itself smooth \cite{JouvetBueler2012}.
 
-We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.
-For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
+We now define the ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).
+For each surface elevation function $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
 \begin{equation}
-\Phi(s) = \begin{cases} \bu|_s \cdot \bn_s, & s > b, \\
+\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                         0, & \text{otherwise}. \end{cases} \label{eq:ido}
@@
 Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
-Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.
+Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.
 
@@
 \end{center}
-\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.
+\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.
 Both the input and output of $\Phi$ are defined on all of $\Omega$.}
@@
 
-FIXME TALK ABOUT IDO.  1 FORM IN SIA CASE. 2 LINEARIZATION DENSE.  3 SPECTRUM OF IDO DETERMINES STABILITY OF EXPLICIT TIME-STEPPING. 4 DIAGONAL OF IDO UNKNOWN BUT POSITIVE?
+We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.
+This operator is \emph{expensive}; evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}.
+Furthermore the operator is clearly \emph{nonlocal} because the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.
+In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different everywhere both solutions are defined: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.
+Thus an FE discretization of $\Phi(s)$ will have a dense Jacobian matrix
 
-Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may immediately re-write the \eqref{eq:strongform} in a simpler strong form, namely an NCP with a hidden dynamical problem:
+By contrast, in the simpler SIA model the operator is a nonlinear, second-order differential operator,
+\begin{equation}
+\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
+\end{equation}
+for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ and $\grad_2 = (\partial_x,\partial_y)$.
+Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.
+However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid.
+
+FIXME THREE IDO REASONS TO CARE: EXPRESS STRONG AND WEAK FORM, DIAGONAL OF IDO UNKNOWN BUT POSITIVE? IN SMOOTHER, SPECTRUM OF IDO DETERMINES STABILITY OF TIME-STEPPING
+
+Now we may re-write \eqref{eq:strongform} more simply as an NCP with a hidden dynamical problem:
 \begin{align}
 s - b &\ge 0  \label{eq:idostrongform} \\
-- \Phi(s) - a &\ge 0 \notag \\
-(s - b) (- \Phi(s) - a) &= 0 \notag
+\Phi(s) - a &\ge 0 \notag \\
+(s - b) (\Phi(s) - a) &= 0 \notag
 \end{align}
-Here the statements hold on all of $\Omega$ and the data $a,b$ are clear.  The solution is the ice surface $z=s(x,y)$, but $\bu,p$ are hidden variables within the evaluation of $\Phi(s)$.
+Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.
+The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.
 
-For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a functional which is nonlinear in $s$:
+For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a ``residual'' functional which is simply the dual of $\Phi$:
 \begin{equation}
-F(s)[r] = - \ip{\Phi(s)}{r} = - \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
+F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
 \end{equation}
-One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.
-Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ is in the dual space of $W^{1,\qq}(\Omega)$, so it possesses minimal regularity.
+One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.
 
-Finally, the SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
+The SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
 \begin{equation}
@@
 \end{equation}
-This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.
-Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.
-In any case one can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
+One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
 In particular the strong SKE \eqref{eq:ske} holds on $s>b$.
 
-Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.
-Speaking geometrically, VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.
-Specifically, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
+Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $\Phi(s)-a$ points directly into $\mathcal{K}$.
+The VI says that the ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
+Also, it is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
 
-It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
+Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.
 The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
 
-The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
+The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The VI problem therefore has three fundamental nonlinearities:
 \renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
@@
 \item the inequality constraint, and
-\item the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
+\item the nonlinearity of free-surface flow as a function of the surface.
 \end{enumerate}
-Regarding (\emph{ii}), observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}.
+Regarding (\emph{ii}), observe that the solution of the classical obstacle problem for the linear Laplacian operator is a nonlinear function of the data \cite{KinderlehrerStampacchia1980}.
 Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.
 
-We will need the (Gateaux) derivative of $F$, and this requires differentiating $\Phi$ as well.
+We will need the (Gateaux) derivative of $F$, thus $\Phi$.
 Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
-(In the following, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional}, if $r\in W_{1,\qq}(\Omega)$ then
+From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then $F'(s)[t,r] = \int_\Omega \Phi'(s)[t]\, r$ where by definition
 \begin{equation}
-F'(s)[t,r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s)[t]\, r \,dx dy \label{eq:sigpfunctionalderiv}
+\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}, \label{eq:idoderiv}
 \end{equation}
-where by definition
-\begin{equation}
-\Phi'(s)[t] = \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}. \label{eq:idoderiv}
-\end{equation}
-In \eqref{eq:idoderiv} note that all quantities are extended to the whole of $\Omega$.
-
-We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.
-Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.
-In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.
-However, to simplify our considerations we define an $s$-independent set
+extended by zero to $\Omega$.
+However, $\Phi'(s)[t]$ is actually a one-sided directional derivative on a portion of $\Omega$.
+That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.
+To simplify we define an $s$-independent set
 \begin{equation}
@@
 \end{equation}
-It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
-We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{W}_+$ has small support.
+It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
 
-In fact the meaning of the finite difference in \eqref{eq:idoderiv} depends on the support of $t$:
+The meaning of the difference in \eqref{eq:idoderiv} depends on the support of $t$:
 \begin{equation}
@@
 Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.
-In physical terms this means that the dynamical part of the SKE \eqref{eq:ske} vanishes for small ice masses on bare ground, which is reasonable to suppose if the ice cannot slide.
+In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for small ice masses on bare ground, a reasonable supposition if the ice cannot slide.
 
 In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.
-When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because an extra solution of Glen-Stokes problem \eqref{eq:glenstokesweak} is required over the icy domain $\Lambda_{s+\eps t}$, but in Section \ref{sec:smoothers} we propose a less-expensive approximation.
+When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, significant performance and stability issues arise.
+These are addressed in Section \ref{sec:smoothers}.
 
@@
 In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.
-The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).
-Our nonlinear variant, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
+The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
 
-Regarding notation for multiple mesh levels, suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, and call it the coarse level.
+Regarding notation for multiple mesh levels, suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, the coarse level.
 For $J\ge 0$, suppose $\{\mathcal{T}^j\}_{j=1}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
-(We assume simple halving of each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and let $\mathcal{V}_0^j \subset \mathcal{V}^j$ be the $P_1$ functions which are zero on $\partial\Omega$.
-If $m_j$ is the number of interior nodes in $\mathcal{T}$ then $\dim(\mathcal{V}_0^j)=m_j$.
-
+(Halve each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$, with subspace $\mathcal{V}_0^j \subset \mathcal{V}^j$ which are zero on $\partial\Omega$.
+If $m_j$ is the number of interior nodes $\{(x_i^j,y_i^j)\}$ in $\mathcal{T}$ then $\dim(\mathcal{V}_0^j)=m_j$.
 We seek an admissible solution to \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.
 Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.
-On this fine level the admissible surface elevations form a closed and convex set
+The admissible surface elevations form a closed and convex set
 \begin{equation}
@@
 \end{equation}
-However, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition of problem \eqref{eq:fesigpweakform} based on a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.
+
+However, we follow \cite{GraeserKornhuber2009} in decomposing a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.
 That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
@@
 \end{equation}
-be the associated defect obstacle, a nonpositive function on $\Omega$.
+be the associated defect obstacle, with $\chi^J \le 0$.
 Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.
-
-Also following \cite{GraeserKornhuber2009}, we define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$.
-Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and denote the level-$j$ nodes as $(x_i^j,y_i^j)$.
-If $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ for coefficients $z^j[i]\in \RR$ then
+Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$ \cite{GraeserKornhuber2009} on $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ by maximizing nodal values over the (open) supports of each each coarser-level hat function:
 \begin{equation}
@@
 \end{equation}
-In words, the coefficients for $\mR z^j \in \mathcal{V}_0^j$ are computed by maximizing the finer-level nodal values over the (open) supports of each each coarser-level hat function.
-
-Now we can recursively define a defect obstacle on each level:
+Now a defect obstacle is defined on each level:
 \begin{equation}
@@
 \end{equation}
-for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences of the defect obstacles,
+for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences,
 \begin{equation}
-\phi^j = \chi^j - \chi^{j-1}.  \label{eq:downobstacles}
+\phi^j = \chi^j - \chi^{j-1},  \label{eq:downobstacles}
 \end{equation}
-In terms of these functions, for $j=0,\dots,J$ we define down-constraint sets
+with $\phi^0=\chi^0$.  Smoothers solve one-dimensional VIs on each level in different admissible sets as one goes down and up in the V-cycle; we define down- and up-constraint sets
 \begin{equation}
-\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}.  \label{eq:downconstraintset}
+\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}, \qquad \mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}. \label{eq:downupsets}
 \end{equation}
-and up-constraint sets
-\begin{equation}
-\mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}.  \label{eq:upconstraintset}
-\end{equation}
-This terminology of down- and up-constraint sets is explained by noting that the smoother will solve one-dimensional VIs on each level, as described in the previous section, where the admissible sets are different as one goes down and then up in a V-cycle.
 

### 2021-06-05 paper/msg.tex
Remove a FIXME.

@@
 
-FIXME THREE IDO REASONS TO CARE: EXPRESS STRONG AND WEAK FORM, DIAGONAL OF IDO UNKNOWN BUT POSITIVE? IN SMOOTHER, SPECTRUM OF IDO DETERMINES STABILITY OF TIME-STEPPING
+We will be concerned with three aspects of the IDO:
+\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
+\begin{enumerate}
+\item Immediately we use $\Phi$ to simplify the SIGP strong form and express its weak form.
+\item Section \ref{sec:smoothers} approximates the diagonal entries of the Jacobian $\Phi'$ to construct smoothers.
+\item Section \ref{sec:evolution} relates the spectrum of $\Phi'$ to the stability of time-stepping methods.
+\end{enumerate}
 
-Now we may re-write \eqref{eq:strongform} more simply as an NCP with a hidden dynamical problem:
+For (\emph{i}) we re-write \eqref{eq:strongform} as a simpler NCP with a hidden dynamical problem:
 \begin{align}

### 2021-06-06 paper/msg.tex
Another Jouvet paper to cite (follows Pralong&Funk).

@@
 However, as a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.
-There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that $s$ is well-defined.
-Furthermore the same concern applies to each time step of an evolving model; actual glaciers can develop overhangs at the margin.
-However, because overhangs are small features in large glaciers and ice sheets, essentially all modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.
-An exception is \cite{PralongFunk2005}, who suggest a potentially well-posed continuum model, a nontrivial extension of the one here, in which serac and ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion.
-Though marginal cliffs may exist in our model (Section \ref{sec:fe}), overhangs are never allowed.
+There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that the function $s$ is well-defined.
+Furthermore the same concern applies to each time step of an evolving model; overhangs can develop at the margin.
+Because overhangs are small features in large glaciers and ice sheets, most modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.
+Exceptions \cite{PralongFunk2005} and \cite{Jouvetetal2011} provide a continuum model in which ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion, and such a model might explain how (essentially) well-defined surface elevation and thickness could arise in a well-posed model.
+However, such a model is a nontrivial extension of the above equations, and though marginal cliffs may exist in our model (Section \ref{sec:fe}), overhangs are never allowed herein.
 
@@
 Recall that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).
-By \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute
+By \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute
 \begin{equation}

### 2021-06-07 py/basesmoother.py
Minor: remove mention of a nonexistent argument.

@@ -48,12 +48,11 @@ class SmootherObstacleProblem(ABC):
         return mesh.l2norm(F)
 
     def smoother(self, iters, mesh, w, ell, phi):
-        '''Apply iters sweeps of obstacle-problem smoother on mesh to modify w in place.  Alternate directions (unless overridden).'''
+        '''Apply iters sweeps of obstacle-problem smoother on mesh to modify w in place.  Alternate directions.'''
         forward = True
         for _ in range(iters):
             self.smoothersweep(mesh, w, ell, phi, forward=forward)
-            if not self.args.sweepsnotalternate:
-                forward = not forward
+            forward = not forward
 
     @abstractmethod
     def residual(self, mesh, w, ell):

### 2021-06-07 paper/msg.tex
Complete first draft of MCDN description.

@@
 In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.
-The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
+The fundamental idea of such a scheme is the basic geometric multigrid idea that smoothing on coarse levels will rapidly reduce the high frequencies in the error, but here we must maintain admissibility on each level in a manner which does not re-introduce high frequencies.
+Our method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
 
@@
 \end{equation}
-Now a defect obstacle is defined on each level:
+Using $\mR$ a defect obstacle is defined on each level:
 \begin{equation}
@@
 \end{equation}
-for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences,
+for $j=1,\dots,J$.  An example is shown in Figure \ref{fig:decompclassical}.  We also compute the differences,
 \begin{equation}
@@
 \end{equation}
-with $\phi^0=\chi^0$.  Smoothers solve one-dimensional VIs on each level in different admissible sets as one goes down and up in the V-cycle; we define down- and up-constraint sets
-\begin{equation}
-\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}, \qquad \mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}. \label{eq:downupsets}
-\end{equation}
+with $\phi^0=\chi^0$ by definition; note $\phi^j\le 0$.
 
@@
 
-FIXME MCD theory
+A point-wise smoother on a given level will solve one-dimensional VIs at each node.  However, different admissible sets are used going down and up in the V-cycle.  Let
+\begin{equation}
+\mathcal{D}^j = \left\{y^j \in \mathcal{V}_0^j \,:\, y^j \ge \phi^j\right\}, \qquad \mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}. \label{eq:downupsets}
+\end{equation}
+The down-admissible constraint sets $\mathcal{D}^j$ sum to an up-admissible constraint set for each level, i.e.~$\mathcal{D}^0 + \mathcal{D}^1 + \dots + \mathcal{D}^j = \mathcal{U}^j$ in the sense that every $z^j \in \mathcal{U}^j$ is the sum of elements, though not uniquely.
+The $\mathcal{U}^j$ are themselves nested in $\mathcal{U}^J= \{z^J \,:\, z^J \ge b^J - s^J\}$, and this forms a multilevel cone approximation, from the inside, of the continuum defect constraint set $\mathcal{K} - s = \{z\,:\,z \ge b - s\}$:
+\begin{equation}
+\mathcal{U}^0 \subset \mathcal{U}^1 \subset \dots \subset \mathcal{U}^J \subset \mathcal{K} - s \subset W_0^{1,\qq}(\Omega). \label{eq:innerconeapprox}
+\end{equation}
 
-FIXME DEBUG THIS PSEUDOCODE nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}
+We seek a solution to VI \eqref{eq:fesigpweakform} on the fine level.
+The original idea of Tai \cite{Tai2003} is to use a multilevel constraint decomposition (MCD) like the above  system $\{\mathcal{D}^j\}$ to compute corrections $y^j \in \mathcal{D}^j$ which sum to a solution of the (nonlinear) fine-level VI,
+\begin{equation}
+F^J(s^J + y^J + \dots + y^j)[v^j - y^j] \ge (a,v^j - y^j) \quad \text{for all } v^j \in \mathcal{D}^j, \label{eq:mcdoriginal}
+\end{equation}
+where $F^J = F^h$ in \eqref{eq:fesigpweakform}.
+In \cite{Tai2003}, VIs \eqref{eq:mcdoriginal} are to be solved on descending levels $j=J$ down to $j=0$, forming a down-slash or V(1,0) cycle.
+(See also \cite[Algorithm 4.7]{GraeserKornhuber2009}.)  Note that $y^J + \dots + y^j \in \mathcal{U}^J$ so $s^J + y^J + \dots + y^j \ge b^J$ is admissible in the original sense, and that \eqref{eq:mcdoriginal} exploits the simplification $(s^J + y^J + \dots + v^j) - (s^J + y^J + \dots + y^j) = v^j - y^j$.
+
+However, there is no practical way to achieve rapid solution of VI \eqref{eq:mcdoriginal} on coarse levels if the latest iterate $s^J + y^J + \dots + y^j$ must be represented; it has fine-level content.
+This is the reason to add the full approximation scheme (FAS) \cite{Trottenbergetal2001} concept of a nonlinear coarse-level ``equation'', which is here a nonlinear VI.
+In order to implement our approach we need to state the FAS coarse-level correction equation based on appropriate restrictions, as follows.
+
+Let $\iR: \mathcal{V}^{j+1} \to \mathcal{V}^j$ be the injection restriction operator \cite{Trottenbergetal2001}.
+Suppose $y^j \in \mathcal{D}^j$ denotes the desired down-correction on the $j$th level.
+For corrections $y^j \in \mathcal{D}^j$, define $t^j \in \mathcal{V}^j$ as the $j$th-level approximation to the corrected fine-level solution (i.e.~$t^j \approx s^J + y^J + \dots + y^j$),
+\begin{equation}
+t^j = \begin{cases} s^J, & j=J \\
+                    \iR(t^{j+1} + y^{j+1}), & j < J.
+      \end{cases}  \label{eq:fassolution}
+\end{equation}
+Note that $t^j$ satisfies $t^j \ge \iR(\dots(\iR(b^J))\dots)$ by construction because $y^j \in \mathcal{D}^j$.
+Though represented on a coarse level, $t^j$ has admissible point values relative to the fine bed $b^J$.
+
+Let $F^j$ represent the FE discretization of $F$ (Section \ref{sec:fe}) on the $j$th level, and let $R: (\mathcal{V}^{j+1})^* \to (\mathcal{V}^j)^*$ be the canonical restriction on functionals \cite{GraeserKornhuber2009}.
+The FAS coarse-level VI is
+\begin{equation}
+F^j(t^j+y^j)[v^j-y^j] \ge \ell^j[v^j-y^j] \quad \text{for all } v^j \in \mathcal{D}^j. \label{eq:fasequation}
+\end{equation}
+where
+\begin{equation}
+\ell^j[v] = \begin{cases} \ip{a}{v}, & j=J \\
+                          F^j(t^j)[v] + R \left(\ell^{j+1} - F^{j+1}(t^{j+1}+y^{j+1})\right)[v], & j < J. \end{cases}  \label{eq:fasell}
+\end{equation}
+On the one hand, all quantities in VI \eqref{eq:fasequation} are represented on the $j$th level, without reference to finer-mesh data.
+On the other hand, for $j<J$ the VI \eqref{eq:fasequation} can be rearranged using \eqref{eq:fasell} to a correction form like the FAS equation for PDEs (e.g.~\cite[equation (5.3.12)]{Trottenbergetal2001}),
+\begin{equation}
+F^j(t^j+y^j)[v^j-y^j] - F^j(t^j)[v^j-y^j] \ge R \left(\ell^{j+1} - F^{j+1}(t^{j+1}+y^{j+1})\right)[v^j-y^j], \label{eq:fasequationtraditional}
+\end{equation}
+for all $v^j \in \mathcal{D}^j$.
+In \eqref{eq:fasequationtraditional} the functional on the left and right side should be smooth, thus explaining \eqref{eq:fasequation}.
+Following the solution of \eqref{eq:fasequation} on all descending levels $j=J,\dots,0$, in fact done by a smoother on each level, a V(1,0) cycle is completed and the new fine-mesh iterate is $s^J + y^J + \dots + y^0$.
+
+The above formulas suffice for a V(1,0) cycle, and represent the FAS extension of Algorithm 4.7 in \cite{GraeserKornhuber2009}, and of the method of Tai \cite{Tai2003}.
+However, for better performance we note that on the ascending part of a V-cycle we can smooth in a less-constrained set, the larger set $\mathcal{U}^j \supset \mathcal{D}^j$.
+Regarding admissibility, the difference between descending and ascending in a V-cycle is that in the former case there are yet-to-be-computed corrections which we must force into small-enough sets $\mathcal{D}^j$ so that their sum is still admissible.
+When ascending we have already summed the coarser-level corrections so we can smooth the result in the larger, nested sets $\mathcal{U}^j$ without destroying the admissibility of the yet-finer corrections to come.
+
+The above ideas define the following MCDN V-cycle Pseudocode \ref{pc:mcdn-vcycle}.
+This algorithm calls a smoothers from Section \ref{sec:smoothers} and it uses the canonical prolongation $P:\mathcal{V}^j \to \mathcal{V}^{j+1}$.
+The default settings correspond to a V(0,1) cycle which we have found to be most efficient in tests on the classical obstacle problem \cite{Bueler2022}.
+The returned value is a correction to the current iterate $s^J$.
 
@@
 \begin{pseudo*}
-\pr{mcdn-vcycle}(J,s^J,b^J,\ell^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
-    $\chi^J = b^J - s^J$ \\
-    $g^J = s^J$ \\
+\pr{mcdn-vcycle}(J,s^J,b^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
+    $\chi^J, \,\ell^J, \,t^J = b^J - s^J, \,\ip{a}{\cdot}, \,s^J$ \\
     for $j=J$ downto $j=1$ \\+
@@
       $y^j = 0$ \\
-      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,g^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\
-      $f = F^j(g^j+y^j) - \ell^j$ \\
-      $g^{j-1} = \iR(g^j + y^j)$ \\
-      $\ell^{j-1} = F^{j-1}(g^{j-1}) - R f$ \\-
+      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,t^j,\ell^j,\phi^j)$ \qquad \qquad \ct{in $\mathcal{D}^j$} \\
+      $t^{j-1} = \iR(t^j + y^j)$ \\
+      $\ell^{j-1} = F^{j-1}(t^{j-1}) + R(\ell^j - F^j(t^j+y^j))$ \\-
     $y^0 = 0$ \\
-    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,g^0,\ell^0,\chi^0)$ \qquad \ct{smoothing in $\mathcal{U}^0$} \\
+    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,t^0,\ell^0,\chi^0)$ \qquad \qquad \ct{in $\mathcal{U}^0$} \\
     $z^0 = y^0$ \\
@@
       $z^j = P z^{j-1} + y^{j}$ \\
-      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,g^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{U}^j$} \\-
+      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,t^j,\ell^j,\chi^j)$ \qquad \qquad \ct{in $\mathcal{U}^j$} \\-
     return $z^J$
 \end{pseudo*}
-\caption{FIXME Nonlinear MCD V-cycle.}
+\caption{MCDN V-cycle.}
 \label{pc:mcdn-vcycle}
@@
 
-Finally, we need to be clear about when an iteration has converged.
-For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
+To our knowledge two aspects of this MCDN V-cycle are new:
+\begin{enumerate}
+\item While Tai \cite{Tai2003} defines and proves the convergence of a nonlinear MCD method, the implementation of this method cannot achieve $O(m_J)$ time for each V-cycle in nonlinear cases because the method refers to the fine level for the residual evaluation on each coarser level.
+Our FAS-type modification achieve $O(m_J)$ time per V-cycle when applied to a local nonlinear problem, for example in the SIA model application in \cite{Bueler2022} in which the smoother is $O(m_j)$ on each level.
+In our current Stokes case the smoother is nonlocal and $O(m_j)$ smoother time is not achieved.
+\item Our V($\alpha$,$\beta$) cycles, with $\alpha$ down-smoother and $\beta$ up-smoother applications, are more efficient when $\beta >0$ than the method proposed in \cite{GraeserKornhuber2009} for $V(1,1)$ cycles because the smoothing occurs in a less-constrained set.
+\end{enumerate}
+
+Finally we consider when iterated V-cycles have converged.
+For any discrete obstacle $\chi^j \in \mathcal{V}^j$, admissible iterate $z^j\in \{r^j \ge \chi^j\} \subset \mathcal{V}^j$, and residual vector $r^j \in (\mathcal{V}^j)'$ let
 \begin{equation}
-\vertiii{F^j}_{(s^j,\varphi^j)} = \left(\sum_{s_i > \varphi_i} |F[\psi_i]|^2 + \sum_{s_i = \varphi_i} |\min\{F[\psi_i],0\}|^2\right)^{1/2} \label{eq:cpnorm}
+\vertiii{r^j}_{(z^j,\chi^j)} = \left(\sum_{z_i > \chi_i} |r^j[\psi_i^j]|^2 + \sum_{z_i = \chi_i} |\min\{r^j[\psi_i^j],0\}|^2\right)^{1/2}, \label{eq:cpnorm}
 \end{equation}
-be the ``CP residual norm'' associated to the finite-dimensional complementarity problem (CP) $s^j \ge \varphi^j$, $F(s^j) \ge 0$, and $(s^j-\varphi^j) F(s^j) = 0$.
-Note $s^j \in \mathcal{V}^j$ solves this CP if and only if $s^j$ is admissible and $\vertiii{F(s^j)}_{(s^j,\varphi^j)}=0$.
-As shown in Pseudocode \ref{pc:mcdn-solver} we will iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.
+where $\psi_i^j$ denotes a $j$th-level nodal basis function.
+This defines a ``CP residual norm'' associated to the finite-dimensional complementarity problem $z^j \ge \chi^j$, $r^j(z^j) \ge 0$, and $(z^j-\chi^j) r^j(z^j) = 0$.
+Note $z^j \in \mathcal{V}^j$ solves this CP if and only if $z^j$ is admissible and $\vertiii{r^j(z^j)}_{(z^j,\chi^j)}=0$.
+
+As shown in Pseudocode \ref{pc:mcdn-solver}, we iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.
 
@@
 \pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
-    function $\rho(s)[r] = F^J(s)[r] - \ip{a}{r}$ \qquad\qquad \ct{residual of \eqref{eq:fesigpweakform}} \\
-    $\rho_0=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
+    $\rho_0=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
     for $k=1,\dots,\id{cyclemax}$ \\+
-        $\chi^J = b^J - s^J$ \qquad\qquad \ct{fine-level defect obstacle} \\
-        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,\chi^J)$ \\
-        $\rho_k=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
+        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,b^J)$ \\
+        $\rho_k=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
         if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ \\+
@@
 
-\section*{Acknowledgments}  Thanks to David Maxwell for suggestions on the formulation and well-posedness of the model.
+\section*{Acknowledgments}  Thanks to David Maxwell for discussions regarding the formulation and well-posedness of the model.
 

### 2021-06-07 paper/msg.tex
Nothing fiddles.

@@
 
+\date{\today}
+
 \begin{abstract} FIXME MCD for steady and evolving geometry with Glen-Stokes dynamics
@@
 
-FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020};  The computations in this paper use the Python FE library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.
+FIXME a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022} and \cite{GraeserKornhuber2009};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020}
 
@@
 \begin{center}
-FIXME %\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
+FIXME %\includegraphics[width=\textwidth]{genfigs/massanalogy.pdf}
 \end{center}
@@
 \begin{pseudo*}
-\pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
+\pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{stol}=10^{-20},\id{max}=100)\text{:} \\+
     $\rho_0=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
     for $k=1,\dots,\id{cyclemax}$ \\+
-        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,b^J)$ \\
+        $z^J = \pr{mcdn-vcycle}(J,s^J,b^J)$ \\
+        $s^J \gets s^J + z^J$ \\
         $\rho_k=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
-        if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ \\+
+        if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ or $\|z^J\| \le \id{stol}$ \\+
             break \\--
 \end{pseudo*}
-\caption{The SIGP is solved by iterating V-cycles (Pseudocode \ref{pc:mcdn-vcycle}) until the CP residual norm \eqref{eq:cpnorm} is small.}
+\caption{The SIGP is solved in-place by iterating V-cycles (Pseudocode \ref{pc:mcdn-vcycle}) until the CP residual norm \eqref{eq:cpnorm} is small.}
 \label{pc:mcdn-solver}
@@
 
-FIXME in this paper the Glen-Stokes problem is solved by Newton linearization and (parallel) direct solution of the Newton step equations; Firedrake and PETSc \cite{Balayetal2020,Bueler2021} used to solve this dynamics problem and thereby evaluate the residual $F$; geometric multigrid schemes exist for the  Glen-Stokes problem \cite{IsaacStadlerGhattas2015}; see also \cite{BrownSmithAhmadia2013} and \cite{Tuminaroetal2016} which solve a first-order approximation of the Glen-Stokes ice sheet problem by geometric and algebraic multigrid, respectively
+FIXME The computations in this section use the Python FE library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.
+Each time the smoother is applied the Glen-Stokes problem is solved, by Newton linearization and (parallel) direct solution of the Newton step equations, in order to evaluate the IDO $\Phi(s)$.
+To replace this non-scalable approach we propose in future work to apply existing geometric \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015} and algebraic \cite{Tuminaroetal2016} multigrid strategies.
 
@@
 
-\section*{Acknowledgments}  Thanks to David Maxwell for discussions regarding the formulation and well-posedness of the model.
+\section*{Acknowledgments}  Thanks to David Maxwell for discussions regarding the formulation of the model.
 

### 2021-06-08 paper/genfigs/stokesdomain.py
Change the IDO figure to show a discontinuity.

@@ -67,7 +67,7 @@ offset = 0.1
 xl, xr = min(x[s>b]), max(x[s>b])
 dx = xr - xl
 xsc = (x[s>b] - xl) / dx
-y = s[s>b] + offset + 0.8 * np.sin(5.0*np.pi*xsc) * np.cos(np.pi*xsc)**4
+y = s[s>b] - offset + 1.2 * np.sin(5.5*np.pi*xsc) * np.cos(np.pi*xsc)**4
 plt.plot(x[s>b], y, color='k', lw=1.0)
 plt.plot(x[x < xl], s[x < xl] + offset, color='k', lw=1.0)
 plt.plot(x[x > xr], s[x > xr] + offset, color='k', lw=1.0)

### 2021-06-08 paper/msg.tex
Edits sections 2-4.

@@
 \begin{equation}
-\nu_\eps = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
+\nu_\eps(|D\bu|) = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
 \end{equation}
@@
 \begin{equation}
-\left(2 \nu_\eps D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
+\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
 \end{equation}
@@
 (s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
-- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
+- \nabla \cdot \left(2 \nu_\eps(|D\bu|)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
 \nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
 \bu &= \bzero && \text{on $\Gamma_0$} \notag \\
-\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
+\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
 \end{align}
-(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is included.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.
+The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.
 
@@
 
-The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here, after which we address the weak form of the SIGP.
+The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.
+Then we address the weak form of the SIGP.
 
-Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.
-Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).
+Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ weak derivatives which are $r$th-power integrable.
+Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of $\partial\Lambda$, and otherwise suppose the boundary is stress-free (Neumann boundary).
 Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.
@@
 \end{equation}
-be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ should satisfy the weak form
+be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ satisfies the weak form
 \begin{equation}
-F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
+F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps(|D\bu|) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
 \end{equation}
-for all $(\bv,q) \in \mathcal{M}_\Lambda$.
+for all $(\bv,q) \in \mathcal{M}_\Lambda$.  If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.
 
 Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.
-In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p$.
+They show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, and that there is a unique pressure $p$.
 
-However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.
+However, we must go beyond the problem on predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.
 Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.
@@
 Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.
-(Though it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, the constraint is surely correct.
-Existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)
-
-Recalling definition \eqref{eq:lambdas}, we further suppose that
+(Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
 \begin{equation}
@@
 \end{equation}
-Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.
-In the SIA model the surface $s$ solves an elliptic equation, so it is smooth on $\{s>b\} \subset \Omega$ if $b$ is itself smooth \cite{JouvetBueler2012}.
+Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, something we cannot offer.
+However, in the SIA model the surface $s$ solves an elliptic equation, and $s$ is $C^1$ on $\{s>b\} \subset \Omega$ if $b$ is continuous \cite{JouvetBueler2012}.
 
-We now define the ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).
-For each surface elevation function $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
+We now define an ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem \eqref{eq:glenstokesweak}; see Figure \ref{fig:idoaction}.
+For each $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
 \begin{equation}
@@
 Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.
-Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.
+Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or be singular at the glacier margin even if $s$ is smooth on $\{s>b\}$.
+(We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.)
 
@@
 \end{center}
-\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.
-Both the input and output of $\Phi$ are defined on all of $\Omega$.}
+\caption{The IDO $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$.
+Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ may be discontinuous as shown.}
 \label{fig:idoaction}
@@
 We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.
-This operator is \emph{expensive}; evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}.
-Furthermore the operator is clearly \emph{nonlocal} because the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.
-In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different everywhere both solutions are defined: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.
-Thus an FE discretization of $\Phi(s)$ will have a dense Jacobian matrix
+This operator is \emph{expensive} and \emph{nonlocal}.
+Evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}, and furthermore the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.
+In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.
+It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix
 
-By contrast, in the simpler SIA model the operator is a nonlinear, second-order differential operator,
+By contrast, in the simpler isothermal, nonsliding SIA model the operator is a nonlinear, second-order differential operator,
 \begin{equation}
@@
 \end{equation}
-for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ and $\grad_2 = (\partial_x,\partial_y)$.
+for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ constant and $\grad_2 = (\partial_x,\partial_y)$.
 Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.
-However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid.
+However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in this sense we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.
 
-We will be concerned with three aspects of the IDO:
+In this section we will use $\Phi$ to simplify the SIGP strong form and express its weak form, but later we will be concerned with two other aspects of the IDO:
 \renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
 \begin{enumerate}
-\item Immediately we use $\Phi$ to simplify the SIGP strong form and express its weak form.
 \item Section \ref{sec:smoothers} approximates the diagonal entries of the Jacobian $\Phi'$ to construct smoothers.
@@
 
-For (\emph{i}) we re-write \eqref{eq:strongform} as a simpler NCP with a hidden dynamical problem:
+For now we re-write \eqref{eq:strongform} as a clean NCP with a hidden dynamical problem:
 \begin{align}
@@
 
-For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a ``residual'' functional which is simply the dual of $\Phi$:
+To show the SIGP weak form we define a functional which is simply the dual of $\Phi$:
 \begin{equation}
@@
 One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.
-
-The SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
+The weak form itself is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
 \begin{equation}
@@
 One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
-In particular the strong SKE \eqref{eq:ske} holds on $s>b$.
+In particular the strong SKE \eqref{eq:ske} holds on $s>b$; it is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.
 
-Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $\Phi(s)-a$ points directly into $\mathcal{K}$.
-The VI says that the ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.
-Also, it is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.
+Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a point where $\Phi(s)-a$ points directly into $\mathcal{K}$.
+(The ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.)  It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge \eqref{eq:sigpweakform} does \emph{not} arise in this way.
+Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.
+This model is also not a minimization, except in the flat-bed case.
 
-Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.
-The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.
-
-The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The VI problem therefore has three fundamental nonlinearities:
+The SIGP weak form \eqref{eq:sigpweakform} is coupled to the Glen-Stokes weak form \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.
+VI problem \eqref{eq:sigpweakform} therefore has three fundamental nonlinearities:
 \renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
@@
 Regarding (\emph{ii}), observe that the solution of the classical obstacle problem for the linear Laplacian operator is a nonlinear function of the data \cite{KinderlehrerStampacchia1980}.
-Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.
+For (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.
 
-We will need the (Gateaux) derivative of $F$, thus $\Phi$.
+We will need the (Gateaux) derivative of $\Phi$.
 Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
-From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then $F'(s)[t,r] = \int_\Omega \Phi'(s)[t]\, r$ where by definition
+From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then by definition
 \begin{equation}
-\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}, \label{eq:idoderiv}
+\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv}
 \end{equation}
-extended by zero to $\Omega$.
-However, $\Phi'(s)[t]$ is actually a one-sided directional derivative on a portion of $\Omega$.
+on $\{s>b\}$.
+However, $\Phi'(s)[t]$ must actually be defined as a one-sided directional derivative on $\{s=b\}$.
 That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.
@@
 \end{equation}
-It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
+It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]=\int_\Omega \Phi'(s)[t] r$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
 
-The meaning of the difference in \eqref{eq:idoderiv} depends on the support of $t$:
+The difference in \eqref{eq:idoderiv} depends on the support of $t$:
 \begin{equation}
@@
 \end{equation}
-Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.
-In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for small ice masses on bare ground, a reasonable supposition if the ice cannot slide.
+Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $\{s=b\}$.
+In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for shrinking ice masses on bare ground, a reasonable supposition if the ice cannot slide.
 
 In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.
-When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, significant performance and stability issues arise.
-These are addressed in Section \ref{sec:smoothers}.
+Substantial performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.
 
@@
 
-Assume $\Omega \subset \RR^d$ is polygonal and that $\mathcal{T}$ is a triangulation of $\Omega$.
+Assume $\Omega \subset \RR^d$ is polygonal with triangulation $\mathcal{T}$.
 (If $d=1$ then $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
@@
 \end{equation}
-Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.
+Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding SIA problem, and that, because of low regularity at the free boundary, $P_1$ is a common choice even for the classical obstacle problem with smooth data \cite{GraeserKornhuber2009}.
 
@@
 
-Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$).
-The triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.
+Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$), so the triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.
 As shown in Figure \ref{fig:extruded}, each icy $T \in \mathcal{T}$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.
 However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.
-An extruded mesh is regenerated each time $F^h(s^h)$ is evaluated, but the base mesh $\mathcal{T}$ is reused.
+The extruded mesh is regenerated each time $F^h(s^h)$ is evaluated, but the base mesh $\mathcal{T}$ is unchanged.
 
@@
 \end{center}
-\caption{We compute the IDO $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.
+\caption{We compute the FE IDO $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.
 In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold).
-In the ``cliffs'' extrusion (right) all elements have a minimum thickness so the trace evaluation may occur above $s^h$ (dotted).}
+In the ``cliffs'' extrusion (right) all elements have a minimum thickness so the near-margin trace evaluation occurs above $s^h$ (dotted).}
 \label{fig:extruded}
@@
 
-We will compare two extrusion modes.
-In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly.
-It follows that the elements at the ice margin are degenerate; they have positive volume but zero height at some nodes or even empty (vertical) facets.
-In the alternate ``cliffs'' extrusion, each extruded element is nondegenerate.
-Specifically, each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$ where $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in computations.
-The surface $s^h$ remains unchanged (and continuous), so these cliffs are only relevent to problem \eqref{eq:glenstokesweak}.
+On the extruded mesh we approximately solve Glen-Stokes problem \eqref{eq:glenstokesweak} using the $P_2\times P_1$ Taylor-Hood mixed space \cite{Elmanetal2014} which we denote $\mathcal{M}^h$.
+Thus $(\bu^h,p^h) \in \mathcal{M}^h$ solves
+\begin{equation}
+F_{\Lambda_{s^h}}(\bu^h,p^h)[\bv^h,q^h] = \int_{\Lambda_{s^h}} 2 \nu_\eps(|D\bu^h|) D\bu^h : D\bv^h - p^h \Div\bv^h - (\Div\bu^h) q^h - \rhoi \bg \cdot \bv^h\,d\bx = 0 \label{eq:feglenstokesweak}
+\end{equation}
+for all $(\bv^h,q^h) \in \mathcal{M}^h$.
+The power-law nonlinearity in \eqref{eq:feglenstokesweak} is resolved by using a Newton iteration, with direct solution of the step equations and back-tracking line search.
+As stated in Section \ref{sec:intro}, our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme are needed to solve the discrete SIGP, \eqref{eq:fesigpweakform} below.
+Overall optimality of the solution algorithm would also require an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
+
+In fact we will compare two extrusion modes.
+In the ``pinched'' extrusion the prism elements tile $\Lambda_{s^h}$ exactly but the elements at the ice margin are degenerate; they have positive volume but zero height at some nodes or even empty (vertical) facets.
+In the alternate ``cliffs'' extrusion, each extruded element is nondegenerate and they tile a larger domain $\tilde\Lambda_{s^h} \supset \Lambda_{s^h}$.
+Specifically, each prism has minimum thickness $H_{\text{min}}/m_z$ where $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in computations.
+The surface $s^h$ remains unchanged (and continuous), and cliffs only appear in problem \eqref{eq:feglenstokesweak}.
 
 Let $\Phi^h:\mathcal{K}^h \to (\mathcal{V}^h)'$ be the FE approximation of the IDO $\Phi$.
-By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.
+By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:feglenstokesweak} on the domain $\Lambda_{s^h}$.
 That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, and then extend by zero to all of $\Omega$.
 In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.
-
-We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}.
-The power-law nonlinearity is resolved by using a Newton iteration, with direct solution of the step equations and back-tracking line search.
-As stated in Section \ref{sec:intro}, our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme are needed to solve the SIGP \eqref{eq:sigpweakform}.
-Overall optimality of the solution algorithm also requires an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
 
@@
 The only role of the extruded mesh in solving \eqref{eq:fesigpweakform} is in the evaluation of the IDO $\Phi^h(s^h)$.
-We believe that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin, and we have no proof.
+We believe that \eqref{eq:fesigpweakform} is well-posed in each extrusion mode, but we have no proof.
 

### 2021-06-08 py/smoother.py
Minor edits to jacobicolorsweep() smoother.

@@ -366,28 +366,27 @@ class SmootherStokes(SmootherObstacleProblem):
         return negd
 
     def jacobicolorsweep(self, mesh1d, s, ella, r,
-                        eps=1.0, dump=False):
+                        eps=1.0, dump=False, cperthickness=3.0):
         '''Do in-place projected nonlinear Jacobi smoothing on s(x)
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
-        by SLOW finite differencing of expensive residual calculations
-        using coloring.
+        by SLOW finite differencing of expensive residual calculations, but
+        using coloring.  Nodes of the same color are separated by cperthickness
+        times the maximum ice thickness.
         If d_i > 0 then
             snew_i <- max(s_i - omega * r_i / d_i, b_i)
         but otherwise
             snew_i <- b_i.
         After snew is completed we do s <- snew.'''
         thkmax = max(s - mesh1d.b)
-        cgoal = 3.0  # coloring every 3 ice thickness, at least
-        c = int(np.ceil(cgoal * thkmax / mesh1d.h))
+        c = int(np.ceil(cperthickness * thkmax / mesh1d.h))
         print('      c = %d' % c)
-        snew = mesh1d.b.copy() - 1.0  # check below for admissibility
+        snew = mesh1d.b.copy() - 1.0  # snew NOT admissible; note check below
         snew[0], snew[mesh1d.m+1] = mesh1d.b[0], mesh1d.b[mesh1d.m+1]
         negd = []
         for k in range(c):
             jlist = np.arange(k+1, mesh1d.m+1, c, dtype=int)
             sperturb = s.copy()
-            for j in jlist:
-                sperturb[j] += eps
+            sperturb[jlist] += eps
             if dump:
                 self.savestatenextresidual(self.args.o + '_jacobi_%d.pvd' % j)
             rperturb = self.residual(mesh1d, sperturb, ella)

### 2021-06-08 py/basesmoother.py
Factor fest: move Stokes solver out of smoother and delete base class.

@@ -1,65 +1,0 @@
-'''Module for abstract base class SmootherObstacleProblem.'''
-
-__all__ = ['SmootherObstacleProblem']
-
-import sys
-from abc import ABC, abstractmethod
-import numpy as np
-
-class SmootherObstacleProblem(ABC):
-    '''Abstact base class for a smoother on an obstacle problem.  Works on
-    any mesh of class MeshLevel1D.'''
-
-    def __init__(self, args, admissibleeps=1.0e-10):
-        self.args = args
-        self.admissibleeps = admissibleeps
-        self.name = 'base'
-
-    def _checkadmissible(self, mesh, w, phi):
-        '''Check admissibility and stop if not.'''
-        for p in range(1, mesh.m+1):
-            if w[p] < phi[p] - self.admissibleeps:
-                print('ERROR: inadmissible w[%d]=%e < phi[%d]=%e (m=%d)' \
-                      % (p, w[p], p, phi[p], mesh.m))
-                sys.exit(0)
-
-    def _sweepindices(self, mesh, forward=True):
-        '''Generate indices for sweep.'''
-        if forward:
-            ind = range(1, mesh.m+1)    # 1,...,m
-        else:
-            ind = range(mesh.m, 0, -1)  # m,...,1
-        return ind
-
-    def shownonzeros(self, z):
-        '''Print a string indicating locations where array z is zero.'''
-        Jstr = ''
-        for k in range(len(z)):
-            Jstr += '_' if z[k] == 0.0 else '*'
-        print('  %d nonzeros: ' % sum(z > 0.0) + Jstr)
-
-    def inactiveresidualnorm(self, mesh, w, r, phi, ireps=0.001):
-        '''Compute the norm of the residual values at nodes where the constraint
-        is NOT active.  Where the constraint is active the residual F(w) in the
-        complementarity problem is allowed to have any positive value; only the
-        residual at inactive nodes is relevant to convergence.'''
-        F = r.copy()
-        F[w < phi + ireps] = np.minimum(F[w < phi + ireps], 0.0)
-        return mesh.l2norm(F)
-
-    def smoother(self, iters, mesh, w, ell, phi):
-        '''Apply iters sweeps of obstacle-problem smoother on mesh to modify w in place.  Alternate directions.'''
-        forward = True
-        for _ in range(iters):
-            self.smoothersweep(mesh, w, ell, phi, forward=forward)
-            forward = not forward
-
-    @abstractmethod
[…]

### 2021-06-08 py/smoother.py

@@ -1,211 +1,60 @@
 '''Module for SmootherStokes class derived from SmootherObstacleProblem.'''
 
 import numpy as np
-import firedrake as fd
-from basesmoother import SmootherObstacleProblem
-from problem import secpera, g, rhoi, nglen, B3
 
-def extend(mesh, f):
-    '''On an extruded mesh extend a function f(x,z), already defined on the
-    base mesh, to the mesh using the 'R' constant-in-the-vertical space.'''
-    Q1R = fd.FunctionSpace(mesh, 'P', 1, vfamily='R', vdegree=0)
-    fextend = fd.Function(Q1R)
-    fextend.dat.data[:] = f.dat.data_ro[:]
-    return fextend
+class SmootherObstacleProblem:
+    '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D.'''
 
-def D(w):
-    return 0.5 * (fd.grad(w) + fd.grad(w).T)
-
-class SmootherStokes(SmootherObstacleProblem):
-    '''Smoother for solving the steady-geometry Stokes problem.  Generates an
-    extruded mesh for each residual evaluation.  Implements projected
-    nonlinear versions of the Richardson and Jacobi smoothers.'''
-
-    def __init__(self, args, admissibleeps=1.0e-10):
-        super().__init__(args, admissibleeps=admissibleeps)
-        # smoother name
-        self.name = 'SmootherStokes'
-        # used in Stokes solver
-        self.Dtyp = 1.0 / secpera        # s-1
-        self.sc = 1.0e-7                 # velocity scale for symmetric scaling
-        # we store the basemesh info and the bed elevation
-        self.basemesh = None
-        self.mx = None
+    def __init__(self, args, solver, admissibleeps=1.0e-10):
+        self.args = args
+        self.solver = solver
+        self.admissibleeps = admissibleeps
+        self.created = False
         self.saveflag = False
-        self.savename = None
+
+    def _checkadmissible(self, mesh1d, w, phi):
+        '''Check admissibility and stop if not.'''
+        for p in range(1, mesh1d.m+1):
+            if w[p] < phi[p] - self.admissibleeps:
+                print('ERROR: inadmissible w[%d]=%e < phi[%d]=%e (m=%d)' \
+                      % (p, w[p], p, phi[p], mesh1d.m))
+                sys.exit(0)
+
+    def _sweepindices(self, mesh1d, forward=True):
+        '''Generate indices for sweep.'''
+        if forward:
[…]

### 2021-06-08 py/steady.py

@@ -25,7 +25,8 @@ from firedrake import *
 
 from meshlevel import MeshLevel1D
 from problem import IceProblem
-from smoother import SmootherStokes
+from stokes import GlenStokes
+from smoother import SmootherObstacleProblem
 from mcdn import mcdnvcycle
 from visualize import showiteratecmb
 
@@ -133,8 +134,8 @@ else:
 ella = finemesh.ellf(problem.source(finemesh.xx()))  # source ell[v] = <a,v>
 s = problem.initial(finemesh.xx())
 
-# set-up smoother
-smooth = SmootherStokes(args)
+# set-up smoother with included Stokes solver
+smooth = SmootherObstacleProblem(args, GlenStokes(args))
 
 # generate .pvd file with initial state (see first residual())
 if args.o:
@@ -173,8 +174,8 @@ if args.o:
     smooth.savestatenextresidual(args.o + '_%d.pvd' % (j+1))
     smooth.residual(finemesh, s, ella)  # extra residual call generates file
     if args.viewperturb is not None:
-        smooth.savename = args.o + '_perturb.pvd'
-        smooth.viewperturb(s, args.viewperturb)
+        smooth.solver.viewperturb(s, finemesh.b, args.viewperturb,
+                                  savename=args.o + '_perturb.pvd')
 
 # output an image
 if args.oimage:

### 2021-06-08 py/stokes.py

@@ -1,0 +1,238 @@
+'''Module for solving the Glen-Stokes problem on an extruded mesh.'''
+
+import numpy as np
+import firedrake as fd
+from problem import secpera, g, rhoi, nglen, B3
+
+def extend(mesh, f):
+    '''On an extruded mesh extend a function f(x,z), already defined on the
+    base mesh, to the mesh using the 'R' constant-in-the-vertical space.'''
+    Q1R = fd.FunctionSpace(mesh, 'P', 1, vfamily='R', vdegree=0)
+    fextend = fd.Function(Q1R)
+    fextend.dat.data[:] = f.dat.data_ro[:]
+    return fextend
+
+def D(w):
+    return 0.5 * (fd.grad(w) + fd.grad(w).T)
+
+class GlenStokes:
+    '''Use Firedrake to solve the fixed-geometry Glen-Stokes problem.  Also report certain quantities about the solution.'''
+
+    def __init__(self, args):
+        self.args = args
+        self.Dtyp = 1.0 / secpera        # s-1
+        self.sc = 1.0e-7                 # velocity scale for symmetric scaling
+        # we store the basemesh info
+        self.basemesh = None
+        self.mx = None
+
+    def _regDu2(self, u):
+        reg = self.args.eps * self.Dtyp**2
+        return 0.5 * fd.inner(D(u), D(u)) + reg
+
+    def stresses(self, mesh, u):
+        ''' Generate effective viscosity and tensor-valued deviatoric stress
+        from the velocity solution.'''
+        Q1 = fd.FunctionSpace(mesh,'Q',1)
+        Du2 = self._regDu2(u)
+        r = 1.0 / nglen - 1.0
+        assert nglen == 3.0
+        nu = fd.Function(Q1).interpolate(0.5 * B3 * Du2**(r/2.0))
+        nu.rename('effective viscosity (Pa s)')
+        TQ1 = fd.TensorFunctionSpace(mesh, 'Q', 1)
+        tau = fd.Function(TQ1).interpolate(2.0 * nu * D(u))
+        tau /= 1.0e5
+        tau.rename('tau (bar)')
+        return nu, tau
+
+    def savestate(self, mesh, u, p, kres, savename=None):
+        ''' Save state and diagnostics into .pvd file.'''
+        nu, tau = self.stresses(mesh, u)
+        u *= secpera
+        u.rename('velocity (m a-1)')
+        p /= 1.0e5
+        p.rename('pressure (bar)')
[…]

### 2021-06-08 py/smoother.py
Clean-up the public interface for SmootherObstacleProblem and GlenStokes.  But now it converges in fewer iterations.  Why?

@@ -3,7 +3,23 @@
 import numpy as np
 
 class SmootherObstacleProblem:
-    '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D.'''
+    '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D and calls a solver of class GlenStokes.  Note the mesh holds the bed
+    elevation (obstacle).
+
+    Implements Richardson plus FD Jacobian GS and Jacobi methods.  These
+    methods do not assume the residual functional has an easily-accessible
+    Jacobian or Jacobian diagonal.
+
+    The public interface implements residual evaluation and application of the
+    in-place smoother:
+        smooth = SmootherObstacleProblem(args, solver)
+        r = smooth.residual(mesh1d, s, ella)
+        smooth.smoothersweep(mesh1d, s, ella)
+    There is also evaluation of the CP norm,
+        irnorm = smooth.inactiveresidualnorm(mesh1d, s, r)
+    and a routine to trigger output:
+        smooth.savestatenextresidual(filename)
+    '''
 
     def __init__(self, args, solver, admissibleeps=1.0e-10):
         self.args = args
@@ -28,28 +44,14 @@ class SmootherObstacleProblem:
             ind = range(mesh1d.m, 0, -1)  # m,...,1
         return ind
 
-    def shownonzeros(self, z):
-        '''Print a string indicating locations where array z is zero.'''
-        Jstr = ''
-        for k in range(len(z)):
-            Jstr += '_' if z[k] == 0.0 else '*'
-        print('  %d nonzeros: ' % sum(z > 0.0) + Jstr)
-
-    def inactiveresidualnorm(self, mesh1d, s, r, b, ireps=0.001):
+    def inactiveresidualnorm(self, mesh1d, s, r, ireps=0.001):
         '''Compute the norm of the residual values at nodes where the constraint
         is NOT active.  Where the constraint is active the residual r=F(s) in
         the complementarity problem is allowed to have any positive value;
         only the residual at inactive nodes is relevant to convergence.'''
         F = r.copy()
-        F[s < b + ireps] = np.minimum(F[s < b + ireps], 0.0)
+        F[s < mesh1d.b + ireps] = np.minimum(F[s < mesh1d.b + ireps], 0.0)
         return mesh1d.l2norm(F)
-
-    def smoother(self, iters, mesh1d, s, ell, b):
-        '''Apply iters sweeps of smoother to modify s in place.  Alternate directions.'''
[…]

### 2021-06-08 py/steady.py

@@ -156,7 +156,7 @@ for j in range(args.cyclemax):
     else:
         s += mcdnvcycle(args, smooth, hierarchy, s, ella)
         r = smooth.residual(finemesh, s, ella)  # needed? return it from MCDN?
-    normF = smooth.inactiveresidualnorm(finemesh, s, r, finemesh.b)
+    normF = smooth.inactiveresidualnorm(finemesh, s, r)
     if args.monitor:
         print('%4d: %.4e' % (j+1, normF))
     if normF < args.irtol * normF0:

### 2021-06-08 py/stokes.py

@@ -16,7 +16,15 @@ def D(w):
     return 0.5 * (fd.grad(w) + fd.grad(w).T)
 
 class GlenStokes:
-    '''Use Firedrake to solve the fixed-geometry Glen-Stokes problem.  Also report certain quantities about the solution.'''
+    '''Use Firedrake to solve the fixed-geometry Glen-Stokes problem.  Also report certain quantities about the solution.  The public interface:
+        stokes = GlenStokes(args)
+        stokes.createbasemesh(...)
+        mesh = stokes.extrudetogeometry(...)
+        u, p, kres = stokes.solve(mesh)
+        r = stokes.extracttop(mesh, kres)
+    There are also two output routines:
+        stokes.savestate(...)
+        stokes.viewperturb(...)'''
 
     def __init__(self, args):
         self.args = args
@@ -26,11 +34,22 @@ class GlenStokes:
         self.basemesh = None
         self.mx = None
 
+    def _kinematical(self, mesh, u):
+        ''' Evaluate kinematic part of residual from given velocity u, namely
+        as a field defined on the whole extruded mesh:
+            kres = u ds/dx - w.
+        Note n_s = <-s_x, 1> so this is <u,w> . n_s.'''
+        _, z = fd.SpatialCoordinate(mesh)
+        kres_ufl = u[0] * z.dx(0) - u[1]
+        Q1 = fd.FunctionSpace(mesh, 'Lagrange', 1)
+        kres = fd.Function(Q1).interpolate(kres_ufl)
+        return kres
+
     def _regDu2(self, u):
         reg = self.args.eps * self.Dtyp**2
         return 0.5 * fd.inner(D(u), D(u)) + reg
 
-    def stresses(self, mesh, u):
+    def _stresses(self, mesh, u):
         ''' Generate effective viscosity and tensor-valued deviatoric stress
         from the velocity solution.'''
         Q1 = fd.FunctionSpace(mesh,'Q',1)
@@ -45,61 +64,45 @@ class GlenStokes:
         tau.rename('tau (bar)')
         return nu, tau
 
-    def savestate(self, mesh, u, p, kres, savename=None):
-        ''' Save state and diagnostics into .pvd file.'''
-        nu, tau = self.stresses(mesh, u)
-        u *= secpera
-        u.rename('velocity (m a-1)')
-        p /= 1.0e5
-        p.rename('pressure (bar)')
-        kres.rename('kinematic residual (a=0)')
[…]

### 2021-06-08 py/steady.py
It should not have run with this bug; the effect was appalling.

@@ -147,7 +147,7 @@ if args.sweepsonly:
 else:
     print('MCDN with smoother "%s" ...' % args.smoother)
 r = smooth.residual(finemesh, s, ella)  # generates file if -o above
-normF0 = smooth.inactiveresidualnorm(finemesh, s, r, finemesh.b)
+normF0 = smooth.inactiveresidualnorm(finemesh, s, r)
 if args.monitor:
     print('   0: %.4e' % normF0)
 for j in range(args.cyclemax):

### 2021-06-08 py/stokes.py
Minor clean.

@@ -42,8 +42,7 @@ class GlenStokes:
         _, z = fd.SpatialCoordinate(mesh)
         kres_ufl = u[0] * z.dx(0) - u[1]
         Q1 = fd.FunctionSpace(mesh, 'Lagrange', 1)
-        kres = fd.Function(Q1).interpolate(kres_ufl)
-        return kres
+        return fd.Function(Q1).interpolate(kres_ufl)
 
     def _regDu2(self, u):
         reg = self.args.eps * self.Dtyp**2

### 2021-06-09 py/problem.py
Clean up IceProblem.

@@ -1,6 +1,5 @@
-'''Dome ice problem.'''
+'''Define a class for a particular ice problem.'''
 
-from abc import ABC
 import numpy as np
 
 # public physical parameters
@@ -11,14 +10,14 @@ nglen = 3.0
 A3 = 1.0e-16 / secpera      # Pa-3 s-1;  EISMINT I ice softness
 B3 = A3**(-1.0/3.0)         # Pa s(1/3);  ice hardness
 
-class IceProblem(ABC):
-    '''Class for a particular ice problem.'''
+class IceProblem:
+    '''Class for an ice problem based on a flat-bed dome.'''
 
     def __init__(self, args):
-        self.args = args
-        # derived constant
-        self.Gamma = 2.0 * A3 * (rhoi * g)**nglen
-        self.Gamma /= (nglen + 2.0)
+        self.Gamma = 2.0 * A3 * (rhoi * g)**nglen / (nglen + 2.0)
+        self.H0 = args.domeH0
+        self.L = args.domeL
+        self.dl = args.domainlength
 
     def bed(self, x):
         '''For now we have a flat bed.'''
@@ -32,24 +31,23 @@ class IceProblem(ABC):
         invn = 1.0 / n
         r1 = 2.0 * n + 2.0                   # e.g. 8
         s1 = (1.0 - n) / n                   #     -2/3
-        C = self.args.domeH0**r1 * self.Gamma # Gamma=A_0 in van der Veen
-        C /= ( 2.0 * self.args.domeL * (1.0 - 1.0 / n) )**n
-        xc = self.args.domainlength / 2.0
-        X = (x - xc) / self.args.domeL          # rescaled coord
+        C = self.H0**r1 * self.Gamma         # Gamma=A_0 in van der Veen
+        C /= ( 2.0 * self.L * (1.0 - 1.0 / n) )**n
+        xc = self.dl / 2.0
+        X = (x - xc) / self.L                # rescaled coord
         m = np.zeros(np.shape(x))
         # usual formula for 0 < |X| < 1
         zzz = (abs(X) > 0.0) * (abs(X) < 1.0)
         if any(zzz):
             Xin = abs(X[zzz])
             Yin = 1.0 - Xin
-            m[zzz] = (C / self.args.domeL) \
-                     * ( Xin**invn + Yin**invn - 1.0 )**(n-1.0) \
[…]

### 2021-06-09 py/smoother.py
Remove -padding and _jacobislowsweep().  _jacobicolorsweep() can do slow.

@@ -117,10 +117,11 @@ class SmootherObstacleProblem:
         return []
 
     def _gsslowsweep(self, mesh1d, s, ella, r,
-                    eps=1.0, dump=False):
+                     eps=1.0, dump=False):
         '''Do in-place projected nonlinear Gauss-Seidel smoothing on s(x)
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
-        by VERY SLOW finite differencing of expensive residual calculations.
+        by VERY SLOW finite differencing of expensive residual calculations
+        at every point.
         If d_i > 0 then
             s_i <- max(s_i - omega * r_i / d_i, b_i)
         but otherwise
@@ -142,47 +143,22 @@ class SmootherObstacleProblem:
             r = self.residual(mesh1d, s, ella)
         return negd
 
-    def _jacobislowsweep(self, mesh1d, s, ella, r,
-                        eps=1.0, dump=False):
-        '''Do in-place projected nonlinear Jacobi smoothing on s(x)
-        where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
-        by VERY SLOW finite differencing of expensive residual calculations.
-        If d_i > 0 then
-            snew_i <- max(s_i - omega * r_i / d_i, b_i)
-        but otherwise
-            snew_i <- b_i.
-        After snew is completed we do s <- snew.'''
-        snew = s.copy()
-        negd = []
-        for j in range(1, len(s)-1): # loop over interior points
-            sperturb = s.copy()
-            sperturb[j] += eps
-            if dump:
-                self.savestatenextresidual(self.args.o + '_jacobi_%d.pvd' % j)
-            rperturb = self.residual(mesh1d, sperturb, ella)
-            d = (rperturb[j] - r[j]) / eps
-            if d > 0.0:
-                snew[j] = max(s[j] - self.args.omega * r[j] / d, mesh1d.b[j])
-            else:
-                snew[j] = mesh1d.b[j]
-                negd.append(j)
-        s[:] = snew[:] # in-place copy
-        return negd
-
     def _jacobicolorsweep(self, mesh1d, s, ella, r,
-                        eps=1.0, dump=False, cperthickness=3.0):
+                          eps=1.0, dump=False):
         '''Do in-place projected nonlinear Jacobi smoothing on s(x)
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
[…]

### 2021-06-09 py/steady.py

@@ -59,6 +59,8 @@ References:
 adda = parser.add_argument
 adda('-coarsesweeps', type=int, default=1, metavar='N',
      help='smoother sweeps on coarsest grid (default=%(default)s)')
+adda('-cperthickness', type=float, default=3.0, metavar='X',
+     help='parameter for -smoother jacobicolor (default=%(default)s m)')
 adda('-cyclemax', type=int, default=100, metavar='N',
      help='maximum number of (multilevel) cycles (default=%(default)s)')
 adda('-domainlength', type=float, default=30.0e3, metavar='L',
@@ -89,8 +91,6 @@ adda('-oimage', metavar='FILE', type=str, default='',
      help='save final image, e.g. .pdf or .png')
 adda('-omega', type=float, metavar='X', default=1.0,  # FIXME sensitive
     help='scale by this factor in smoother iteration (default=%(default)s)')
-adda('-padding', action='store_true', default=False,
-     help='put Hmin thickness of ice in ice-free locations')
 adda('-printwarnings', action='store_true', default=False,
      help='print pointwise feasibility warnings')
 adda('-smoother', choices=['richardson', 'gsslow', 'jacobislow', 'jacobicolor'],
@@ -109,8 +109,6 @@ if args.steadyhelp:
     parser.print_help()
     sys.exit(0)
 
-if args.padding:
-    assert args.Hmin > 0.0, 'padding requires minimum positive thickness'
 if args.viewperturb is not None:
     assert len(args.o) > 0, 'use -view perturb with -o'
 

### 2021-06-09 py/stokes.py

@@ -71,14 +71,10 @@ class GlenStokes:
     def extracttop(self, mesh, field):
         '''On an extruded mesh with some ice-free (i.e. empty) columns, loop
         over the base mesh finding top cells where ice is present, then top
-        nodes, and evaluate the field there.  On an extruded mesh with padding,
-        just get the top Dirichlet nodes.  Only works for Q1 fields.
+        nodes, and evaluate the field there.  Only works for Q1 fields.
         (Thanks Lawrence Mitchell.)'''
         assert self.basemesh is not None
         Q1 = fd.FunctionSpace(mesh, 'Lagrange', 1)
-        if self.args.padding:
-            topbc = fd.DirichletBC(Q1, 1.0, 'top')
-            return field.dat.data_ro[topbc.nodes]
         # get the cells from basemesh and mesh
         bmP1 = fd.FunctionSpace(self.basemesh, 'Lagrange', 1)
         bmcnm = bmP1.cell_node_map().values
@@ -107,34 +103,24 @@ class GlenStokes:
         '''Generate extruded mesh over self.basemesh, to height s.  The icy
         columns get their height from s, with minimum height args.Hmin.  By
         default the extruded mesh has empty (0-element) columns if ice-free
-        according to s.  If args.padding==True then the whole extruded mesh has
-        the same layer count.  Optional reporting of mesh stats.'''
+        according to s.  Optional reporting of mesh stats.'''
         assert self.basemesh is not None
-        if report:
-            print('mesh: base of %d elements (intervals)' \
-                  % self.mx)
         # extrude to temporary total height 1.0
         mz = self.args.mz
-        if self.args.padding:
-            assert self.args.Hmin > 0.0, \
-                'padding requires minimum positive thickness'
-            mesh = fd.ExtrudedMesh(self.basemesh, mz, layer_height=1.0 / mz)
-            if report:
-                print('      extruded is padded, has %d x %d elements' \
-                      % (self.mx, mz))
-        else:
-            layermap = np.zeros((self.mx, 2), dtype=int)  # [[0,0], ..., [0,0]]
-            thk = s - b
-            thkelement = ( (thk[:-1]) + (thk[1:]) ) / 2.0
-            icyelement = (thkelement > self.args.Hmin + 1.0e-3)
-            layermap[:,1] = mz * np.array(icyelement, dtype=int)
-            # FIXME: in parallel we must provide local, haloed layermap
[…]

### 2021-06-09 py/smoother.py
Minor: Better comments.

@@ -6,9 +6,10 @@ class SmootherObstacleProblem:
     '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D and calls a solver of class GlenStokes.  Note the mesh holds the bed
     elevation (obstacle).
 
-    Implements Richardson plus FD Jacobian GS and Jacobi methods.  These
-    methods do not assume the residual functional has an easily-accessible
-    Jacobian or Jacobian diagonal.
+    Implements nonlinear Richardson plus nonlinear GS and Jacobi methods based
+    on the computation of the Jacobian diagonal entries by finite-differencing.
+    (These methods do not assume the residual has an easily-accessible
+    Jacobian diagonal.)
 
     The public interface implements residual evaluation and application of the
     in-place smoother:
@@ -122,10 +123,11 @@ class SmootherObstacleProblem:
         where the diagonal entry d_i = F'(s)[psi_i,psi_i] is computed
         by VERY SLOW finite differencing of expensive residual calculations
         at every point.
-        If d_i > 0 then
+        For each i, compute r = F(s).  If d_i > 0 then
             s_i <- max(s_i - omega * r_i / d_i, b_i)
         but otherwise
-            s_i <- phi_i.'''
+            s_i <- phi_i.
+        Note s_i is updated immediately.'''
         negd = []
         for j in range(1, len(s)-1): # loop over interior points
             sperturb = s.copy()
@@ -151,7 +153,7 @@ class SmootherObstacleProblem:
         using coloring.  Nodes of the same color are separated by cperthickness
         times the maximum ice thickness.  Set -cperthickness 1.0e10 (for
         example) to use VERY SLOW finite differencing without coloring.
-        If d_i > 0 then
+        First r = F(s).  Then for each i, if d_i > 0 then
             snew_i <- max(s_i - omega * r_i / d_i, b_i)
         but otherwise
             snew_i <- b_i.
@@ -164,6 +166,7 @@ class SmootherObstacleProblem:
         snew[0], snew[mesh1d.m+1] = mesh1d.b[0], mesh1d.b[mesh1d.m+1]
         negd = []
         for k in range(c):
+            # note jlist = [k+1,] (singleton) if k+1+c >= mesh1d.m+1
             jlist = np.arange(k+1, mesh1d.m+1, c, dtype=int)
             sperturb = s.copy()
             sperturb[jlist] += eps

### 2021-06-09 py/smoother.py
Start drafting a Newton-LU based smoother, derived from tri.py in p4pdes.

@@ -2,6 +2,12 @@
 
 import numpy as np
 
+# one smoother type sets up a PETSc Mat and KSP system
+import sys
+import petsc4py
+petsc4py.init(sys.argv)  # must come before "import PETSc"
+from petsc4py import PETSc
+
 class SmootherObstacleProblem:
     '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D and calls a solver of class GlenStokes.  Note the mesh holds the bed
     elevation (obstacle).
@@ -22,17 +28,17 @@ class SmootherObstacleProblem:
         smooth.savestatenextresidual(filename)
     '''
 
-    def __init__(self, args, solver, admissibleeps=1.0e-10):
+    def __init__(self, args, solver):
         self.args = args
         self.solver = solver
-        self.admissibleeps = admissibleeps
+        self.ieps = 0.001  # if s[j] > b[j] + ieps then node j is inactive
         self.created = False
         self.saveflag = False
 
     def _checkadmissible(self, mesh1d, w, phi):
         '''Check admissibility and stop if not.'''
         for p in range(1, mesh1d.m+1):
-            if w[p] < phi[p] - self.admissibleeps:
+            if w[p] < phi[p]:
                 print('ERROR: inadmissible w[%d]=%e < phi[%d]=%e (m=%d)' \
                       % (p, w[p], p, phi[p], mesh1d.m))
                 sys.exit(0)
@@ -45,13 +51,14 @@ class SmootherObstacleProblem:
             ind = range(mesh1d.m, 0, -1)  # m,...,1
         return ind
 
-    def inactiveresidualnorm(self, mesh1d, s, r, ireps=0.001):
+    def inactiveresidualnorm(self, mesh1d, s, r):
         '''Compute the norm of the residual values at nodes where the constraint
         is NOT active.  Where the constraint is active the residual r=F(s) in
         the complementarity problem is allowed to have any positive value;
         only the residual at inactive nodes is relevant to convergence.'''
         F = r.copy()
-        F[s < mesh1d.b + ireps] = np.minimum(F[s < mesh1d.b + ireps], 0.0)
+        F[s < mesh1d.b + self.ieps] = np.minimum(F[s < mesh1d.b + self.ieps],
+                                                 0.0)
         return mesh1d.l2norm(F)
 
     def savestatenextresidual(self, name):
@@ -145,22 +152,27 @@ class SmootherObstacleProblem:
             r = self.residual(mesh1d, s, ella)
[…]

### 2021-06-09 py/smoother.py
Minor: rename.

@@ -1,4 +1,4 @@
-'''Module for SmootherStokes class derived from SmootherObstacleProblem.'''
+'''Module for ObstacleSmoother.'''
 
 import numpy as np
 
@@ -8,7 +8,7 @@ import petsc4py
 petsc4py.init(sys.argv)  # must come before "import PETSc"
 from petsc4py import PETSc
 
-class SmootherObstacleProblem:
+class ObstacleSmoother:
     '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D and calls a solver of class GlenStokes.  Note the mesh holds the bed
     elevation (obstacle).
 
@@ -19,7 +19,7 @@ class SmootherObstacleProblem:
 
     The public interface implements residual evaluation and application of the
     in-place smoother:
-        smooth = SmootherObstacleProblem(args, solver)
+        smooth = ObstacleSmoother(args, solver)
         r = smooth.residual(mesh1d, s, ella)
         smooth.smoothersweep(mesh1d, s, ella)
     There is also evaluation of the CP norm,

### 2021-06-09 py/steady.py

@@ -26,7 +26,7 @@ from firedrake import *
 from meshlevel import MeshLevel1D
 from problem import IceProblem
 from stokes import GlenStokes
-from smoother import SmootherObstacleProblem
+from smoother import ObstacleSmoother
 from mcdn import mcdnvcycle
 from visualize import showiteratecmb
 
@@ -133,7 +133,7 @@ ella = finemesh.ellf(problem.source(finemesh.xx()))  # source ell[v] = <a,v>
 s = problem.initial(finemesh.xx())
 
 # set-up smoother with included Stokes solver
-smooth = SmootherObstacleProblem(args, GlenStokes(args))
+smooth = ObstacleSmoother(args, GlenStokes(args))
 
 # generate .pvd file with initial state (see first residual())
 if args.o:

### 2021-06-09 py/stokes.py
Fix bug with -viewperturb (not covered by test ...)

@@ -193,8 +193,7 @@ class GlenStokes:
         assert self.basemesh is not None
         # solve the Glen-Stokes problem on the unperturbed extruded mesh
         meshs = self.extrudetogeometry(s, b)
-        us, ps = self.solve(meshs)
-        kress = self.kinematical(meshs, us)
+        us, ps, kress = self.solve(meshs)
         # solve on the PERTURBED extruded mesh
         sP = s.copy()
         for k in klist:
@@ -207,8 +206,7 @@ class GlenStokes:
                 print('WARNING viewperturb(): skipping bare-ground node k=%d' \
                       % k)
         meshP = self.extrudetogeometry(sP, b)
-        uP, pP = self.solve(meshP)
-        kresP = self.kinematical(meshP, uP)
+        uP, pP, kresP = self.solve(meshP)
         # compute difference as a function on the unperturbed mesh
         V = fd.VectorFunctionSpace(meshs, 'Lagrange', 2)
         W = fd.FunctionSpace(meshs, 'Lagrange', 1)

### 2021-06-09 py/smoother.py
Initial implementation of Newton reduced-space smoother.  (It does set up and solve a system ...)

@@ -106,10 +106,10 @@ class ObstacleSmoother:
             negd = self._richardsonsweep(mesh1d, s, ella, currentr)
         elif self.args.smoother == 'gsslow':
             negd = self._gsslowsweep(mesh1d, s, ella, currentr)
-        elif self.args.smoother == 'jacobislow':
-            negd = self._jacobislowsweep(mesh1d, s, ella, currentr)
         elif self.args.smoother == 'jacobicolor':
             negd = self._jacobicolorsweep(mesh1d, s, ella, currentr)
+        elif self.args.smoother == 'newtonrslu':
+            negd = self._newtonrslu(mesh1d, s, ella, currentr)
         mesh1d.WU += 1
         if self.args.monitor and len(negd) > 0:
             print('      negative diagonal entries: ', end='')
@@ -152,8 +152,9 @@ class ObstacleSmoother:
             r = self.residual(mesh1d, s, ella)
         return negd
 
-    def _coloringgap(self, mesh1d, s):
-        '''Compute c which is the gap between nodes with the same color.
+    def _colors(self, mesh1d, s):
+        '''Compute c which is the number of colors (and the gap between nodes
+        with the same color).
         Nodes of the same color are separated by cperthickness
         times the maximum ice thickness.  Set e.g. -cperthickness 1.0e10
         to use VERY SLOW finite differencing without coloring.'''
@@ -172,8 +173,8 @@ class ObstacleSmoother:
         but otherwise
             snew_i <- b_i.
         After snew is completed we do s <- snew.'''
-        c = self._coloringgap(mesh1d, s)
-        print('      c = %d' % c)
+        c = self._colors(mesh1d, s)
+        print('      c = %d colors' % c)
         snew = mesh1d.b.copy() - 1.0  # snew NOT admissible; note check below
         snew[0], snew[mesh1d.m+1] = mesh1d.b[0], mesh1d.b[mesh1d.m+1]
         negd = []
@@ -198,58 +199,82 @@ class ObstacleSmoother:
         s[:] = snew[:] # in-place copy
         return negd
 
-    def _newtonlusweep(self, mesh1d, s, ella, r,
-                       eps=1.0, dump=False, band=1):
-        '''Do one in-place projected Newton step on s(x), with an approximated
-        Jacobian using coloring, and using LU to solve the linear step
-        equations.  The approximated Jacobian is band-limited to [-band,band]
[…]

### 2021-06-09 py/steady.py

@@ -93,7 +93,7 @@ adda('-omega', type=float, metavar='X', default=1.0,  # FIXME sensitive
     help='scale by this factor in smoother iteration (default=%(default)s)')
 adda('-printwarnings', action='store_true', default=False,
      help='print pointwise feasibility warnings')
-adda('-smoother', choices=['richardson', 'gsslow', 'jacobislow', 'jacobicolor'],
+adda('-smoother', choices=['richardson', 'gsslow', 'jacobicolor', 'newtonrslu'],
      metavar='X', default='jacobicolor',
      help='smoother (default=%(default)s)')
 adda('-steadyhelp', action='store_true', default=False,

### 2021-06-09 paper/msg.tex
Some edits section 6.

@@
 In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.
-The fundamental idea of such a scheme is the basic geometric multigrid idea that smoothing on coarse levels will rapidly reduce the high frequencies in the error, but here we must maintain admissibility on each level in a manner which does not re-introduce high frequencies.
-Our method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
+The fundamental idea of such a scheme, in fact the basic geometric multigrid idea, is that smoothing on coarse levels will rapidly reduce the high frequencies in the error.
+However, here we must maintain admissibility on each level in a manner which does not re-introduce high frequencies.
+Our method is derived from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}), but our nonlinear variant (MCDN) uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
 
 Regarding notation for multiple mesh levels, suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, the coarse level.
-For $J\ge 0$, suppose $\{\mathcal{T}^j\}_{j=1}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
-(Halve each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$, with subspace $\mathcal{V}_0^j \subset \mathcal{V}^j$ which are zero on $\partial\Omega$.
-If $m_j$ is the number of interior nodes $\{(x_i^j,y_i^j)\}$ in $\mathcal{T}$ then $\dim(\mathcal{V}_0^j)=m_j$.
+For $J\ge 0$, which gives the number of finer levels, suppose $\{\mathcal{T}^j\}_{j=1}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.
+(Halve each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$, with subspace $\mathcal{V}_0^j = \mathcal{V}^j \cap W_0^{1,\qq}(\Omega)$.
+If $m_j$ is the number of interior nodes $\{(x_i^j,y_i^j)\}$ in $\mathcal{T}^j$ then $\dim(\mathcal{V}_0^j)=m_j$.
+
 We seek an admissible solution to \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.
 Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.
-The admissible surface elevations form a closed and convex set
+The admissible fine-level surface elevations form a closed and convex set
 \begin{equation}
@@
 \end{equation}
-
-However, we follow \cite{GraeserKornhuber2009} in decomposing a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.
-That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
+However, we follow \cite{GraeserKornhuber2009} in formulating a ``defect obstacle'' decomposition, instead of direct solution over the set $\mathcal{K}^J$.
+Suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
 \begin{equation}
@@
 \end{equation}
-be the associated defect obstacle, with $\chi^J \le 0$.
+be the associated defect obstacle, so that $\chi^J \le 0$.
 Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.
-Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$ \cite{GraeserKornhuber2009} on $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ by maximizing nodal values over the (open) supports of each each coarser-level hat function:
+
+Now we define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$ \cite{GraeserKornhuber2009} on $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$, with coefficients $z^j[i]\in\RR$, by maximizing nodal values over the (open) supports of each coarser-level hat function:
 \begin{equation}
@@
 \end{equation}
-Using $\mR$ a defect obstacle is defined on each level:
+We use this nonlinear operator to define a defect obstacle on each level:
 \begin{equation}
@@
 \end{equation}
-The down-admissible constraint sets $\mathcal{D}^j$ sum to an up-admissible constraint set for each level, i.e.~$\mathcal{D}^0 + \mathcal{D}^1 + \dots + \mathcal{D}^j = \mathcal{U}^j$ in the sense that every $z^j \in \mathcal{U}^j$ is the sum of elements, though not uniquely.
-The $\mathcal{U}^j$ are themselves nested in $\mathcal{U}^J= \{z^J \,:\, z^J \ge b^J - s^J\}$, and this forms a multilevel cone approximation, from the inside, of the continuum defect constraint set $\mathcal{K} - s = \{z\,:\,z \ge b - s\}$:
+Note that $0 \in \mathcal{D}^j \cap \mathcal{U}^j$; a zero perturbation is always admissible.
+The down-admissible constraint sets $\mathcal{D}^j$ sum to an up-admissible constraint set for each level.
+The $\mathcal{U}^j$ are themselves nested,
 \begin{equation}
-\mathcal{U}^0 \subset \mathcal{U}^1 \subset \dots \subset \mathcal{U}^J \subset \mathcal{K} - s \subset W_0^{1,\qq}(\Omega). \label{eq:innerconeapprox}
+\mathcal{U}^0 \subset \mathcal{U}^1 \subset \dots \subset \mathcal{U}^J \subset W_0^{1,\qq}(\Omega). \label{eq:innerconeapprox}
 \end{equation}
@@
 \begin{equation}
-F^J(s^J + y^J + \dots + y^j)[v^j - y^j] \ge (a,v^j - y^j) \quad \text{for all } v^j \in \mathcal{D}^j, \label{eq:mcdoriginal}
+F^J(s^J + y^J + \dots + y^j)[v^j - y^j] \ge (a,v^j - y^j) \quad \text{for all } v^j \in \mathcal{D}^j. \label{eq:mcdoriginal}
 \end{equation}
-where $F^J = F^h$ in \eqref{eq:fesigpweakform}.
+(Here $F^J = F^h$ in \eqref{eq:fesigpweakform}.)  Observe that \eqref{eq:mcdoriginal} exploits the simplification $(s^J + y^J + \dots + v^j) - (s^J + y^J + \dots + y^j) = v^j - y^j$.
+Also note that $y^J + \dots + y^j \in \mathcal{U}^J$ so the updated fine-level iterate is admissible in the original sense: $s^J + y^J + \dots + y^j \ge b^J$.
+
 In \cite{Tai2003}, VIs \eqref{eq:mcdoriginal} are to be solved on descending levels $j=J$ down to $j=0$, forming a down-slash or V(1,0) cycle.
-(See also \cite[Algorithm 4.7]{GraeserKornhuber2009}.)  Note that $y^J + \dots + y^j \in \mathcal{U}^J$ so $s^J + y^J + \dots + y^j \ge b^J$ is admissible in the original sense, and that \eqref{eq:mcdoriginal} exploits the simplification $(s^J + y^J + \dots + v^j) - (s^J + y^J + \dots + y^j) = v^j - y^j$.
+(See also \cite[Algorithm 4.7]{GraeserKornhuber2009}.)  However, there is no practical way to achieve rapid solution of VI \eqref{eq:mcdoriginal} on coarse levels if the latest iterate $s^J + y^J + \dots + y^j$ must be directly represented; it has fine-level content.
+This is the reason to add a full approximation scheme (FAS) \cite{Trottenbergetal2001} coarse-level ``equation'', which is here a nonlinear VI.
+In order to implement our approach we will need to state the FAS coarse-level correction equation using appropriate linear restriction and prolongation operators, as follows.
 
-However, there is no practical way to achieve rapid solution of VI \eqref{eq:mcdoriginal} on coarse levels if the latest iterate $s^J + y^J + \dots + y^j$ must be represented; it has fine-level content.
-This is the reason to add the full approximation scheme (FAS) \cite{Trottenbergetal2001} concept of a nonlinear coarse-level ``equation'', which is here a nonlinear VI.
-In order to implement our approach we need to state the FAS coarse-level correction equation based on appropriate restrictions, as follows.
-
-Let $\iR: \mathcal{V}^{j+1} \to \mathcal{V}^j$ be the injection restriction operator \cite{Trottenbergetal2001}.
+Let $\iR: \mathcal{V}^{j+1} \to \mathcal{V}^j$ be the injection operator \cite{Trottenbergetal2001}.
 Suppose $y^j \in \mathcal{D}^j$ denotes the desired down-correction on the $j$th level.

### 2021-06-10 py/steady.py
Do not forget this idea.

@@ -2,6 +2,7 @@
 '''Solve steady-geometry Stokes obstacle problem by a multilevel constraint decomposition method.'''
 
 # TODO:
+#   0. fix the FD Jacobian scheme so that Hmin does not conflict
 #   1. widen the default sheet?; consider -domainlength 60.0e3 -domeL 25.0e3
 #   2. replace meshlevel.py with use of Firedrake interval mesh (requires
 #      new implementation of monotone restriction); this would allow

### 2021-06-10 talk/talk.tex
Work on talk.

@@
 
-\begin{frame}{the steady ice geometry problem (SIGP)}
+\begin{frame}{the ice geometry problem (IGP)}
 
@@
 \begin{itemize}
-\item a glacier which is in balance with its climate
-    \begin{itemize}
-    \item the simplest such problem (grounded, nonsliding, isothermal)
-    \item re~\emph{steady}: solve this and you can handle implicit steps too
-    \end{itemize}
-\item steady data given on fixed $\Omega \subset \RR^2$:
+\item glacier geometry which evolves according to climate
+\item take simplest such problem (grounded, nonsliding, isothermal)
+\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
     \begin{itemize}
@@
     \end{itemize}
-\item determine: surface elevation $s(x,y)$ on $\Omega$, domain $\Lambda_s \subset \RR^3$
+\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
 \end{itemize}
@@
 \begin{itemize}
-\item glaciers flow, so there is an equation for the surface elevation
+\item glaciers flow
+\item $\therefore$ must solve an equation for the surface elevation
 \item the surface kinematical equation (SKE) holds \emph{on} the ice
-    $$\bu|_s \cdot \bn_s + a=0 \qquad \text{\emph{steady state}}$$
+\begin{align*}
+\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a & &\text{\emph{evolving}} \\
+\bu|_s \cdot \bn_s + a &= 0 & &\text{\emph{steady state}}
+\end{align*}
 
@@
     \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
+    \item \emph{not} on the ice: $a\le 0$ for $t>0$ (or in steady state)
     \end{itemize}
-\item \emph{off} the ice we know $a\le 0$ in steady state
 \end{itemize}
@@
 \begin{columns}
-    \column{0.6\textwidth}
+    \column{0.55\textwidth}
         \begin{itemize}
-        \item for all of this talk:
+        \item assumed for this talk:
             \begin{itemize}
-            \item $s$ is assumed to be well-defined (no overhangs)
-            \item $s=b$ where ice is not present, so $s$ is defined on all of $\Omega$
+            \item $s$ well-defined (no overhangs)
+            \item $s=b$ where ice is not present
+            \item $s$ is defined on all of $\Omega$
             \end{itemize}
         \end{itemize}
-    \column{0.4\textwidth}
-        \includegraphics[width=1.05\textwidth]{figs/simpledomain.png}
+    \column{0.45\textwidth}
+        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
 \end{columns}
@@
 
-\begin{frame}{SIGP strong form}
+\begin{frame}{steady IGP strong form}
 
 \begin{itemize}
-\item the SIGP free-boundary problem determines $s$ on $\Omega$
+\item the steady IGP free-boundary problem for $s(x,y)$ on $\Omega$
 \item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}
@@
 
+\begin{frame}{\alert{implicit} IGP strong form}
+
+\begin{itemize}
+\item the evolving IGP free-boundary problem for $s(x,y,t)$ on $\Omega \times [0,T]$
+\item consider one step of backward Euler scheme
+    \begin{itemize}
+    \item solve for $s,\bu,p$ at new time $t^\ell$
+    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$, and write $s$ for $s^\ell$
+    \end{itemize}
+\item again: NCP for $s$ coupled to a Stokes problem for $\bu,p$:
+
+\vspace{-5mm}
+
+\begin{align*}
+s - b &\ge 0 && \text{on $\Omega$} \\
+\alert{s - s^{\ell-1} - \Delta t\, \bu|_s \cdot \bn_s - \Delta t\, a} &\ge 0 && \text{''} \\
+(s - b) (\alert{s - s^{\ell-1} - \Delta t\, \bu|_s \cdot \bn_s - \Delta t\, a}) &= 0 && \text{''} \\
+- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
+\nabla \cdot \bu &= 0 && \text{''} \\
+\bu &= \bzero && \text{on $\Gamma_0$} \\
+\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
+\end{align*}
+
+    \begin{itemize}
+    \item \alert{pretty similar \dots mostly suffices to consider steady IGP}
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
 \begin{frame}{existing ice sheet models}
@@
 \begin{itemize}
-\item almost no one is solving the SIGP directly
+\item almost no one is solving the implicit IGP
     \begin{itemize}
-    \item or doing implicit time-stepping (\emph{basically the same as SIGP})
-    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} (\emph{semi-coupled, nonscalable, artificial ice layer})
+    \item except Brinkerhoff et al (2017) (\emph{upwinding limits time-step?})
+    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
+    \item neither scalable (\emph{single level})
     \end{itemize}
@@
     \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
-    \item mostly with shallow approximations, e.g.~PISM run below
-    \end{itemize}
-\end{itemize}
+    \item often with shallow approximations, e.g.~PISM run below
 
@@
 \begin{columns}
-    \column{0.8\textwidth}
-        \hfill \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
-    \column{0.2\textwidth}
-        \scriptsize \emph{show clip of Alpine run by Julien Seguinot}
+    \column{0.7\textwidth}
+        \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
+    \column{0.3\textwidth}
+        \scriptsize \emph{by Julien Seguinot}
 \end{columns}
+
+\medskip
+    \item my goal is to do better
+    \end{itemize}
+\end{itemize}
 \end{frame}
@@
 
-\begin{block}{definition: ice dynamics operator}
+\begin{block}{ice dynamics operator}
 $\Phi : s \mapsto - \bus \cdot \bn_s$
@@
 
-\hfill \includegraphics[width=0.75\textwidth]{figs/idoaction.png}
+\hfill \includegraphics[width=0.7\textwidth]{figs/idoaction.png}
 \end{block}
@@
 extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
-\item regarding $F_{\Lambda_s}(\bu,p) = 0$:
+\item regarding Stokes problem for fixed geometry:
     \begin{itemize}
-    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz \cite{JouvetRappaz2011})
-    \item optimal solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
+    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
+    \item optimal solver exists (Isaac et al 2015)
     \end{itemize}
+\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
 \end{itemize}
@@
 
-\begin{frame}{SIGP weak form is a variational inequality}
+\begin{frame}{IGP weak form is a variational inequality}
 
 \begin{itemize}
-\item using $\Phi$ gives a cleaner NCP over $\Omega$:
+\item using $\Phi$ gives a cleaner NCP over $\Omega$ for steady IGP:
 \begin{align*}
@@
 
-\begin{block}{SIGP weak form}
+\begin{block}{steady IGP weak form}
 find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
-    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
-for all $r \in \mathcal{K}$, where $F(s)[w] = \int_\Omega \Phi(s)\, w \,dx dy$
+    $$\ip{\Phi(s)-a}{r-s} \ge 0$$
+for all $r \in \mathcal{K}$
 \end{block}
@@
     \begin{itemize}
-    \item but existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
+    \item but existence holds for SIA version (Jouvet \& Bueler, 2012)
     \end{itemize}
@@
 \end{itemize}
-
 \end{frame}
 
-\begin{frame}{goal: robust and optimal SIGP solver}
+\begin{frame}{what kind of beast is $\Phi$?}
+
+\begin{center}
+\mbox{\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}}
+\end{center}
+
+\begin{itemize}
+\small
+\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
+\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
+    \begin{itemize}
+    \item $\Phi$ is some kind of integral operator? (\emph{convolve with Stokeslets})
+    \item $\Phi$ is short range? (\emph{concentrated near the diagonal})
+    \end{itemize}
+\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
+    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
+
+    \begin{itemize}
+    \item $\Phi$ is diffusive in the large!
+    \item $\Phi$ is degenerate at the ice margin!
+    \end{itemize}
+\item yet common to say $\Phi$ is hyperbolic because SKE is transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
+\end{itemize}
+\end{frame}
+
+\begin{frame}{goal: robust and optimal IGP solver}
 
 \begin{block}{numerical goal}
-Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.
-Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$.
-Solve VI for $s^h \in \mathcal{K}^h$:
+given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, and convex set $\mathcal{K}^h = \{r^h \ge b^h\} \subset \mathcal{V}^h$, solve VI for $s^h \in \mathcal{K}^h$:
 
-$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
+$$\ip{\Phi^h(s^h) - a}{r^h - s^h} \ge 0 \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
 \end{block}
@@
 \begin{itemize}
+\item[]
+    \begin{itemize}
+    \item $P_1$ because low regularity at free boundary
+    \item uses numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$
+    \end{itemize}
 \item seeking a solver which is:
@@
     \item \alert{robust}
-    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
+    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$ 
         \begin{itemize}
-        \item[{\color{black} $\circ$}] optimal requires multigrid
+        \item[{\color{black} $\circ$}] optimal means work is $O(m)$ \dots requires multigrid
         \end{itemize}
@@
 
-\begin{frame}{what thing is $\Phi$?}
-
-\begin{block}{SIGP weak form}
-find $s \in \mathcal{K}$ so $F(s)[r - s] \ge \ip{a}{r-s}$ for all $r \in \mathcal{K}$
-\end{block}
+\begin{frame}{Stokes on an extruded mesh}
 
 \begin{itemize}
-\item $F(s)[w] = \int_\Omega \Phi(s)\, w$ is dual of $\Phi(s)=-\bus\cdot\bn_s$
-\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$
-\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
-    \begin{itemize}
-    \item some kind of integral operator (\emph{convolve with bdry Stokeslets})
-    \item $\Phi$ is concentrated near the diagonal, i.e.~short range?
+\item FIXME
 \end{itemize}
-\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
-\small
-    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
-\normalsize
-\end{itemize}
+
+\includegraphics[width=0.8\textwidth]{figs/extruded.png}
 \end{frame}
@@
 \begin{itemize}
-\item FIXME \cite{BuelerMitchell2022}
+\item FIXME
 \end{itemize}
 
-\includegraphics[width=0.8\textwidth]{figs/extruded.png}
+\includegraphics[width=0.5\textwidth]{figs/decompclassical.png}
 
 \input{tikz/vcycle.tex}
+\end{frame}
+
+\begin{frame}{smoothers}
+
+\begin{itemize}
+\item FIXME
+\end{itemize}
 \end{frame}
@@
 
-\begin{frame}{References}
+\begin{frame}{Some references}
 %\bibliography{../paper/msg.bib}
@@
 
+\hrulefill
+
+\bibitem{Brinkerhoffetal2017}
+D.~Brinkerhoff, M.~Truffer, and A.~Aschwanden. \textbf{Sediment transport drives tidewater glacier periodicity.} {\em Nature Communications}, 8(1):1--8, 2017.
+
 \bibitem{IsaacStadlerGhattas2015}
 T.~Isaac, G.~Stadler, and O.~Ghattas.  \textbf{Solution of nonlinear Stokes equations \dots with application to ice sheet dynamics.} {\em SIAM J. Sci. Comput.}, 37(6):B804--B833, 2015.
-
-\bibitem{JouvetBueler2012}
-G.~Jouvet and E.~Bueler. \textbf{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation.} {\em SIAM J. Appl. Math.}, 72(4):1292--1314, 2012.
-
-\bibitem{JouvetRappaz2011}
-G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary Stokes problem arising in glaciology.} {\em Adv. Numer. Anal.}, 2011.
 

### 2021-06-11 talk/talk.tex
Edits on slides.

@@
     \begin{itemize}
-    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
-    \item bed elevation $b(x,y)$
+    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
+    \item bed elevation $b$
     \end{itemize}
@@
 \begin{align*}
-\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a & &\text{\emph{evolving}} \\
-\bu|_s \cdot \bn_s + a &= 0 & &\text{\emph{steady state}}
+\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{evolving}} \\
+-\bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{steady state}}
 \end{align*}
@@
     \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
-    \item \emph{not} on the ice: $a\le 0$ for $t>0$ (or in steady state)
+    \item where \emph{not} on the ice: $a\le 0$
     \end{itemize}
@@
 \begin{itemize}
-\item the steady IGP free-boundary problem for $s(x,y)$ on $\Omega$
-\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}
+\item the steady IGP free-boundary problem, for $s(x,y)$ on $\Omega$, is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}
 
@@
     \begin{itemize}
-    \item plus Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
+    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
       $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
@@
 \begin{itemize}
-\item the evolving IGP free-boundary problem for $s(x,y,t)$ on $\Omega \times [0,T]$
-\item consider one step of backward Euler scheme
+\item solve one step of backward Euler for $s,\bu,p$ at new time $t^\ell$
     \begin{itemize}
-    \item solve for $s,\bu,p$ at new time $t^\ell$
-    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$, and write $s$ for $s^\ell$
+    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$ in evolving SKE, and write $s$ for $s^\ell$
     \end{itemize}
-\item again: NCP for $s$ coupled to a Stokes problem for $\bu,p$:
+\item again get an NCP coupled to a Stokes problem:
 
@@
     \begin{itemize}
-    \item \alert{pretty similar \dots mostly suffices to consider steady IGP}
+    \item \alert{similar \dots mostly suffices to consider steady IGP}
     \end{itemize}
@@
     \begin{itemize}
-    \item except Brinkerhoff et al (2017) (\emph{upwinding limits time-step?})
+    \item except Brinkerhoff et al (2017) (\emph{upwinding, semi-coupled})
     \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
-    \item neither scalable (\emph{single level})
+    \item neither scalable (\emph{single level}) nor unconditionally stable
     \end{itemize}
@@
 \medskip
-    \item my goal is to do better
+    \item \emph{my goal}: unconditionally-stable implicit with Stokes dynamics
     \end{itemize}
@@
 \item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
+    \begin{itemize}
+    \item a regularity conjecture
+    \end{itemize}
 \end{itemize}
@@
 \begin{center}
-\mbox{\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}}
+\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.15\textwidth]{figs/elephant.png}
 \end{center}

### 2021-06-11 paper/genfigs/extruded.py
Do not show original s in marginal columns.

@@ -38,9 +38,9 @@ def plotextruded(x, b, s, mz=5, Hmin=0.0):
                 else:
                     plt.plot([x[j], x[j+1]], [zl, zr], 'k-')
             # show original s near cliffs
-            if Hmin > 0.0 and \
-                    (not isgap(b[j],s[j]) or not isgap(b[j+1],s[j+1])):
-                plt.plot([x[j], x[j+1]], [s[j], s[j+1]], 'k:', lw=3.0)
+            #if Hmin > 0.0 and \
+            #        (not isgap(b[j],s[j]) or not isgap(b[j+1],s[j+1])):
+            #    plt.plot([x[j], x[j+1]], [s[j], s[j+1]], 'k:', lw=3.0)
         else:
             plt.plot([x[j], x[j+1]], [s[j], s[j+1]], 'k-', lw=3.0)
     plt.axis('off')

### 2021-06-11 talk/talk.tex
Include Greenland cross section.  Edits.

@@
     \end{itemize}
-\item again get an NCP coupled to a Stokes problem:
+\item at each time step, solve NCP coupled to Stokes:
 
@@
     \begin{itemize}
-    \item \alert{similar \dots mostly suffices to consider steady IGP}
+    \item very similar to steady IGP
     \end{itemize}
@@
     \begin{itemize}
-    \item except Brinkerhoff et al (2017) (\emph{upwinding, semi-coupled})
-    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
-    \item neither scalable (\emph{single level}) nor unconditionally stable
+    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
+    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
+    \item none are scalable (\emph{single level})
     \end{itemize}
@@
 \medskip
-    \item \emph{my goal}: unconditionally-stable implicit with Stokes dynamics
+    \item \emph{my goals}: Stokes, unconditionally-stable steps, scalable
     \end{itemize}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.15\textwidth]{figs/elephant.png}
+\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}
 \end{center}
@@
     \end{itemize}
-\item yet common to say $\Phi$ is hyperbolic because SKE is transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
+\item yet common to say $\Phi$ is hyperbolic because SKE is equivalent to a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
 \end{itemize}
@@
     \begin{itemize}
-    \item $P_1$ because low regularity at free boundary
-    \item uses numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$
+    \item $P_1$ because of low regularity at free boundary
+    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ using numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
     \end{itemize}
@@
     \item \alert{robust}
-    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$ 
+    \item \alert{optimal} $O(m)$ work over $m$ degrees of freedom in $\mathcal{V}^h$
         \begin{itemize}
-        \item[{\color{black} $\circ$}] optimal means work is $O(m)$ \dots requires multigrid
+        \item[{\color{black} $\circ$}] requires multigrid
         \end{itemize}
-    \item<2> \emph{for now these are aspirations \dots}
+    \item \emph{for now these are aspirations \dots}
     \end{itemize}
@@
 
-\begin{frame}{Stokes on an extruded mesh}
+\begin{frame}{smoothness and multilevel methods}
 
+\begin{columns}
+\column{0.35\textwidth}
     \begin{itemize}
-\item FIXME
+    \small
+    \item which solution variable?
+    \item choice between surface elevation ($s$) and thickness ($H$)
+    \item interior smoothness means $s$ is better for multigrid
+        \begin{itemize}
+        \item need to \emph{not} have fake ice where $s=b$
         \end{itemize}
-
-\includegraphics[width=0.8\textwidth]{figs/extruded.png}
+    \end{itemize}
+\column{0.65\textwidth}
+    \includegraphics[width=\textwidth]{figs/giscross.png}
+\end{columns}
 \end{frame}
@@
 \input{tikz/vcycle.tex}
+\end{frame}
+
+\begin{frame}{Stokes on an extruded mesh}
+
+\begin{itemize}
+\item FIXME
+\end{itemize}
+
+\includegraphics[width=0.8\textwidth]{figs/extruded.png}
 \end{frame}

### 2021-06-11 py/smoother.py
Update tests.

@@ -174,7 +174,10 @@ class ObstacleSmoother:
             snew_i <- b_i.
         After snew is completed we do s <- snew.'''
         c = self._colors(mesh1d, s)
-        print('      c = %d colors' % c)
+        if c >= mesh1d.m + 1:
+            print('      [coloring off]')
+        else:
+            print('      c = %d colors' % c)
         snew = mesh1d.b.copy() - 1.0  # snew NOT admissible; note check below
         snew[0], snew[mesh1d.m+1] = mesh1d.b[0], mesh1d.b[mesh1d.m+1]
         negd = []

### 2021-06-12 py/smoother.py
Credible NewtonRS implementation.  Needs thoughtful testing.

@@ -12,18 +12,26 @@ class ObstacleSmoother:
     '''A smoother on an obstacle problem.  Works on a mesh of class MeshLevel1D and calls a solver of class GlenStokes.  Note the mesh holds the bed
     elevation (obstacle).
 
-    Implements nonlinear Richardson plus nonlinear GS and Jacobi methods based
-    on the computation of the Jacobian diagonal entries by finite-differencing.
-    (These methods do not assume the residual has an easily-accessible
-    Jacobian diagonal.)
+    Implements:
+        * nonlinear Richardson (pointwise)
+        * nonlinear Gauss-Seidel (pointwise)
+        * nonlinear Jacobi (pointwise, coloring)
+        * Newton reduced-space (coloring)
+    In all cases the Jacobian entries are computed by finite-differencing,
+    but only the diagonal is used in the pointwise smoothers.  The coloring
+    smoothers get multiple rows of the Jacobian matrix by assigning different
+    colors to points which are more than a certain number of ice thicknesses
+    apart.
 
     The public interface implements residual evaluation and application of the
     in-place smoother:
         smooth = ObstacleSmoother(args, solver)
         r = smooth.residual(mesh1d, s, ella)
         smooth.smoothersweep(mesh1d, s, ella)
+    Note smoothersweep() calls a smoother from the dictionary self.smoothers.
+
     There is also evaluation of the CP norm,
-        irnorm = smooth.inactiveresidualnorm(mesh1d, s, r)
+        irnorm = smooth.cpresidualnorm(mesh1d, s, r)
     and a routine to trigger output:
         smooth.savestatenextresidual(filename)
     '''
@@ -31,9 +39,12 @@ class ObstacleSmoother:
     def __init__(self, args, solver):
         self.args = args
         self.solver = solver
-        self.ieps = 0.001  # if s[j] > b[j] + ieps then node j is inactive
         self.created = False
         self.saveflag = False
+        self.smoothers = {'richardson':  self._richardsonsweep,
+                          'gsslow':      self._gsslowsweep,
+                          'jacobicolor': self._jacobicolorsweep,
+                          'newtonrs':    self._newtonrs}
 
     def _checkadmissible(self, mesh1d, w, phi):
         '''Check admissibility and stop if not.'''
@@ -51,14 +62,13 @@ class ObstacleSmoother:
             ind = range(mesh1d.m, 0, -1)  # m,...,1
         return ind
[…]

### 2021-06-12 py/steady.py

@@ -76,6 +76,8 @@ adda('-eps', type=float, metavar='X', default=1.0e-2,  # FIXME sensitive
     help='regularization used in viscosity (default=%(default)s)')
 adda('-Hmin', type=float, metavar='X', default=0.0,
     help='minimum ice thickness (default=%(default)s)')
+adda('-initialzero', action='store_true', default=False,
+     help='initialize with s=0 (vs default of dome)')
 adda('-irtol', type=float, default=1.0e-3, metavar='X',
      help='reduce norm of inactive residual (default=%(default)s)')
 adda('-jcoarse', type=int, default=0, metavar='J',
@@ -94,7 +96,7 @@ adda('-omega', type=float, metavar='X', default=1.0,  # FIXME sensitive
     help='scale by this factor in smoother iteration (default=%(default)s)')
 adda('-printwarnings', action='store_true', default=False,
      help='print pointwise feasibility warnings')
-adda('-smoother', choices=['richardson', 'gsslow', 'jacobicolor', 'newtonrslu'],
+adda('-smoother', choices=['richardson', 'gsslow', 'jacobicolor', 'newtonrs'],
      metavar='X', default='jacobicolor',
      help='smoother (default=%(default)s)')
 adda('-steadyhelp', action='store_true', default=False,
@@ -131,7 +133,10 @@ else:
 
 # fine-level data
 ella = finemesh.ellf(problem.source(finemesh.xx()))  # source ell[v] = <a,v>
-s = problem.initial(finemesh.xx())
+if args.initialzero:
+    s = finemesh.zeros()
+else:
+    s = problem.initial(finemesh.xx())
 
 # set-up smoother with included Stokes solver
 smooth = ObstacleSmoother(args, GlenStokes(args))
@@ -146,7 +151,7 @@ if args.sweepsonly:
 else:
     print('MCDN with smoother "%s" ...' % args.smoother)
 r = smooth.residual(finemesh, s, ella)  # generates file if -o above
-normF0 = smooth.inactiveresidualnorm(finemesh, s, r)
+normF0 = smooth.cpresidualnorm(finemesh, s, r)
 if args.monitor:
     print('   0: %.4e' % normF0)
 for j in range(args.cyclemax):
@@ -155,7 +160,7 @@ for j in range(args.cyclemax):
     else:
         s += mcdnvcycle(args, smooth, hierarchy, s, ella)
[…]

### 2021-06-12 talk/talk.tex
minor edit
I could give this talk if necessary.

@@
-\documentclass[usepdftitle=false]{beamer}
+\documentclass[usepdftitle=false,usenames,dvipsnames]{beamer}
 
@@
 
-\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/titleextruded.png} \\ \vspace{-3mm} \includegraphics[width=0.15\textwidth]{figs/uafbw.png}}
+\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/titleextruded.png} \\ \vspace{-8mm} \includegraphics[width=0.2\textwidth]{figs/uafbw.png}}
 
@@
 
-\date{\phantom{foo} \bigskip \bigskip \bigskip \\ SIAM GS21 \\ with assistance from Lawrence Mitchell}
+\date{\phantom{foo} \bigskip \bigskip \bigskip \\ SIAM GS21}
 
@@
 	\maketitle
+\end{frame}
+
+\begin{frame}{overview}
+\Large
+\begin{enumerate}
+\item explain the problem I'm obsessed with
+\item discuss the current barriers to success
+\end{enumerate}
+
+\normalsize
+\vspace{10mm}
+\begin{itemize}
+\item<2> work in progress, with no results yet
+    \begin{itemize}
+    \item<2> low expectations, please
+    \end{itemize}
+\item<2> I am giving this talk at 4am my time
+    \begin{itemize}
+    \item<2> ditto
+    \end{itemize}
+\end{itemize}
 \end{frame}
@@
     \item $P_1$ because of low regularity at free boundary
-    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ using numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
+    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ by solving Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
     \end{itemize}
@@
     \item \alert{robust}
-    \item \alert{optimal} $O(m)$ work over $m$ degrees of freedom in $\mathcal{V}^h$
+    \item \alert{near optimal} $O(m^{1+\eps})$ work over $m$ degrees of freedom in $\mathcal{V}^h$
         \begin{itemize}
-        \item[{\color{black} $\circ$}] requires multigrid
+        \item[{\color{black} $\circ$}] requires multigrid \dots more below
         \end{itemize}
-    \item \emph{for now these are aspirations \dots}
+    \item \emph{these are aspirations}
     \end{itemize}
@@
     \item which solution variable?
-    \item choice between surface elevation ($s$) and thickness ($H$)
+    \item surface elevation ($s$) or thickness ($H$)?
     \item interior smoothness means $s$ is better for multigrid
-        \begin{itemize}
-        \item need to \emph{not} have fake ice where $s=b$
-        \end{itemize}
+    \item bed roughness means \emph{not} having ``fake ice'' where ice-free ($s=b$)
     \end{itemize}
@@
 
-\begin{frame}{multilevel constraint decomposition}
+\begin{frame}{multilevel constraint decomposition (MCD)}
+
+\begin{block}{strategy}
+decompose $\mathcal{K}^h$ using mesh hierarchy and then apply V-cycles
+\end{block}
+
+\bigskip
+\begin{columns}
+\column{0.52\textwidth}
+\begin{itemize}
+\item mesh levels over $\Omega$:
+
+$\mathcal{T}^0 \text{ (coarse) to } \mathcal{T}^J$
+\item given fine-level iterate $s^J$
+\item \emph{defect constraint} $\chi^J = b^J - s^J$
+\item \emph{monotone restriction} $R^\oplus$
+
+\item defect constraints on levels:
+
+$\chi^j = R^\oplus \chi^{j+1}$
+\item admissible perturbations:
+
+$z^j \ge \chi^j \iff s^j + z^j \ge b^j$
+\end{itemize}
+\column{0.48\textwidth}
+\includegraphics[width=1.05\textwidth]{figs/decompclassical.png}
+\end{columns}
+\end{frame}
+
+\begin{frame}{multilevel constraint decomposition (MCD)}
+
+\begin{block}{strategy}
+decompose $\mathcal{K}^h$ using mesh hierarchy and then apply V-cycles
+\end{block}
 
 \begin{itemize}
-\item FIXME
+\item this strategy gives $O(m\log m)$ solver for Laplacian obstacle problem (Tai, 2003)
+    \begin{itemize}
+    \item using Gauss-Seidel smoother
+    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{{\color{ForestGreen} a multilevel VI strategy is the only hope for (near) optimality}}
+    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
     \end{itemize}
+\item for IGP a single residual evaluation requires solving Stokes
+    \begin{itemize}
+    \item expensive but implemented {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
+    \item Stokes is nonlocal
 
-\includegraphics[width=0.5\textwidth]{figs/decompclassical.png}
+\smallskip
+    \item Jacobian dense
 
-\input{tikz/vcycle.tex}
+\smallskip
+    \item pointwise smoothers (GS, Jacobi) use diagonal entries
+        \begin{itemize}
+        \item<2>[{\color{black} $\circ$}] \alert{which apparently fail {\Large \, :-(}}
+        \end{itemize}
+    \end{itemize}
+\end{itemize}
 \end{frame}
@@
 
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figs/extruded.png}
+\end{center}
+
 \begin{itemize}
-\item FIXME
+\item for each base mesh level $\mathcal{T}^j$ in FAS V-cycle,
+\item evaluate the residual $\Phi(s^j) - a^j$ in the VI:
+    \begin{itemize}
+    \item source $a^j$ arises from FAS formula
+    \item extrude to current geometry $s^j$ (with or without cliffs)
+    \item no extrusion where ice free\footnote{thanks to Lawrence Mitchell for enabling this Firedrake capability}
+    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$ (Isaac et al, 2015)
+    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bu_{s^j}$ and subtract $a^j$
+    \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{smoother options}
+
+\begin{itemize}
+\item \only<1>{now all the evil is in the smoother}\only<2>{\alert{now all the evil is in the smoother}}
+\item options from weaker to stronger:
+    \begin{itemize}
+    \item nonlinear Richardson?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] not promising because degenerate
+        \end{itemize}
+    \item nonlinear pointwise smoothers (GS, Jacobi)?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] \only<1>{extremely expensive because residual nonlocal} \only<2>{\alert{extremely expensive because residual nonlocal}}
+        \item[{\color{black} $\circ$}] seem to fail anyway
+        \end{itemize}
+    \item nonlinear \texttt{PCPatch} smoothers?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] what does this even mean?
+        \end{itemize}
+    \item exploit SIA analogy?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
+        \end{itemize}
+    \item reduced-space Newton using banded Jacobian approximation?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
+        \end{itemize}
     \end{itemize}
 
-\includegraphics[width=0.8\textwidth]{figs/extruded.png}
+\bigskip
+\item<2> status report: \alert{nothing works well so far}
+\end{itemize}
 \end{frame}
 
-\begin{frame}{smoothers}
+\begin{frame}{conclusion}
 
 \begin{itemize}
-\item FIXME
+\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
+    \begin{itemize}
+    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
+    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
+    \end{itemize}
+\item \emph{GOAL 2}:\, beat explicit time-stepping and enable paleo {\Large \strut} \only<3>{\Large \alert{:-(}}
+    \begin{itemize}
+    \item<2-3> problem is
+        \begin{itemize}
+        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
+        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
+        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
+        \end{itemize}
+    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for the VI}
+    \item<3> \alert{stuck on finding a robust optimal smoother}
+    \end{itemize}
+
+\normalsize
+\bigskip\bigskip
+\item<3> \emph{thanks for your attention \dots questions?}
 \end{itemize}
@@
 
-\begin{frame}{Some references}
+\begin{frame}{Some references 1}
+%\nocite{*}
 %\bibliography{../paper/msg.bib}
@@
 \begin{thebibliography}{1}
-\bibitem{BuelerMitchell2022}
-E.~Bueler and L.~Mitchell. \textbf{Multilevel computation of glacier geometry from Stokes dynamics.} in preparation
-
-\hrulefill
-
 \bibitem{Brinkerhoffetal2017}
@@
 
+\bibitem{Bueler2016}
+E.~Bueler. \textbf{Stable finite volume element schemes for the shallow-ice approximation.} {\em J. Glaciol.}, 62 (232):230--242, 2016.
+
 \bibitem{IsaacStadlerGhattas2015}
 T.~Isaac, G.~Stadler, and O.~Ghattas.  \textbf{Solution of nonlinear Stokes equations \dots with application to ice sheet dynamics.} {\em SIAM J. Sci. Comput.}, 37(6):B804--B833, 2015.
+
+\bibitem{JouvetBueler2012}
+G.~Jouvet and E.~Bueler. \textbf{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation.} {\em SIAM J. Appl. Math.}, 72(4):1292--1314, 2012.
+
+\bibitem{JouvetRappaz2011}
+G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary {S}tokes problem arising in glaciology.} {\em Advances in Numerical Analysis}, 2011.
+\end{thebibliography}
+\end{frame}
+
+\begin{frame}{Some references 2}
+\begin{thebibliography}{1}
+\bibitem{Tai2003}
+X.-C. Tai. \textbf{Rate of convergence for some constraint decomposition methods for nonlinear variational inequalities.} {\em Numer. Math.}, 93(4):755--786, 2003.
 

### 2021-06-12 talk/talk.tex
Few more edits.

@@
 \item glacier geometry which evolves according to climate
-\item take simplest such problem (grounded, nonsliding, isothermal)
-\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
+    \begin{itemize}
+    \item simplest such problem (grounded, nonsliding, isothermal)
+    \end{itemize}
+\item assume time-independent data on fixed $\Omega \subset \RR^2$:
     \begin{itemize}
@@
             \item $s=b$ where ice is not present
-            \item $s$ is defined on all of $\Omega$
+
+                \begin{itemize}
+                \item[{\color{black} $\circ$}] $s$ is defined on all of $\Omega$
+                \end{itemize}
             \end{itemize}
@@
       $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
-    \item \emph{count the 3 nonlinearities}
+    \item \alert{\emph{the 3 nonlinearities}}
     \end{itemize}
@@
     \end{itemize}
-\item what are people doing instead?
+\item what are people doing instead, to get science?
     \begin{itemize}
@@
     \begin{itemize}
-    \item a regularity conjecture
+    \item which is a regularity conjecture
     \end{itemize}
@@
     \begin{itemize}
-    \item $\Phi$ is some kind of integral operator? (\emph{convolve with Stokeslets})
-    \item $\Phi$ is short range? (\emph{concentrated near the diagonal})
+    \item $\Phi$ is some kind of integral operator (\emph{convolve with Stokeslets})
+    \item $\Phi$ is short range (\emph{concentrated near the diagonal})
     \end{itemize}
@@
     \begin{itemize}
-    \item $\Phi$ is diffusive in the large!
-    \item $\Phi$ is degenerate at the ice margin!
+    \item $\Phi$ is diffusive in the large
+    \item $\Phi$ is degenerate at the ice margin
     \end{itemize}
-\item yet common to say $\Phi$ is hyperbolic because SKE is equivalent to a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
+\item yet common to say
+    \begin{itemize}
+    \item $\Phi$ is hyperbolic because SKE is a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
+    \end{itemize}
 \end{itemize}

### 2021-06-13 paper/msg.tex
Attempt to put ideas in some order.

@@
 
-In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations which are variants of the Gauss-Seidel and Jacobi methods, but made suitable for VIs \cite{KinderlehrerStampacchia1980}.
-Given a current surface elevation $s^h\in \mathcal{K}^h$, these iterations sweep through the $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems.
-While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.
+In this section we consider single-level solvers for \eqref{eq:fesigpweakform}, namely certain nonlinear iterations which are made suitable for VIs \cite{KinderlehrerStampacchia1980}.
+While these iterations are, in some cases, adequate smoothers in a multilevel method (next section), they converge slowly if used directly as solvers.
 Because of the non-locality and density of the IDO $\Phi^h$ (Section \ref{sec:weakido}), we must make careful approximations in order to retain adequate per-iteration performance.
+These iterative solvers, \emph{smoothers} from now on, can be categorized by the number of nodes which are adjusted at each iteration: pointwise, patch-type, and global.
 
-Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.
-Let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.
-Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.
-Note that $\psi_i \in \mathcal{W}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
-We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.
+Suppose there are $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$, denote a vertex in $\mathcal{T}$ by $(x_j,y_j)$, and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$.
+Let $s^h\in \mathcal{K}^h$ be the current surface elevation iterate.
+The Richardson pointwise smoother sweeps through the nodes and subtracts-off the residual: $s^h \gets s^h - \alpha (F^h(s^h)-a)$ where $\alpha>0$ must be chosen so that the iteration converges.
+By contrast the Gauss-Seidel, and Jacobi pointwise smoothers sweep through the nodes solving one-dimensional VI problems for a value $c\in \RR$ (below).
+If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node before doing any updates.
+In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}.
 
-The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for some scalars $\gamma,c$.
-Specifically, we seek $c \ge \beta_i^{s^h}$ such that
+To be more precise, let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.
+If $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, note that $\psi_i \in \mathcal{W}_+^h$, and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.
+(One might call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.)  The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} by replacements $s^h \to s^h+c \psi_i$, $r^h \to r^h + \gamma \psi_i$ for some scalars $c,\gamma$.
+We then seek $c \ge \beta_i^{s^h}$ such that
 \begin{equation}
@@
 \end{equation}
-be the point-wise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
+be the pointwise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
 \begin{equation}
@@
 \end{equation}
-The smoother will approximately solve \eqref{eq:fepointwisevi} for $c$ the smoother, and then update $s^h \gets s^h + c \psi_i$.
+The smoother approximately solves \eqref{eq:fepointwisevi} for $c$ and then updates $s^h \gets s^h + c \psi_i$ either in serial or parallel.
 Observe that admissibility is preserved because $s^h+c \psi_i \in \mathcal{K}^h$ when $c \ge \beta_i^{s^h}$.
 
-If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.
+If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to VI \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.
 In fact $\rho_i(s^h; c)$ is nonlinear, and we will linearize
 \begin{equation}
-\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0), \label{eq:rhoapprox}
+\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0). \label{eq:rhoapprox}
 \end{equation}
-and at each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.
-If $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.
+At each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.
+However, if $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.
 % FIXME TOO AGGRESSIVE?
 
-Because the interior PDE in the shallow, flat-bed model is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}.
-However, for the full Glen-Stokes model there is no such guarantee to our knowledge.
+Using a single Newton step and approximation \eqref{eq:rhofd}, GS is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.
+It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.
+Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.
 
+\begin{pcode}[ht]
+\begin{pseudo*}
+\pr{pngs}(s^h,b^h,\id{omega}=1.0)\text{:} \\+
+    \ct{check admissibility: $s^h \ge b^h$} \\
+    for $i = 0,\dots,m-1$ \\+
+        $\alpha_i = \rho_i'(s^h; 0)$  \qquad\qquad \ct{Jacobian diagonal entry} \\
+        if $\alpha_i > 0$ \\+
+            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
+            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
+        else \\+
+            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
+\end{pseudo*}
+\caption{Projected nonlinear GS iteration, a conceptual in-place, pointwise smoother which solves one-dimensional VIs \eqref{eq:fepointwisevi} in serial.}
+\label{pc:pngsslow}
+\end{pcode}
+
+Patch-type smoothers \cite{Farrelletal2019} have to our knowledge not been used for VI problems before, but they would sweep through the nodes solving local VIs of small dimension $\ell>1$, and again there would be serial and parallel versions.
+However, we first consider the opposite extreme from a pointwise smoother, namely a global smoother in which all the degrees of freedom are coupled together into an algebraic system.
+
+FIXME reduced-space Newton solver using a banded Jacobian approximation
+
+\section{Approximate Jacobians} \label{sec:jacobians}
+
+The above pseudocodes assume the accessibility of the diagonal of the Jacobian (pointwise smoothers) or of arbitrary entries (Newton solver).
+
+Because the interior PDE in the SIA is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}, but for the full Glen-Stokes model there is no such guarantee to our knowledge.
 Note that the derivative $\rho_i'(s^h; 0)$, which is also a diagonal entry of the Jacobian of $F^h$, can be approximated by a finite difference,
@@
 
-If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node $i$ before doing any updates.
-In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely that GS imposes the expense of yet more residual evaluations.
-Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.
-It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.
-Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.
+A glaring difference between the GS and Jacobi smoothers now dominates.  Namely, GS imposes the expense of far more residual evaluations.
 
-\begin{pcode}[ht]
-\begin{pseudo*}
-\pr{pngs\_slow}(s^h,b^h,\id{eps}=1.0,\id{omega}=1.0)\text{:} \\+
-    \ct{check admissibility: $s^h \ge b^h$} \\
-    for $i = 0,\dots,m-1$ \\+
-        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{Jacobian diagonal entry} \\
-        if $\alpha_i > 0$ \\+
-            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
-            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
-        else \\+
-            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
-\end{pseudo*}
-\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.}
-\label{pc:pngsslow}
-\end{pcode}
+FIXME finite-difference approximation using pseudo-coloring based on ice thicknesses or rather longitudinal coupling length \cite{KambEchelmeyer1986}
 
@@
 
-FIXME SMOOTHER POSSIBILITIES.  1 COLORING (PSEUDO-COLORING WITH LATERAL SEPARATION OF A FEW ICE THICKNESSES).  2 PATCH-TYPE WHICH SOLVE LOW-DIMENSIONAL VI.
+FIXME PSEUDO-COLORING WITH LATERAL SEPARATION OF A FEW ICE THICKNESSES
 
@@
 
+\section{Performance models} \label{sec:perfmodels}
+
+FIXME
+
 \section{Results for steady geometry} \label{sec:results}

### 2021-06-13 talk/talk.tex
Fiddle around with language.

@@
 
-\begin{frame}{coupling via operator}
+\begin{frame}{rewrite coupling as operator}
 
@@
     \begin{itemize}
-    \item which is a regularity conjecture
+    \item which is a regularity conjecture for the coupled theory
     \end{itemize}
@@
     \begin{itemize}
-    \item $\Phi$ is diffusive in the large
+    \item $\Phi$ is elliptic in the large
     \item $\Phi$ is degenerate at the ice margin
@@
     \begin{itemize}
-    \item $\Phi$ is hyperbolic because SKE is a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
+    \item $\Phi$ is advective because SKE is write-able as a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
     \end{itemize}
@@
 
-\begin{block}{numerical goal}
+\begin{block}{numerical problem}
 given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, and convex set $\mathcal{K}^h = \{r^h \ge b^h\} \subset \mathcal{V}^h$, solve VI for $s^h \in \mathcal{K}^h$:
@@
     \item which solution variable?
-    \item surface elevation ($s$) or thickness ($H$)?
-    \item interior smoothness means $s$ is better for multigrid
-    \item bed roughness means \emph{not} having ``fake ice'' where ice-free ($s=b$)
+        \begin{itemize}
+        \item surface elevation $s$?
+        \item thickness $H$?
+        \end{itemize}
     \end{itemize}
@@
 \end{columns}
+
+\bigskip
+\begin{itemize}
+\item \alert{smoothness suggests $s$ is better for multigrid}
+    \begin{itemize}
+    \item \emph{corollary}. do not put ``fake ice'' where $s=b$
+    \item only solve for ice velocity where there is ice!
+    \end{itemize}
+\end{itemize}
 \end{frame}
@@
 \begin{itemize}
-\item this strategy gives $O(m\log m)$ solver for Laplacian obstacle problem (Tai, 2003)
+\item in V-cycle, on each level solve a VI:
+    $$\ip{\Phi^j(s^j + z^j) - a^j}{v^j - z^j} \ge 0 \quad \text{ for all $v^j \ge \chi^j$}$$
+
+\item this strategy gives an $O(m\log m)$ solver for the Laplacian obstacle problem (Tai, 2003)
     \begin{itemize}
-    \item using Gauss-Seidel smoother
-    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{{\color{ForestGreen} a multilevel VI strategy is the only hope for (near) optimality}}
+    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{\alert{a multilevel VI strategy is the only hope for (near) optimality}}
+    \end{itemize}
+\item for IGP:
+    \begin{itemize}
     \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
-    \end{itemize}
-\item for IGP a single residual evaluation requires solving Stokes
-    \begin{itemize}
-    \item expensive but implemented {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
-    \item Stokes is nonlocal
-
-\smallskip
-    \item Jacobian dense
-
-\smallskip
-    \item pointwise smoothers (GS, Jacobi) use diagonal entries
-        \begin{itemize}
-        \item<2>[{\color{black} $\circ$}] \alert{which apparently fail {\Large \, :-(}}
-        \end{itemize}
+    \item each residual is nonlocal and expensive {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
+    \item<2> {\color{ForestGreen} strategy implemented}
     \end{itemize}
@@
 
-\begin{frame}{Stokes on an extruded mesh}
+\begin{frame}{Stokes on a Firedrake extruded mesh}
 
@@
 
+\begin{columns}
+\column{0.77\textwidth}
 \begin{itemize}
-\item for each base mesh level $\mathcal{T}^j$ in FAS V-cycle,
-\item evaluate the residual $\Phi(s^j) - a^j$ in the VI:
+\item V-cycle based on mesh hierarchy $\{\mathcal{T}^j\}$ over $\Omega$
+\item to evaluate the residual $\Phi(s^j) - a^j$ in the VI:
     \begin{itemize}
-    \item source $a^j$ arises from FAS formula
-    \item extrude to current geometry $s^j$ (with or without cliffs)
-    \item no extrusion where ice free\footnote{thanks to Lawrence Mitchell for enabling this Firedrake capability}
-    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$ (Isaac et al, 2015)
-    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bu_{s^j}$ and subtract $a^j$
+    \item extrude base mesh $\mathcal{T}^j$ to current geometry $s^j$
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] with or without small cliffs
+        \end{itemize}
+    \item no extrusion where ice free
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] thanks to Lawrence Mitchell for this capability
+        \end{itemize}
+    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$
+    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bn_{s^j}$ and subtract $a^j$
     \end{itemize}
 \end{itemize}
+\column{0.23\textwidth}
+\input{tikz/vcycle.tex}
+\end{columns}
 \end{frame}
@@
 \begin{itemize}
-\item \only<1>{now all the evil is in the smoother}\only<2>{\alert{now all the evil is in the smoother}}
-\item options from weaker to stronger:
+\item now all the evil is in the smoother
+\item smoother options from weaker to stronger:
     \begin{itemize}
@@
         \begin{itemize}
-        \item[{\color{black} $\circ$}] \only<1>{extremely expensive because residual nonlocal} \only<2>{\alert{extremely expensive because residual nonlocal}}
-        \item[{\color{black} $\circ$}] seem to fail anyway
-        \end{itemize}
-    \item nonlinear \texttt{PCPatch} smoothers?
-        \begin{itemize}
-        \item[{\color{black} $\circ$}] what does this even mean?
+        \item[{\color{black} $\circ$}] \only<1>{GS extremely expensive because residual nonlocal} \only<2>{\alert{GS extremely expensive because residual nonlocal}}
+        \item[{\color{black} $\circ$}] seem to fail
         \end{itemize}
@@
         \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
+        \end{itemize}
+    \item nonlinear and nonlocal \texttt{PCPatch} smoothers?
+        \begin{itemize}
+        \item[{\color{black} $\circ$}] what does this even mean?
         \end{itemize}
@@
 
-\bigskip
-\item<2> status report: \alert{nothing works well so far}
+\medskip
+\item<2> status report:
+    \begin{itemize}
+    \item \alert{nothing works well so far}
+    \item NewtonRS and SIA-based smoothers most promising?
+    \end{itemize}
 \end{itemize}
@@
     \end{itemize}
-\item \emph{GOAL 2}:\, beat explicit time-stepping and enable paleo {\Large \strut} \only<3>{\Large \alert{:-(}}
+\item \emph{GOAL 2}:\, beat explicit time-stepping and enable long-time paleo simulations {\Large \strut} \only<3>{\Large \alert{:-(}}
     \begin{itemize}
@@
         \end{itemize}
-    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for the VI}
+    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for solving the VI}
     \item<3> \alert{stuck on finding a robust optimal smoother}

### 2021-06-21 talk/talk.tex
Edit talk.

@@
 \begin{itemize}
-\item<2> work in progress, with no results yet
+\item<2-3> work in progress
+\item<2-3> I am giving this talk at 4am my time
     \begin{itemize}
-    \item<2> low expectations, please
-    \end{itemize}
-\item<2> I am giving this talk at 4am my time
-    \begin{itemize}
-    \item<2> ditto
+    \item<3> low expectations, please
     \end{itemize}
@@
 \begin{itemize}
-\item glacier geometry which evolves according to climate
+\item glacier geometry evolution according to the climate?
     \begin{itemize}
@@
     \begin{itemize}
-    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
-    \item bed elevation $b$
+    \item (net) climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for accumulation}
+    \item bed elevation $b(x,y)$
     \end{itemize}
-\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
+\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$ \emph{and} icy domain $\Lambda_s \subset \RR^3$
 \end{itemize}
@@
 \begin{itemize}
-\item glaciers flow
-\item $\therefore$ must solve an equation for the surface elevation
-\item the surface kinematical equation (SKE) holds \emph{on} the ice
-\begin{align*}
-\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{evolving}} \\
--\bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{steady state}}
-\end{align*}
+\item glaciers flow, so one must solve the surface kinematical equation (SKE) \emph{on} the ice
+    $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a = 0$$
 
@@
 \begin{itemize}
-\item the steady IGP free-boundary problem, for $s(x,y)$ on $\Omega$, is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}
+\item the \alert{steady} IGP is a \alert{nonlinear complementarity problem (NCP) for $s$}, a free-boundary problem, \alert{which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}
 
@@
 
-\begin{frame}{\alert{implicit} IGP strong form}
+\begin{frame}{implicit IGP strong form}
 
@@
 \begin{itemize}
-\item almost no one is solving the implicit IGP
+\item almost no one is solving such a implicit or steady IGP
     \begin{itemize}
@@
     \end{itemize}
-\item what are people doing instead, to get science?
+\item what are people doing instead, for production science?
     \begin{itemize}
@@
     \item often with shallow approximations, e.g.~PISM run below
+    \end{itemize}
 
@@
 \end{columns}
-
-\medskip
-    \item \emph{my goals}: Stokes, unconditionally-stable steps, scalable
-    \end{itemize}
 \end{itemize}
@@
 extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
-\item regarding Stokes problem for fixed geometry:
+\item regarding this Stokes problem (for \emph{fixed} geometry):
     \begin{itemize}
@@
     \begin{itemize}
-    \item which is a regularity conjecture for the coupled theory
+    \item which is a regularity conjecture
     \end{itemize}
@@
 
-\item well-posedness is completely open
+\item well-posedness is almost completely open
     \begin{itemize}
-    \item but existence holds for SIA version (Jouvet \& Bueler, 2012)
+    \item existence holds for SIA version (Jouvet \& Bueler, 2012)
     \end{itemize}
@@
     \begin{itemize}
-    \item $\Phi$ is elliptic in the large
+    \item $\Phi$ is elliptic ``in the large''
     \item $\Phi$ is degenerate at the ice margin

### 2021-06-21 paper/genfigs/stokesdomain.py

@@ -59,6 +59,7 @@ writeout('stokesdomain.pdf')
 plt.figure(figsize=(16,4))
 x, s, _ = genbasicfig(xshift=0.0)
 plt.text(x[300], s[300] + 1.2, r'$s$', fontsize=bigfsize, color='k')
+plt.text(x[500], s[500] - 1.2, r'$\Lambda_s$', fontsize=bigfsize, color='k')
 plt.arrow(10.0, 1.5, 1.0, 0.0,
           lw=1.5, head_width=0.1, color='k')
 plt.text(10.4, 1.6, r'$\Phi$', fontsize=bigfsize, color='k')

### 2021-06-21 talk/talk.tex
More.

@@
     \begin{itemize}
-    \item \emph{corollary}. do not put ``fake ice'' where $s=b$
-    \item only solve for ice velocity where there is ice!
+    \item but \emph{do not} put ``fake ice'' where $s=b$; only solve for ice velocity where there is ice!
     \end{itemize}
@@
 
-\begin{block}{strategy}
-decompose $\mathcal{K}^h$ using mesh hierarchy and then apply V-cycles
-\end{block}
-
 \begin{itemize}
@@
     \begin{itemize}
-    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{\alert{a multilevel VI strategy is the only hope for (near) optimality}}
+    \item \alert{a multilevel VI strategy is the only hope for (near) optimality}
     \end{itemize}

### 2021-06-22 talk/talk.tex
Simplifying edits after practice run (23 min).

@@
 \begin{enumerate}
-\item explain the problem I'm obsessed with
-\item discuss the current barriers to success
+\item explain the problem I'm working on
+\item mention the (current) barriers to success
 \end{enumerate}
@@
 \begin{itemize}
-\item<2-3> work in progress
-\item<2-3> I am giving this talk at 4am my time
+\item work in progress
+\item I am giving this talk at 4am my time
     \begin{itemize}
-    \item<3> low expectations, please
+    \item low expectations, please
     \end{itemize}
@@
 \begin{itemize}
-\item glacier geometry evolution according to the climate?
-    \begin{itemize}
-    \item simplest such problem (grounded, nonsliding, isothermal)
-    \end{itemize}
+\item how does glacier geometry evolve in response to the climate?
+\item only the simplest version (grounded, nonsliding, isothermal)
 \item assume time-independent data on fixed $\Omega \subset \RR^2$:
@@
     \end{itemize}
-\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$ \emph{and} icy domain $\Lambda_s \subset \RR^3$
+\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$
+
+\hspace{20.5mm} icy domain $\Lambda_s \subset \RR^3$
 \end{itemize}
@@
     \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
+
+\vspace{4mm}
     \item where \emph{not} on the ice: $a\le 0$
@@
         \begin{itemize}
-        \item assumed for this talk:
+        \item assumed:
             \begin{itemize}
@@
                 \begin{itemize}
-                \item[{\color{black} $\circ$}] $s$ is defined on all of $\Omega$
+                \item[{\color{black} $\therefore$}] $s$ is defined on all of $\Omega$
                 \end{itemize}
@@
     \begin{itemize}
-    \item very similar to steady IGP
+    \item \alert{very similar to steady IGP}
     \end{itemize}
@@
     \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
-    \item often with shallow approximations, e.g.~PISM run below
+    \item usually shallow approximations, e.g.~PISM run below
     \end{itemize}
@@
     \end{itemize}
-\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
+\item[$\therefore$] $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
     \begin{itemize}
@@
 \begin{itemize}
-\item using $\Phi$ gives a cleaner NCP over $\Omega$ for steady IGP:
+\item using $\Phi$ gives a cleaner NCP over $\Omega$ for the steady IGP:
 \begin{align*}
@@
 
-\item NCP (strong form) $\leftrightarrow$ VI (weak form)
+    \begin{itemize}
+    \item recall NCP (strong form) $\leftrightarrow$ VI (weak form)
+    \end{itemize}
 
@@
 find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
-    $$\ip{\Phi(s)-a}{r-s} \ge 0$$
-for all $r \in \mathcal{K}$
+    $$\ip{\Phi(s)-a}{r-s} \ge 0 \qquad \text{ for all } r \in \mathcal{K}$$
 \end{block}
@@
 \item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
-\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
+\item the Stokes stress balance is nonlocal
     \begin{itemize}
-    \item $\Phi$ is some kind of integral operator (\emph{convolve with Stokeslets})
-    \item $\Phi$ is short range (\emph{concentrated near the diagonal})
+    \item $\Phi$ is integral operator (\emph{convolve with Stokeslets})
+    \item $\Phi$ is short range (\emph{dense but concentrated near the diagonal})
     \end{itemize}
@@
     \begin{itemize}
-    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
-    \item each residual is nonlocal and expensive {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
-    \item<2> {\color{ForestGreen} strategy implemented}
+    \item for nonlinear VI, need additional FAS strategy {\Large\strut} {\color{ForestGreen} \, {\Large $\checkmark$}}
+    \item \dots but each residual is nonlocal and expensive
     \end{itemize}
@@
     \item nonlinear Richardson?
-        \begin{itemize}
-        \item[{\color{black} $\circ$}] not promising because degenerate
-        \end{itemize}
-    \item nonlinear pointwise smoothers (GS, Jacobi)?
-        \begin{itemize}
-        \item[{\color{black} $\circ$}] \only<1>{GS extremely expensive because residual nonlocal} \only<2>{\alert{GS extremely expensive because residual nonlocal}}
-        \item[{\color{black} $\circ$}] seem to fail
-        \end{itemize}
     \item exploit SIA analogy?
@@
         \end{itemize}
-    \item nonlinear and nonlocal \texttt{PCPatch} smoothers?
+    \item nonlinear pointwise smoothers (GS, Jacobi)?
         \begin{itemize}
-        \item[{\color{black} $\circ$}] what does this even mean?
+        \item[{\color{black} $\circ$}] \alert{GS extremely expensive because residual nonlocal}
+        \item[{\color{black} $\circ$}] seem to fail
         \end{itemize}
+    %\item nonlinear and nonlocal \texttt{PCPatch} smoothers?
+    %    \begin{itemize}
+    %    \item[{\color{black} $\circ$}] what does this even mean?
+    %    \end{itemize}
     \item reduced-space Newton using banded Jacobian approximation?
@@
     \end{itemize}
+\end{itemize}
+\end{frame}
 
-\medskip
-\item<2> status report:
+\begin{frame}{status report}
+
 \begin{itemize}
 \item \alert{nothing works well so far}
-    \item NewtonRS and SIA-based smoothers most promising?
-    \end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
+\item \emph{GOAL 1}:\, no shallow approximations in flow physics
     \begin{itemize}
-    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
-    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
+    \item[{\Large \color{ForestGreen} $\checkmark$}] Stokes solver working in Firedrake on extruded mesh
+    \item[{\Large \color{ForestGreen} $\checkmark$}] solver choices by (Isaac et al, 2015) putatively optimal
     \end{itemize}
-\item \emph{GOAL 2}:\, beat explicit time-stepping and enable long-time paleo simulations {\Large \strut} \only<3>{\Large \alert{:-(}}
+\item \emph{GOAL 2}:\, beat explicit time-stepping
     \begin{itemize}
-    \item<2-3> problem is
+    \item[{\Large \color{ForestGreen} $\checkmark$}] observe that problem is
         \begin{itemize}
-        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
-        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
-        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
+        \item[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
+        \item[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
+        \item[{\color{black} $\circ$}] nonlocal (Stokes)
         \end{itemize}
-    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for solving the VI}
-    \item<3> \alert{stuck on finding a robust optimal smoother}
+    \item[{\Large \color{ForestGreen} $\checkmark$}] MCD FAS path to optimality for solving the VI
+    \item[{\Large \alert{:-(}}] \alert{stuck on finding a robust, efficient smoother}
+    \end{itemize}
 \end{itemize}
 
-\normalsize
-\bigskip\bigskip
-\item<3> \emph{thanks for your attention \dots questions?}
-\end{itemize}
+\begin{block}{ultimate goal}
+for better science, enable long-time (paleo) simulations without shallow approximations
+\end{block}
+
+\emph{thanks for your attention \dots questions?}
 \end{frame}

### 2021-12-26 paper/msg.tex
Edits.

@@
 
-The standard model for determining the geometry of glaciers is based in part upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.
-This Glen-Stokes ice-dynamics model can only determine the free-surface glacier geometry when it is combined with the surface kinematical equation (SKE) and a climate input, namely the balance of the rates of accumulation of snow and of ablation, i.e.~melting and runoff.
+The standard model for determining the geometry of glaciers is based upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.
+However, this Glen-Stokes ice-dynamics model only determines the free-surface glacier geometry when combined with the surface kinematical equation (SKE) and a climate input, namely the balance of the rates of snow accumulation and melting/runoff (ablation).
 In this section we state the strong form of the resulting ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).
-The emphasis here is on the complementarity-problem nature of the SIGP, which is often not explicit in the glaciers literature (e.g.~as observed by \cite{SchoofHewitt2013}).
-We then define its weak form, noting along the way several unknown aspects of the theory.
-Section \ref{sec:evolution} gives the straightfoward extension of the steady case to a time-discretized evolving-geometry model, the implicit ice geometry problem (IIGP).
+The emphasis here is on the complementarity-problem nature of the SIGP, which is often not explicit in the glaciers literature (but see \cite{SchoofHewitt2013}).
+We then define the weak form, noting along the way several unresolved aspects of the theory.
+The steady model has a straightforward extension to a time-discretized evolving-geometry model, the implicit ice geometry problem (IIGP; Section \ref{sec:evolution}).
 
-The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ for $d=1,2$ (Figure \ref{fig:stokesdomain}).
-The coordinates on $\Omega$ are denoted as $x$ when $d=1$ and $x,y$ when $d=2$.
-On $\Omega$ we assume that a climatic mass balance (CMB) \cite{Cogleyetal2011} function $a(x,y)$ is defined at every point, whether or not ice is present at that location.
+The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ for $d=1,2$ (Figure \ref{fig:stokesdomain}; coordinates on $\Omega$ are denoted as $x$ when $d=1$ and $x,y$ when $d=2$).
+On $\Omega$ we assume that a time-independent climatic mass balance (CMB) \cite{Cogleyetal2011} function $a(x,y)$ is defined at every point, whether or not ice is present at that location.
 In ice-free areas this function is the ``potential'' CMB, namely the annual balance of snow accumulation and melt if ice were present, e.g.~as computed by energy balance in a climate or weather model.
-Here we assume $a$ has units of ice thickness per time, compatible with our assumption of constant ice density (below).
-Next we assume there is a bed elevation function $b(x,y)$ defined everywhere on $\Omega$, which determines the topography upon which the glacier sits.
+We assume $a$ has units of ice thickness per time, compatible with our assumption of constant ice density.
+Also, a bed elevation function $b(x,y)$ is defined everywhere on $\Omega$; it gives the topography upon which the glacier sits.
 The functions $a$ and $b$ are the data (inputs) of the SIGP.
@@
 
-The Glen-Stokes dynamical model, the so-called stress balance, only applies in the icy domain in $\RR^{d+1}$.
-We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption this this icy domain has a well-defined upper surface elevation, a function $s(x,y)$.
+Now, the Glen-Stokes stress balance equations (below) only apply in a certain icy domain in $\RR^{d+1}$.
+We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption that this domain has a well-defined upper surface elevation, a function $s(x,y)$.
 That is, we assume there are \emph{no overhangs}, and that the under-side of the ice is in contact with the bed $b$.
@@
 
-However, as a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.
-There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that the function $s$ is well-defined.
-Furthermore the same concern applies to each time step of an evolving model; overhangs can develop at the margin.
-Because overhangs are small features in large glaciers and ice sheets, most modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.
-Exceptions \cite{PralongFunk2005} and \cite{Jouvetetal2011} provide a continuum model in which ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion, and such a model might explain how (essentially) well-defined surface elevation and thickness could arise in a well-posed model.
-However, such a model is a nontrivial extension of the above equations, and though marginal cliffs may exist in our model (Section \ref{sec:fe}), overhangs are never allowed herein.
-
-Based on a well-defined upper surface, the icy domain is the open set
+Based on this well-defined upper surface, the icy domain is the open set
 \begin{equation}
@@
 where $z$ is vertically-upward.
-(Coordinates are denote $(x,z)$ on $\Lambda_s$ when $d=1$.)  The domain $\Lambda_s$ has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).
-Again note $\Lambda_s$ is part of the solution.
+The domain $\Lambda_s$, which is a \emph{solution} in our model, has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).
 
-We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.
-Allowing any Glen exponent $\nn\ge 1$, the equations inside $\Lambda_s$ are
+We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid in $\Lambda_s$.  Allowing any Glen exponent $\nn\ge 1$ \cite{GreveBlatter2009}, the equations are
 \begin{align}
@@
 The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.
-Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation.
+Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation \cite[Chapter 1]{FowlerNg2021}.
 
-Regarding tensors and their notation, recall that the (Cauchy) stress tensor $\sigma$ decomposes into the deviatoric part $\tau$ minus the pressure, i.e.~$\sigma = \tau - p\,I$, so equation \eqref{eq:forcebalance} simply says $-\Div \sigma = \rhoi \bg$.
+Regarding tensors and their notation, recall that the Cauchy stress tensor $\sigma$ decomposes into the deviatoric part $\tau$ minus the pressure, i.e.~$\sigma = \tau - p\,I$, so equation \eqref{eq:forcebalance} simply says $-\Div \sigma = \rhoi \bg$.
 The strain rate tensor $D\bu$ is the symmetric part of $\grad \bu$, $D\bu = \frac{1}{2} \left(\grad\bu + \grad\bu^\top\right)$, and the tensor norm used in \eqref{eq:viscflowlaw} satisfies $|D\bu|^2 = \frac{1}{2} (D\bu)_{ij} (D\bu)_{ij}$.
-Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} then implies that $\tau$ is also symmetric with trace zero, thus that $p=-(d+1)^{-1} \trace \sigma$.
+Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} then implies that $\tau$ is also symmetric with trace zero, thus that $p=-\trace \sigma / (d+1)$.
 
@@
 Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.
+
+As a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.
+There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, generates an overhang, violating the assumption that the function $s$ is well-defined.
+Furthermore the same concern applies to each time step of an evolving model; overhangs can develop at the margin.
+Because overhangs are small features in large glaciers and ice sheets, most modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.
+The exceptional literature \cite{Jouvetetal2011,PralongFunk2005} provides a model in which ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion, which might explain how a (nearly) well-defined surface elevation and thickness could arise in a more-complete, and well-posed, model.
+However, such a model is a nontrivial extension of the above equations.
+Neither marginal cliffs nor overhangs are allowed in our theory.
 

### 2021-12-26 paper/msg.tex
Edits.

@@
 \end{center}
-\caption{In the SIGP the CMB $a$ (arrows; down for accumulation) and bed elevation $b$ (solid) are given data on a map-plane region $\Omega \subset \RR^d$.
+\caption{In the SIGP the CMB $a$ (arrows; down for accumulation) and bed elevation $b$ (solid) are given data on a map-plane region $\Omega \subset \RR^d$ for $d=1,2$.
 The solution includes a surface elevation $s$ (dashed) on $\Omega$, such that $s\ge b$, plus the ice velocity $\bu$ and pressure $p$ on the icy domain $\Lambda_s = \{b < z < s\} \subset \RR^{d+1}$.}
@@
 
-In computations we will use constants $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.
-The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.
+In computations we will use the following constants: Glen exponent $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.
+The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996}.
 
-For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.
-This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
+For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ suggest an effective viscosity function of $|D\bu|$.
+This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the following regularized effective viscosity using $\eps>0$,
 \begin{equation}
@@
 
-Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding glaciers.
+Our SIGP model uses dynamic boundary conditions for an isolated, grounded, and non-sliding glaciers.
 (Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.
@@
 \end{equation}
-On the other sub-aerial surfaces we set a condition of zero applied stress,
+On the sub-aerial surfaces we set a condition of zero applied stress,
 \begin{equation}
-\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
+\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{upper ice surface}} \label{eq:topbc}
 \end{equation}
-where $\bn$ is any normal to $\overline{\partial} \Lambda$.
-(A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, but that $\grad s$ may become singular there.
+where $\bn$ is any normal to $\partial \Lambda_s$.
+(A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached.
+Generally $\grad s$ is singular at a glacier margin.
 
 The simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.
-However, the above equations make no reference to the climate input function $a$, so we need another ``equation'', one which is actually an inequality.
+However, the above equations make no reference to the climate input function $a$, so we need another ``equation'', one which is fundamentally an inequality as well.
 Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we set
@@
 \end{equation}
-as an (un-normalized) top-surface normal almost everywhere on $\Omega$.  Furthermore we extend the surface value of the velocity to all of $\Omega$:
+as an (un-normalized) top-surface normal defined almost everywhere on $\Omega$.  Furthermore we extend the surface value of the velocity to all of $\Omega$:
 \begin{equation}
@@
 \end{equation}
-A discontinuity of this function is allowed at the ice margin.
+A discontinuity of this function is (generally) expected at the ice margin.
 
-The steady-state surface kinematical equation (SKE) \cite[equation (5.21)]{GreveBlatter2009} is the needed additional aspect of mass conservation:
+The steady-state surface kinematical equation (SKE) \cite[equation (5.21)]{GreveBlatter2009} is the needed additional equation:
 \begin{equation}
@@
 \end{equation}
-Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time, while some references use a distinct quantity, namely the thickness added perpendicularly to the ice surface, significantly different if $\grad s$ is large.
-
+Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time \cite{GreveBlatter2009}.
 Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.
 In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.
-An NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
+By definition, an NCP on a vector space $V$ combines the three statements
 \begin{equation}
@@
 However, as the domain on which $\bu,p$ are defined is only known via the solution, \eqref{eq:strongform} is at best an incomplete description.
-The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will partially address this concern.
-
-However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unknown theory regarding well-posedness and solution regularity.
+(The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will address this concern.)  However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unresolved theory regarding well-posedness and solution regularity.
 Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.
@@
 After stating the SIGP weak form next in Section \ref{sec:weakido}, in Sections \ref{sec:fe}--\ref{sec:results} we will construct and demonstrate a multilevel scheme for the problem.
-As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.
-By contrast, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.
+Noting that numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.
 
@@
 The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.
-Then we address the weak form of the SIGP.
+Then we address the weak form of the geometry-determining SIGP.
 
@@
 \end{equation}
-for all $(\bv,q) \in \mathcal{M}_\Lambda$.  If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.
-
+for all $(\bv,q) \in \mathcal{M}_\Lambda$.
+If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.
 Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.
@@
 
-However, we must go beyond the problem on predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.
+However, we must go beyond predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.
 Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.
@@
 Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.
-(Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
+(Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds that a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, is in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
 \begin{equation}
@@
 Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or be singular at the glacier margin even if $s$ is smooth on $\{s>b\}$.
-(We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.)
+We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.
 
@@
 \caption{The IDO $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$.
-Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ may be discontinuous as shown.}
+Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ is generally discontinuous.}
 \label{fig:idoaction}
@@
 Evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}, and furthermore the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.
-In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.
-It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix
+In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}; generically $\tilde\bu - \bu \ne 0$ a.e.~on $\Lambda_s$.
+It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix.
 
-By contrast, in the simpler isothermal, nonsliding SIA model the operator is a nonlinear, second-order differential operator,
+By contrast, in the simpler isothermal, nonsliding, $d=2$ SIA model the operator is a nonlinear, second-order differential operator,
 \begin{equation}
@@
 Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.
-However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in this sense we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.
+However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in fact for ice sheets we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.
 
@@
 
-For now we re-write \eqref{eq:strongform} as a clean NCP with a hidden dynamical problem:
+For now we re-write \eqref{eq:strongform} as a cleaner NCP with a hidden dynamical problem:
 \begin{align}
@@
 \end{align}
-Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.
-The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.
+These statements hold on all of $\Omega$, and the roles of the data $a,b$ are evident, and the solution is the ice surface $z=s(x,y)$, but the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.
 
-To show the SIGP weak form we define a functional which is simply the dual of $\Phi$:
+To reveal the SIGP weak form we define a functional which is simply the dual of $\Phi$:
 \begin{equation}
-F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
+F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy \label{eq:sigpfunctional}
 \end{equation}
+% in here one could integrate-by-parts on the vertical velocity using incompressibility:
+% u|_s . n|_s = u|_s s_x + v|_s s_y + w|_s = u|_s s_x + v|_s s_y - \int_b^s (u_x + v_y) dz
+% and then integrate by parts in plane
+where $r \in W^{1,q}(\Omega)$.
 One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.
@@
 \end{equation}
-One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
-In particular the strong SKE \eqref{eq:ske} holds on $s>b$; it is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.
+One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:ske} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.
+Note that the strong SKE \eqref{eq:ske}, which holds on $s>b$, is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.
 
@@
 (The ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.)  It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge \eqref{eq:sigpweakform} does \emph{not} arise in this way.
-Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.
-This model is also not a minimization, except in the flat-bed case.
+Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, and this SIA model is not a minimization except in the flat-bed case.
 
@@
 
-We will need the (Gateaux) derivative of $\Phi$.
-Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.
-From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then by definition
+We will need the (Gateaux) derivative of $\Phi$.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  Then
 \begin{equation}
@@
 on $\{s>b\}$.
-However, $\Phi'(s)[t]$ must actually be defined as a one-sided directional derivative on $\{s=b\}$.
-That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.
+However, $\Phi'(s)[t]$ is a one-sided directional derivative on $\{s=b\}$.
+That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, thus $t\ge 0$, on the (active) set where $s=b$.
 To simplify we define an $s$-independent set
@@
 \end{equation}
-It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]=\int_\Omega \Phi'(s)[t] r$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
+Then $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, and $F'(s)[t,r]=\int_\Omega \Phi'(s)[t] r$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.
 
@@
 In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.
-Substantial performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.
+Performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.
 
@@
 
-Assume $\Omega \subset \RR^d$ is polygonal with triangulation $\mathcal{T}$.
-(If $d=1$ then $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
+Assume $d=2$ and that $\Omega \subset \RR^d$ is polygonal with triangulation $\mathcal{T}$.
+(If $d=1$ then $\mathcal{T}$ would denote an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
 \begin{equation}
@@
 \begin{equation}
-\mathcal{K}^h = \{r^h \in \mathcal{V}^h \,:\, r^h \ge b^h \text{ and } r^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
+\mathcal{K}^h = \{s^h \in \mathcal{V}^h \,:\, s^h \ge b^h \text{ and } s^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
 \end{equation}
-Each $s^h\in \mathcal{K}^h$ defines a numerical icy domain $\Lambda_{s^h} \subset \RR^{d+1}$ as in \eqref{eq:lambdas}.
+(Compare \eqref{eq:Kconstraintset}.)  Each $s^h\in \mathcal{K}^h$ defines a numerical icy domain $\Lambda_{s^h} \subset \RR^{d+1}$ as in \eqref{eq:lambdas}.
 A triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in $\Lambda_{s^h}$ if $s^h=b^h$ at every vertex of $T$, and icy otherwise.
 
-Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$), so the triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.
+Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$); the triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.
 As shown in Figure \ref{fig:extruded}, each icy $T \in \mathcal{T}$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.
@@
 In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold).
-In the ``cliffs'' extrusion (right) all elements have a minimum thickness so the near-margin trace evaluation occurs above $s^h$ (dotted).}
+In the ``cliffs'' extrusion (right) all elements have a minimum thickness (dotted).}
 \label{fig:extruded}
@@
 Thus $(\bu^h,p^h) \in \mathcal{M}^h$ solves
+\begin{align}
+G_{\Lambda_{s^h}}(\bu^h,p^h)[\bv^h,q^h] &= 0  \label{eq:feglenstokesweak}
+\end{align}
+for all $(\bv^h,q^h) \in \mathcal{M}^h$, where
 \begin{equation}
-F_{\Lambda_{s^h}}(\bu^h,p^h)[\bv^h,q^h] = \int_{\Lambda_{s^h}} 2 \nu_\eps(|D\bu^h|) D\bu^h : D\bv^h - p^h \Div\bv^h - (\Div\bu^h) q^h - \rhoi \bg \cdot \bv^h\,d\bx = 0 \label{eq:feglenstokesweak}
+G_{\Lambda_{s^h}} = \int_{\Lambda_{s^h}} 2 \nu_\eps(|D\bu^h|) D\bu^h : D\bv^h - p^h \Div\bv^h - (\Div\bu^h) q^h - \rhoi \bg \cdot \bv^h\,d\bx.  \label{eq:feglenstokesfunctional}
 \end{equation}
-for all $(\bv^h,q^h) \in \mathcal{M}^h$.
+
 The power-law nonlinearity in \eqref{eq:feglenstokesweak} is resolved by using a Newton iteration, with direct solution of the step equations and back-tracking line search.
 As stated in Section \ref{sec:intro}, our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme are needed to solve the discrete SIGP, \eqref{eq:fesigpweakform} below.
-Overall optimality of the solution algorithm would also require an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
+Overall optimality of the solution algorithm would also require an optimal solution of Glen-Stokes problem \eqref{eq:feglenstokesweak}, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.
 
-In fact we will compare two extrusion modes.
+In fact we will compare two extrusion modes when solving problem \eqref{eq:feglenstokesweak}.
 In the ``pinched'' extrusion the prism elements tile $\Lambda_{s^h}$ exactly but the elements at the ice margin are degenerate; they have positive volume but zero height at some nodes or even empty (vertical) facets.
@@
 That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, and then extend by zero to all of $\Omega$.
-In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.
+In the ``cliffs'' extrusion the velocity trace is computed at the top of each column, and thus only nearby the original surface $s^h$, but the normal direction uses $s^h$ itself.
 

### 2022-10-15 paper/msg.tex
correct a published ref

@@
 
-Patch-type smoothers \cite{Farrelletal2019} have to our knowledge not been used for VI problems before, but they would sweep through the nodes solving local VIs of small dimension $\ell>1$, and again there would be serial and parallel versions.
+Patch-type smoothers \cite{Farrelletal2021} have to our knowledge not been used for VI problems before, but they would sweep through the nodes solving local VIs of small dimension $\ell>1$, and again there would be serial and parallel versions.
 However, we first consider the opposite extreme from a pointwise smoother, namely a global smoother in which all the degrees of freedom are coupled together into an algebraic system.
